---
syncID: 5f6d1e96deb747b2a527077b7303abf9
title: "Time Series 06: Create Plots with Multiple Panels, Grouped by Time Using ggplot Facets"
description: "This tutorial covers how to plot subsetted time series data (e.g., plot by season) using facets() and ggplot2. It also covers how to plot multiple metrics in one display panel."
dateCreated: 2015-10-22
authors: Megan A. Jones, Marisa Guarinello, Courtney Soderberg, Leah Wasser
contributors:
estimatedTime:
packagesLibraries: ggplot2, scales, gridExtra, grid, dplyr, reshape2
topics: time-series, phenology
languagesTool: R
dataProduct:
code1: /R/dc-tabular-time-series/06-Plotting-Facets-And-Group-By-Time-In-R.R
tutorialSeries: tabular-time-series 
urlTitle: dc-time-series-plot-facets-ndvi-r
---

This tutorial covers how to plot subsetted time series data using the `facets()`
function (e.g., plot by season) and to plot time series data with NDVI.

<div id="ds-objectives" markdown="1">

## Learning Objectives
After completing this tutorial, you will be able to:

 * Use `facets()` in the `ggplot2` package.
 * Combine different types of data into one plot layout.

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and, preferably, RStudio loaded on
your computer to complete this tutorial.

### Install R Packages
* **ggplot2:** `install.packages("ggplot2")`
* **scales:** `install.packages("scales")`
* **gridExtra:** `install.packages("gridExtra")`
* **grid:** `install.packages("grid")`
* **dplyr:** `install.packages("dplyr")`
* **reshape2:** `install.packages("reshape2")`
* **zoo:** `install.packages("zoo")`

<a href="https://www.neonscience.org/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

### Download Data 
<h3> <a href="https://ndownloader.figshare.com/files/3701572" > NEON Teaching Data Subset: Meteorological Data for Harvard Forest</a></h3>

The data used in this lesson were collected at the 
<a href="https://www.neonscience.org/" target="_blank"> National Ecological Observatory Network's</a> 
<a href="https://www.neonscience.org/field-sites/field-sites-map/HARV" target="_blank"> Harvard Forest field site</a>. 
These data are proxy data for what will be available for 30 years on the
 <a href="http://data.neonscience.org/" target="_blank">NEON data portal</a>
for the Harvard Forest and other field sites located across the United States.

<a href="https://ndownloader.figshare.com/files/3701572" class="link--button link--arrow"> Download Dataset</a>





****

**Set Working Directory:** This lesson assumes that you have set your working 
directory to the location of the downloaded and unzipped data subsets. 

<a href="https://www.neonscience.org/set-working-directory-r" target="_blank"> An overview
of setting the working directory in R can be found here.</a>

**R Script & Challenge Code:** NEON data lessons often contain challenges that reinforce 
learned skills. If available, the code for challenge solutions is found in the
downloadable R script of the entire lesson, available in the footer of each lesson page.


****

### Recommended Tutorials
This tutorial uses both `dplyr` and `ggplot2`. If you are new to either of these 
R packages, we recommend the following NEON Data Skills tutorials before 
working through this one.

* <a href="https://www.neonscience.org/dc-time-series-subset-dplyr-r" target="_blank">*Subset & Manipulate Time Series Data with dplyr* tutorial</a>.
* <a href="https://www.neonscience.org/dc-time-series-plot-ggplot-r" target="_blank">*Plotting Time Series with ggplot in R* tutorial</a>. 


</div>

## Plot Data Subsets Using Facets
In this tutorial we will learn how to create a panel of individual plots - known
as facets in `ggplot2`. Each plot represents a particular `data_frame`
time-series subset, for example a year or a season.

### Load the Data
We will use the daily micro-meteorology data for 2009-2011 from the Harvard
Forest. If you do not have this data loaded into an R `data_frame`, please 
load them and convert date-time columns to a `date-time class` now.

```{r load-data}

# Remember it is good coding technique to add additional libraries to the top of
# your script 
library(lubridate) # for working with dates
library(ggplot2)  # for creating graphs
library(scales)   # to access breaks/formatting functions
library(gridExtra) # for arranging plots
library(grid)   # for arrangeing plots
library(dplyr)  # for subsetting by season

# set working directory to ensure R can find the file we wish to import
# setwd("working-dir-path-here")

# daily HARV met data, 2009-2011
harMetDaily.09.11 <- read.csv(
  file="NEON-DS-Met-Time-Series/HARV/FisherTower-Met/Met_HARV_Daily_2009_2011.csv",
  stringsAsFactors = FALSE
  )

# covert date to Date  class
harMetDaily.09.11$date <- as.Date(harMetDaily.09.11$date)

```

## ggplot2 Facets

Facets allow us to plot subsets of data in one cleanly organized panel. We use
`facet_grid()` to create a plot of a particular **variable** subsetted by a
particular **group**.
 
Let's plot air temperature as we did previously. We will name the `ggplot`
object `AirTempDaily`.

```{r plot-airt }

AirTempDaily <- ggplot(harMetDaily.09.11, aes(date, airt)) +
           geom_point() +
           ggtitle("Daily Air Temperature\n NEON Harvard Forest\n 2009-2011") +
           xlab("Date") + ylab("Temperature (C)") +
           scale_x_date(labels=date_format ("%m-%y"))+
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

AirTempDaily
```

<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i> **Data Tip:** If you are working with a date & time
class (e.g. POSIXct), you can use `scale_x_datetime` instead of `scale_x_date`.
</div>

This plot tells us a lot about the annual increase and decrease of temperature
at the NEON Harvard Forest field site. However, what if we want to plot each
year's worth of data individually?

We can use the `facet()` element in `ggplot` to create facets or a panel of 
plots that are grouped by a particular category or time period. To create a 
plot for each year, we will first need a **year** column in our data to use as a
subset factor. We created a year column using the `year` function in the 
`lubridate` package in the 
<a href="https://www.neonscience.org/time-series-subset-dplyr" target="_blank">*Subset and Manipulate Time Series Data with dplyr* tutorial</a>.

```{r plot-by-year}

# add year column to daily values
harMetDaily.09.11$year <- year(harMetDaily.09.11$date)

# view year column head and tail
head(harMetDaily.09.11$year)
tail(harMetDaily.09.11$year)

```

## Facet by Year
Once we have a column that can be used to group or subset our data, we can 
create a faceted plot - plotting each year's worth of data in an individual, 
labelled panel.

```{r plot-facet-year }
# run this code to plot the same plot as before but with one plot per season
AirTempDaily + facet_grid(. ~ year)
```

Oops - what happened? The plot did not render because we added the `year` column
after creating the `ggplot` object `AirTempDaily`. Let's rerun the plotting code
to ensure our newly added column is recognized.

```{r plot-facet-year-2}

AirTempDaily <- ggplot(harMetDaily.09.11, aes(date, airt)) +
           geom_point() +
           ggtitle("Daily Air Temperature\n NEON Harvard Forest") +
            xlab("Date") + ylab("Temperature (C)") +
            scale_x_date(labels=date_format ("%m-%y"))+
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

# facet plot by year
AirTempDaily + facet_grid(. ~ year)
```

The faceted plot is interesting, however the x-axis on each plot is formatted
as: month-day-year starting in 2009 and ending in 2011. This means that the data
for 2009 is on the left end of the x-axis and the data for 2011 is on the right 
end of the x-axis of the 2011 plot. 

Our plots would be easier to visually compare if the days were formatted in 
Julian or year days rather than date. We have Julian days stored in our
`data_frame` (`harMetDaily.09.11`) as `jd`.

<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i> **Data Tip:** If you are unfamiliar with Julian day or
year day, see the 
<a href="https://www.neonscience.org/julian-day-conversion-r)" target="_blank">Convert to Julian Day tutorial </a>.
</div>

```{r plot-precip-jd }

AirTempDaily_jd <- ggplot(harMetDaily.09.11, aes(jd, airt)) +
           geom_point() +
           ggtitle("Air Temperature\n NEON Harvard Forest Field Site") +
           xlab("Julian Day") + ylab("Temperature (C)") +
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

# create faceted panel
AirTempDaily_jd + facet_grid(. ~ year)

```

Using Julian day, our plots are easier to visually compare. Arranging our plots 
this way, side by side, allows us to quickly scan for differences along the
y-axis. Notice any differences in min vs max air temperature across the three
years?

## Arrange Facets

We can rearrange the facets in different ways, too.

```{r rearrange-facets }

# move labels to the RIGHT and stack all plots
AirTempDaily_jd + facet_grid(year ~ .)

```

If we use `facet_wrap` we can specify the number of columns.

```{r rearrange-facets-columns }

# display in two columns
AirTempDaily_jd + facet_wrap(~year, ncol = 2)

```

## Graph Two Variables on One Plot
Next, let's explore the relationship between two variables - air temperature
and soil temperature. We might expect soil temperature to fluctuate with changes
in air temperature over time.  

We will use `ggplot()` to plot `airt` and `s10t` (soil temperature 10 cm below
the ground). 

```{r plot-airt-soilt }

airSoilTemp_Plot <- ggplot(harMetDaily.09.11, aes(airt, s10t)) +
           geom_point() +
           ggtitle("Air vs. Soil Temperature\n NEON Harvard Forest Field Site\n 2009-2011") +
           xlab("Air Temperature (C)") + ylab("Soil Temperature (C)") +
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

airSoilTemp_Plot

```

The plot above suggests a relationship between the air and soil temperature as
we might expect. However, it clumps all three years worth of data into one plot.

Let's create a stacked faceted plot of air vs. soil temperature grouped by year.

Lucky for us, we can do this quickly with one line of code while reusing the
plot we created above.

```{r faceted-temp-plots }
# create faceted panel
airSoilTemp_Plot + facet_grid(year ~ .)
```

Have a close look at the data. Are there any noticeable min/max temperature 
differences between the three years?

<div id="ds-challenge" markdown="1">
### Challenge: Faceted Plot

Create a faceted plot of air temperature vs soil temperature by *month* rather 
than year.

HINT: To create this plot, you will want to add a month column to our
`data_frame`. We can use lubridate `month` in the same way we used `year` to add
a year column. 

</div>

```{r challenge-answer-temp-month, echo=FALSE }

# add month column to daily values
harMetDaily.09.11$month <- month(harMetDaily.09.11$date)

# recreate plot
airSoilTemp_Plot <- ggplot(harMetDaily.09.11, aes(airt, s10t)) +
           geom_point() +
           ggtitle("Air vs. Soil Temperature\n NEON Harvard Forest Field Site\n 2009-2011") +
           xlab("Air Temperature (C)") + ylab("Soil Temperature (C)") +
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

# create faceted panel
airSoilTemp_Plot + facet_wrap(~month, nc=3)

```

## Faceted Plots & Categorical Groups

In the challenge above, we grouped our data by *month* - specified by a numeric
value between 1 (January) and 12 (December). However, what if we wanted to 
organize our plots using a categorical (character) group such as month name?
Let's do that next.

If we want to group our data by month *name*, we first need to create a month
name column in our `data_frame`. We can create this column using the following
syntax:

`format(harMetDaily.09.11$date,"%B")`, 

which tells R to extract the month name (`%B`) from the date field.

```{r extract-month-name }
# add text month name column
harMetDaily.09.11$month_name <- format(harMetDaily.09.11$date,"%B")

# view head and tail
head(harMetDaily.09.11$month_name)
tail(harMetDaily.09.11$month_name)

# recreate plot
airSoilTemp_Plot <- ggplot(harMetDaily.09.11, aes(airt, s10t)) +
           geom_point() +
           ggtitle("Air vs. Soil Temperature \n NEON Harvard Forest Field Site\n 2009-2011") +
            xlab("Air Temperature (C)") + ylab("Soil Temperature (C)") +
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

# create faceted panel
airSoilTemp_Plot + facet_wrap(~month_name, nc=3)

```

Great! We've created a nice set of plots by month. However, how are the plots 
ordered? It looks like R is ordering things alphabetically, yet we know
that months are ordinal not character strings. To account for order, we can 
reassign the `month_name` field to a `factor`. This will allow us to specify
an order to each factor "level" (each month is a level).

The syntax for this operation is 

1. Turn field into a factor: `factor(fieldName) `.
2. Designate the `levels` using a list `c(level1, level2, level3)`.

In our case, each level will be a month.

```{r factor }
# order the factors
harMetDaily.09.11$month_name = factor(harMetDaily.09.11$month_name, 
                                      levels=c('January','February','March',
                                               'April','May','June','July',
                                               'August','September','October',
                                               'November','December'))
```

Once we have specified the factor column and its associated levels, we can plot 
again. Remember, that because we have modified a column in our `data_frame`, we
need to rerun our `ggplot` code.

```{r plot-by-month-levels }

# recreate plot
airSoilTemp_Plot <- ggplot(harMetDaily.09.11, aes(airt, s10t)) +
           geom_point() +
           ggtitle("Air vs. Soil Temperature \n NEON Harvard Forest Field Site\n 2009-2011") +
            xlab("Air Temperature (C)") + ylab("Soil Temperature (C)") +
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

# create faceted panel
airSoilTemp_Plot + facet_wrap(~month_name, nc=3)

```

## Subset by Season - *Advanced Topic*
Sometimes we want to group data by custom time periods. For example, we might
want to group by season. However, the definition of various seasons may vary by 
region which means we need to manually define each time period.

In the next coding section, we will add a season column to our data using a
manually defined query. Our field site is Harvard Forest (Massachusetts),
located in the northeastern portion of the United States. Based on the climate
of this region, we can divide the year into 4 seasons as follows: 

 * Winter: December - February
 * Spring: March - May 
 * Summer: June - August
 * Fall: September - November 
 
In order to subset the data by season we will use the `dplyr` package.  We
can use the numeric month column that we added to our data earlier in this
tutorial.

```   {r subsetting-by-season-1}

# add month to data_frame - note we already performed this step above.
harMetDaily.09.11$month  <- month(harMetDaily.09.11$date)

# view head and tail of column
head(harMetDaily.09.11$month)
tail(harMetDaily.09.11$month)
```

We can use `mutate()` and a set of `ifelse` statements to create a new
categorical variable called `season` by grouping three months together. 

Within `dplyr` `%in%` is short-hand for "contained within". So the syntax

`ifelse(month %in% c(12, 1, 2), "Winter",`

can be read as "if the `month` column value is 12 or 1 or 2, then assign the
value "Winter"". 

Our `ifelse` statement ends with

`ifelse(month %in% c(9, 10, 11), "Fall", "Error")`

which we can translate this as "if the `month` column value is 9 or 10 or 11,
then assign the value "Winter"."

The last portion `, "Error"` tells R that if a `month` column value does not 
fall within any of the criteria laid out in previous `ifelse` statements, 
assign the column the value of "Error". 


``` {r  subsetting-by-season-2}

harMetDaily.09.11 <- harMetDaily.09.11 %>% 
  mutate(season = 
           ifelse(month %in% c(12, 1, 2), "Winter",
           ifelse(month %in% c(3, 4, 5), "Spring",
           ifelse(month %in% c(6, 7, 8), "Summer",
           ifelse(month %in% c(9, 10, 11), "Fall", "Error")))))


# check to see if this worked
head(harMetDaily.09.11$month)
head(harMetDaily.09.11$season)
tail(harMetDaily.09.11$month)
tail(harMetDaily.09.11$season)

```

Now that we have a season column, we can plot our data by season!

```{r plot-by-season}

# recreate plot
airSoilTemp_Plot <- ggplot(harMetDaily.09.11, aes(airt, s10t)) +
           geom_point() +
           ggtitle("Air vs. Soil Temperature\n 2009-2011\n NEON Harvard Forest Field Site") +
            xlab("Air Temperature (C)") + ylab("Soil Temperature (C)") +
           theme(plot.title = element_text(lineheight=.8, face="bold",
                 size = 20)) +
           theme(text = element_text(size=18))

# run this code to plot the same plot as before but with one plot per season
airSoilTemp_Plot + facet_grid(. ~ season)

```

Note, that once again, we re-ran our `ggplot` code to make sure our new column
is recognized by R. We can experiment with various facet layouts next.

``` {r plot-by-season2}

# for a landscape orientation of the plots we change the order of arguments in
# facet_grid():
airSoilTemp_Plot + facet_grid(season ~ .)

```

Once again, R is arranging the plots in an alphabetical order not an order
relevant to the data. 

<div id="ds-challenge" markdown="1">
### Challenge: Create Plots by Season
The goal of this challenge is to create plots that show the relationship between
air and soil temperature across the different seasons with seasons arranged in
an ecologically meaningful order.  

1. Create a *factor* class season variable by converting the season column that
we just created to a factor, then organize the seasons chronologically as
follows: Winter, Spring, Summer, Fall. 

2. Create a new faceted plot that is 2 x 2 (2 columns of plots). HINT: One can
neatly plot multiple variables using facets as follows: 
`facet_grid(variable1 ~ variable2)`. 

3. Create a plot of air vs soil temperature grouped by year and season.
</div>

``` {r assigning-level-to-season, include=TRUE, results="hide", echo=FALSE  }
# 1
# create factor / assign levels
harMetDaily.09.11$season<- factor(harMetDaily.09.11$season, 
                                  level=c("Winter","Spring","Summer","Fall")) 

# check to make sure it worked
str(harMetDaily.09.11$season)

# rerun original air & soil temp code to incorporate the levels. 
airSoilTemp_Plot_season <- ggplot(harMetDaily.09.11,aes(airt, s10t)) + 
  geom_point(na.rm=TRUE) +    #removing the NA values
  ggtitle("Seasonal Air vs Soil Temperature\n 2009-2011 \n NEON Harvard Forest Site") +
  theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
  theme(text = element_text(size=20)) +
  xlab("Air Temperature (C)") + ylab("Soil Temperature (C)")

# 2   air & soil temp by season         
# new facetted plots
airSoilTemp_Plot_season + facet_wrap(~ season, nc=2)

# 3  air & soil temp by season & year
# new facetted plots
airSoilTemp_Plot_season + facet_grid(year ~ season)
```

## Work with Year-Month Data: base R and zoo Package 
Some data will have month formatted in `Year-Month` 
(e.g. `met_monthly_HARV$date`). 

(Note: You will load this file in the Challenge below)
``` {r view-year-month-data, echo=FALSE}

met_monthly_HARV <- read.csv(
  "NEON-DS-Met-Time-Series/HARV/FisherTower-Met/hf001-04-monthly-m.csv",
  stringsAsFactors = FALSE
  )

# view header of date
head(met_monthly_HARV$date)
```

For many analyses, we might want to summarize this data into a yearly total. 
Base R does NOT have a distinct year-month date class. Instead to work with a 
year-month field using base R, we need to convert to a Date class, which 
necessitates adding an associated day value. The syntax would be:

`as.Date(paste(met_monthly_HARV$date,"-01", sep=""))`

The syntax above creates a `Date` column from the `met_montly_HARV$date` column.
We then add the arbitrary date - the first (`"-01"`). The final bit of code
(`sep=""`) designates the character string used to separate the month, day,
and year portions of the returned string (in our case nothing, as we have 
included the hyphen with our inserted date value).

Alternatively, to work directly with a year-month data we could use the `zoo`
package and its included year-month date class - `as.yearmon`. With `zoo` the
syntax would be:

`as.Date(as.yearmon(met_monthly_HARV$date))`

<div id="ds-challenge" markdown="1">
### Challenge: Convert Year-Month Data
The goal of this challenge is to use both the base R and the `zoo` package
methods for working with year-month data.  

Load the `NEON-DS-Met-Time-Series/HARV/FisherTower-Met/hf001-04-monthly-m.csv`
file and give it the name `met_monthly_HARV`. Then:

1. Convert the date field into a date/time class using both base R and the 
`zoo` package. Name the new fields `date_base` and `ymon_zoo` respectively.  
2. Look at the format and check the class of both new date fields.
3. Convert the `ymon_zoo` field into a new Date class field (`date_zoo`) so it
can be used in base R, ggplot, etc.  

HINT: be sure to load the `zoo` package, if you have not already.  

</div>

```{r challenge-code-convert-monthly-data, results="hide", echo=FALSE, message=FALSE}

# read in the data
met_monthly_HARV <- read.csv(
  "NEON-DS-Met-Time-Series/HARV/FisherTower-Met/hf001-04-monthly-m.csv",
  stringsAsFactors = FALSE
  )
# base R
# convert to date time - add a day "01" to each date to support this in base R
met_monthly_HARV$date_base <- as.Date(paste(met_monthly_HARV$date,"-01",sep=""))
head(met_monthly_HARV$date_base)
class(met_monthly_HARV$date_base)

# zoo
# load the zoo package
library(zoo)

# convert to yearmon class 
met_monthly_HARV$ymon_zoo <- as.yearmon(met_monthly_HARV$date)
head(met_monthly_HARV$ymon_zoo)
class(met_monthly_HARV$ymon_zoo)

# convert yearmon to Date class
met_monthly_HARV$date_zoo<- as.Date(met_monthly_HARV$ymon_zoo)
head(met_monthly_HARV$date_zoo)
class(met_monthly_HARV$date_zoo)

```

Do you prefer to use base R or `zoo` to convert these data to a date/time
class?

<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i> **Data Tip:** `zoo` date/time classes cannot be used
directly with ggplot2. If you deal with date formats that make sense to
primarily use `zoo` date/time classes, you can use ggplot2 with the addition of 
other functions. For details see the 
<a href="http://search.r-project.org/library/zoo/html/ggplot2.zoo.html" target="_blank" > ggplot2.zoo documentation.</a>
</div>

<div id="ds-challenge" markdown="1">
### Challenge: Plot Year-Month Data
Using the date field `date_base` (or `date_zoo`) that you created in the
previous challenge, create a faceted plot of annual air temperature for each
year (2001-2015) with month as the x-axis for the NEON Harvard Forest field
site.

</div>

``` {r challenge-code-plot-yearmonth-data, include=TRUE, results="hide", echo=FALSE  }
# add year- for facetted plot
met_monthly_HARV$year <- year(met_monthly_HARV$date_base)

# add month - x-axis in plot
met_monthly_HARV$month <- factor(month(met_monthly_HARV$date_base))

# create plot
long_term_temp <- ggplot(met_monthly_HARV,aes(month, airt)) +
        geom_point(na.rm=TRUE) +    #removing the NA values
        ggtitle("Air Temperature 2001-2015 \n NEON Harvard Forest Site") +
        theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
        theme(text = element_text(size=20)) +
        xlab("Month") + ylab("Air Temperature (C)")

# new facetted plots
long_term_temp + facet_wrap(~ year, nc=3)

```
