---
title: Content for download and explore tutorial
output: html_fragment
dateCreated: '2024-09-20'
---

This tutorial covers downloading NEON data, using the Data Portal and 
either the neonUtilities R package or the neonutilities Python package, 
as well as basic instruction in beginning to explore and work with the 
downloaded data, including guidance in navigating data documentation. 
We will explore data of 3 different types, and make a simple figure from 
each.

## NEON data
There are 3 basic categories of NEON data:

1. Remote sensing (AOP) - Data collected by the airborne observation platform, 
e.g. LIDAR, surface reflectance
1. Observational (OS) - Data collected by a human in the field, or in an 
analytical laboratory, e.g. beetle identification, foliar isotopes
1. Instrumentation (IS) - Data collected by an automated, streaming sensor, e.g. 
net radiation, soil carbon dioxide. This category also includes the 
surface-atmosphere exchange (SAE) data, which are processed and structured in 
a unique way, distinct from other instrumentation data 
(see the introductory <a href="https://www.neonscience.org/eddy-data-intro" target="_blank">eddy flux data tutorial</a> for details).

This lesson covers all three types of data. The download procedures are 
similar for all types, but data navigation differs significantly by type.

<div id="ds-objectives" markdown="1">

## Objectives

After completing this activity, you will be able to:

* Download NEON data using the neonUtilities package.
* Understand downloaded data sets and load them into R or Python for analyses.

## Things You’ll Need To Complete This Tutorial
You can follow either the R or Python code throughout this tutorial.
* For R users, we recommend using R version 4+ and RStudio.
* For Python users, we recommend using Python 3.9+.

```{r setup, include=FALSE}
library(reticulate)
knitr::opts_chunk$set(dev="png")
```

## Set up: Install Packages {.tabset}

Packages only need to be installed once, you can skip this step after the first time:

### R

* **neonUtilities**: Basic functions for accessing NEON data
* **neonOS**: Functions for common data wrangling needs for NEON observational data.
* **terra**: Spatial data package; needed for working with remote sensing data.

```{r install, eval=FALSE}

install.packages("neonUtilities")
install.packages("neonOS")
install.packages("terra")

```

### Python

* **neonutilities**: Basic functions for accessing NEON data
* **rasterio**: Spatial data package; needed for working with remote sensing data.

```{python p-install, eval=FALSE}

pip install neonutilities
pip install rasterio

```

## {-}

### Additional Resources

* <a href="https://github.com/NEONScience/NEON-Utilities/neonUtilities" target="_blank">GitHub repository for neonUtilities R package</a>
* <a href="https://github.com/NEONScience/NEON-utilities-python" target="_blank">GitHub repository for neonutilities Python package</a>
* <a href="https://www.neonscience.org/sites/default/files/cheat-sheet-neonUtilities_0.pdf" target="_blank">neonUtilities cheat sheet</a>. A quick reference guide for users. Focuses on the R package, but applicable to Python as well.

</div>


## Set up: Load packages {.tabset}

### R

```{r R-library, results="hide", message=FALSE}

library(neonUtilities)
library(neonOS)
library(terra)

```

### Python

```{python p-import}

import neonutilities as nu
import os
import rasterio
import pandas as pd
import matplotlib.pyplot as plt

```

## {-}


## Getting started: Download data from the Portal

Go to the 
<a href="http://data.neonscience.org" target="_blank">NEON Data Portal</a> 
and download some data! To follow the tutorial exactly, download 
Photosynthetically active radiation (PAR) (DP1.00024.001) data from 
September-November 2019 at Wind River Experimental Forest (WREF). The 
downloaded file should be a zip file named NEON_par.zip.

If you prefer to explore a different data product, you can still follow this 
tutorial. But it will be easier to understand the steps in the tutorial, 
particularly the data navigation, if you choose a sensor data product for 
this section.

Once you've downloaded a zip file of data from the portal, switch over to R 
or Python to proceed with coding.

## Stack the downloaded data files: stackByTable() {.tabset}

The `stackByTable()` (or `stack_by_table()`) function will unzip and join 
the files in the downloaded zip file.

### R

```{r stacking-portal, eval=FALSE}

# Modify the file path to match the path to your zip file
stackByTable("~/Downloads/NEON_par.zip")

```

### Python

```{python p-stacking-portal, eval=FALSE}

# Modify the file path to match the path to your zip file
nu.stack_by_table(os.path.expanduser("~/Downloads/NEON_par.zip"))

```

## {-}

In the directory where the zipped file was saved, you should now have 
an unzipped folder of the same name. When you open this you will see a 
new folder called **stackedFiles**, which should contain at least seven 
files: **PARPAR_30min.csv**, **PARPAR_1min.csv**, **sensor_positions.csv**, 
**variables_00024.csv**, **readme_00024.txt**, **issueLog_00024.csv**, 
and **citation\_00024\_RELEASE-202X.txt**.

## Navigate data downloads: IS {.tabset}

Let's start with a brief description of each file. This set of files is 
typical of a NEON IS data product.

* **PARPAR_30min.csv**: PAR data at 30-minute averaging intervals
* **PARPAR_1min.csv**: PAR data at 1-minute averaging intervals
* **sensor_positions.csv**: The physical location of each sensor collecting PAR measurements. There is a PAR sensor at each level of the WREF tower, and this table lets you connect the tower level index to the height of the sensor in meters.
* **variables_00024.csv**: Definitions and units for each data field in the PARPAR_#min tables.
* **readme_00024.txt**: Basic information about the PAR data product.
* **issueLog_00024.csv**: A record of known issues associated with PAR data.
* **citation\_00024\_RELEASE-202X.txt**: The citation to use when you publish a paper using these data, in BibTeX format.

We'll explore the 30-minute data. To read the file, use the function 
`readTableNEON()` or `read_table_neon()`, which uses the variables file 
to assign data types to each column of data:

### R

```{r read-par-hidden, results="hide", message=FALSE, warning=FALSE}

par30 <- readTableNEON(
  dataFile="~/Downloads/NEON_par_R/stackedFiles/PARPAR_30min.csv", 
  varFile="~/Downloads/NEON_par_R/stackedFiles/variables_00024.csv")
head(par30)

```
```{r read-par, eval=FALSE}

par30 <- readTableNEON(
  dataFile="~/Downloads/NEON_par/stackedFiles/PARPAR_30min.csv", 
  varFile="~/Downloads/NEON_par/stackedFiles/variables_00024.csv")
head(par30)

```

### Python

```{python p-read-par, results="hide", message=FALSE, warning=FALSE}

par30 = nu.read_table_neon(
  data_file=os.path.expanduser(
    "~/Downloads/NEON_par/stackedFiles/PARPAR_30min.csv"), 
  var_file=os.path.expanduser(
    "~/Downloads/NEON_par/stackedFiles/variables_00024.csv"))
# Open the par30 table in the table viewer of your choice

```

## {-}

The first four columns are added by `stackByTable()` when it merges 
files across sites, months, and tower heights. The column 
`publicationDate` is the date-time stamp indicating when the data 
were published, and the `release` column indicates which NEON data release 
the data belong to. For more information about NEON data releases, see the 
<a href="https://www.neonscience.org/data-samples/data-management/data-revisions-releases" target="_blank">Data Product Revisions and Releases</a> page.

Information about each data column can be found in the variables file, 
where you can see definitions and units for each column of data.

### Plot PAR data {.tabset}

Now that we know what we're looking at, let's plot PAR from the top 
tower level. We'll use the mean PAR from each averaging interval, and we 
can see from the sensor positions file that the vertical index 080 
corresponds to the highest tower level. To explore the sensor positions 
data in more depth, see the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/neon-spatial-data-basics" target="_blank">spatial data</a> tutorial.

#### R

```{r plot-par, dpi=72, eval=TRUE}

plot(PARMean~endDateTime, 
     data=par30[which(par30$verticalPosition=="080"),],
     type="l")

```

#### Python

```{python p-plot-par, dpi=72, eval=TRUE}

par30top = par30[par30.verticalPosition=="080"]
fig, ax = plt.subplots()
ax.plot(par30top.endDateTime, par30top.PARMean)
plt.show()

```

### {-}

Looks good! The sun comes up and goes down every day, and some days 
are cloudy.

### Plot more PAR data {.tabset}

To see another layer of data, add PAR from a lower tower level to the 
plot.

#### R

```{r plot-par-lower, dpi=72, eval=TRUE}

plot(PARMean~endDateTime, 
     data=par30[which(par30$verticalPosition=="080"),],
     type="l")

lines(PARMean~endDateTime, 
     data=par30[which(par30$verticalPosition=="020"),],
     col="orange")

```

#### Python

```{python p-plot-par-lower, dpi=72, eval=TRUE}

par30low = par30[par30.verticalPosition=="020"]
fig, ax = plt.subplots()
ax.plot(par30top.endDateTime, par30top.PARMean)
ax.plot(par30low.endDateTime, par30low.PARMean)
plt.show()

```

### {-}

We can see there is a lot of light attenuation through the canopy.

## Download files and load directly to R: loadByProduct() {.tabset}

At the start of this tutorial, we downloaded data from the NEON data portal. 
NEON also provides an API, and the `neonUtilities` packages provide methods 
for downloading programmatically.

The steps we carried out above - downloading from the portal, stacking the 
downloaded files, and reading in to R or Python - can all be carried out in 
one step by the neonUtilities function `loadByProduct()`.

To get the same PAR data we worked with above, we would run this line of code 
using `loadByProduct()`:

### R

```{r run-loadByProduct-par-hidden, results="hide", include=FALSE}

parlist <- loadByProduct(dpID="DP1.00024.001", 
                         site="WREF", 
                         startdate="2019-09",
                         enddate="2019-11",
                         check.size=F)

```
```{r run-loadByProduct-par, eval=FALSE}

parlist <- loadByProduct(dpID="DP1.00024.001", 
                         site="WREF", 
                         startdate="2019-09",
                         enddate="2019-11")

```

### Python

```{python p-run-loadByProduct-par-hidden, results="hide", include=FALSE}

parlist = nu.load_by_product(dpid="DP1.00024.001", 
                site="WREF", 
                startdate="2019-09",
                enddate="2019-11",
                check_size=False)

```
```{python p-run-loadByProduct-par, eval=FALSE}

parlist = nu.load_by_product(dpid="DP1.00024.001", 
                site="WREF", 
                startdate="2019-09",
                enddate="2019-11")

```

## {-}

### Explore loaded data

The object returned by `loadByProduct()` in R is a named list, and the 
object returned by `load_by_product()` in Python is a dictionary. The 
objects contained in the list or dictionary are the same set of tables 
we ended with after stacking the data from the portal above. You can 
see this by checking the names of the tables in `parlist`:

#### R

```{r par-list-names}

names(parlist)

```

#### Python

```{python p-par-list-names}

parlist.keys()

```

### {-}


Now let's walk through the details of the inputs and options in 
`loadByProduct()`.

This function downloads data from the NEON API, merges the site-by-month 
files, and loads the resulting data tables into the programming environment, 
assigning each data type to the appropriate class. This is a popular 
choice for NEON data users because it ensures you're always working with 
the latest data, and it ends with ready-to-use tables. However, if 
you use it in a workflow you run repeatedly, keep in mind it will re-download 
the data every time. See below for suggestions on saving the data locally 
to save time and compute resources.

`loadByProduct()` works on most observational (OS) and sensor (IS) data, 
but not on surface-atmosphere exchange (SAE) data and remote sensing (AOP) 
data. For functions that download AOP data, see the final 
section in this tutorial. For functions that work with SAE data, see 
the <a href="https://www.neonscience.org/eddy-data-intro" target="_blank">NEON eddy flux data tutorial</a>.

The inputs to `loadByProduct()` control which data to download and how 
to manage the processing. The list below shows the R syntax; in Python, 
the inputs are the same but all lowercase (e.g. `dpid` instead of `dpID`) 
and `.` is replaced by `_`.

* `dpID`: the data product ID, e.g. DP1.00002.001
* `site`: defaults to "all", meaning all sites with available data; 
can be a vector of 4-letter NEON site codes, e.g. 
`c("HARV","CPER","ABBY")` (or `["HARV","CPER","ABBY"]` in 
Python)
* `startdate` and `enddate`: defaults to NA, meaning all dates 
with available data; or a date in the form YYYY-MM, e.g. 
2017-06. Since NEON data are provided in month packages, finer 
scale querying is not available. Both start and end date are 
inclusive.
* `package`: either basic or expanded data package. Expanded data 
packages generally include additional information about data 
quality, such as chemical standards and quality flags. Not every 
data product has an expanded package; if the expanded package is 
requested but there isn't one, the basic package will be 
downloaded.
* `timeIndex`: defaults to "all", to download all data; or the 
number of minutes in the averaging interval. Only applicable to IS 
data.
* `release`: Specify a NEON data release to download. Defaults to 
the most recent release plus provisional data. See the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/release-provisional-tutorial" target="_blank">release tutorial</a> 
for more information.
* `include.provisional`: T or F: should Provisional data be included in 
the download? Defaults to F to return only Released data, which are 
citable by a DOI and do not change over time. Provisional data are 
subject to change.
* `check.size`: T or F: should the function pause before downloading 
data and warn you about the size of your download? Defaults to T; if 
you are using this function within a script or batch process you 
will want to set it to F.
* `token`: Optional NEON API token for faster downloads. See <a href="https://www.neonscience.org/resources/learning-hub/tutorials/neon-api-tokens-tutorial" target="_blank">this tutorial</a> 
for instructions on using a token.
* `progress`: Set to F to turn off progress bars.
* `cloud.mode`: Can be set to T if you are working in a cloud 
environment; enables more efficient data transfer from NEON's 
cloud storage.

The `dpID` is the data product identifier of the data you want to 
download. The DPID can be found on the 
<a href="http://data.neonscience.org/data-products/explore" target="_blank">
Explore Data Products page</a>.
It will be in the form DP#.#####.###

### Download observational data {.tabset}

To explore observational data, we'll download aquatic plant chemistry data 
(DP1.20063.001) from three lake sites: Prairie Lake (PRLA), Suggs Lake (SUGG), 
and Toolik Lake (TOOK).

#### R

```{r run-loadByProduct, results="hide"}

apchem <- loadByProduct(dpID="DP1.20063.001", 
                  site=c("PRLA","SUGG","TOOK"), 
                  package="expanded",
                  release="RELEASE-2024",
                  check.size=F)

```

#### Python

```{python p-run-loadByProduct, results="hide"}

apchem = nu.load_by_product(dpid="DP1.20063.001", 
                  site=["PRLA", "SUGG", "TOOK"], 
                  package="expanded",
                  release="RELEASE-2024",
                  check_size=False)

```

### {-}

## Navigate data downloads: OS {.tabset}

As we saw above, the object returned by `loadByProduct()` is a named list of 
data frames. Let's check out what's the same and what's different from the IS 
data tables.

### R

```{r loadBy-list, eval=T}

names(apchem)

```

### Python

```{python p-loadBy-list, eval=T}

apchem.keys()

```

## {-}

### Explore tables {.tabset}

As with the sensor data, we have some data tables and some metadata tables. 
Most of the metadata files are the same as the sensor data: **readme**, 
**variables**, **issueLog**, and **citation**. These files contain the same 
type of metadata here that they did in the IS data product. Let's look at the 
other files:

* **apl_clipHarvest**: Data from the clip harvest collection of aquatic plants
* **apl_biomass**: Biomass data from the collected plants
* **apl_plantExternalLabDataPerSample**: Chemistry data from the collected plants
* **apl_plantExternalLabQA**: Quality assurance data from the chemistry analyses
* **asi_externalLabPOMSummaryData**: Quality metrics from the chemistry lab
* **validation_20063**: For observational data, a major method for ensuring data quality is to control data entry. This file contains information about the data ingest rules applied to each input data field.
* **categoricalCodes_20063**: Definitions of each value for categorical data, such as growth form and sample condition

You can work with these tables from the named list object, but many people find 
it easier to extract each table from the list and work with it as an 
independent object. To do this, use the `list2env()` function in R or 
`globals().update()` in Python:

#### R

``` {r env, eval=T}

list2env(apchem, .GlobalEnv)

```

#### Python

``` {python p-env, eval=T}

globals().update(apchem)

```

### {-}

### Save data locally {.tabset}

Keep in mind that using `loadByProduct()` will re-download the data every 
time you run your code. In some cases this may be desirable, but it can be 
a waste of time and compute resources. To come back to these data without 
re-downloading, you'll want to save the tables locally. The most efficient 
option is to save the named list in total - this will also preserve the 
data types.

#### R

```{r save-files, eval=F}

saveRDS(apchem, 
        "~/Downloads/aqu_plant_chem.rds")

```

#### Python

```{p p-save-files, eval=F}

# There are a variety of ways to do this in Python; NEON
# doesn't currently have a specific recommendation. If 
# you don't have a data-saving workflow you already use, 
# we suggest you check out the pickle module.

```

### {-}

Then you can re-load the object to a programming environment any time.

Other options for saving data locally:

1. Similar to the workflow we started this tutorial with, but 
using `neonUtilities` to download instead of the Portal: 
Use `zipsByProduct()` and `stackByTable()` 
instead of `loadByProduct()`. With this option, use the function 
`readTableNEON()` to read the files, to get the same column type 
assignment that `loadByProduct()` carries out. Details
can be found in our <a href="https://www.neonscience.org/neonDataStackR" target="_blank">neonUtilities tutorial</a>.
2. Try out the community-developed `neonstore` package, which is designed for 
maintaining a local store of the NEON data you use. The `neonUtilities` 
function `stackFromStore()` works with files downloaded by `neonstore`. See 
the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/neonstore-stackfromstore-tutorial" target="_blank">neonstore tutorial</a> 
for more information.

Now let's explore the aquatic plant data. OS data products 
are simple in that the data are generally tabular, and data volumes are 
lower than the other NEON data types, but they are complex in that 
almost all consist of multiple tables containing information collected 
at different times in different ways. For example, samples collected 
in the field may be shipped to a laboratory for analysis. Data 
associated with the field collection will appear in one data table, 
and the analytical results will appear in another. Complexity in 
working with OS data usually involves bringing data together from 
multiple measurements or scales of analysis.

As with the IS data, the variables file can tell you more about 
the data tables.

OS data products each come with a Data Product User Guide, 
which can be downloaded with the data, or accessed from the 
document library on the Data Portal, or the <a href="https://data.neonscience.org/data-products/DP1.20063.001" target="_blank">Product Details</a> 
page for the data product. The User Guide is designed to give 
a basic introduction to the data product, including a brief 
summary of the protocol and descriptions of data format and 
structure.

### Explore isotope data {.tabset}

To get started with the aquatic plant chemistry data, let's 
take a look at carbon isotope ratios in plants across the three 
sites we downloaded. The chemical analytes are reported in the 
`apl_plantExternalLabDataPerSample` table, and the table is in 
long format, with one record per sample per analyte, so we'll 
subset to only the carbon isotope analyte:

#### R

```{r 13C-by-site, dpi=72, message=FALSE, warning=FALSE}

boxplot(analyteConcentration~siteID, 
        data=apl_plantExternalLabDataPerSample, 
        subset=analyte=="d13C",
        xlab="Site", ylab="d13C")

```

#### Python

```{python p-13C-by-site, dpi=72, results="hide", message=FALSE, warning=FALSE}

apl13C = apl_plantExternalLabDataPerSample[
         apl_plantExternalLabDataPerSample.analyte=="d13C"]
grouped = apl13C.groupby("siteID")["analyteConcentration"]
fig, ax = plt.subplots()
ax.boxplot(x=[group.values for name, group in grouped],
           tick_labels=grouped.groups.keys())
plt.show()

```

### {-}

We see plants at Suggs and Toolik are quite low in 13C, with more 
spread at Toolik than Suggs, and plants at Prairie Lake are relatively 
enriched. Clearly the next question is what species these data represent. 
But taxonomic data aren't present in the `apl_plantExternalLabDataPerSample` 
table, they're in the `apl_biomass` table. We'll need to join the two 
tables to get chemistry by taxon.

Every NEON data product has a Quick Start Guide (QSG), and for OS 
products it includes a section describing how to join the tables in the 
data product. Since it's a pdf file, `loadByProduct()` doesn't bring it in, 
but you can view the Aquatic plant chemistry QSG on the 
<a href="https://data.neonscience.org/data-products/DP1.20063.001" target="_blank">Product Details</a> 
page. In R, the `neonOS` package uses the information from the QSGs to provide 
an automated table-joining function, `joinTableNEON()`.

### Explore isotope data by species {.tabset}

#### R

```{r table-merge, eval=TRUE, results="hide", message=F}

apct <- joinTableNEON(apl_biomass, 
            apl_plantExternalLabDataPerSample)

```

Using the merged data, now we can plot carbon isotope ratio 
for each taxon.

```{r set-graph-param, eval=TRUE, echo=F, results='hide', message=F, warning=F}

par(mar=c(12,4,0.25,1))

```

```{r plot-13C-by-tax, dpi=72, eval=TRUE}

boxplot(analyteConcentration~scientificName, 
        data=apct, subset=analyte=="d13C", 
        xlab=NA, ylab="d13C", 
        las=2, cex.axis=0.7)

```

#### Python

There is not yet an equivalent to the `neonOS` package in 
Python, so we will code the table join manually, based on 
the info in the Quick Start Guide:

```{python p-table-merge, eval=TRUE, results="hide", message=F}

apct = pd.merge(apl_biomass, 
            apl_plantExternalLabDataPerSample,
            left_on=["siteID", "chemSubsampleID"],
            right_on=["siteID", "sampleID"],
            how="outer")

```

Using the merged data, now we can plot carbon isotope ratio 
for each taxon.

```{python p-plot-13C-by-tax, dpi=72, eval=TRUE, results="hide"}

apl13Cspp = apct[apct.analyte=="d13C"]
grouped = apl13Cspp.groupby("scientificName")["analyteConcentration"]
fig, ax = plt.subplots()
ax.boxplot(x=[group.values for name, group in grouped],
           tick_labels=grouped.groups.keys())
ax.tick_params(axis='x', labelrotation=90)
plt.show()

```

### {-}

And now we can see most of the sampled plants have carbon 
isotope ratios around -30, with just a few species accounting 
for most of the more enriched samples.


## Download remote sensing data: byFileAOP() and byTileAOP() {.tabset}

Remote sensing data files are very large, so downloading them 
can take a long time. `byFileAOP()` and `byTileAOP()` enable 
easier programmatic downloads, but be aware it can take a very 
long time to download large amounts of data.

Input options for the AOP functions are:

* `dpID`: the data product ID, e.g. DP1.00002.001
* `site`: the 4-letter code of a single site, e.g. HARV
* `year`: the 4-digit year to download
* `savepath`: the file path you want to download to; defaults to the 
working directory
* `check.size`: T or F: should the function pause before downloading 
data and warn you about the size of your download? Defaults to T; if 
you are using this function within a script or batch process you 
will want to set it to F.
* `easting`: `byTileAOP()` only. Vector of easting UTM coordinates whose 
corresponding tiles you want to download
* `northing`: `byTileAOP()` only. Vector of northing UTM coordinates 
whose corresponding tiles you want to download
* `buffer`: `byTileAOP()` only. Size in meters of buffer to include 
around coordinates when deciding which tiles to download
* `token`: Optional NEON API token for faster downloads.
* `chunk_size`: Only in Python. Set the chunk size of chunked 
downloads, can improve efficiency in some cases. Defaults to 1 MB.

Here, we'll download one tile of Ecosystem structure (Canopy Height 
Model) (DP3.30015.001) from WREF in 2017.

### R

```{r aop-tile, results="hide", eval=FALSE, comment=NA}

byTileAOP(dpID="DP3.30015.001", site="WREF", 
          year=2017,easting=580000, 
          northing=5075000, 
          savepath="~/Downloads")

```

### Python

```{python p-aop-tile, results="hide", eval=FALSE, comment=NA}

nu.by_tile_aop(dpid="DP3.30015.001", site="WREF", 
               year=2017,easting=580000, 
               northing=5075000, 
               savepath=os.path.expanduser(
                 "~/Downloads"))

```

## {-}

In the directory indicated in `savepath`, you should now have a folder 
named `DP3.30015.001` with several nested subfolders, leading to a tif 
file of a canopy height model tile.


## Navigate data downloads: AOP {.tabset}

To work with AOP data, the best bet in R is the `terra` package. 
It has functionality for most analyses you might want to do. In 
Python, we'll use the rasterio package here; explore 
NEON remote sensing tutorials for more guidance.

First let's read in the tile we downloaded:

### R

```{r read-aop, eval=TRUE}

chm <- rast("~/Downloads/DP3.30015.001/neon-aop-products/2017/FullSite/D16/2017_WREF_1/L3/DiscreteLidar/CanopyHeightModelGtif/NEON_D16_WREF_DP3_580000_5075000_CHM.tif")

```

### Python

```{python p-read-aop, eval=TRUE}

chm = rasterio.open(os.path.expanduser("~/Downloads/DP3.30015.001/neon-aop-products/2017/FullSite/D16/2017_WREF_1/L3/DiscreteLidar/CanopyHeightModelGtif/NEON_D16_WREF_DP3_580000_5075000_CHM.tif"))

```

## {-}

### Plot canopy height model {.tabset}

#### R

```{r reset-graph-param, eval=TRUE, echo=F, results='hide', message=F, warning=F}

par(mar=c(5,4,4,1))

```

```{r plot-aop, dpi=72, eval=TRUE}

plot(chm, col=topo.colors(6))

```

#### Python

```{python p-plot-aop, dpi=36, eval=TRUE}

plt.imshow(chm.read(1))
plt.show()

```

### {-}

Now we can see canopy height across the downloaded tile; 
the tallest trees are over 60 meters, not surprising in 
the Pacific Northwest. There is a clearing or clear 
cut in the lower right quadrant.

## Next steps

Now that you've learned the basics of downloading and understanding NEON data, 
where should you go to learn more? There are many more NEON tutorials to 
explore, including how to align remote sensing and ground-based measurements, 
a deep dive into the data quality flagging in the sensor data products, and 
much more. For a recommended suite of tutorials for new users, check out 
the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/get-started-neon-data-series-data-tutorials" target="_blank">Getting Started with NEON Data</a> 
tutorial series.


