---
syncID: 
title: "Using the neonPlants Package to Estimate Woody Plant Biomass and Production"
description: "Use the functions in the neonPlants package to estimate biomass and production of woody plants."
dateCreated: 2026-10-05
authors: Claire K. Lunch, Courtney Meier
contributors: 
estimatedTime: 60 minutes
packagesLibraries: neonOS, neonUtilities, neonPlants
topics: biomass, production, trees
languageTool: R
dataProduct: 
code1: 
tutorialSeries:
urlTitle: woody-biomass-production

---

NEON collects a wide variety of plant measurements designed for estimation of 
terrestrial plant biomass and primary production. For full details, see the 
<a href="https://data.neonscience.org/api/v0/documents/NEON.DOC.000914vC" target="_blank">TOS Science Design for Plant Biomass and Productivity</a>. In this tutorial, we will focus on biomass and 
production of trees. For biomass and production of herbaceous plants, see the 
<a href="TBD" target="_blank">herbaceous tutorial</a>. For aquatic plant 
biomass, see the <a href="TBD" target="_blank">aquatic plant tutorial</a>.

If you are new to using NEON data, or new to the `neonUtilities` R package, you 
may want to start with the 
<a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> tutorial, 
which provides an overview of NEON data access and the structure of NEON 
data and metadata files.

<div id="ds-objectives" markdown="1">

## Objectives

After completing this activity, you will be able to:

 * Use the `neonPlants` R package to estimate biomass and production of trees
 * Understand the algorithms used in the `neonPlants` R package to estimate biomass and production of trees
 * Understand the assumptions and exceptions in the algorithm

## Things You’ll Need To Complete This Tutorial

You will need a version of R (4.0 or higher. This code may work with 
earlier versions but it hasn't been tested) and, preferably, `RStudio` 
loaded on your computer to complete this tutorial.

## Additional Resources

* NEON <a href="http://data.neonscience.org" target="_blank"> Data Portal </a>
* <a href="https://www.neonscience.org/resources/code-hub" target="_blank">NEON Code Hub</a>
* <a href="TBD" target="_blank">NEON plant data tutorial series</a>

</div>


## Set up R environment and download data

First install and load the necessary packages. As of June 2026, NEON requires 
an API token for data downloads, to reduce bot scraping and improve user 
support. Tokens can be generated in NEON data portal user accounts - log in to 
your account or create one, and go to the API Tokens section. For best 
practices in storing and using tokens, follow the instructions <a href="https://www.neonscience.org/resources/learning-hub/tutorials/api-token-setup" target="_blank">here</a>.

```{r setup, eval=FALSE}

# install packages. you can skip this step if 
# the packages are already installed
install.packages("neonUtilities")
install.packages("neonOS")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("pak")
pak::pkg_install("NEONScience/neonPlants")

# load packages
library(neonUtilities)
library(neonOS)
library(neonPlants)
library(ggplot2)
library(dplyr)

# load token
token <- Sys.getenv("NEON_TOKEN")

```

```{r libraries, include=FALSE}

library(neonUtilities)
library(neonOS)
library(neonPlants)
library(ggplot2)
library(dplyr)

token <- Sys.getenv("NEON_TOKEN")

```

The productivity functions in `neonPlants` are designed to work with a single 
site at a time, and it's best to include all available data for a given site, 
to maximize the time series and reduce the impacts of any missing data. We'll 
use the San Joaquin Experimental Range (SJER) as the example site 
throughout this tutorial. SJER is an oak savanna site in California.

Use the `neonUtilities` function `loadByProduct()` to download and read 
in the data. Include provisional data to ensure the most recent data are 
included.

``` {r load-data, results="hide", message=FALSE}

# CKL: UPDATE TO REMOVE LATEST FOR PUBLICATION

veg <- loadByProduct(dpID="DP1.10098.001", 
                     site="SJER", 
                     include.provisional=T,
                     release="LATEST",
                     check.size=F,
                     token=token)

```

Copy each of the downloaded tables into the R environment.

``` {r env, results="hide"}

list2env(veg, .GlobalEnv)

```


## Explore data tables

For woody vegetation biomass and production, there are three data tables we 
will use:

* `vst_mappingandtagging`: Location and identification of mapped trees
* `vst_perplotperyear`: Record of which plots were sampled in which years
* `vst_apparentindividual`: Measurements of status, height, and diameter of monitored trees

Before proceeding, let's check each of these tables for duplicates.

``` {r remove-dups-all}

vegmap <- removeDups(data=vst_mappingandtagging, 
                          variables=variables_10098)

vegplot <- removeDups(data=vst_perplotperyear, 
                          variables=variables_10098)

vegai <- removeDups(data=vst_apparentindividual, 
                          variables=variables_10098)

```

There were no duplicates found in the `vst_mappingandtagging`  or 
`vst_perplotperyear` data tables. There were 50 duplicates in 
`vst_apparentindividual`, out of 13499 total records. Let's take a look at 
those records.

``` {r view-flag-values}

View(vegai[which(vegai$duplicateRecordQF!=0),])

```

One of the duplicate record pairs is of a single-bole tree, and we can see one 
record lists "tagRemoved" and the other contains stem and height measurements. 
We'll keep the record with the measurements.

``` {r drop-duplicate, message=FALSE}

vegai <- vegai |>
  filter_out(individualID=="NEON.PLA.D17.SJER.00014" &
               tagStatus=="tagRemoved")

```

The other duplicate records are shrubs, and we are focused on trees in this 
tutorial, so those duplicates will not affect our analysis.

## Estimate biomass

Use the `estimateWoodMass()` function from the `neonPlants` package to estimate 
the biomass of trees. Possible inputs to this function are:

* `inputDataList`: The list returned by `loadByProduct()`, which we named `veg` here
* `inputPerPlot`: The `vst_perplotperyear` table
* `inputMapTag`: The `vst_mappingandtagging` table
* `inputIndividual`: The `vst_apparentindvidual` table
* `inputNonWoody`: The `vst_non-woody` table
* `plotSubset`: The field plots to include in the analysis. Can be "all", "towerAll", "towerAnnualSubset", or "distributed".
* `growthFormSubset`: The growth forms to include in the analysis. Can be "all" or "tree".

The function can accept either a list of tables (`inputDataList`) or the 
individual tables. Since we modified the `vst_apparentindvidual` table and 
removed a duplicate, we'll input the individual tables rather than the 
original list returned by `loadByProduct()`.

We are interested in trees from all plots, so we enter `plotSubset="all` and 
`growthFormSubset="tree"`. We don't need to include the `vst_non-woody` table 
since we are not estimating non-woody mass.

``` {r biomass}

biom <- estimateWoodMass(inputIndividual=vegai,
                         inputMapTag=vegmap,
                         inputPerPlot=vegplot,
                         plotSubset="all",
                         growthFormSubset="tree")

```

What is returned by this function?

``` {r biomass-tables}

names(biom)

```

We have five tables, containing above-ground biomass estimates for individual 
trees, at the plot level, and at the site level, for each year of measurement. 
There is also a table with a list of lost or downed trees, and a `variables` 
table that provides definitions for the data fields in the other tables.

INSERT DISCUSSION OF ALLOMETRY ALGORITHM

Let's check out the biomass trajectory over the years measured. We'll plot 
live biomass by year. Note that total biomass includes standing dead trees! 
These can be very interesting, but be careful how you evaluate them in analyses.

We'll also include error bars for the standard deviation.

```{r plot-biomass-site}

gg <- ggplot(biom$vst_AGB_site, aes(year, woodLiveMassMean_Mgha)) +
  geom_line() +
  geom_point() +
  geom_linerange(aes(ymin=woodLiveMassMean_Mgha-woodLiveMassSD_Mgha,
                     ymax=woodLiveMassMean_Mgha+woodLiveMassSD_Mgha))
  xlab("Year") +
  ylab("Live wood mass")
gg

```

There are some ups and downs here, and the standard deviations are 
large. But consider that different sets of sampling plots are measured in 
different years. Let's look at biomass by plot by year:

```{r plot-biomass-plot}

gg <- ggplot(biom$vst_AGB_plot, aes(year, agbLive_Mgha, 
                                    group=plotID,
                                    color=plotID)) +
  
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Live wood mass")
gg

```

Here we can see most plots are quite consistent in biomass over time. The 
year-to-year variation in the site-level biomass estimate, and the high 
uncertainty, are driven by the variation in biomass between plots. This is 
important to consider when structuring your analyses. If you want to evaluate 
biomass over time, you may want to work at the plot level, or work with a 
subset of plots that are measured at the same frequency (e.g. the annually 
measured tower plots).

And keep in mind this can also manifest differently at different sites. SJER 
is an open woodland with scattered trees, so it's not surprising that plots 
vary substantially in biomass. This level of variation might be less likely at 
a more densely forested site.

Let's dig into the data a little more to look for evidence that this is what's 
driving the variation - are there different numbers of trees in different 
plots, or trees of very different biomass in different plots?

Let's get the most recent individual-level data for each plot, and plot the 
biomass of each tree at that time.

```{r plot-individuals-plot}

agb_recent <- biom$vst_AGB_indiv |>
  group_by(plotID, individualID) |>
  filter(year==max(year))

gg <- ggplot(agb_recent, aes(plotID, agb_kg, 
                                    group=individualID,
                                    color=plotID)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("") +
  ylab("Live wood mass")
gg

```

Here we can see both things are true: the number of trees varies quite a bit 
between plots, and the size of trees in each plot also varies a lot.

What next? We started out looking at site-level biomass over time. If we're 
interested in the trajectory of biomass, we are probably interested in 
productivity.

## Estimate production

Your first thought might be to simply calculate the change in biomass from one 
measurement point to the next, perhaps in the plot-level data above. But the 
calculation is a bit more complicated than that.

The `estimateWoodProd()` function in the `neonPlants` package follows the 
paradigm laid out in Clark et al 2001 (INSERT LINK). Essentially, we calculate 
the growth increment for each living tree between two time points, subtract 
the biomass of any trees that died, and add in the increment for any new, 
young trees.

Again, you may be thinking, doesn't that work out mathematically to be the same 
as the change in total live biomass? Not quite. In particular, new trees have 
to be calculated carefully. In common with all forest monitoring projects, NEON 
uses a threshold for minimum tree size - in NEON's case, 10cm diameter. If a 
tree was 9cm in 2022 and 11cm in 2023, its contribution to the biomass estimate 
in 2022 was zero, but its growth increment from 2022 to 2023 was not 11cm. To 
estimate productivity accurately, we estimate the time when its diameter was 
10cm, and then calculate the growth increment since then.

Similarly, trees that died may have grown between measurement periods before 
dying, and subtracting out their biomass at last measurement may underestimate 
the amount of live biomass lost. As of late 2026, `estimateWoodProd()` does not 
yet make this adjustment, but we plan to add it in a later version of the 
package. In most cases, this adjustment is much smaller than the adjustment for 
new trees.

Let's use `estimateWoodProd()` to estimate production at SJER. The data inputs 
are the same, except we don't need to specify the growth form, since this 
function is only configured to work for trees.

``` {r production}

prod <- estimateWoodProd(inputIndividual=vegai,
                         inputMapTag=vegmap,
                         inputPerPlot=vegplot,
                         plotSubset="all")

```

What is returned by this function?

``` {r production-tables}

names(prod)

```

As with the biomass function, we have tables of production estimates by 
individual, plot, and site; and table of variable definitions. We also have 
tables for duplicates, flagged, and missing. These tables contain information 
about possible data errors that may impact analyses. Details on each one:

* `vst_ANPP_duplicates`: If there are duplicate records in the `vst_apparentindividual` table (i.e. more than one measurement of the same stem in the same year), the code will remove them and copy them to this table. Note that NEON staff work to identify and remove duplicates internally; if you see duplicates here they may be removed in future data releases.
* `vst_ANPP_flagged`: There are a few scenarios that `estimateWoodProd()` will flag. (1) A tree is recorded as dead or not found in a sampling bout that is bracketed by sampling bouts where it is recorded as alive. In these cases we infer that it was alive at the intermediate point. (2) There is an implausibly large change in the mass of a tree from one measurement to the next, or (3) there is an implausibly large change in the stem diameter of a tree from one measurement to the next. The default behavior of `estimateWoodProd()` is to retain these records in the biomass calculation and copy all records relating to affected trees into this table for reference. To exclude records of type 2 and 3 from biomass calculations, add `flagged=filter` to the function call.
* `vst_ANPP_missing`: Sometimes a given tree is measured in one or more sampling bouts and then never observed in the plot again, alive or dead. Records associated with trees affected by this issue are copied into this table. The default behavior of `estimateWoodProd()` is to exclude these trees from biomass calculations; to include the trees and assume they have died, add `missing=retain` to the function call.

Let's take a look at productivity at the site level over time:

```{r plot-production-site}

gg <- ggplot(prod$vst_ANPP_site, aes(year, woodProd_Mghayr)) +
  geom_line() +
  geom_point() +
  geom_linerange(aes(ymin=woodProd_Mghayr-woodProdSD_Mghayr,
                     ymax=woodProd_Mghayr+woodProdSD_Mghayr)) +
  xlab("Year") +
  ylab("ANPP (Mg per hectare per year)")
gg

```

Some ups and downs and some variability, but nothing alarming. Note that ANPP 
in the first year is set to zero, because we can't evaluate productivity 
without a prior biomass measurement (SHOULD IT BE LIKE THIS? NA?).

Let's take a look at the plot level.

```{r plot-production-plot}

gg <- ggplot(prod$vst_ANPP_plot, aes(year, woodProd_Mghayr, 
                                    group=plotID,
                                    color=plotID)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("ANPP (Mg per hectare per year)")
gg

```

It might seem surprising that a number of plots saw negative ANPP between 
2015 and 2016. Remember that  
<a href="https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022JG007234" target="_blank">California experienced a severe drought</a> in 
2012-2016, and that negative ANPP typically results from tree mortality. 

Let's take a look at the frequency of recorded mortality and recruitment 
(growth of a new tree) events in 2016 versus the other measurement years.

```{r mortality-percent}

mortality_pct <- prod$vst_ANPP_indiv |>
  group_by(year) |>
  summarize(mortality_num=sum(transitionStatus=="mortality", na.rm=T),
            recruitment_num=sum(transitionStatus=="recruitment", na.rm=T),
            n=length(unique(individualID)))

mortality_pct <- mortality_pct |>
  mutate(mortality_pct=100*mortality_num/n,
         recruitment_pct=100*recruitment_num/n)

gg <- ggplot(mortality_pct, aes(year, mortality_pct)) +
  geom_point() +
  geom_line() +
  xlab("") +
  ylab("Percent mortality")
gg

```

The mortality rate was fairly high in 2016, but even higher in 2019, which 
doesn't match the pattern in productivity. Let's calculate the biomass lost to 
mortality in each year.

```{r mortality-mass}

mortality_mass <- prod$vst_ANPP_indiv |>
  group_by(year, transitionStatus) |>
  summarize(status_mass=sum(agb_kg, na.rm=T))

mortality_mass_pct <- mortality_mass |>
  tidyr::pivot_wider(names_from=transitionStatus,
              values_from=status_mass)

mortality_mass_pct <- mortality_mass_pct |>
  mutate(mortality_mass_pct=100*mortality/sum(`NA`, 
                                              noChange, 
                                              recruitment, 
                                              mortality,
                                              na.rm=T),
         recruitment_mass_pct=100*recruitment/sum(`NA`, 
                                              noChange, 
                                              recruitment, 
                                              mortality,
                                              na.rm=T))


```
