---
syncID: 
title: "Using the neonPlants Package to Estimate Woody Plant Biomass and Production"
description: "Use the functions in the neonPlants package to estimate biomass and production of woody plants."
dateCreated: 2026-10-05
authors: Claire K. Lunch, Courtney Meier
contributors: 
estimatedTime: 60 minutes
packagesLibraries: neonOS, neonUtilities, neonPlants
topics: biomass, production, trees
languageTool: R
dataProduct: 
code1: 
tutorialSeries:
urlTitle: woody-biomass-production

---

NEON collects a wide variety of plant measurements designed for estimation of 
terrestrial plant biomass and primary production. For full details, see the 
<a href="https://data.neonscience.org/api/v0/documents/NEON.DOC.000914vC" target="_blank">TOS Science Design for Plant Biomass and Productivity</a>. In this tutorial, we will focus on biomass and 
production of trees. For biomass and production of herbaceous plants, see the 
<a href="TBD" target="_blank">herbaceous tutorial</a>. For aquatic plant 
biomass, see the <a href="TBD" target="_blank">aquatic plant tutorial</a>.

If you are new to using NEON data, or new to the `neonUtilities` R package, you 
may want to start with the 
<a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> tutorial, 
which provides an overview of NEON data access and the structure of NEON 
data and metadata files.

<div id="ds-objectives" markdown="1">

## Objectives

After completing this activity, you will be able to:

 * Use the `neonPlants` R package to estimate biomass and production of trees
 * Understand the algorithms used in the `neonPlants` R package to estimate biomass and production of trees
 * Understand the assumptions and exceptions in the algorithm

## Things You’ll Need To Complete This Tutorial

You will need a version of R (4.0 or higher. This code may work with 
earlier versions but it hasn't been tested) and, preferably, `RStudio` 
loaded on your computer to complete this tutorial.

## Additional Resources

* NEON <a href="http://data.neonscience.org" target="_blank"> Data Portal </a>
* <a href="https://www.neonscience.org/resources/code-hub" target="_blank">NEON Code Hub</a>
* <a href="TBD" target="_blank">NEON plant data tutorial series</a>

</div>


## Set up R environment and download data

First install and load the necessary packages. As of June 2026, NEON requires 
an API token for data downloads, to reduce bot scraping and improve user 
support. Tokens can be generated in NEON data portal user accounts - log in to 
your account or create one, and go to the API Tokens section. For best 
practices in storing and using tokens, follow the instructions <a href="https://www.neonscience.org/resources/learning-hub/tutorials/api-token-setup" target="_blank">here</a>.

```{r setup, eval=FALSE}

# install packages. you can skip this step if 
# the packages are already installed
install.packages("neonUtilities")
install.packages("neonOS")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("pak")
pak::pkg_install("NEONScience/neonPlants")

# load packages
library(neonUtilities)
library(neonOS)
library(neonPlants)
library(ggplot2)
library(dplyr)

# load token
token <- Sys.getenv("NEON_TOKEN")

```

```{r libraries, include=FALSE}

library(neonUtilities)
library(neonOS)
library(neonPlants)
library(ggplot2)
library(dplyr)

token <- Sys.getenv("NEON_TOKEN")

```

The productivity functions in `neonPlants` are designed to work with a single 
site at a time, and it's best to include all available data for a given site, 
to maximize the time series and reduce the impacts of any missing data. We'll 
use the Wind River Experimental Forest (WREF) as the example site 
throughout this tutorial. WREF is a conifer forest in southwestern Washington 
State. 

Use the `neonUtilities` function `loadByProduct()` to download and read 
in the data. Include provisional data to ensure the most recent data are 
included.

``` {r load-data, results="hide", message=FALSE}

# CKL: UPDATE TO REMOVE LATEST FOR PUBLICATION

veg <- loadByProduct(dpID="DP1.10098.001", 
                     site="SJER", 
                     include.provisional=T,
                     release="LATEST",
                     check.size=F,
                     token=token)

```

Copy each of the downloaded tables into the R environment.

``` {r env, results="hide"}

list2env(veg, .GlobalEnv)

```


## Explore data tables

For woody vegetation biomass and production, there are three data tables we 
will use:

* `vst_mappingandtagging`: Location and identification of mapped trees
* `vst_perplotperyear`: Record of which plots were sampled in which years
* `vst_apparentindividual`: Measurements of status, height, and diameter of monitored trees

Before proceeding, let's check each of these tables for duplicates.

``` {r remove-dups-all}

vegmap <- removeDups(data=vst_mappingandtagging, 
                          variables=variables_10098)

vegplot <- removeDups(data=vst_perplotperyear, 
                          variables=variables_10098)

vegai <- removeDups(data=vst_apparentindividual, 
                          variables=variables_10098)

```

There were no duplicates found in the `vst_mappingandtagging`  or 
`vst_perplotperyear` data tables. There were 50 duplicates in 
`vst_apparentindividual`, out of 13499 total records. Let's take a look at 
those records.

``` {r view-flag-values}

View(vegai[which(vegai$duplicateRecordQF!=0),])

```

One of the duplicate record pairs is of a single-bole tree, and we can see one 
record lists "tagRemoved" and the other contains stem and height measurements. 
We'll keep the record with the measurements.

``` {r drop-duplicate, message=FALSE}

vegai <- vegai |>
  filter_out(individualID=="NEON.PLA.D17.SJER.00014" &
               tagStatus=="tagRemoved")

```

The other duplicate records are shrubs, and we are focused on trees in this 
tutorial, so those duplicates will not affect our analysis.

## Estimate biomass

Use the `estimateWoodMass()` function from the `neonPlants` package to estimate 
the biomass of trees. Possible inputs to this function are:

* `inputDataList`: The list returned by `loadByProduct()`, which we named `veg` here
* `inputPerPlot`: The `vst_perplotperyear` table
* `inputMapTag`: The `vst_mappingandtagging` table
* `inputIndividual`: The `vst_apparentindvidual` table
* `inputNonWoody`: The `vst_non-woody` table
* `plotSubset`: The field plots to include in the analysis. Can be "all", "towerAll", "towerAnnualSubset", or "distributed".
* `growthFormSubset`: The growth forms to include in the analysis. Can be "all" or "tree".

The function can accept either a list of tables (`inputDataList`) or the 
individual tables. Since we modified the `vst_apparentindvidual` table and 
removed a duplicate, we'll input the individual tables rather than the 
original list returned by `loadByProduct()`.

We are interested in trees from all plots, so we enter `plotSubset="all` and 
`growthFormSubset="tree"`. We don't need to include the `vst_non-woody` table 
since we are not estimating non-woody mass.

``` {r biomass}

biom <- estimateWoodMass(inputIndividual=vegai,
                         inputMapTag=vegmap,
                         inputPerPlot=vegplot,
                         plotSubset="all",
                         growthFormSubset="tree")

```

What is returned by this function?

``` {r biomass-tables}

names(biom)

```

We have five tables, containing above-ground biomass estimates for individual 
trees, at the plot level, and at the site level, for each year of measurement. 
There is also a table with a list of lost or downed trees, and a `variables` 
table that provides definitions for the data fields in the other tables.

Let's check out the biomass trajectory over the years measured. We'll plot 
live biomass by year. Note that total biomass includes standing dead trees! 
These can be very interesting, but be careful how you evaluate them in analyses.

```{r plot-biomass-site}

gg <- ggplot(biom$vst_AGB_site, aes(year, woodLiveMassMean_Mgha)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Live wood mass")
gg

```

There are some surprising ups and downs here. But consider that different sets 
of sampling plots are measured in different years. Let's look at biomass by 
plot by year:

```{r plot-biomass-plot}

gg <- ggplot(biom$vst_AGB_plot, aes(year, agbLive_Mgha, 
                                    group=plotID,
                                    color=plotID)) +
  
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Live wood mass")
gg

```

Here we can see most plots are quite consistent in biomass over time.




