---
syncID: 
title: "Using the neonPlants Package to Estimate Woody Plant Biomass and Production"
description: "Use the functions in the neonPlants package to estimate biomass and production of woody plants."
dateCreated: 2026-10-05
authors: Claire K. Lunch, Courtney Meier
contributors: 
estimatedTime: 60 minutes
packagesLibraries: neonOS, neonUtilities, neonPlants
topics: biomass, production, trees
languageTool: R
dataProduct: 
code1: 
tutorialSeries:
urlTitle: woody-biomass-production

---

NEON collects a wide variety of plant measurements designed for estimation of 
terrestrial plant biomass and primary production. For full details, see the 
<a href="https://data.neonscience.org/api/v0/documents/NEON.DOC.000914vC" target="_blank">TOS Science Design for Plant Biomass and Productivity</a>. In this tutorial, we will focus on biomass and 
production of trees. For biomass and production of herbaceous plants, see the 
<a href="TBD" target="_blank">herbaceous tutorial</a>. For aquatic plant 
biomass, see the <a href="TBD" target="_blank">aquatic plant tutorial</a>.

If you are new to using NEON data, or new to the `neonUtilities` R package, you 
may want to start with the 
<a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> tutorial, 
which provides an overview of NEON data access and the structure of NEON 
data and metadata files.

<div id="ds-objectives" markdown="1">

## Objectives

After completing this activity, you will be able to:

 * Use the `neonPlants` R package to estimate biomass and production of trees
 * Understand the algorithms used in the `neonPlants` R package to estimate biomass and production of trees
 * Understand the assumptions and exceptions in the algorithm

## Things You’ll Need To Complete This Tutorial

You will need a version of R (4.0 or higher. This code may work with 
earlier versions but it hasn't been tested) and, preferably, `RStudio` 
loaded on your computer to complete this tutorial.

## Additional Resources

* NEON <a href="http://data.neonscience.org" target="_blank"> Data Portal </a>
* <a href="https://www.neonscience.org/resources/code-hub" target="_blank">NEON Code Hub</a>
* <a href="TBD" target="_blank">NEON plant data tutorial series</a>

</div>


## Set up R environment and download data

First install and load the necessary packages. As of June 2026, NEON requires 
an API token for data downloads, to reduce bot scraping and improve user 
support. Tokens can be generated in NEON data portal user accounts - log in to 
your account or create one, and go to the API Tokens section. For best 
practices in storing and using tokens, follow the instructions <a href="https://www.neonscience.org/resources/learning-hub/tutorials/api-token-setup" target="_blank">here</a>.

```{r setup, eval=FALSE}

# install packages. you can skip this step if 
# the packages are already installed
install.packages("neonUtilities")
install.packages("neonOS")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("pak")
pak::pkg_install("NEONScience/neonPlants")

# load packages
library(neonUtilities)
library(neonOS)
library(neonPlants)
library(ggplot2)
library(dplyr)

# load token
token <- Sys.getenv("NEON_TOKEN")

```

```{r libraries, include=FALSE}

library(neonUtilities)
library(neonOS)
library(neonPlants)
library(ggplot2)
library(dplyr)

token <- Sys.getenv("NEON_TOKEN")

```

The productivity functions in `neonPlants` are designed to work with a single 
site at a time, and it's best to include all available data for a given site, 
to maximize the time series and reduce the impacts of any missing data. We'll 
use the San Joaquin Experimental Range (SJER) as the example site 
throughout this tutorial. SJER is an oak savanna site in California.

Use the `neonUtilities` function `loadByProduct()` to download and read 
in the data. Include provisional data to ensure the most recent data are 
included.

``` {r load-data, results="hide", message=FALSE}

# CKL: UPDATE TO REMOVE LATEST FOR PUBLICATION

veg <- loadByProduct(dpID="DP1.10098.001", 
                     site="SJER", 
                     include.provisional=T,
                     release="LATEST",
                     check.size=F,
                     token=token)

```

Copy each of the downloaded tables into the R environment.

``` {r env, results="hide"}

list2env(veg, .GlobalEnv)

```


## Explore data tables

For woody vegetation biomass and production, there are three data tables we 
will use:

* `vst_mappingandtagging`: Location and identification of mapped trees
* `vst_perplotperyear`: Record of which plots were sampled in which years
* `vst_apparentindividual`: Measurements of status, height, and diameter of monitored trees

Before proceeding, let's check each of these tables for duplicates.

``` {r remove-dups-all}

vegmap <- removeDups(data=vst_mappingandtagging, 
                          variables=variables_10098)

vegplot <- removeDups(data=vst_perplotperyear, 
                          variables=variables_10098)

vegai <- removeDups(data=vst_apparentindividual, 
                          variables=variables_10098)

```

There were no duplicates found in the `vst_mappingandtagging`  or 
`vst_perplotperyear` data tables. There were 50 duplicates in 
`vst_apparentindividual`, out of 13499 total records. Let's take a look at 
those records.

``` {r view-flag-values, eval=FALSE}

vegai |>
  filter(duplicateRecordQF!=0) |>
  View()

```

One of the duplicate record pairs is of a single-bole tree, and we can see one 
record lists "tagRemoved" and the other contains stem and height measurements. 
We'll keep the record with the measurements.

``` {r drop-duplicate, message=FALSE}

vegai <- vegai |>
  filter_out(individualID=="NEON.PLA.D17.SJER.00014" &
               tagStatus=="tagRemoved")

```

The other duplicate records are shrubs, and we are focused on trees in this 
tutorial, so those duplicates will not affect our analysis. Also note that NEON 
staff work to identify and remove duplicates internally; the duplicates 
identified here may be removed in future data releases.

## Estimate biomass

Use the `estimateWoodMass()` function from the `neonPlants` package to estimate 
the biomass of trees. Possible inputs to this function are:

* `inputDataList`: The list returned by `loadByProduct()`, which we named `veg` here
* `inputPerPlot`: The `vst_perplotperyear` table
* `inputMapTag`: The `vst_mappingandtagging` table
* `inputIndividual`: The `vst_apparentindvidual` table
* `inputNonWoody`: The `vst_non-woody` table
* `plotSubset`: The field plots to include in the analysis. Can be "all", "towerAll", "towerAnnualSubset", or "distributed".
* `growthFormSubset`: The growth forms to include in the analysis. Can be "all" or "tree".

The function can accept either a list of tables (`inputDataList`) or the 
individual tables. Since we modified the `vst_apparentindvidual` table and 
removed a duplicate, we'll input the individual tables rather than the 
original list returned by `loadByProduct()`.

We are interested in trees from all plots, so we enter `plotSubset="all` and 
`growthFormSubset="tree"`. We don't need to include the `vst_non-woody` table 
since we are not estimating non-woody mass.

``` {r biomass}

biom <- estimateWoodMass(inputIndividual=vegai,
                         inputMapTag=vegmap,
                         inputPerPlot=vegplot,
                         plotSubset="all",
                         growthFormSubset="tree")

```

What is returned by this function?

``` {r biomass-tables}

names(biom)

```

We have five tables, containing above-ground biomass estimates for individual 
trees, at the plot level, and at the site level, for each year of measurement. 
There is also a table with a list of lost or downed trees, and a `variables` 
table that provides definitions for the data fields in the other tables.

INSERT DISCUSSION OF ALLOMETRY ALGORITHM

Let's check out the biomass trajectory over the years measured. We'll plot 
live biomass by year. Note that total biomass includes standing dead trees! 
These can be very interesting, but be careful how you evaluate them in analyses.

We'll also include error bars for the standard deviation.

```{r plot-biomass-site}

gg <- ggplot(biom$vst_AGB_site, aes(year, woodLiveMassMean_Mgha)) +
  geom_line() +
  geom_point() +
  geom_linerange(aes(ymin=woodLiveMassMean_Mgha-woodLiveMassSD_Mgha,
                     ymax=woodLiveMassMean_Mgha+woodLiveMassSD_Mgha))
  xlab("Year") +
  ylab("Live wood mass")
gg

```

There are some ups and downs here, and the standard deviations are 
large. But consider that different sets of sampling plots are measured in 
different years. Let's look at biomass by plot by year:

```{r plot-biomass-plot}

gg <- ggplot(biom$vst_AGB_plot, aes(year, agbLive_Mgha, 
                                    group=plotID,
                                    color=plotID)) +
  
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Live wood mass")
gg

```

Here we can see most plots are quite consistent in biomass over time. The 
year-to-year variation in the site-level biomass estimate, and the high 
uncertainty, are driven by the variation in biomass between plots. This is 
important to consider when structuring your analyses. If you want to evaluate 
biomass over time, you may want to work at the plot level, or work with a 
subset of plots that are measured at the same frequency (e.g. the annually 
measured tower plots).

And keep in mind this can also manifest differently at different sites. SJER 
is an open woodland with scattered trees, so it's not surprising that plots 
vary substantially in biomass. This level of variation might be less likely at 
a more densely forested site.

Let's dig into the data a little more to look for evidence that this is what's 
driving the variation - are there different numbers of trees in different 
plots, or trees of very different biomass in different plots?

Let's get the most recent individual-level data for each plot, and plot the 
biomass of each tree at that time.

```{r plot-individuals-plot}

agb_recent <- biom$vst_AGB_indiv |>
  group_by(plotID, individualID) |>
  filter(year==max(year))

gg <- ggplot(agb_recent, aes(plotID, agb_kg, 
                                    group=individualID,
                                    color=plotID)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("") +
  ylab("Live wood mass")
gg

```

Here we can see both things are true: the number of trees varies quite a bit 
between plots, and the size of trees in each plot also varies a lot.

What next? We started out looking at site-level biomass over time. If we're 
interested in the trajectory of biomass, we are probably interested in 
productivity.

## Estimate production

Your first thought might be to simply calculate the change in biomass from one 
measurement point to the next, perhaps in the plot-level data above. But the 
calculation is a bit more complicated than that.

The `estimateWoodProd()` function in the `neonPlants` package follows the 
paradigm laid out in Clark et al 2001 (INSERT LINK). Essentially, we calculate 
the growth increment for each living tree between two time points and add in 
the increment for any new, young trees.

Again, you may be thinking, doesn't that work out mathematically to be the same 
as the change in total live biomass? Not quite. See the Clark paper for 
details, but in particular, new trees have to be accounted for carefully. In 
common with all forest monitoring projects, NEON uses a threshold for minimum 
tree size - in NEON's case, 10cm diameter. If a tree was 9cm in 2022 and 11cm 
in 2023, its contribution to the biomass estimate in 2022 was zero, but its 
growth increment from 2022 to 2023 was not 11cm. To estimate productivity 
accurately, we estimate the growth increment since it reached the 10cm 
threshold.

Similarly, trees that died between sampling bouts are ignored when calculating 
the increment for the growth interval during which they died, whereas they are 
included in the live biomass data until they die.

Let's use `estimateWoodProd()` to estimate production at SJER. The data inputs 
are the same, except we don't need to specify the growth form, since this 
function is only configured to work for trees.

``` {r production}

prod <- estimateWoodProd(inputIndividual=vegai,
                         inputMapTag=vegmap,
                         inputPerPlot=vegplot,
                         plotSubset="all")

```

What is returned by this function?

``` {r production-tables}

names(prod)

```

As with the biomass function, we have tables of production estimates by 
individual, plot, and site; and table of variable definitions. We also have 
tables for duplicates, flagged, and missing. These tables contain information 
about possible data errors that may impact analyses. Details on each one:

* `vst_ANPP_duplicates`: If there are duplicate records in the `vst_apparentindividual` table (i.e. more than one measurement of the same stem in the same year), the code will remove them and copy them to this table. The criteria for identifying duplicates here are more stringent than in the `removeDups()` function used above, and are specific to the tree growth form, which is why the two pairs of duplicates we see here weren't identified above.
* `vst_ANPP_flagged`: There are a few scenarios that `estimateWoodProd()` will flag. (1) A tree is recorded as dead or not found in a sampling bout that is bracketed by sampling bouts where it is recorded as alive. In these cases we infer that it was alive at the intermediate point. (2) There is an implausibly large change in the mass of a tree from one measurement to the next, or (3) there is an implausibly large change in the stem diameter of a tree from one measurement to the next. The default behavior of `estimateWoodProd()` is to retain these records in the productivity calculation and copy all records relating to affected trees into this table for reference. To exclude records of type 2 and 3 from productivity calculations, add `flagged=filter` to the function call.
* `vst_ANPP_missing`: Sometimes a given tree is measured in one or more sampling bouts and then never observed in the plot again, alive or dead. Records associated with trees affected by this issue are copied into this table. The default behavior of `estimateWoodProd()` is to exclude these trees from biomass calculations; to include the trees and assume they have died, add `missing=retain` to the function call.

Let's take a look at productivity at the site level over time:

```{r plot-production-site}

gg <- ggplot(prod$vst_ANPP_site, aes(year, woodProd_Mghayr)) +
  geom_line() +
  geom_point() +
  geom_linerange(aes(ymin=woodProd_Mghayr-woodProdSD_Mghayr,
                     ymax=woodProd_Mghayr+woodProdSD_Mghayr)) +
  xlab("Year") +
  ylab("ANPP (Mg per hectare per year)")
gg

```

Some ups and downs and some variability. Note that ANPP 
in the first year is set to zero, because we can't evaluate productivity 
without a prior biomass measurement (COURTNEY IS UPDATING THIS TO NA).

Let's take a look at the plot level.

```{r plot-production-plot}

gg <- ggplot(prod$vst_ANPP_plot, aes(year, woodProd_Mghayr, 
                                    group=plotID,
                                    color=plotID)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("ANPP (Mg per hectare per year)")
gg

```

A number of plots saw negative ANPP between 2015 and 2016. Production is a 
measurement of growth, and should always be positive. Negative values are 
sometimes observed when the diameter of a tree shrinks between measurement 
intervals, and for small shrinkage, this can be within the uncertainty of the 
method - trees shrink and swell with water content, for example.

Let's take a look at the negative values here.

``` {r view-neg-values, eval=FALSE}

prod$vst_ANPP_indiv |>
  filter(agbIncr_kg<0) |>
  View()

```

There are plenty of records with small reductions in size, but there are some 
with large changes. They are especially prevalent in 2016, and they are 
generally flagged. Let's re-run the calculation, excluding flagged records, 
and view plot-level production again.

```{r run-prod-flagged}

prodf <- estimateWoodProd(inputIndividual=vegai,
                          inputMapTag=vegmap,
                          inputPerPlot=vegplot,
                          plotSubset="all",
                          flagged="filter")

gg <- ggplot(prodf$vst_ANPP_plot, aes(year, woodProd_Mghayr, 
                                    group=plotID,
                                    color=plotID)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("ANPP (Mg per hectare per year)")
gg
```

Far fewer negative values, and a little noisy, but that is expected. At the 
site level, things are looking smoother:

```{r plot-production-site-flagged}

gg <- ggplot(prodf$vst_ANPP_site, aes(year, woodProd_Mghayr)) +
  geom_line() +
  geom_point() +
  geom_linerange(aes(ymin=woodProd_Mghayr-woodProdSD_Mghayr,
                     ymax=woodProd_Mghayr+woodProdSD_Mghayr)) +
  xlab("Year") +
  ylab("ANPP (Mg per hectare per year)")
gg

```

## Compare sites

Now that we have a picture of how the functions work and how to understand the 
outputs, let's take a look at the other two terrestrial sites in California, 
Soaproot Saddle (SOAP) and Lower Teakettle (TEAK).

Note that we could download these sites' data together, into a single set of 
outputs, but since `estimateWoodProd()` can only process a single site at a 
time, downloading each site separately will make the code steps easier.

``` {r load-data-other, results="hide", message=FALSE}

# CKL: UPDATE TO REMOVE LATEST FOR PUBLICATION

vegsoap <- loadByProduct(dpID="DP1.10098.001", 
                     site="SOAP", 
                     include.provisional=T,
                     release="LATEST",
                     check.size=F,
                     token=token)

vegteak <- loadByProduct(dpID="DP1.10098.001", 
                     site="TEAK", 
                     include.provisional=T,
                     release="LATEST",
                     check.size=F,
                     token=token)

```

Let's check the relevant tables for duplicates.

``` {r remove-dups-sites}

vegmapsoap <- removeDups(data=vegsoap$vst_mappingandtagging, 
                         table="vst_mappingandtagging",
                         variables=variables_10098)

vegplotsoap <- removeDups(data=vegsoap$vst_perplotperyear, 
                          table="vst_perplotperyear",
                          variables=variables_10098)

vegaisoap <- removeDups(data=vegsoap$vst_apparentindividual, 
                        table="vst_apparentindividual",
                        variables=variables_10098)

vegmapteak <- removeDups(data=vegteak$vst_mappingandtagging, 
                         table="vst_mappingandtagging",
                         variables=variables_10098)

vegplotteak <- removeDups(data=vegteak$vst_perplotperyear, 
                          table="vst_perplotperyear",
                          variables=variables_10098)

vegaiteak <- removeDups(data=vegteak$vst_apparentindividual, 
                        table="vst_apparentindividual",
                        variables=variables_10098)

```

Again we see no duplicates in the plot or mapping data, but a few in the annual 
measurements. Let's take a look.

``` {r view-flag-values-sites, eval=FALSE}

vegaisoap |>
  filter(duplicateRecordQF!=0) |>
  View()

vegaiteak |>
  filter(duplicateRecordQF!=0) |>
  View()

```

Among the full-sized trees, there are two pairs of duplicates at SOAP and five 
at TEAK.

Code not shown, because this is a manual/visual process: we examine all the 
records in `vst_apparentindividual` over the years for each of the affected 
individuals, and conclude the records indicated below are inconsistent with 
other records for the individual and should be dropped:

``` {r drop-duplicates-sites, message=FALSE}

vegaisoap <- vegaisoap |>
  filter_out(individualID=="NEON.PLA.D17.SOAP.03218" &
               growthForm=="multi-bole tree")

vegaisoap <- vegaisoap |>
  filter_out(individualID=="NEON.PLA.D17.SOAP.03510" &
               plantStatus=="Downed")

vegaiteak <- vegaiteak |>
  filter_out(individualID %in% c("NEON.PLA.D17.TEAK.00324",
                                 "NEON.PLA.D17.TEAK.01139",
                                 "NEON.PLA.D17.TEAK.01558") &
               plantStatus=="Standing dead")

vegaiteak <- vegaiteak |>
  filter_out(individualID=="NEON.PLA.D17.SOAP.01844" &
               growthForm=="sapling")

vegaiteak <- vegaiteak |>
  filter_out(individualID=="NEON.PLA.D17.SOAP.03269" &
               plantStatus=="Lost, presumed dead")

```

And now let's calculate biomass and production:

``` {r biomass-sites}

biosoap <- estimateWoodMass(inputIndividual=vegaisoap,
                         inputMapTag=vegmapsoap,
                         inputPerPlot=vegplotsoap,
                         plotSubset="all",
                         growthFormSubset="tree")

bioteak <- estimateWoodMass(inputIndividual=vegaiteak,
                         inputMapTag=vegmapteak,
                         inputPerPlot=vegplotteak,
                         plotSubset="all",
                         growthFormSubset="tree")

prodsoap <- estimateWoodProd(inputIndividual=vegaisoap,
                         inputMapTag=vegmapsoap,
                         inputPerPlot=vegplotsoap,
                         plotSubset="all",
                          flagged="filter")

prodteak <- estimateWoodProd(inputIndividual=vegaiteak,
                         inputMapTag=vegmapteak,
                         inputPerPlot=vegplotteak,
                         plotSubset="all",
                          flagged="filter")

```

Although we processed them separately, let's stitch the sites together for 
easier plotting.

``` {r stack-tables}

bioplots <- rbind(biom$vst_AGB_plot, 
                  biosoap$vst_AGB_plot, 
                  bioteak$vst_AGB_plot)

prodsites <- rbind(prodf$vst_ANPP_site,
                   prodsoap$vst_ANPP_site,
                   prodteak$vst_ANPP_site)

```



