---
syncID:
title: "Exploring Fish Species Richness"
description: "Use dplyr and ggplot2 to get a visual summary of species richness for all NEON aquatic sites."
dateCreated: 2020-05-26
authors: Garrett M. Williams
contributors: 
estimatedTime: 
packagesLibraries: neonUtilities, dplyr, ggplot2
topics: organisms, data-viz
languagesTool: R
dataProduct: DP1.20107.001
code1: 
tutorialSeries: 
urlTitle: fish-species-richness
---

<div id="ds-objectives" markdown="1">

## Learning Objectives 
After completing this tutorial you will be able to: 

* Download and navigate NEON's electrofishing data
* Get the data you need with the `select()` and `filter()` funtions from the 
  dplyr package


## Things You’ll Need To Complete This Tutorial

### R Programming Language
You will need a current version of R to complete this tutorial. We also recommend 
the RStudio IDE to work with R.

### The `dplyr` Package
You will need a basic understanding of dplyr, especially the piping operator, 
`%>%`. This can be found in our tutorial <a href="https://www.neonscience.org/grepl-filter-piping-dplyr-r">here</a>.

### R Packages to Install
Prior to starting the tutorial ensure that the following packages are installed. 
* **neonUtilites:** `install.packages("neonUtilities")`
* **dplyr:** `install.packages("dplyr")`
* **ggplot2:** `install.packages("ggplot2")`


<a href="https://www.neonscience.org/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.
 
</div>

## Getting Started

Don't forget to load packages to your library!

```{r library-packages}
library(dplyr)
library(neonUtilities)
library(ggplot2)
```

First we need to download the data we want to work with. For simplicity let's look at the most recent electrofishing bout available at the time this was written: Fall 2019.

```{r load-data }
electrofishing_fall2019 <- loadByProduct(dpID = "DP1.20107.001",
                                         startdate = "2019-07",
                                         enddate = "2019-12",
                                         package = "expanded",
                                         check.size = FALSE)
```

If you want, you can omit the `check.size = FALSE` argument. If you do, 
`loadByProduct` will check the size of the data you want to download and ask your 
permission to proceed.

Let's take a look at our data!
```{r structure}
class(electrofishing_fall2019)
names(electrofishing_fall2019)
```

The `electrofishing_fall2019` variable is a list of 8 tables. We can reference 
these by using the `$` operator (eg. `electrofishing_fall2019$fsh_bulkCount`) to 
look at each table, but this is cumbersome. To save ourselves some typing, we can 
list each of the tables to our global environment so we can call them directly. R 
has a built-in function for this: `list2env()`.

```{r list-to-environment}
list2env(electrofishing_fall2019, .GlobalEnv)
```

Your environment panel should now contain an entry for each table in 
`electrofishing_fall2019`. Now we can call `fsh_bulkCount`, etc. directly.

## Counting Species

Let's start by getting a handle on what we're working with. The `fsh_perFish` 
table lists all of the fish NEON collected individual data for. If there are too 
many individuals of a single taxon for NEON technicians to measure efficiently, 
the first 50 individuals are processed, and the rest are counted in bulk. This 
information is included in the fsh_bulkCount table.

When dealing with a new dataset, it can be useful to `View()` the variables table 
(named `variables_20107` for this data product.) RStudio will open this in a new 
tab.

Since we're just counting unique species for now, we can just use the 
`fsh_perFish` table to get what we need. We can select the column we want, 
`scientificName`, with the `select()` function. Let's see how many individual 
fish measurements NEON made in our Fall 2019 bout:

```{r individual-observations}
fsh_perFish %>%
    select(scientificName) %>%
    summarise(n())            # summarise() is used to provide summary statistics
                              # n() as an argument means we just want a count
```

That's a lot of fish! Let's narrow it down with the `distinct()` which will 
eliminate any repeats.

```{r total-species-1}
fsh_perFish %>%
     select(scientificName) %>%
     distinct(scientificName) %>%
     summarise(n())
```

If you wish, you can view the list of species by replacing `summarise(n())` with `View()`.

It is important to keep in mind that many of these species are difficult to 
identify in the field. Information regarding uncertainty of the ID is recorded in 
the `identificationQualifier` variable. If an individual couldn't be IDed to 
species level, the most specific taxon possible was recorded. The variable 
`taxonRank` contains this information.

Let's see how many species we have that were IDed to species level with no qualifiers regarding that ID. We can do this by piping our data through two `filter()` functions.

```{r total-species-2}
fsh_perFish %>%
        filter(is.na(identificationQualifier)) %>%     # make sure identificationQualifier is 'NA'
        filter(taxonRank == "species") %>%             # NEVER USE `variable == NA` for this
        select(scientificName) %>%                     # `is.na(variable)` is much more reliable
        distinct(scientificName) %>%
        summarise(n())       # NOTE: filters on these columns must be run before
                             # select() removes those columns from our data
```

## Grouping by Site and Graphing

Now let's see how many species were found at each site. We'll run the same code 
as before and store it to a variable, but this time we'll include the `siteID` 
variable and run `group_by(siteID)` so the `summarise()` function will find a 
count for each site individually. 

```{r richness-by-site}
richness_by_site <- fsh_perFish %>%
        filter(is.na(identificationQualifier)) %>%
        filter(taxonRank == "species") %>%
        select(siteID, scientificName) %>%
        group_by(siteID) %>%                # groups by site
        distinct(scientificName) %>%
        summarise(num_species = n()) %>%
        arrange(-num_species)               # puts rows in decreasing order
```

Now we can make a graph with ggplot!

```{r richness-by-site-graph}
ggplot(richness_by_site, aes(x = siteID, y = num_species)) +
        geom_bar(stat = "identity") +
        scale_x_discrete(limits = richness_by_site$siteID) +
        theme(axis.text.x = element_text(angle = 90)) +
        scale_y_continuous(breaks = seq(from = 0, to = 30, by = 2)) +
        labs(
                title = "Total Species Richness by Site",
                x = "Site ID",
                y = "Number of Species"
        )
```

Looks like <a href="https://www.neonscience.org/field-sites/field-sites-map/MAYF">Mayfield Creek</a> in Domain 8 has far and away the
highest species richness, but what about other measures of diversity? For that 
we'll need an idea of population abundance. Since NEON uses a three-pass 
electrofishing method, we will need to do a bit of processing to get the 
quantities we need.