---
title: "Teakettle Canopy Height Analysis"
author: "Naupaka Zimmerman"
date: "June 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyis of Teakettle LiDAR-derived Canopy Height

Out goals are to:

1. Import a raster — A lidar canopy height model (lidar/Teak_lidarCHM.tif)
1. For the CHM, set values == 0 to NA (not trees)
1. Classify the raster according to some distribution – low medium and tall trees. This could be done using a histogram potentially or we could just decide that <2m is generally grasses / understory, <6m small trees,and the rest are tall trees. Visualize histogram/density and plot vertical cutoff lines.
1. Plot the reclassified raster and add a legend for each “class”. 
1. Export the plot figure to a pdf in the outputs folder – publishable
1. Export the classified raster as a geotiff with NaFlagg = -9999 to an outputs folder.


```{r code}
# load library
library("raster")

# set parameters
chm_file <- "../NEONdata/D17-California/TEAK/2013/lidar/TEAK_lidarCHM.tif"
chosen.breaks <- c(6, 30, 50, 100)

# Load raster into memory
chm <- raster(chm_file)

# quick sanity check
plot(chm)

# Fix 0 values for R
chm[chm == 0] <- NA
plot(chm)

##########################################
##########################################
# Function to reclassify raster values

create_height_class_matrix <- function(breaks){
	# first create a matrix of values that represent the classification ranges
	# Lowest height class = 1
	
	# Get length of bins vector to figure out number of classes
	br.len <- length(breaks)
	
	# initialize height class vector
	height.class.m <- c(0)
	
	for (i in 1:br.len) {
		height.class.m <- c(height.class.m, breaks[i - 1], breaks[i], i)
	}
	
	# for input of breaks = c(6, 30, 50, 100) we want to make something like this:
	# height.class.m <- c(0, 6, 1, 
	# 					6, 30, 2, 
	# 					30, 50, 3, 
	# 					50, 100, 4)

	
	# reshape the object into a matrix with columns and rows
	reclass.height.mat <- matrix(height.class.m, 
                ncol = 3, 
                byrow = TRUE)

	reclass.height.mat
}

reclass.height.mat <- create_height_class_matrix(chosen.breaks)



# plot density of chm heights with cutoff lines
plot_chm_density <- function(rast.in, title, bins) {
	
	density(rast.in, main = title, xlab = "Height (m)")

	# cutoffs for bins from the reclassification matrix (second column)
	sapply(bins, function(x) abline(v = x, col = "red"))

}

plot_chm_density(chm, "Canopy heights at Teakettle site\n with bin cutoffs in red", chosen.breaks)



##########################################
##########################################
# Write density plot out to pdf

pdf(file = "output/chm_density_plot.pdf", width = 6, height = 5)

plot_chm_density(chm, "Canopy heights at Teakettle site\n with bin cutoffs in red", chosen.breaks)

dev.off()




# reclassify the raster using the reclass object - reclass.height.mat
reclassified.chm <- reclassify(chm, 
                     reclass.height.mat)



# function to plot the reclassified raster

plot_reclassified_raster <- function(rast.in, site.name, breaks){
	# this is a tricky bit because we need to out the legend
	# outside of the plot region
	
	# Get colors for plotting
	bin.colors <- rev(terrain.colors(length(breaks)))
	
	# make room for a legend
	 
	par(xpd = FALSE, mar = c(5.1, 4.1, 4.1, 4.5))
	
	# plot
	plot(rast.in,
	     col = bin.colors,
	     main = paste("Canopy height classes \n", site.name),
	     legend = FALSE)
	
	# allow legend to plot outside of bounds
	par(xpd = TRUE)
	
	# legend x
	leg.x <- par()$usr[2] + 20
	
	# legend y
	leg.y <- par()$usr[4] + 50 - (abs(par()$usr[3] - par()$usr[4]) / 2) 
	
	# create legend text
	height.mat <- create_height_class_matrix(breaks)
	
	# initialize legend text
	legend.text <- c()
	
	for (i in 1:length(breaks)) {
		
		legend.text <- c(legend.text, 
						paste0(height.mat[i, 1], "-", 
							   height.mat[i, 2], " m"))
	}
	
	
	# create the legend
	legend(leg.x, leg.y,  # set x,y legend location
	       legend = legend.text,  # make sure the order matches colors
	       fill = bin.colors,
	       bty = "n") # turn off border

	# turn off plotting outside of bounds
	par(xpd = FALSE)
}

plot_reclassified_raster(reclassified.chm, "Teakettle", breaks = chosen.breaks)



##########################################
##########################################
# Write out to pdf

pdf(file = "output/reclassified_chm.pdf", height = 5, width = 6)

plot_reclassified_raster(reclassified.chm, "Teakettle", breaks = chosen.breaks)

dev.off()

##########################################
##########################################


# export geotiff
writeRaster(reclassified.chm,
            filename = "output/reclassified_lidar_CHM.tif",
            format = "GTiff",
            options = "COMPRESS=LZW",
            overwrite = TRUE,
            NAflag = -9999)


```
