---
title: Content for NEON hydrology discharge tutorial
output: html_fragment
dateCreated: '2026-03-24'
---

::: {#ds-objectives markdown="1"}
## Learning Objectives

After completing this activity, you will be able to:

-   Download and explore the contents of NEON hydrologic data products
    from the observational and instrumented subsystems.
-   Navigate to tools and data sources for more derived data products
    such as geospatial data, bundled eddy-covariance data, or data from
    the airborne observation platform.
-   Understand the similarities and linkages between different NEON data
    products.
-   Join and plot hydrologic data sets from the instrumented subsystem
    across a terrestrial-aquatic gradient at NEON co-located sites.

You can follow either the R or Python code throughout this tutorial.

-   For R users, we recommend using R version 4+ and RStudio.
-   For Python users, we recommend using Python 3.9+.

```{r setup, include=FALSE}
library(arrow)
library(reticulate)
if(dir.exists("C:/Users/nickerson/")){
  # Windows-specific Python configuration for reticulate
  if (Sys.info()["sysname"] == "Windows") {
    # Method 1: Try explicit Python path
    python_exe <- "C:/Users/nickerson/AppData/Local/Programs/Python/Python312/python.exe"
    if (file.exists(python_exe)) {
      # Configure environment to help with DLL loading
      python_dir <- dirname(python_exe)
      # Set environment variables that can help with DLL issues
      Sys.setenv(
        PYTHONHOME = python_dir,
        PYTHONPATH = file.path(python_dir, "Lib"),
        PATH = paste(python_dir, file.path(python_dir, "Scripts"), Sys.getenv("PATH"), sep = ";")
      )
      # Use the Python executable  
      reticulate::use_python(python_exe, required = TRUE)
      message("✓ Configured Python: ", python_exe)
    } else {
      # Method 2: Try py launcher
      py_launcher <- Sys.which("py")
      if (py_launcher != "") {
        reticulate::use_python(py_launcher)
        message("✓ Using py launcher: ", py_launcher)
      } else {
        warning("Python not found. Install Python 3.9+ and ensure it's in PATH")
      }
    }
  } else {
    message("Non-Windows system - using automatic Python detection")
  }
  # Show final configuration
  message("\n=== Python Configuration ===")
  tryCatch({
    config <- py_config()
    message("Python: ", config$python)
    message("Version: ", config$version)
  }, error = function(e) {
    message("Configuration check failed: ", e$message)
  })
}
reticulate::py_config()
# reticulate::py_install(c("neonutilities"))
# py_run_string("import neonutilities; print(neonutilities.__version__)")
# reticulate::py_discover_config()
# reticulate::py_module_available("neonutilities")
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

## Set up: Install Packages {.tabset}

Packages only need to be installed once, you can skip this step after
the first time:

### R

-   **neonUtilities**: Basic functions for accessing NEON data
-   **tidyverse**: Collection of R packages designed for data science
-   **geosphere**: Compute distances between latitude/longitude coordinates
-   **plotly**: Functions for producing interactive plots

```{r install, eval=FALSE}
install.packages("neonUtilities")
install.packages("tidyverse")
install.packages("geosphere")
install.packages("plotly")
```

### Python

-   **os**: Module allowing interaction with user's operating system
-   **pandas**: Module for working with data frames
-   **neonutilities**: Basic functions for accessing NEON data
-   **matplotlib**: Functions for plotting
-   **geopy**: Compute distances between latitude/longitude coordinates
-   **plotly**: Functions for producing interactive plots
-   **statsmodels**: Functions for the estimation of statistical models

```{python p-install, eval=FALSE}
pip install os
pip install pandas
pip install neonutilities
pip install matplotlib
pip install geopy
pip install plotly
pip install statsmodels
```

## Additional Resources

-   <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Tutorial
    for using neonUtilities from both R and Python environments.</a>
-   <a href="https://github.com/NEONScience/NEON-Utilities/neonUtilities" target="_blank">GitHub
    repository for neonUtilities</a>
-   <a href="https://www.neonscience.org/sites/default/files/cheat-sheet-neonUtilities_0.pdf" target="_blank">neonUtilities
    cheat sheet</a>. A quick reference guide for users.
-   <a href="https://www.hydroshare.org/resource/53718072f33646fa920f1b72d7b403eb/" target="_blank">HydroShare
    Collection - NEON Hydrologic Data Products: Site-Level
    Resources</a>.
-   <a href="https://www.youtube.com/watch?v=kJYWra3J6RU&feature=youtu.be" target="_blank">CUAHSI
    Cyberseminar Series: Introduction to NEON for Hydrology - An
    Overview of Hydrologic Data Products and Tools</a>. :::

## Set up: Load Packages {.tabset}

### R

```{r R-library, results="hide", message=FALSE, warning=FALSE}
library(neonUtilities)
library(tidyverse)
library(geosphere)
library(plotly)
```

### Python

```{python p-import, results="hide", message=FALSE, warning=FALSE}
import os
import pandas as pd
import neonutilities as nu
import matplotlib.pyplot as plt
import matplotlib.dates as mdates
from geopy.distance import geodesic
import plotly.graph_objects as go
import statsmodels.api as sm
import numpy as np
```

## Set NEON Data Portal API Token {.tabset}

It is recommended that NEON data users have a NEON Data Portal API token
set as an environment variable. See
<a href="https://www.neonscience.org/resources/learning-hub/tutorials/neon-api-tokens-tutorial" target="_blank">this
tutorial.</a> for instructions on obtaining a NEON API token.

### R

```{r R-envvar, results="hide", message=FALSE, warning=FALSE}
Sys.setenv(NEON_PAT="YOUR_API_TOKEN_HERE")
```

### Python

```{python p-envvar, results="hide", message=FALSE, warning=FALSE}
os.environ.setdefault('NEON_PAT',"YOUR_API_TOKEN_HERE")
```

## Download & Explore: Introduction

In this tutorial, we will focus on one pair of co-located NEON sites
from Domain 07 - Appalachians & Cumberland Plateau:

-   <a href="https://www.neonscience.org/field-sites/ornl" target="_blank">Oak
    Ridge National Laboratory (ORNL) - Terrestrial</a>
-   <a href="https://www.neonscience.org/field-sites/walk" target="_blank">Walker
    Branch (WALK) - Aquatic</a>

But, the workflow can be replicated for any pair of co-located sites
across the observatory that contain all the data products used in this
tutorial, which is defined by the following criteria:

-   Aquatic site has published discharge data and has groundwater wells
    installed.
-   Terrestrial site has published precipitation data.
    - *Note*: Precipitation data are available from two different data products
    depending on the collection method at a site. Check the following data
    products to ensure you are downloading the correct data product for a site:
        - <a href="https://data.neonscience.org/data-products/DP1.00044.001" target="_blank">Precipitation - weighing gauge (DP1.00044.001)</a>
        - <a href="https://data.neonscience.org/data-products/DP1.00045.001" target="_blank">Precipitation - tipping bucket (DP1.00045.001)</a>
-   Terrestrial and aquatic site are within 10-km of each other.

The following site pairs meet that criteria:

| NEON Domain | Aquatic Site | Terrestrial Site |
|-------------|--------------|------------------|
| D02         | LEWI         | BLAN             |
| D02         | POSE         | SCBI             |
| D03         | FLNT         | JERC             |
| D06         | KING         | KONA             |
| D07         | WALK         | ORNL             |
| D08         | TOMB         | LENO             |
| D08         | BLWA         | DELA             |
| D08         | MAYF         | TALL             |
| D12         | BLDE         | YELL             |
| D13         | COMO         | NIWO             |
| D16         | MART         | WREF             |
| D17         | BIGC         | SOAP             |

In this tutorial, we will focus on one water year of data: Water Year 2024, 
defined as 2023-10-01 through 2024-09-30. And, for each data product used in
this tutorial, we will download data published in RELEASE-2026. To learn more 
about the differences between released and provisional data, see the 
<a href="https://www.neonscience.org/resources/learning-hub/tutorials/release-provisional-tutorial" target="_blank">Understanding Releases and Provisional Data</a> tutorial on the NEON website.

## Download & Explore: Instrumented Data Products {.tabset}

For this exercise, we will download and explore one data product from the
instrumented subsystem: <a href="https://data.neonscience.org/data-products/DP1.00045.001" target="_blank">Precipitation - tipping bucket (DP1.00045.001)</a>.
This data product is published from NEON's terrestrial sites; thus, we will use the 
NEON site code 'ORNL'.

### R

```{r R-download-is, results="hide", message=FALSE, warning=FALSE}
# Download precipitation data for a single water year
ptp_r <- neonUtilities::loadByProduct(dpID="DP1.00045.001",
                                      site="ORNL",
                                      startdate="2023-10",
                                      enddate="2024-09",
                                      release="RELEASE-2026",
                                      package='expanded',
                                      check.size = F,
                                      token=Sys.getenv("NEON_PAT"))
```

### Python

```{python p-download-is, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# Download precipitation data for a single water year
ptp_py = nu.load_by_product(dpid="DP1.00045.001",
                            site="ORNL",
                            startdate="2023-10",
                            enddate="2024-09",
                            release="RELEASE-2026",
                            package="expanded",
                            check_size=False,
                            token=os.environ.get("NEON_PAT"))
```


Downloads from the NEON Utilities packages contain multiple files, including 
data tables, metadata, and data product documentation. Let's explore each set of
files in turn.

### Files Associated with Downloads {.tabset}

The data we've downloaded comes as an object that is a named list/dictionary of 
objects. Let's view the contents of the download package.

#### R

```{r R-files-is, results="hide", message=FALSE, warning=FALSE}
# Get all file names in the download package
names(ptp_r)
```

#### Python

```{python p-files-is, results="hide", message=FALSE, warning=FALSE}
# Get all file names in the download package
ptp_py.keys()
```


In this tutorial, we downloaded the `expanded` download package. What are the 
files contained in this download package and why are they useful?

-   `TIPPRE_1min` and `TIPPRE_30min`: Includes the primary data tables of 
    the Precipitation - tipping bucket data product. We will dive deeper into 
    data tables in the next section.
-   `sensor_positions_00045`: Reports the geolocation of each sensor included
    in the download.
-   `science_review_flags_00045`: Lists each science review flag (SRF) date 
    range, flag value, and justification applied to the data included in this 
    download.                                                                   #NH comment: justification applied to the data or the flag?
-   `issueLog_00045`: Reports issues that may impact data quality, or changes
    to a data product that affects one or more sites.
-   `variables_00045`: This file contains all the variables found in the data
    table(s) included in this download. This includes full definitions, units, 
    and other important information.
-   `readme_00130`: The readme file provides important information relevant to
    the data product and the specific instance of downloading the data.
-   `citation_00045_RELEASE-2026`: Formatted citation and DOI for the data
    included in this download.

### Explore Data Tables {.tabset}

The expanded download package for DP1.00045.001 contains two data tables
that report precipitation time series data, each reporting data at a different
temporal resolution:

-   `TIPPRE_1min`: Tipping bucket precipitation reported at a 1-min resolution
-   `TIPPRE_30min`: Tipping bucket precipitation averaged to a 30-min resolution

Below, we will explore the first few rows of `TIPPRE_30min`. Add to the code
below to also view other tables included in the expanded download package.

#### R

```{r R-tables-is, results="hide", message=FALSE, warning=FALSE}
# Print the first 5 records in the time series data
print("First 5 rows of TIPPRE_30min")
head(ptp_r$TIPPRE_30min)
```

#### Python

```{python p-tables-is, results="hide", message=FALSE, warning=FALSE}
# Print the first 5 records in the time series data
print("First 5 rows of TIPPRE_30min")
print(ptp_py['TIPPRE_30min'].head())
```

### Explore Variables {.tabset}

The `variables_00045` file provides insight into the structure of each data 
table and associated variables included in a download package. view the 
variables file and familiarize yourself with the different fields, data types, 
and units used in this data product.

#### R

```{r R-variables-is, results="hide", message=FALSE, warning=FALSE}
# View variables file to understand data table structure
View(ptp_r$variables_00045)
```

#### Python

```{python p-variables-is, results="hide", message=FALSE, warning=FALSE}
# View variables file to understand data table structure
print(ptp_py['variables_00045'])
```

## Download & Explore: Observational Data Products {.tabset}

Let us now explore a hydrologic data product from the observational subsystem.
For this exercise, we move to the surface component of the hydrologic cycle. 
We will download and explore hydrologic data derived from surface water 
grab samples: <a href="https://data.neonscience.org/data-products/DP1.20206.001" target="_blank">Stable isotopes in surface water (DP1.20206.001)</a>.
This data product is published from NEON's aquatic sites; thus, we will use the NEON
site code 'WALK'.

### R

```{r R-download-os, results="hide", message=FALSE, warning=FALSE}
# Download precipitation data for a single water year
asi_r <- neonUtilities::loadByProduct(dpID="DP1.20206.001",
                                      site="WALK",
                                      startdate="2023-10",
                                      enddate="2024-09",
                                      release="RELEASE-2026",
                                      package='expanded',
                                      check.size = F,
                                      token=Sys.getenv("NEON_PAT"))
```

### Python

```{python p-download-os, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# Download precipitation data for a single water year
asi_py = nu.load_by_product(dpid="DP1.20206.001",
                            site="WALK",
                            startdate="2023-10",
                            enddate="2024-09",
                            release="RELEASE-2026",
                            package="expanded",
                            check_size=False,
                            token=os.environ.get("NEON_PAT"))
```


Let's do the sample exploration of the download as we did for the instrumented
data product and see what is similar and different.

### Files Associated with Downloads {.tabset}

#### R

```{r R-files-os, results="hide", message=FALSE, warning=FALSE}
# Get all file names in the download package
names(asi_r)
```

#### Python

```{python p-files-os, results="hide", message=FALSE, warning=FALSE}
# Get all file names in the download package
asi_py.keys()
```

When we view the content of the observational data product download, we notice
similarities and differences relative to the instrumented data product. For example, both
data products include `citation`, `variables`, `issuelog`, and `readme`
files. What do we notice that is different?

-   The observational data product does not contain `science_review_flags` or
    `sensor_positions` files. Those files are specific to instrumented data 
    products.
-   Files specific to observational data products are included:
      - `categoricalCodes_20206`: Some variables in the data tables are 
      published as strings and constrained to a standardized list of values 
      (LOV). This file shows all the LOV options for variables published in this
      data product.
      - `validation_20206`: If any fields require validation prior to
      publication, those validation rules are reported in this table.
-   There are many more data tables published in this observational data product.
    Let's explore that in the next section.

### Explore Data Tables {.tabset}

For this sample-based observational data product, there are many more tables
published than the previous instrumented data product we explored. That is 
because data are collected and published at each point along the lifetime of a
sample, from collection to analysis. Let's break down the table structure for this
stable isotopes data product.

-   `asi_fieldSuperParent`: Field data associated with the 'superparent' water
    sample, which is a 4-L grab samples that, once subsampled, results in
    multiple observational data products, including this stable isotopes
    data product.
-   `asi_fieldData`: Field data associated with the stable isotopes subsample. 
-   `asi_externalLabH2OIsotopes`: Results of hydrogen-2 and oxygen-18 stable
    isotope ratio analysis in filtered surface water samples.
-   `asi_externalLabSummaryData`: Accuracy and precision data for the instrument
    used in the analysis of H2O stable isotopes.
-   `asi_POMExternalLabDataPerSample`: Results of carbon-13 and nitrogen-15
    stable isotope ratio analysis in particulate organic matter (POM) filtered
    out of surface water samples.
-   `asi_externalLabPOMSummaryData`: Accuracy and precision data for the 
    instrument used in the analysis of POM stable isotopes.

For this exercise, let's just explore the first few rows of 
`asi_externalLabH2OIsotopes`. Add to the code below to also view other tables 
included in the expanded download package.

#### R

```{r R-tables-os, results="hide", message=FALSE, warning=FALSE}
# Print the first 5 records in H2O stable isotope lab data
print("First 5 rows of asi_externalLabH2OIsotopes")
head(asi_r$asi_externalLabH2OIsotopes)
```

#### Python

```{python p-tables-os, results="hide", message=FALSE, warning=FALSE}
# Print the first 5 records in H2O stable isotope lab data
print("First 5 rows of asi_externalLabH2OIsotopes")
print(asi_py['asi_externalLabH2OIsotopes'].head())
```

### Explore Variables {.tabset}

#### R

```{r R-variables-os, results="hide", message=FALSE, warning=FALSE}
# View variables file to understand data table structure
View(asi_r$variables_20206)
```

#### Python

```{python p-variables-os, results="hide", message=FALSE, warning=FALSE}
# View variables file to understand data table structure
print(asi_py['variables_20206'])
```

## Download & Explore: Higher-Level Hydrologic Data Products

NEON data products are processed at progressive levels. The precipitation and 
stable isotopes data products are Level 1 data products, which is the lowest
level of data processing required for a NEON data product. Higher level
hydrologic data products exist that include additional processing in the form of
spatial and/or temporal interpolation, or the incorporation of algorithms or 
scientific theory to derive higher-order quantities. 

<a href="https://www.neonscience.org/data-samples/data-management/data-processing" target="_blank">More information on NEON data processing levels</a>.

For this exercise, we will introduce three high-level hydrologic data products
and show how to download them.

### Stream morphology maps {.tabset}

The <a href="https://data.neonscience.org/data-products/DP4.00131.001" target="_blank">Stream morphology maps (DP4.00131.001)</a> 
data product is a Level 4 aquatic data product published at all NEON stream 
sites. The data product includes many data tables with post-processed survey 
data and links to geospatial data and site maps stored in the cloud. Let's 
download the data product and fetch the geospatial data from the cloud.

#### R

```{r R-download-geo, results="hide", message=FALSE, warning=FALSE}
# Download stream morpholog data for the lifetime of a site
geo_r <- neonUtilities::loadByProduct(dpID="DP4.00131.001",
                                      site="WALK",
                                      release="RELEASE-2026",
                                      package='basic',
                                      check.size = F,
                                      token=Sys.getenv("NEON_PAT"))

# URL to download geospatial data from the cloud is stored in geo_surveySummary
# Get the URL for the most recent geomorphology survey
print(max(geo_r$geo_surveySummary$dataFilePath[
  geo_r$geo_surveySummary$surveyBoutTypeID=="geomorphology"
]))

# Copy and paste the URL to your browser to retrieve the data package
```

#### Python

```{python p-download-geo, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# Download stream morpholog data for the lifetime of a site
geo_py = nu.load_by_product(dpid="DP4.00131.001",
                            site="WALK",
                            release="RELEASE-2026",
                            package="basic",
                            check_size=False,
                            token=os.environ.get("NEON_PAT"))

# URL to download geospatial data from the cloud is stored in geo_surveySummary
# Get the URL for the most recent geomorphology survey
print(max(geo_py['geo_surveySummary']['dataFilePath'][
    geo_py['geo_surveySummary']['surveyBoutTypeID']=="geomorphology"
]))

# Copy and paste the URL to your browser to retrieve the data package
```

### Net Surface-Atmosphere Exchange (Eddy Covariance) {.tabset}

The net surface-atmosphere exchange data products are available for all 
terrestrial sites and are bundled together in a single Level 4 data product:
<a href="https://data.neonscience.org/data-products/DP4.00200.001" target="_blank">Bundled data products - eddy covariance (DP4.00200.001)</a>.
The data packages do not contain comma separated tabular data. Rather, the data
and metadata are stored as HDF5 files.

To download and view bundled eddy covariance data, you cannot use the standard
R-loadByProduct() or Python-load_by_product() function. You must use a 
combination of other functions available in the NEON Utilities package.

#### R

```{r R-download-sae, results="hide", message=FALSE, warning=FALSE}
# Download precipitation data for a single water year
neonUtilities::zipsByProduct(dpID="DP4.00200.001",
                             site="ORNL",
                             startdate="2023-10",
                             enddate="2024-09",
                             release="RELEASE-2026",
                             package='basic',
                             check.size = F,
                             token=Sys.getenv("NEON_PAT"))
# Stack the data download, parse to data frames and read into environment 
# defaults to stacking only L4 products
sae_r <- neonUtilities::stackEddy(filepath = "filesToStack00200")

# The data are stored by site name - print the header of the 'ORNL' table
head(sae_r$ORNL)
```

#### Python

```{python p-download-sae, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# Download precipitation data for a single water year
nu.zips_by_product(dpid="DP4.00200.001",
                   site="ORNL",
                   startdate="2023-10",
                   enddate="2024-09",
                   release="RELEASE-2026",
                   package="basic",
                   check_size=False,
                   token=os.environ.get("NEON_PAT"))
# Stack the data download, parse to data frames and read into environment 
# defaults to stacking only L4 products
# sae_py = nu.stack_eddy(filepath = "filesToStack00200")

# The data is stored by site name - print the header of the 'ORNL' table
# print(sae_py["ORNL"].head())
```

### Canopy Water Indices - Mosaic

The <a href="https://data.neonscience.org/data-products/DP3.30019.001" target="_blank">Canopy water indices - mosaic (DP3.30019.001)</a>
is a Level 3 (spatially-interpolated) data product published from the Airborne
Observation Platform (AOP) subsystem. 

Remote sensing data products are large, but NEON has developed many tools 
to aid users in downloading and interpreting AOP data. We will not download AOP 
data in this exercise. Rather, follow the links below for guides on downloading 
AOP data in R, Python, and Google Earth Engine (GEE).

-   <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data#download-remote-sensing-data-byfileaop-and-bytileaop" target="_blank">Download and Explore NEON Data</a>.
    -   Directly links to 'Download remote sensing data: byFileAOP() and byTileAOP()' section.
-   <a href="https://www.neonscience.org/resources/learning-hub/tutorials/intro-aop-data-google-earth-engine-gee-tutorial-series" target="_blank">Intro to AOP Data in Google Earth Engine (GEE) Tutorial Series</a>.
-   <a href="https://www.neonscience.org/resources/learning-hub/tutorials/aop-data-management-releases" target="_blank">Understanding AOP Data Releases and Best Practices for AOP Data Management</a>.

##

## Merge & Visualize: The Water Cycle at Co-Located Sites

In this exercise, we will merge together three hydrologic data products, each
from a different section of the water cycle at NEON co-located sites:

-   <a href="https://data.neonscience.org/data-products/DP1.00045.001" target="_blank">Precipitation - tipping bucket (DP1.00045.001)</a>.
-   <a href="https://data.neonscience.org/data-products/DP1.20100.001" target="_blank">Elevation of groundwater (DP1.20100.001)</a>.
-   <a href="https://data.neonscience.org/data-products/DP4.00130.001" target="_blank">Continuous discharge (DP4.00130.001)</a>.

We will download each data product, identify how each product can be related,
then merge the three data streams into a single data frame.

### Download Data Products {.tabset}

This time, we will download the basic download packages.

#### R

```{r R-download-wc, results="hide", message=FALSE, warning=FALSE}
# Download precipitation data for a single water year
ptp_r <- neonUtilities::loadByProduct(dpID="DP1.00045.001",
                                      site="ORNL", # Terrestrial data product
                                      startdate="2023-10",
                                      enddate="2024-09",
                                      release="RELEASE-2026",
                                      package='basic',
                                      check.size = F,
                                      token=Sys.getenv("NEON_PAT"))

# Download groundwater elevation data for a single water year
egw_r <- neonUtilities::loadByProduct(dpID="DP1.20100.001",
                                      site="WALK", # Aquatic data product
                                      startdate="2023-10",
                                      enddate="2024-09",
                                      release="RELEASE-2026",
                                      package='basic',
                                      check.size = F,
                                      token=Sys.getenv("NEON_PAT"))

# Download discharge data for a single water year
csd_r <- neonUtilities::loadByProduct(dpID="DP4.00130.001",
                                      site="WALK", # Aquatic data product
                                      startdate="2023-10",
                                      enddate="2024-09",
                                      release="RELEASE-2026",
                                      package='basic',
                                      check.size = F,
                                      token=Sys.getenv("NEON_PAT"))
```

#### Python

```{python p-download-wc, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# Download precipitation data for a single water year
ptp_py = nu.load_by_product(dpid="DP1.00045.001",
                             site="ORNL", # Terrestrial data product
                             startdate="2023-10",
                             enddate="2024-09",
                             release="RELEASE-2026",
                             package='basic',
                             check_size=False,
                             token=os.environ.get("NEON_PAT"))

# Download groundwater elevation data for a single water year
egw_py = nu.load_by_product(dpid="DP1.20100.001",
                             site="WALK", # Aquatic data product
                             startdate="2023-10",
                             enddate="2024-09",
                             release="RELEASE-2026",
                             package='basic',
                             check_size=False,
                             token=os.environ.get("NEON_PAT"))

# Download discharge data for a single water year
csd_py = nu.load_by_product(dpid="DP4.00130.001",
                             site="WALK", # Aquatic data product
                             startdate="2023-10",
                             enddate="2024-09",
                             release="RELEASE-2026",
                             package='basic',
                             check_size=False,
                             token=os.environ.get("NEON_PAT"))
```

### Identify Relational Data & Merge {.tabset}

Due to the standardized spatial and temporal designs of NEON data products, 
these three instrumented data products can be related and merged in a relatively
easy fashion.

**Temporal Relationships**

-   All three data products have the same temporal structure. They all have the
    columns `startDateTime` and `endDateTime` in the data tables, and the 
    columns are all formatted the same in published data: YYYY-MM-DD HH:MM:SS 
    (UTC).
-   All three data products are published at a similar temporal frequency. The 
    precipitation and groundwater elevation products are each published at a
    30-min resolution and the discharge product is published at a 15-min
    resolution.
-   Therefore, merging the three tables by one of the datetime columns will
    ensure the data will be temporally related.

**Spatial Relationships**

-   For this pair of co-located sites, the precipitation and discharge data 
    products are published at one location, but the groundwater elevation
    data product is published at multiple locations. 
-   For this exercise, we will use the groundwater well location that is closest
    to the discharge locations. This information can easily be parsed by
    comparing sensor location coordinates in the `sensor_positions` file
    included in each data download.

#### R

```{r R-merge-wc, results="hide", message=FALSE, warning=FALSE}
# In this download, there are 3 well locations that publish elevation
# There is only 1 location for discharge
# Use `geosphere` to identify which well location is closest to discharge
egw_coords <- egw_r$sensor_positions_20100%>%
  dplyr::distinct(locationReferenceLatitude,locationReferenceLongitude,
                  .keep_all = T)
csd_coords <- csd_r$sensor_positions_00130
dist <- geosphere::distHaversine(egw_coords[,c('locationReferenceLongitude',
                                               'locationReferenceLatitude')],
                                 csd_coords[,c('locationReferenceLongitude',
                                               'locationReferenceLatitude')])

# Which well is closest to the discahrge location (horizontal position - HOR)?
close_loc <- egw_coords$HOR.VER[which.min(dist)]

# Let's use only the data from the closest well (subset by HOR)
egw_df <- egw_r$EOG_30_min[
  egw_r$EOG_30_min$horizontalPosition== substr(close_loc,0,3),# First 3 digits = HOR
]

# Merge 3 data streams into a single data frame
# Keep the relevant data needed to plot timeseries and examine relationships
ptp_df <- ptp_r$TIPPRE_30min%>%
  dplyr::select(endDateTime,precipBulk,finalQF)
egw_df <- egw_df%>%
  dplyr::select(endDateTime,groundwaterElevMean,gWatElevFinalQF)
csd_df <- csd_r$csd_15_min%>%
  dplyr::select(endDateTime,dischargeContinuous,dischargeFinalQF)

wc_df <- dplyr::full_join(ptp_df,egw_df)
wc_df <- dplyr::full_join(wc_df,csd_df)
wc_df <- wc_df[order(wc_df$endDateTime),]
```

#### Python

```{python p-merge-wc, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# In this download, there are 3 well locations that publish elevation
# There is only 1 location for discharge
# Use `geopy` to identify which well location is closest to discharge
from geopy.distance import geodesic
egw_coords = egw_py['sensor_positions_20100'].drop_duplicates(subset=['locationReferenceLatitude', 'locationReferenceLongitude'])
csd_coords = csd_py['sensor_positions_00130']
dist = egw_coords.apply(lambda row: geodesic((row['locationReferenceLatitude'], row['locationReferenceLongitude']), 
                                              (csd_coords.iloc[0]['locationReferenceLatitude'], csd_coords.iloc[0]['locationReferenceLongitude'])).meters, axis=1)

# Which well is closest to the discharge location (horizontal position - HOR)?
close_loc = egw_coords.loc[dist.idxmin(), 'HOR.VER']

# Let's use only the data from the closest well (subset by HOR)
egw_df = egw_py['EOG_30_min'][egw_py['EOG_30_min']['horizontalPosition'] == close_loc[:3]]  # First 3 digits = HOR

# Merge 3 data streams into a single data frame
# Keep the relevant data needed to plot timeseries and examine relationships
ptp_df = ptp_py['TIPPRE_30min'][['endDateTime', 'precipBulk', 'finalQF']]
egw_df = egw_df[['endDateTime', 'groundwaterElevMean', 'gWatElevFinalQF']]
csd_df = csd_py['csd_15_min'][['endDateTime', 'dischargeContinuous', 'dischargeFinalQF']]

wc_df = pd.merge(ptp_df, egw_df, on='endDateTime', how='outer')
wc_df = pd.merge(wc_df, csd_df, on='endDateTime', how='outer')
wc_df = wc_df.sort_values('endDateTime')
```

### Plot & Download Merged Interactive Timeseries {.tabset}

Now, let's plot the three data streams in a single plotting field. We will use 
the `plotly` package to give us the ability to interact with the plot. Check 
your current working directory for the HTML file containing the plot.

#### R

```{r R-plot-wc, results="hide", message=FALSE, warning=FALSE}
# Format each y-axis
y1 <- list(side='left',
           automargin=T,
           title="Discharge (L s-1)",
           tickfont=list(size=16),
           titlefont=list(size=18),
           showgrid=F,
           zeroline=F)
y2 <- list(side='right',
           overlaying="y",
           automargin=T,
           title="Groundwater Elevation (m)",
           tickfont=list(size=16,color = '#CC79A7'),
           titlefont=list(size=18,color = '#CC79A7'),
           showgrid=F,
           zeroline=F)
y3 <- list(side='right',
           overlaying="y",
           automargin=T,
           title="Precipitation (mm)",
           tickfont=list(size=16,color = "#0072B2"),
           titlefont=list(size=18,color = "#0072B2"),
           showgrid=F,
           zeroline=F,
           anchor="free",
           position=0.98)

# Build plot layout
ts <- plotly::plot_ly(data=wc_df)%>%
  plotly::layout(
    yaxis = y1, yaxis2 = y2, yaxis3 = y3,
    xaxis=list(domain=c(0,.9),
               tick=14,
               automargin=T,
               title="Date",
               tickfont=list(size=16),
               titlefont=list(size=18)),
    legend=list(orientation = "h",
                y=-0.15,
                font=list(size=14)),
    updatemenus=list(
      list(
        type='buttons',
        showactive=FALSE,
        buttons=base::list(
          list(label='Scale Discharge\n- Linear -',
               method='relayout',
               args=list(list(yaxis=list(type='linear',
                                         title="Discharge (L s-1)",
                                         tickfont=list(size=16),
                                         titlefont=list(size=18),
                                         showgrid=F,
                                         zeroline=F)))),
          list(label='Scale Discharge\n- Log -',
               method='relayout',
               args=list(list(yaxis=list(type='log',
                                         title="Discharge (L s-1) - log",
                                         tickfont=list(size=16),
                                         titlefont=list(size=18),
                                         showgrid=F,
                                         zeroline=F))))))))

# Plot traces
ts <- ts%>%
  # H and Q Series
  plotly::add_trace(x=~endDateTime,y=~dischargeContinuous, 
                    name="Discharge",type='scatter',mode='line',
                    line = list(color = "black"))%>%
  plotly::add_trace(x=~endDateTime,y=~groundwaterElevMean,
                    yaxis="y2", name="GW Elevation",type='scatter',mode='line',
                    line = list(color = '#CC79A7'))%>%
  plotly::add_trace(x=~endDateTime,y=~precipBulk, yaxis="y3",
                    name="Precipitation",type='scatter',mode='line',
                    line = list(color = '#0072B2'))

htmlwidgets::saveWidget(plotly::as_widget(ts),
                        "NEON.D07.P.H.Q.WY2024.html")
```

#### Python

```{python p-plot-wc, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# Create figure
wc_df_ts = wc_df.dropna(subset=['precipBulk'])
fig = go.Figure()

# Format axes and layout
fig.update_layout(
    margin=dict(l=90, r=110, t=40, b=60),
    
    xaxis=dict(
        domain=[0, 0.9],
        tickmode='auto',
        nticks=14,
        automargin=True,
        title="Date",
        tickfont=dict(size=16)
    ),

    yaxis=dict(
        title="Discharge (L s-1)",
        tickfont=dict(size=16),
        showgrid=False,
        zeroline=False
    ),

    yaxis2=dict(
        title="Groundwater Elevation (m)",
        tickfont=dict(size=16, color='#CC79A7'),
        showgrid=False,
        zeroline=False,
        overlaying='y',
        side='right'
    ),

    yaxis3=dict(
        title="Precipitation (mm)",
        tickfont=dict(size=16, color='#0072B2'),
        showgrid=False,
        zeroline=False,
        overlaying='y',
        side='right',
        anchor='free',
        position=0.98
    ),

    legend=dict(
        orientation="h",
        y=-0.15,
        font=dict(size=14)
    ),

    updatemenus=[
        dict(
            type='buttons',
            showactive=False,
            buttons=[
                dict(
                    label='Scale Discharge\n- Linear -',
                    method='relayout',
                    args=[{
                        'yaxis.type': 'linear',
                        'yaxis.title': "Discharge (L s-1)"
                    }]
                ),
                dict(
                    label='Scale Discharge\n- Log -',
                    method='relayout',
                    args=[{
                        'yaxis.type': 'log',
                        'yaxis.title': "Discharge (L s-1) - log"
                    }]
                )
            ]
        )
    ]
)

# Add traces
fig.add_trace(go.Scatter(
    x=wc_df_ts['endDateTime'],
    y=wc_df_ts['dischargeContinuous'],
    mode='lines',
    name='Discharge',
    line=dict(color='black')
))

fig.add_trace(go.Scatter(
    x=wc_df_ts['endDateTime'],
    y=wc_df_ts['groundwaterElevMean'],
    mode='lines',
    name='GW Elevation',
    line=dict(color='#CC79A7'),
    yaxis='y2'
))

fig.add_trace(go.Scatter(
    x=wc_df_ts['endDateTime'],
    y=wc_df_ts['precipBulk'],
    mode='lines',
    name='Precipitation',
    line=dict(color='#0072B2'),
    yaxis='y3'
))

fig.write_html("NEON.D07.P.H.Q.WY2024.html")
```

### Further Exploration: Cumulative Precipitation & Discharge {.tabset}

#### R

```{r R-plot-wc-2, results="hide", message=FALSE, warning=FALSE}
# Plot cumulative precipitation & discharge together using ggplot with 2 y-axes
wc_df_subset <- wc_df%>%
  filter(!is.na(precipBulk))
wc_df_subset$cumulativeP <- cumsum(wc_df_subset$precipBulk)
wc_df_subset$cumulativeQ <- cumsum(wc_df_subset$dischargeContinuous)
cumsum <- wc_df_subset%>%
  ggplot(aes(x = endDateTime)) +
  geom_smooth(aes(y = cumulativeP), method="loess", color = "#0072B2") +
  geom_smooth(aes(y = cumulativeQ/150), method="loess", color = "black") +
  scale_y_continuous(
    name = "Cumulative Precipitation (mm)",
    sec.axis = sec_axis(~ .*150, name = "Cumulative Discharge (L s-1)")
  ) +
  labs(x = "Date") +
  theme_minimal() +
  theme(
    axis.title.y.left = element_text(color = "#0072B2", size = 14),
    axis.title.y.right = element_text(color = "black", size = 14)
  )
cumsum
```

#### Python

```{python p-plot-wc-2, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# Plot cumulative precipitation & discharge together using ggplot with 2 y-axes
wc_df_subset = wc_df[~wc_df['precipBulk'].isna()].copy()
wc_df_subset['cumulativeP'] = wc_df_subset['precipBulk'].cumsum()
wc_df_subset['cumulativeQ'] = wc_df_subset['dischargeContinuous'].cumsum()
x = wc_df_subset['endDateTime']
x_numeric = x.astype(np.int64)  # nanoseconds since epoch
lowess = sm.nonparametric.lowess
smoothP = lowess(wc_df_subset['cumulativeP'], x_numeric, frac=0.3)
smoothQ = lowess(wc_df_subset['cumulativeQ'] / 150, x_numeric, frac=0.3)
fig, ax1 = plt.subplots(figsize=(10, 6))
ax1.plot(x, smoothP[:,1], color='#0072B2', label='Cumulative Precipitation (mm) – LOESS')
ax1.set_xlabel('Date', fontsize=12)
ax1.set_ylabel('Cumulative Precipitation (mm)', color='#0072B2', fontsize=12)
ax1.tick_params(axis='y', labelcolor='#0072B2')
ax2 = ax1.twinx()
ax2.plot(x, smoothQ[:,1], color='black', label='Cumulative Discharge (L s-1) – LOESS')
ax2.set_ylabel('Cumulative Discharge (L s-1)', color='black', fontsize=12)
ax2.tick_params(axis='y', labelcolor='black')
fig.tight_layout()
plt.show()
```

### Further Exploration: Correlation of Groundwater Elevation & Discharge {.tabset}

#### R

```{r R-plot-wc-3, results="hide", message=FALSE, warning=FALSE}
# Plot scatterplots of one variable to another to assess correlation
# Create a continuous color scale by date to add the time-of-year dimension
corr <- wc_df %>%
  ggplot(aes(x = groundwaterElevMean, y = dischargeContinuous, color = as.integer(endDateTime))) +
  geom_point(aes(color = as.Date(endDateTime))) +
  scale_color_date(low="blue",high="darkorange") +
  labs(x = "Groundwater Elevation (m)", y = "Discharge (L s-1)", color = "Date") +
  theme_minimal()
corr
```

#### Python

```{python p-plot-wc-3, eval=TRUE, results="hide", message=FALSE, warning=FALSE}
# Plot scatterplots of one variable to another to assess correlation
# Create a continuous color scale by date to add the time-of-year dimension
dates = pd.to_datetime(wc_df['endDateTime'])
date_nums = mdates.date2num(dates)
fig, ax = plt.subplots(figsize=(10, 6))
scatter = ax.scatter(
    wc_df['groundwaterElevMean'],
    wc_df['dischargeContinuous'],
    c=date_nums,
    cmap='cool',
    alpha=0.6
)
ax.set_xlabel('Groundwater Elevation (m)', fontsize=12)
ax.set_ylabel('Discharge (L s-1)', fontsize=12)
cbar = plt.colorbar(scatter, ax=ax)
cbar.set_label('Date', fontsize=12)
tick_locs = cbar.get_ticks()
cbar.ax.set_yticklabels([mdates.num2date(t).strftime('%Y-%m-%d') for t in tick_locs])
plt.tight_layout()
plt.show()
```


:::