---
title: "Teakettle Canopy Height Analysis"
author: "Naupaka Zimmerman"
date: "June 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyis of Teakettle LiDAR-derived Canopy Height

Out goals are to:

1. Import a raster — A lidar canopy height model (lidar/Teak_lidarCHM.tif)
1. For the CHM, set values == 0 to NA (not trees)
1. Classify the raster according to some distribution – low medium and tall trees. This could be done using a histogram potentially or we could just decide that <2m is generally grasses / understory, <6m small trees,and the rest are tall trees. Visualize histogram/density and plot vertical cutoff lines.
1. Plot the reclassified raster and add a legend for each “class”. 
1. Export the plot figure to a pdf in the outputs folder – publishable
1. Export the classified raster as a geotiff with NaFlagg = -9999 to an outputs folder.


```{r initialize-set-params}
# load library
library("raster")

# load functions
source("scripts/functions.R")

# set parameters
chm_file <- "../NEONdata/D17-California/TEAK/2013/lidar/TEAK_lidarCHM.tif"
chosen.breaks <- c(6, 30, 50, 100)
site_name <- "Teakettle"
site_file_code <- "TEAK"
output_dir_path <- "../outputs/tuesday-pm/"
```

```{r create-output-dir}
# create the output directory we want to use if it doesn't yet exist
ifelse(!dir.exists(file.path(output_dir_path)), 
	   dir.create(file.path(output_dir_path), showWarnings = TRUE))
```

```{r quick-sanity-plots}
# Load raster into memory
chm <- raster(chm_file)

# quick sanity check
plot(chm)

# Fix 0 values for R
chm[chm == 0] <- NA
plot(chm)
```


```{r density-plot-with-breaks}
# create the matrix to reclassify our raster
reclass.height.mat <- create_height_class_matrix(chosen.breaks)

# plot the density of values with the selected breaks
plot_chm_density(chm, site_name, chosen.breaks)

# Write density plot out to pdf
make_pdf(plot_chm_density(chm, site_name, chosen.breaks),
		 file = paste0(output_dir_path, site_file_code, "chm_density_plot.pdf"), width = 6, height = 5)

```


```{r reclassify-plot-and-save}
# reclassify the raster using the reclass object - reclass.height.mat
reclassified.chm <- reclassify(chm, 
                     reclass.height.mat)

# plot to make sure it looks ok
plot_reclassified_raster(reclassified.chm, site_name, breaks = chosen.breaks)

# Write out to pdf
make_pdf(plot_reclassified_raster(reclassified.chm, site_name, breaks = chosen.breaks),
	file = paste0(output_dir_path, site_file_code, "reclassified_chm.pdf"), height = 5, width = 6)

# export geotiff
writeRaster(reclassified.chm,
            filename = paste0(output_dir_path, site_file_code, "reclassified_lidar_CHM.tif"),
            format = "GTiff",
            options = "COMPRESS=LZW",
            overwrite = TRUE,
            NAflag = -9999)

```
