---
title: "Teakettle Canopy Height Analysis"
author: "Naupaka Zimmerman"
date: "June 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyis of Teakettle LiDAR-derived Canopy Height

Out goals are to:

1. Import a raster — A lidar canopy height model (lidar/Teak_lidarCHM.tif)
1. For the CHM, set values == 0 to NA (not trees)
1. Classify the raster according to some distribution – low medium and tall trees. This could be done using a histogram potentially or we could just decide that <2m is generally grasses / understory, <6m small trees,and the rest are tall trees. Visualize histogram/density and plot vertical cutoff lines
1. Plot the reclassified raster and add a legend for each “class”
1. Export the plot figure to a pdf in the outputs folder
1. Export the classified raster as a geotiff with NaFlagg = -9999 to an outputs folder


```{r code}
# load library
library("raster")

# load data
chm_file <- "../NEONdata/D17-California/TEAK/2013/lidar/TEAK_lidarCHM.tif"
chm <- raster(chm_file)

# quick sanity check
plot(chm)

# Fix 0 values for R
chm[chm == 0] <- NA
plot(chm)

##########################################
##########################################
# Reclassify raster values

# first create a matrix of values that represent the classification ranges
# Lowest height class = 1
height.class.m <- c(0, 6, 1,
					6, 30, 2,
					30, 50, 3,
					50, 100, 4)
height.class.m

# reshape the object into a matrix with columns and rows
reclass.height.mat <- matrix(height.class.m, 
                ncol = 3, 
                byrow = TRUE)

reclass.height.mat

# reclassify the raster using the reclass object - reclass.height.mat
reclassified.chm <- reclassify(chm, 
                     reclass.height.mat)

# plot density of chm heights with cutoff lines
density(chm, main = "Canopy heights at Teakettle site\n with bin cutoffs in red", xlab = "Height (m)")

# cutoffs for bins from the reclassification matrix (second column)
sapply(reclass.height.mat[, 2], function(x) abline(v = x, col = "red"))


##########################################
##########################################
# Write density plot out to pdf

pdf(file = "output/chm_density_plot.pdf", width = 6, height = 5)

# plot density of chm heights with cutoff lines
density(chm, main = "Canopy heights at Teakettle site\n with bin cutoffs in red", xlab = "Height (m)")

# cutoffs for bins from the reclassification matrix (second column)
sapply(reclass.height.mat[, 2], function(x) abline(v = x, col = "red"))

dev.off()





# this is a tricky bit because we need to out the legend
# outside of the plot region

# make room for a legend
# 
par(xpd = FALSE, mar = c(5.1, 4.1, 4.1, 4.5))

# plot
plot(reclassified.chm,
     col = c("yellow", "orange", "red", "blue"), # hard code colors
     main = "Canopy height classes \nLower Teakettle",
     legend = FALSE)

# allow legend to plot outside of bounds
par(xpd = TRUE)

# legend x
leg.x <- par()$usr[2] + 20

# legend y
leg.y <- par()$usr[4] + 50 - (abs(par()$usr[3] - par()$usr[4]) / 2) 

# create the legend
legend(leg.x, leg.y,  # set x,y legend location
       legend = c("0-6 m", "6-30 m", "30-50 m", "50-100 m"),  # make sure the order matches colors
       fill = c("yellow","orange", "red", "blue"),
       bty = "n") # turn off border


##########################################
##########################################
# Write out to pdf

pdf(file = "output/reclassified_chm.pdf", height = 5, width = 6)

# make room for a legend
par(xpd = FALSE, mar = c(5.1, 4.1, 4.1, 4.5))

# plot
plot(reclassified.chm,
     col = c("yellow", "orange", "red", "blue"), # hard code colors
     main = "Canopy height classes \nLower Teakettle",
     legend = FALSE)

# allow legend to plot outside of bounds
par(xpd = TRUE)

# legend x
leg.x <- par()$usr[2] + 20

# legend y
leg.y <- par()$usr[4] + 50 - (abs(par()$usr[3] - par()$usr[4]) / 2) 

# create the legend
legend(leg.x, leg.y,  # set x,y legend location
       legend = c("0-6 m", "6-30 m", "30-50 m", "50-100 m"),  # make sure the order matches colors
       fill = c("yellow","orange", "red", "blue"),
       bty = "n") # turn off border

dev.off()

##########################################
##########################################


# export geotiff
writeRaster(reclassified.chm,
            filename = "output/reclassified_lidar_CHM.tif",
            format = "GTiff",
            options = "COMPRESS=LZW",
            overwrite = TRUE,
            NAflag = -9999)


```
