---
layout: post
title: "Spatial Intro 08: Cleaning Data -- Missing and Bad Data Values"
date:   2015-10-23
authors: []
contributors: Alison Dernbach
dateCreated: 2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: rgdal, ggplot2, raster
category: [self-paced-tutorial]
topics: spatial-data-gis
languagesTool: R
mainTag: spatial-data-management-series
workshopSeries: [spatial-data-management-series]
description: "This tutorial covers spatial data cleaning - specifically dealing with missing
(NA / NAN) values and bad values when working with spatial data in R."
code1: /R/dc-spatio-temporal-intro/08-clean-data-bad-and-missing-values.R
image:
feature: NEONCarpentryHeader_2.png
credit: A collaboration between the National Ecological Observatory Network 
(NEON) and Data Carpentry
creditlink:
permalink: R/missing-bad-data
comments: true
---


## About
This tutorial covers spatial data cleaning - specifically dealing with missing
(NA / NAN) values and bad values when working with spatial data in `R`.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

#Goals / Objectives

After completing this activity, you will:

* Recognize different `NoData` values used
* Understand how to check for `NoData` values in spatial data
* Know how to check for bad data values

## Things You’ll Need To Complete This Lesson
To complete this lesson you will need the most current version of R, and
preferably, RStudio loaded on your computer.

### Install R Packages

* **rgdal:** `install.packages("rgdal")`
* **ggplot2:** `install.packages("ggplot2")`
* **raster:** `install.packages("raster")`

<a href="https://www.neonscience.org/packages-in-r" target="_blank"> More on 
Packages in R </a>– Adapted from Software Carpentry.

### Download Data
<h3> <a href="https://ndownloader.figshare.com/files/3701578"> NEON Teaching Data Subset: Airborne Remote Sensing Data </a></h3>

The LiDAR and imagery data used to create this raster teaching data subset 
were collected over the 
<a href="https://www.neonscience.org/" target="_blank"> National Ecological Observatory Network's</a> 
<a href="https://www.neonscience.org/field-sites/field-sites-map/HARV" target="_blank" >Harvard Forest</a>
and 
<a href="https://www.neonscience.org/field-sites/field-sites-map/SJER" target="_blank" >San Joaquin Experimental Range</a>
field sites and processed at NEON headquarters. 
The entire dataset can be accessed by request from the 
<a href="http://data.neonscience.org" target="_blank"> NEON Data Portal</a>.

<a href="https://ndownloader.figshare.com/files/3701578" class="link--button link--arrow"> Download Dataset</a>


****

**Set Working Directory:** This lesson assumes that you have set your working 
directory to the location of the downloaded and unzipped data subsets. 

<a href="https://www.neonscience.org/set-working-directory-r" target="_blank"> An overview
of setting the working directory in R can be found here.</a>

**R Script & Challenge Code:** NEON data lessons often contain challenges that reinforce 
learned skills. If available, the code for challenge solutions is found in the
downloadable R script of the entire lesson, available in the footer of each lesson page.


**Spatial-Temporal Data & Data Management Lesson Series:** This lesson is part
of a lesson series introducing 
#####UPDATE URL BEFORE PUBLISHING#####
<a href="https://www.neonscience.org/resources/series/NEEDS.URL" target="_blank"> *spatial data and data management in `R`*</a>.
It is also part of a larger
<a href="https://www.neonscience.org/sp-temp-data-usgs-20170412" target="_blank"> *Spatio-Temporal Data Carpentry Workshop* </a>
that includes working with
<a href="https://www.neonscience.org/resources/series/introduction-working-raster-data-r" target="_blank"> *raster data in `R`*</a>,
<a href="https://www.neonscience.org/vector-data-series" target="_blank"> *vector data in `R`*</a>, 
and<a href="https://www.neonscience.org/tabular-time-series" target="_blank"> *tabular time series in `R`*</a>.

****

### Additional Resources

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">
Read more about the `raster` package in `R`</a>
* <a href="https://www.neonscience.org/raster-data-r" target="_blank">
NEON Data Skills: *Raster Data in R - The Basics*</a>
* <a href="https://www.neonscience.org/image-raster-data-r" target="_blank">
NEON Data Skills: *Image Raster Data in R - An Intro*</a>

</div>

### Clean Data
No dataset is perfect. It is common to encounter large files containing obviously
erroneous data (bad data).  It is also common to encounter `NoData` values that 
we need to account for when analyzing our data.

## NoData Values (NA, NAN)
If we are lucky when working with external data, the `NoData` value is clearly 
specified in the metadata. Sometimes this value is `NA` (not available) or `NAN` 
(not a number). However, `NA` isn't always used. Text values can make data storage 
difficult for some programs and thus, sometimes you'll encounter a large negative 
value such as `-9999` used as the `NoData` value. At other times, we might see blank 
values in a data file which designate `NoData`. Blanks are particularly problematic 
because we can't be certain if a data value is purposefully missing (not measured 
that day or a bad measurement) or if someone unintentionally deleted it.

<i class="fa fa-star"></i> **Data Tip:**`-9999` is a common value used in
both the remote sensing field and the atmospheric fields. It is also
the standard used by the <a href="https://www.neonscience.org/" target="_blank">
National Ecological Observatory Network (NEON)</a>.

Because the actual value used to designate missing data can vary depending upon
what data we are working with, it is important to always check the metadata for
the files associated `NoData` value. If the value is `NA`, we are in luck, `R`
will recognize and flag this value as `NoData`. If the value is numeric (e.g.,
`-9999`),then we might need to assign this value to `NA`.

<i class="fa fa-star"></i> **Data Tip:** `NA` values will be ignored when
performing calculations in `R`. However a `NoData` value of `-9999` will be
recognized as an integer and processed accordingly. If you encounter a numeric
`NoData` value be sure to assign it to `NA` in `R`:
`objectName[objectName==-9999] <- NA`


### Check for NA values
We can quickly check for `NoData` values in our raster data using the`is.na()`
function. By asking for the `sum()` of `is.na()` we can see how many NA / missing
values we have in a dataframe.

```{r missing-values}
# load libraries
library(rgdal)
library(ggplot2)
library(raster)

# set working directory to ensure R can find the file we wish to import and where 
# we want to save our files. Be sure to move the unzipped download into your working directory! 
wd <- "~/Documents/data/" #This will depend on your local environment 
setwd(wd)

# create a data frame
RGB_HARV <- raster(paste0(wd,"NEON-DS-Airborne-Remote-Sensing/HARV/RGB_Imagery/HARV_Ortho_wNA.tif"))
RGB_HARV_df <- as.data.frame(RGB_HARV, xy = TRUE)

# check for NA values 
sum(is.na(RGB_HARV_df$HARV_Ortho_wNA))

```

The result above tells us there are `NoData` values in the data.

### Deal with NoData Values
When we encounter `NoData` values (blank, NAN, -9999, etc.) in our data we need 
to decide how to deal with them. By default `R` treats `NoData` values designated 
with a `NA` as a missing value rather than a zero. This is good, as a value of
zero (no rain today) is not the same as missing data (e.g. we didn't measure the
amount of rainfall today).

How we deal with `NoData` values will depend on:

* the data type we are working with
* the analysis we are conducting
* the significance of the gap or missing value

Let's visualize where the `NoData` values are in the `RGB_HARV` raster by highlighting them in pink.

``` {r raster-plot, fig.cap="Monochrome non-rectangular landscape with NA values highlighted in pink on all sides of the landscape to create a rectangular image."}
#plot the raster
ggplot() +
  geom_raster(data = RGB_HARV_df, aes(x = x, y = y, fill = HARV_Ortho_wNA)) +
  scale_fill_gradient(na.value = "pink") +
  ggtitle("HARV Orthographic Image")
```

The shape of a raster is always rectangular. If we have a dataset that isn't 
rectangular, the pixels at the edge of the raster will have `NoData` values. This 
can happen when the data were collected by an airplane that only flew over some 
part of a defined region. Highlighting `NoData` values in a different color helps 
identify where the missing data is.

Sometimes we might need to "gap fill" our data. This means we will interpolate
or estimate missing values often using statistical methods. Gap filling can be
complex and is beyond the scope of this lesson. The take away from this lesson
is simply that it is important to acknowledge missing values in your data and to
carefully consider how you wish to account for them during analysis.

Other resources:

* <a href="http://www.statmethods.net/input/missingdata.html" target="_blank">Quick-R: Missing Data </a>
-- `R` code for dealing with missing data
* The Institute for Digital Research and Education has an <a href="https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/" target="_blank"> `R` FAQ on Missing Values</a>

## Bad Data Values
Bad data values are different from `NoData` values. Bad data values are values that 
fall outside of the applicable range of a dataset.

Examples of bad data values:

* The normalized difference vegetation index (NDVI), which is a measure of
greenness, has a valid range of -1 to 1. Any value outside of that range would
be considered a "bad" or miscalculated value.
* If we are using a Julian day (0-365/366) to represent the days of the year. A value
of 1110 is clearly not correct.

### Find Bad Data Values
Sometimes a dataset's metadata will tell us the range of expected values for a 
variable or common sense dictates the expected value (as in the Julian day example 
above). Values outside of this range are suspect and we need to consider them 
when we analyze the data. Sometimes, we need to use some common sense and scientific 
insight as we examine the data - just as we would for field data to identify 
questionable values.

We can explore the distribution of values contained within our data using the 
`geom_histogram()` function. Histograms are often useful in identifying outliers 
and bad data values in our raster data. Additionally, use `summary()` to look at 
the min and max values and compare them with the expected values.

<i class="fa fa-star"></i> **Data Tip:** Remotely sensed images often have 
values from 0-255, representing a string of 8 binary numbers that correspond to brightness 
shades from white to black. Each pixel in an image is assigned one of these values. 
For more information, see <a href="https://earthobservatory.nasa.gov/features/RemoteSensing/remote_06.php" target="_blank">NASA's Earth Observatory page on Pixels and Bits</a> 
and an <a href="https://desktop.arcgis.com/en/arcmap/latest/manage-data/geodatabases/arcgis-field-data-types.htm" target="_blank">ArcGIS article on data types</a>.

```{r good-histogram, fig.cap="Histogram with a good bell curve and some values clustered at zero."}

# create and view histogram
ggplot() +
  geom_histogram(data = RGB_HARV_df, aes(HARV_Ortho_wNA), bins = 40)

# compare min and max values with the expected range
summary(RGB_HARV_df)

```

<i class="fa fa-star"></i> **Data Tip:** In `geom_histogram()`, 30 is the default 
number of bins for the data. We can define the number of bins by using the `bins` 
value. 

The shape of the histogram above looks reasonable and the min and max values are 
within the expected 0-255 range. There are likely no bad data values in this raster. 

As mentioned earlier in this tutorial, `NoData` values can also be represented as 
the integer -9999. Below, we will manually change the `NoData` values in `RGB_HARV` 
from `NA` to `-9999` and see how that changes the histogram and summary. 

``` {r bad-histogram, fig.cap="Histogram with most values in two columns between 0 and 255 but some values are in a column at -9999."}

# convert NA values to -9999
RGB_HARV[is.na(RGB_HARV)]=-9999

# update dataframe 
RGB_HARV_df <- as.data.frame(RGB_HARV, xy = TRUE)

# plot new histogram
ggplot() +
  geom_histogram(data = RGB_HARV_df, aes(HARV_Ortho_wNA), bins = 40)

# look at the new min and max values
summary(RGB_HARV_df)

```

This histogram contains bad data. The range of values should be 0-255 but the 
histogram and summary table show many values are `-9999`. 

### Deal with Bad Data Values
Sometimes bad data values are actually `NoData` values, like second histogram above. 
To correct these values we can assign the `NoData` integer (-9999 in this case) 
to `NA`. The `NoData` value can be found in the data documentation or by using the 
`GDALinfo` function.

``` {r assign-NA, fig.cap="Corrected histogram showing a bell curve-shaped distribution with some values clustered at zero."}

# assign NoData integer to NA
NA_Value <- -9999

# assign NA in dataset
RGB_HARV[RGB_HARV==NA_Value] <- NA

# update dataframe again
RGB_HARV_df <- as.data.frame(RGB_HARV, xy = TRUE)

# plot newest histogram
ggplot() +
  geom_histogram(data = RGB_HARV_df, aes(HARV_Ortho_wNA), bins = 40)

# look at the newest min and max values
summary(RGB_HARV_df)

```

Bad data values are not always a `NoData` integer and are sometimes actual data 
errors (like the 1110 value in the Julian day example above). In these cases, 
`NA` can be used as a place holder for bad data so those values do not show in 
the raster plot. For this example, we will assume that the valid range of values 
for this hypothetical raster is from 0 - 0.6, and we will assign a value of `NA` to 
any cells that are greater than 0.6.

``` {r bad-data, fig.cap=c("Raster with 100 cells and each cell is one of several colors that represents a number from 0 to 1.","Raster with 100 cells like the previous raster but many of the cells are empty indicating an NA value.")} 
# create a raster
my_raster <- raster(ncol=10, nrow=10)

# add random values to raster
values(my_raster) <- runif(ncell(my_raster))

# plot the raster
plot(my_raster, main="Raster with 100 cells")

# assign NA to cells with invalid values (greater than 0.6)
my_raster[my_raster > 0.6] = NA

# plot raster again
plot(my_raster, main="Raster with bad data removed")
```

In the example above, values above 0.6 are considered to be bad data in the raster 
dataset. By assigning those values to `NA`, they appear blank in the final raster 
plot. 

Other resources on handling bad data:

* <a href="https://towardsdatascience.com/all-about-missing-data-handling-b94b8b5d2184" target="_blank">An article</a> describing many ways to handle missing or bad data values. 
* More information on manipulating raster data can be found 
<a href="https://rspatial.org/raster/spatial/8-rastermanip.html" target="_blank">here</a>.
* A discussion about errors in GIS data from <a href="https://www.gislounge.com/gis-data-a-look-at-accuracy-precision-and-types-of-errors/" target="_blank">GIS Lounge</a>. 