---
layout: post
title: "Working with Lidar Data"
date:   2015-10-22 20:49:52
authors: "Keely Roth, Mike Alonzo, Jonah Duckles"
dateCreated:  2015-10-22 20:49:52
lastModified: 2015-10-22 14:28:52
category: remote-sensing-workshop
tags: [module-1]
mainTag: GIS-Spatial-Data
description: "This post explains the fundamental principles, functions and metadata that you need to work with raster data in R."
code1:
image:
  feature: lidar_GrandMesa.png
  credit: LiDAR data collected over Grand Mesa, Colorado - National Ecological Observatory Network (NEON)
  creditlink: http://www.neoninc.org
permalink: /R/Raster-Data-In-R/
code1: /R/2015-07-22-Introduction-to-Raster-Data-In-R.R
comments: true

---

<section id="table-of-contents" class="toc">
  <header>
    <h3>Contents</h3>
  </header>
<div id="drawer" markdown="1">
*  Auto generated table of contents
{:toc}
</div>
</section><!-- /#table-of-contents -->


## About
* Exploring Spatial patterns in Lidar Data: histograms,
* Summarizing lidar data: binary binning/thresholding, etc.

## R Skills Level
Intermediate (link to other group's lessons)

### Background

## Learning Objectives
* Understand the process of lidar measurement the resulting attributed point cloud
* Understand the process of creating gridded (raster surface) products from the point cloud
* Be able to open, plot, and explore the pixel values in a canopy height model (CHM)
* Understand the ecological meaning and value of the canopy height model
* Be able to crop out two study regions and display the subsets
* Be able to analyze and threshold height distributions using histograms
* Be able to visualize the spatial variation in the CHM using standard deviation in a moving window

## Things You'll Need to Complete this Lesson

### R libraries
* `raster`
* `rgdal`
* `rhdf5`

``` {r}
install.packages('raster')
install.packages('rgdal')
install.packages('rhdf5')

library('raster')
library('rgdal')
library('rhdf5')
```
### Tools

* R
* RStudio

### Data
* Lidar canopy height model (1 m resolution)

``` {r}
#download and save the chm
```

## Recommended Pre Lesson Reading

## Lesson 1: Open, crop, display, and explore the CHM

To get started, we'll open the CHM and display it. 

``` {r}
# Open the CHM
# Display the image

```
Now we're going to crop the lidar chm so that the extents match those of the hyperspectral and the 
  RGB camera data. With the cropped image, click a few locations in the image. What are the values you're seeing?
  FIXME: don't know what the extents will be

``` {r}
# Crop so that extents are the same as the RGB and hyperspectral data
# Extract values at a few locations to see the height values

```

Challenge: Create a histogram of the height values and discuss the shape of the distribution.
  What does it mean to have so many 0 values?

## Lesson 2: Subset the CHM to two Landsat pixels and summarize the data

We now want to look at the structural variation of the vegetation within two 30 m Landsat pixels. To 
  do this we will use the crop function based on outlines of the two pixels stored in a shapefile. Within these
  pixel boundaries we will summarize the height distributions using a histogram. Finally, we will use the 
  histogram to inform a threshold value that separates trees from low vegetation and ground.
  FIXME: do we want two subsets that will be summarized separately because they represent structurally unique
    areas?
  
``` {r}
# Crop again but using the shapefile of the two landsat pixels
# Use the tool (manipulate) that puts a slider on a histogram. Moving the slider will change the threshold
# directly on the image
```
Challenge:

## Lesson 3: Compute local standard deviation using moving window

With this lidar CHM we have the opportunity to go beyond simple summary statistics based on the data distribution.
  It may be useful to understand the spatial variation in vegetation structure. For instance, evenly spaced trees
  across the landscape could be indicative of a community that is limited by water availability.
  
In this lesson we will examine the spatial variation of canopy heights using local standard deviation computed with
  a moving X by X window. We can output a map of the variability and simple summary statistics.
  FIXME: clarify what are the useful summary stats after playing with the moving window method
  
``` {r}
# code for 2-D standard deviation using some window size that makes sense for the data
# sum and/or means of the two landsat pixel subsets to compare
```

Challenge: Vary the window size of the moving window and generate summary statistics. FIXME: need to play around
  with moving window
  
