---
layout: post
title: "Hyperspectral Remote Sensing"
date:   2015-10-22 20:49:52
authors: "Keely Roth, Mike Alonzo, Jonah Duckles"
dateCreated:  2015-10-22 20:49:52
lastModified: 2015-10-22 14:28:52
category: remote-sensing-workshop
tags: [module-3]
mainTag: GIS-Remote-Sensing-Data
description: "Overview of hyperspectral analysis in R"
code1:
image:
  feature: lidar_GrandMesa.png
  credit: LiDAR data collected over Grand Mesa, Colorado - National Ecological Observatory Network (NEON)
  creditlink: http://www.neoninc.org
permalink: /R/Raster-Data-In-R/
code1: /R/2015-07-22-Introduction-to-Raster-Data-In-R.R
comments: true


---

<section id="table-of-contents" class="toc">
  <header>
    <h3>Contents</h3>
  </header>
<div id="drawer" markdown="1">
*  Auto generated table of contents
{:toc}
</div>
</section><!-- /#table-of-contents -->



## About

## R Skills Level: need to know RS basics


## Goals and Objectives
* understanding of spectral concepts: bands and wavelengths
* open & display multispectral and hyperspectral data
* geographically clip raster imagery
* plot and compare single-pixel spectra
* Understanding of information content of visible, multispectral and hyperspectral data
* recognize general spectral signatures
* calculate and understand the physical basis for spectral metrics
* create and threshold based on histogram visualizations of spectral information
* standard deviation moving window

## Things You'll Need to Complete this Lesson

### R libraries
* `raster`
* `rgdal`
* `rhdf5`
* `sp`
* `manipulate`
* `plyr`*


``` {r}
install.packages('raster')
install.packages('rgdal')
install.packages('rhdf5')
install.packages('sp')
install.packages('manipulate')
install.packages('plyr')

library('raster')
library('rgdal')
library('rhdf5')
library('sp')
library('manipulate')
library('plyr')
```

### Tools

* R
* RStudio

### Data
* Visible High Res 10cm resolution
* Hyperspectral HDF5 1m resolution
* Multispectral Data ~30m resolution

``` {r}
#download and save each datasets
```


## Recommended Pre Lesson Reading

understand what reflectance is


## Notes for Formatting Lessons
fill in outline for each module (think about story arc):
with lesson obj
lesson text/concept blocks
chunk in programming steps
Challenges
organization:
Objective
Tutorial
Functions
Challenge





## NEON Data Resolutions
|             |  Spatial             | Spectral         | Temporal |
|-------------|----------------------|------------------|----------|
|NEON Visible       | Very High (eg 10cm)  | 3 broad bands    | 1-2/year|
|Landsat Multispectral | Moderate (eg 30m)    | 6 broad bands    | ~22/year|
|NEON hyperspectral | High (eg 1m)         | 424 FIXME narrow bands| 1-2/year|

FIXME: add spectral profile graph and spectral sampling graphs
FIXME: need to add the cropping portions of the lesson

## Lesson 1: Opening and Exploring Multispectral and Hyperspectral Imagery

To get started, we'll explore each of the three image datasets used in this module.
First we'll open and display single bands from the high resolution RGB camera data. Then we'll combine these in a raster
stack and display them in RGB.

``` {r}
# Open RGB camera image
# Display single bands 
# Create RGB stack
# Display RGB stack
```

We can also examine the red, green and blue digital number values (0-255) for a pixel and compare values for different pixels.
Using the code below, click on four different locations (try to find some trees, some grass, some dirt, something impervious) that are fairly far apart.
``` {r}
# Click and save point locations
# FIXME:Assign wavelength values to each band
# Extract data values from these pixels and examine resulting table
# Plot the RGB values for these pixels
# FIXME: Plot the locations of these pixels on the RGB image
```
We call each of these lines the "spectral profile" of the pixel. How might you expect these to look different? How different do they look?

Now we'll do the same thing with multispectral Landsat data. We'll read this in a bit differently since each band is in its own file. We'll read in just 6 bands (Landsat 8 has 11 bands total). Use the band centers as the wavelength values.
|Band Number|Wavelength Region (Band Range)|Band Center|
|___________|______________________________|___________|
|2          | blue (450-510 nm)            | 480 nm    |
|3          | green (530-590 nm)           | 560 nm    |
|4          | red (640-670 nm)             | 665 nm    |
|5          | NIR (850-880 nm)             | 865 nm    |
|6          | SWIR1 (1570-1650 nm)         | 1610 nm   |
|7          | SWIR2 (2110-2290 nm)         | 2200 nm   |

``` {r}
# Read in single band GeoTIFFS (nice to have loop to read in from dir list)
```

Challenge: Create a stack of all 6 Landsat bands and display an RGB image. Plot the spectral profiles of the same four 
locations as you did in the RGB camera data.

Finally we will open and display the hyperspectral imagery.
``` {r}
# read in HDF5 file
# FIXME: examine metadata
# FIXME: how to display an RGB composite AND extract spectral data from all bands
```

Challenge: Explore spectral variation within a tree crown, or within a field of grass. Select a new set of pixel locations (save to new variable) and plot the signatures. 

## Lesson 2: Multispectral vs. Hyperspectral Profiles & Spectral Signatures

Now we'd like to more directly compare multispectral and hyperspectral signatures. However, our Landsat data is at a much coarser spatial resolution than the IS data, making a direct comparison difficult. So instead, we'll simulate the 6 Landsat
bands from the IS data using the "convolve" function provided below. This function uses something called a "sensor response
function", which basically says how much of the light at each wavelength in a band range is recorded by the sensor (add
figure and link to more information)

``` {r}
# Convolve IS signatures using pre-defined function
# Plot the 8 spec signatures from each of the images (4 convolved IS-to-Landsat and 4 from IS)
```
These spectra differ in four ways (spectral resolution, number of bands, and location of bands). [Add info; refer to 
table/fig]

We see lots more info... 
Physics that drive these signatures (nice graphics here)

Basic spectral signatures (building off the physics above)

Challenge: Extract, plot and identify the mystery signatures from the provided coordinates. We'll use these spectra
moving forward.

## Lesson 3: Spectral Metrics

Now that we understand the driving forces behind the spectrum, we can explore different ways of capturing that information.
Spectral indices are a popular method for doing this. NDVI example. Because we have finer spectral resolution in IS data, we
call these "narrowband indices" when calculated with IS data.

``` {r}
# Calculate NDVI with the "Landsat" spectra
# Calculate narrowband NDVI with the IS spectra
# Compare values.
```
Challenge: Here's the equation for INDEX-X. Calculate this for multispectral and IS spectra.

One of the major benefits of IS data is that we can capture information from very specific bands.
Red edge wavelength example. Note that we CAN calculate this w/multispec data, but not very precise.

```{r}
# Calc red edge w/IS spectra
   # Subset IS spectra to wavelength region of interest
   # Calculate first derivative
   # Find wavelength at which maximum derivative value occurs
   # Compare resulting values
```

Challenge: Using the same approach, we'll find the wavelength of max absorption (i.e.,min value) in a lignin-cellulose feature (xx-xx nm)

Another major benefit of IS data is creating a metric that reflects an actual physical measurement (e.g., how much water
in this pixel). Equivalent Water Thickness. Note that we can't even try to calc this using this approach w/multispec
``` {r}
# Adapt IDL code here into a function
# Calculate EWT for the IS spectra
```

## Lesson 4: Mapping Spectral Metrics

Now we'll make a map of these metrics and use it to answer some questions.

``` {r}
# Same approach as in lidar. Choose one index, calculate for entire image
# Display index raster
# Explore histogram thresholding to create a map.
```

Challenge: Create a map using another metric.

## BELOW HERE IS JUST NOTES:

###  viz
Open
Display one band of image (for each band)
Create stack and display rgb composite
Open vector and overlay
Clip raster using the vector (using crop?)

```{r}
#load 10 cm vis dataset

```

Use image(red) image(green) image(blue)
SJER_camera.tif
import(raster)
plotrgb()


### multi
Open the geotiffs
Display one band of image
Create 6-band stack and display a true color composite
Open vector and overlay
Clip stack using the vector (using crop?)

```{r}
#load landsat subset here
```

###hyper
Open HDF5
(maybe look at some metadata)
Slice out one band of image
Find roughly the same wavelenght as done above 
Slice out true color wavelengths
Create 3-band stack and display
Open vector and overlay
Clip the full set of bands using the vector (using crop?)



```{r}
#load hyperspectral data as h5



```



### Compare
3-by raster plot
3-by spectral plot

## Clip

Why do we clip?

Apply a clip across all three rasters to about two landsat pixels

## Understanding hyperspectral signatures
Physics of light and reflectance

Static diagram with spectral regions

Fire trucks (that are yellow/green)!

### Plot example spectral signatures

* Plot some spectral signatures

### Compare spectral signatures
* Compare
* plot of several materials (trees, dirt, grass, water)

### Spectral metrics
* lines at 550 for green leaves
* lines at 684 for chlorophyl absorption
* Calculate spectral metrics
  * Naroww band NDVI
  * Water metric
  * Stress metric
* Single pixel calculation
* Creating a new raster based on the metrics with histogram
* Threshold the new computed raster



##About
This activity will walk you through the fundamental principles of working
with raster data in R.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives">

<h3>Goals / Objectives</h3>
After completing this activity, you will know:
<ol>
<li>What a raster dataset is and its fundamental attributes.</li>
<li>How to import rasters into `R` using the raster library.</li>
<li>How to perform raster calculations in `R`.</li>
</ol>

<h3>Things You'll Need To Complete This Lesson</h3>

<h3>R Libraries to Install:</h3>
<ul>
<li><strong>raster:</strong> <code> install.packages("raster")</code></li>
<li><strong>rgdal:</strong> <code> install.packages("rgdal")</code></li>

</ul>
<h4>Tools To Install</h4>

Please be sure you have the most current version of `R` and preferably
R studio to write your code.


<h4>Data to Download</h4>

Download the raster and *in situ* collected vegetation structure data:
<ul>
<li><a href="http://neonhighered.org/Data/LidarActivity/CHM_InSitu_Data.zip" class="btn btn-success"> DOWNLOAD Sample NEON LiDAR data in Raster Format & Vegetation Sampling Data</a></li>
<li><a href="{{ site.baseurl }}/data/rasterLayers_tif.zip" class="btn btn-success"> DOWNLOAD NEON imagery data (tiff format) California Domain 17 (D17)</a></li>
</ul>

<p>The LiDAR and imagery data used to create the rasters in this dataset were
collected over the San Joaquin field site located in California (NEON Domain 17)
and processed at <a href="http://www.neoninc.org" target="_blank" >NEON </a>
headquarters. The entire dataset can be accessed by request from the NEON website.</p>  

<h4>Recommended Pre-Lesson Reading</h4>
<ul>
<li>
<a href="{{ site.baseurl }}/GIS-Spatial-Data/Working-With-Rasters/" target="_blank">
The Relationship Between Raster Resolution, Spatial extent & Number of Pixels - in R</a>
</li>
<li>
<a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">
Read more about the `raster` package in R.</a>
</li>
</ul>
</div>

#About Raster Data
Raster or "gridded" data are data that are saved in pixels. In the spatial world,
each pixel represents an area on the Earth's surface. For example in the raster
below, each pixel represents a particular land cover class that would be found in
that location in the real world.
<a href="{{ site.baseurl }}/GIS-Spatial-Data/Working-With-Rasters/"> More on
rasters here</a>.

```{r chunk-1}
#code will go here!

```
