---
layout: post
title: "Fusing Information from Lidar & Hyperspectral"
date:   2015-10-23 13:43
authors: "Keely Roth, Mike Alonzo, Jonah Duckles"
dateCreated:  2014-11-26 20:49:52
lastModified: 2015-07-23 14:28:52
category: time-series-workshop
tags: [module-4]
mainTag: GIS-Spatial-Data
description: "This post explains the fundamental principles, functions and metadata that you need to work with raster data in R."
code1: 
image:
  feature: lidar_GrandMesa.png
  credit: LiDAR data collected over Grand Mesa, Colorado - National Ecological Observatory Network (NEON)
  creditlink: http://www.neoninc.org
permalink: /R/Raster-Data-In-R/
code1: /R/2015-07-22-Introduction-to-Raster-Data-In-R.R
comments: true

---

<section id="table-of-contents" class="toc">
  <header>
    <h3>Contents</h3>
  </header>
<div id="drawer" markdown="1">
*  Auto generated table of contents
{:toc}
</div>
</section><!-- /#table-of-contents -->

##About
This activity will walk you through a simple example of combining information from both lidar and hyperspectral data.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives">

##Goals / Objectives
After completing this activity, you will know:
* load and display raster products from both lidar and hyperspectral modules
* plot hyperspectral metrics on CHM (3D surface)
* pixel-based correlation & scatter plot
* combine the two data products to answer key questions
* create summary stat report (total area statistics)
* create landscape metric report (simple LE mets/spatial stats)



##Things You'll Need To Complete This Lesson

###R Libraries to Install:


###Tools To Install

Please be sure you have the most current version of `R` and preferably
R studio to write your code.


###Data to Download


###Recommended Pre-Lesson Reading:

### Lesson 1: Open & Visualize Data Products

In this section, you will work with the rasters used in the previous modules (add link to modules), specifically, the
lidar canopy height model, the spectral metric(s) rasters.

```{r}
# display the CHM and metric rasters side by side for each study region 1
# display and compare region 2
```

Now let's create a 3-D surface for one of the spectral metrics using the height data from the CHM
```{r}
# add code to create and rotate surface
```

We can also compare pixel values from the different data types using a scatterplot.
```{r}
# create an xy scatterplot from CHM and a spectral metric
# [would be cool to have a conditional slider to help them visualize pixels]
```
Challenge: Compare CHM values and a different spectral metric using this approach

### Lesson 2: Create a fused data product

Here we will combine the binary maps created in (linked modules) from the lidar and hyperspectral data. [The details here
depend on exploring the data a bit more to create the story arc].

```{r}
# add raster math for mapping something specific (based on story arc)
```
Challenge: Add another layer to your fusion map (i.e., incorporate another metric).

### Lesson 3: Spatial Analysis

In this section you will run some simple landscape ecology metrics to analyze your fusion maps and compare study regions.
[Explain metric example]

```{r}
# Calculate LE metric and display output for both study regions
# Compare study regions
```

Challenge: Calculate [another LE metric] and compare your study regions.

