---
syncID: 
title: "Summarizing Fish Capture Data"
description: "This tutorial will provide an introduction to discovering, accessing and summarizing NEON fish count data using R"
dateCreated: 2026-02-20
authors: Madaline Ritter
contributors: Madaline Ritter, Ryan Williams, Claire Lunch, Garrett Williams
estimatedTime: 1.5 hrs
packagesLibraries: neonUtilities, dplyr, ggplot2
topics: data-analysis, data-visualization
languagesTool: R
dataProduct: DP1.20107.001
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/...
tutorialSeries: NA
urlTitle: fish-summary
---

<div id="ds-objectives" markdown="1">

## Learning Objectives 
After completing this tutorial you will be able to: 

* Download NEON fish data. 
* Generate simple abundance metrics.
* Visualize diversity across sites

## Things You’ll Need To Complete This Tutorial

### R Programming Language
You will need a current version of R to complete this tutorial. We also recommend 
the RStudio IDE to work with R. 

### R Packages To Install
Start by installing any packages that are used during the course of the tutorial (if necessary) and setting options. Installation can be run once, then periodically to get package updates.

* **dplyr:** `install.packages("dplyr")`
* **neonUtilities:** `install.packages("neonUtilities")`
* **ggplot2:** `install.packages("ggplot2")`

<a href="/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

</div>

## 1. Setup

In this first section we will get our R session set up so that we are prepared to work with and analyze the data.  This involves loading the necessary R packages, downloading the data, setting a working directory if manual data upload is needed, and a review of the NEON data citation policies.

### Load R Packages

Once the packages have been installed they need to be loaded to the session. This needs to be done every time you run code. 

```{r load-packages, , message=FALSE, warning=FALSE, results="hide"}
library(dplyr)
library(neonUtilities)
library(ggplot2)
```

### Download NEON Fish Data
Download the fish electrofishing, gill netting, and fyke netting counts using the `loadByProduct()` function in
the `neonUtilities` package. Inputs needed to the function are:

* `dpID`: data product ID; fish electrofishing, gill netting, and fyke netting counts = DP1.20107.001
* `site`: (vector of) 4-letter site codes; PROP, PRLA, BLUE, CUPE, CRAM
* `package`: basic or expanded; we'll download basic here
* `check.size`: should this function prompt the user with an estimated download size? Set to `FALSE` here for ease of processing as a script, but good to leave as default `TRUE` when downloading a dataset for the first time.

Refer to the <a href="https://www.neonscience.org/sites/default/files/cheat-sheet-neonUtilities.pdf" target="_blank">cheat sheet</a> 
for the `neonUtilities` package for more details if desired.

```{r fshdat, results='hide', message=FALSE}
fshdat <- neonUtilities::loadByProduct(
  dpID="DP1.20107.001",
  site=c("PRPO", "PRLA", "BLUE", "CRAM", "CUPE"),
  package="basic",
  check.size = FALSE,
  startdate = "2024-01",
  enddate = "2024-12")
```

### Set Working Directory if Using Locally-Saved Data
If the data are not loaded directly into the R session with loadByProduct (e.g., if wifi is not available), this lesson assumes that you have set your working directory to the location of the zipped data downloaded from the NEON portal. 

<a href="https://www.neonscience.org/set-working-directory-r" target="_blank"> An overview
of setting the working directory in R can be found here.</a>

```{r set directory, results="hide", message=FALSE}
# This section of code should only be run if the section above using loadByProduct is not used.

# Change the filepath below to match the location of locally saved zipped files:
wd<-"C:/Users/ritterm1/Documents/data/"

# Set working directory
setwd(wd)

# Read downloaded files into R
fshdat <- neonUtilities::stackByTable(
  filepath = "NEON_count-fish.zip", 
  savepath = "envt")
```

### NEON Data Citation
The use of NEON data should be cited according to our
<a href="https://www.neonscience.org/data-samples/data-policies-citation" target="_blank"> Data Policies & Citation Guidelines</a>.

The data used in this tutorial were collected at the 
<a href="http://www.neonscience.org" target="_blank"> National Ecological Observatory Network's</a> 
<a href="/field-sites/field-sites-map" target="_blank"> field sites</a>.  

* NEON (National Ecological Observatory Network). Fish electrofishing, gill netting, and fyke netting counts (DP1.20107.001). https://data.neonscience.org (accessed on `r Sys.Date()`)

## 2. Compiling the NEON Fish Data
The data are downloaded into a list of separate tables. Before working with the data the tables are added to the R environment

```{r download-overview, message=FALSE, warning=FALSE}
# View all tables in the list of downloaded fish data:
names(fshdat)

# Extract the items from the list and add as dataframes in the R environment:
list2env(fshdat, envir=.GlobalEnv)
```

* The `categoricalCodes` file provides controlled lists used in the data

* The `issueLog` and `readme` have the same information that you will find on the data product landing page of the data portal.

* The `fsh_fieldData` table includes the date and time for all reach sampling efforts and will include an eventID value to indicate a unique bout of sampling in the 2026 release.

* The `fsh_perPass` table includes a record for each sampling pass conducted, and is linked to the field table through eventID and namedLocation.

* The `fsh_perFish` table includes records for each individually processed fish.

* The `fsh_bulkCount` table includes a count per species of fish captured after the specified number of individual fish are processed for the unique taxon or morphospecies.

* The `validation` file provides the rules that constrain data upon ingest into the NEON database.

* The `variables` file describes each field in the returned data tables.

### Quality Verification
For bulk count data there should only be one bulk count record for each taxonID and pass. Before summarizing total captures across tables it is helpful to ensure bulk count records are present when expected.

```{r quality checks}
bulkCount_count <- fsh_bulkCount %>% 
  select(eventID, boutEndDate,  passNumber, scientificName, taxonID, namedLocation, bulkFishCount) %>%
  group_by(eventID, boutEndDate, passNumber, namedLocation, scientificName, taxonID) %>%
  count()

if(unique(bulkCount_count$n) == 1){
  cat('There are no taxa with >1 bulk count record per pass')
}else{
  cat('>1 Bulk count record exists for the following records:\n')
  print(bulkCount_count %>% filter(n != 1))
}
```

### Subset the data
First the bulk count data should be subset to only the necessary columns, to help keep the final summary output concise.

```{r subset data, results="hide"}
bulkCount_sub <- fsh_bulkCount %>% 
  select(eventID, passStartTime, passEndTime, boutEndDate,
         passNumber, namedLocation, barrierSubReach, 
         scientificName, taxonID, bulkFishCount)
```

### Calculate the number of individual fish processed per pass
Before joining individual fish data with per pass data to calculate total rates of capture, the total number of individual fish in the per fish table must be calculated for each pass. Per fish data is grouped by eventID, passNumber, namedLocation, barrierSubReach, and taxonID before tallying records for each taxon.

```{r tally per fish individuals, results="hide"}
perFish_total <- fsh_perFish %>%
  select(eventID, passStartTime, passEndTime, boutEndDate, passNumber, 
         namedLocation, barrierSubReach, scientificName, taxonID) %>%
  group_by(eventID, passStartTime, passEndTime, boutEndDate, passNumber, 
         namedLocation, barrierSubReach, scientificName, taxonID) %>%
  count(name = 'individualFishCount')
```

### Join individual and bulk count tables
Many analyses of NEON data will require the joining of two or more data tables that contain different sets of information. For instance, the information about individual fish captures and samples from a bout is found in the fsh_perFish data table, while information about additional fish that were not processed individually and instead counted in bulk is found in the fsh_bulkCount data table. These two tables can be joined so that capture data can be easily summarized by pass and bout.  

Details about which tables can be joined together and what variables should be used to link tables can be found in the "Table joining" section of the Quick Start Guide on each data product landing page.

```{r table join, results="hide"}
fsh_all <- perFish_total %>%
  full_join(., bulkCount_sub, by=c('eventID', 'passStartTime', 'passEndTime', 
                                   'boutEndDate', 'passNumber', 
                                   'namedLocation', 'barrierSubReach', 
                                   'scientificName', 'taxonID'))
```

### Calculate total captures
There are now two columns in the new `fsh_all` data frame that contain capture numbers for each unique pass - one with total number of fish in the perFish table, and one with total number of fish in the bulkCount table. These two values can be added to calculate the total number of fish captured per pass.

```{r total captures, results="hide"}
fsh_all <- fsh_all %>%  
  mutate(
    totalFishCount = rowSums(across(c(individualFishCount, bulkFishCount)), na.rm = TRUE))
```


## 3. Visualizing Species Richness
Here we will calculate species richness for each unique eventID, to examine the number of  species captured during each sampling bout. 

### Calculate richness for each eventID

```{r richness-by-site, message=FALSE}
richness_by_event <- fsh_all %>%
  select(eventID, scientificName) %>%
  group_by(eventID) %>%
  distinct(scientificName) %>%
  summarise(numSpecies = n()) %>%
  arrange(-numSpecies)
```

Now we can visualize the results.

```{r richness-by-site-graph}
ggplot(richness_by_event, aes(x = eventID, y = numSpecies)) +
        geom_bar(stat = "identity") +
        scale_x_discrete(limits = richness_by_event$eventID) +
        theme(axis.text.x = element_text(angle = 90)) +
        scale_y_continuous(breaks = seq(from = 0, to = 30, by = 2)) +
        labs(
                title = "Total Species Richness by EventID",
                x = "Event ID",
                y = "Number of Species"
        )
```

Looks like <a href="https://www.neonscience.org/field-sites/field-sites-map/BLUE">Blue River</a> in Domain 11 has far and away the
highest species richness, but to get a deeper understanding of Blue's 
fish populations we'll need an idea of relative and absolute abundance.