---
syncID: 963d7dcebaa54e09a297bc43ecd9ae2b
title: "Exploring diel carbon flux cycles"
description: Download and extract NEON eddy flux data to evaluate diel carbon flux cycles
dateCreated:  2023-04-09
authors: [David Durden]
contributors: 
estimatedTime: 1 hour
packagesLibraries: [rhdf5, neonUtilities, ggplot2]
topics: HDF5, eddy-covariance, eddy-flux, carbon
languageTool: R
dataProducts: DP4.00200.001
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/eddy-covariance/diel-cycle-flux/flux-diel-cycle.R 
tutorialSeries: 
urlTitle: eddy-diel-cycle
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

This data tutorial provides an overview of exploring NEON carbon 
flux data, using the `neonUtilities` R package. If you are just starting to work with NEON data it is recommended that you also checkout the more general tutorials, such as the 
<a href="https://www.neonscience.org/neonDataStackR" target="_blank">neonUtilities tutorial</a> 
or the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/eddy-data-intro" target="_blank">Introduction to working with NEON eddy flux data</a> for a more flux centric introductory tutorial. This tutorial builds off the data exploration techniques described in the aforementioned tutorials, and will explore carbon dynamics at two forested ecosystem sites with very different land management practices in NEON's Domain 5 - Great Lakes, <a href="https://www.neonscience.org/field-sites/stei" target="_blank">STEI</a> and <a href="https://www.neonscience.org/field-sites/tree" target="_blank">TREE</a>. The STEI and TREE sites are only located approximately 1 mile from each, so the observed differences in fluxes can be attributed to land management practices associated with the forests.

This tutorial assumes general familiarity with eddy-covariance data, and will briefly touch discuss both turbulent (`turb`) and storage (`stor`) fluxes in relation to the net surface atmosphere exchange (`nsae`). The `nsae` for CO<sub>2</sub> fluxes often referred to as Net Ecosystem Exchange (NEE).

## 1. Setup

Start by installing and loading packages and setting options. The function below will evaluate if the required packages (`packReq`) already exist in you R library, and will install and load these packages. The `rhdf5` package  is hosted on Bioconductor requiring a BiocManager to be installed to allow `rhdf5` to be pulled from the Bioconductor repository:


```{r install-load, results='hide'}

#Required R packages
packReq <- c("BiocManager", "rhdf5", 'neonUtilities', 'ggplot2','tidyverse', "lubridate")

#Install and load all required packages
lapply(packReq, function(x) {
  if(require(x, character.only = TRUE) == FALSE) {
    install.packages(x)
    library(x, character.only = TRUE)
  }})

options(stringsAsFactors=F)

```

Now that all the required packages are loaded, let's define all the necessary function arguments to download and stack our flux data using the `neonUtilities` package's `zipsByProduct()` and `stackEddy()` functions. I will define the input arguments as variables to make it easy to modify if I want to explore additional sites or months of data. For this analysis, we will focus on the 2022 growing season at STEI and TREE. Please be aware that the data download size is ~ 1 GB. The code can be modified to shorten or extend this window, but to get a good look at the carbon dynamics at the site it was important to capture most of the growing season.

Inputs to the `zipsByProduct()` function:

* `dpID`: DP4.00200.001, the bundled eddy covariance product
* `package`: basic (the expanded package has additional quality metrics and advanced footprint matrices that are not relevant to this tutoiral)
* `site`: STEI and TREE
* `startate`: 2022-04
* `enddate`: 2022-09 
* `savepath`: modify this to something logical on your machine or use the R `tempdir()`
* `check.size`: TRUE if you want to see file size before downloading, otherwise FALSE

The download may take a while, especially if you're on a slow network. 
For faster downloads, consider using an <a href="https://www.neonscience.org/resources/learning-hub/tutorials/neon-api-tokens-tutorial" target="_blank">API token</a>.


```{r download, results="hide", message=FALSE}

#Target dates
startDate <- "2022-04"
endDate <- "2022-09"

#Site
site <- c("STEI", "TREE")

#File directory
dirFile <- tempdir()

zipsByProduct(dpID="DP4.00200.001", package="basic", 
              site=site, 
              startdate=startDate, enddate=endDate,
              savepath=dirFile, 
              check.size=FALSE)

```

## 2. Stacking Level 4 Flux Data

There are five levels of data contained in the eddy flux bundle. For full 
details, refer to the <a href="https://data.neonscience.org/documents/10179/2403599/NEON.DOC.004571vC/4c72353a-35fb-1136-ef9f-cbdc514711ad" target="_blank">NEON algorithm document</a>.

In this tutorial we will only be focusing on Level 4 (`dp04`) flux data products; however, additional data products from Level 0' (`dp0p`) calibrated raw data to Level 3 (`dp03`) spatially interpolated vertical profiles used to derive the storage flux are available in the EC bundled HDF5 files. Information can be found on the <a href="<a href="https://data.neonscience.org/data-products/DP4.00200.001" target="_blank">NEON algorithm document</a> webpage. The  <a href="https://www.neonscience.org/resources/learning-hub/tutorials/eddy-data-intro" target="_blank">Introduction to working with NEON eddy flux data</a> tutorial dives into additional detail regarding the other data product levels, but this tutoiral will focus exclusively on flux data.

To extract the `dp04` data from the HDF5 files and merge them into a 
single table, we'll use the `stackEddy()` function. We provide the function to input arguments, `filepath` and `level`. The `filepath` will be the file directory (`dirFile`) used for the data download via `zipsByProduct()` combined with the `filestoStack00200` folder created by the function. To grab just the flux data products `level = dp04`:



```{r stack-dp04, results="hide"}

flux <- neonUtilities::stackEddy(filepath=paste0(dirFile,"/filesToStack00200"),
                  level="dp04")

```

We now have an object called `flux`. It's a named list containing four 
tables: one table for each site's data, and `variables` and `objDesc` 
tables. One of the advantages to the HDF5 files is all the metadata and data descriptions can be compiled into file as additional data tables or as attributes. The `stackEddy()` function stacks this contextual data in the `variables` and `objDesc` tables to help you interpret the column headers in the data table (though it's not comprehensive currently). 

The terms of interest for our analysis in `dp04` the three flux quantities under
`fluxCo2`:

* `turb`: Turbulent flux
* `stor`: Storage
* `nsae`: Net surface-atmosphere exchange

Note the `variables` table contains the units for each field:

```{r see-variables}

flux$variables

```



## 3.  Plotting CO<sub>2</sub> Flux Data

Let's look at some data! First, we will combine the data from the two sites into a single dataframe with an additional factor variable for the `site` to allow us to utilize `tidyverse` and `ggplot2` packages for some data munging and to create some interesting data exploration visualizations. In this tutorial we utilize the piping operater `%>%` from the `dplyr` to allow us to quickly subset, filter, sort, and plot data using just our combined dataframe `dfFlux`. We start by looking at the turbulent CO<sub>2</sub> flux timeseries at TREE with the points colored by `qfFinl`:


```{r plot-fluxes}

flux$STEI$Site <- "STEI"
flux$TREE$Site <- "TREE"

dfFlux <- rbind.data.frame(flux$STEI, flux$TREE)
dfFlux$Site <- as.factor(dfFlux$Site)

dfFlux %>% 
    ggplot(aes(timeBgn, data.fluxCo2.turb.flux, color=factor(qfqm.fluxCo2.turb.qfFinl) )) +
    geom_point() + 
    scale_color_brewer(palette="Set2", name="qfFinal") +
    facet_grid(~Site) 
    

```

When we plot the data turbulent CO<sub>2</sub> flux data, we see a good agreement
between the STEI and TREE sites. We also see some outliers that look to be due to 
erroneous data, especially at the STEI site. Fortunately, it looks like final 
quality flag (qfFinal = `qfFinl`) caught the largest outliers, and  we can use it to remove
these data from the analysis. Let's first dig into a little background information 
about NEON bundled EC quality flags.

### Quality flags

The NEON Surface-atmosphere exchange (SAE) data products all come accompanied by
a qfFinal. Information the defining quality flag (qf), quality metric (qm), 
and the framework to derive qfFinal are detailed in <a href="https://data.neonscience.org/documents/10179/2403599/NEON.DOC.004571vC/4c72353a-35fb-1136-ef9f-cbdc514711ad" target="_blank">NEON EC bundled data products algorithm document</a> and 
<a href="https://data.neonscience.org/api/v0/documents/NEON.DOC.001113vA" target="_blank">NEON Quality Flags and Quality Metrics for TIS
Data Products algorithm document</a>. Ultimately, for the `dp04` flux data 
products (e.g. `data.fluxCo2.turb.flux`), qfFinal of `dp01` constituent data 
streams derived from plausibility tests (i.e. range test, step test, spike test,
and persistence test) and sensor diagnostic flags, `fluxCo2` = `rtioMoleDryCo2` 
and `veloZaxsErth` (dp01 data outside the scope of tutorial), are propagated and 
combined with theoretical eddy-covariance specific quality tests (i.e., stationarity,
integral turbulence characteristics tests)  to determine whether the data product
is flagged as valid (qfFinal = 0) or invalid (qfFinal = 1).

Let's summarize our `qfqm.fluxCo2.turb.qfFinl` to see the total flagged percentage
at our 2 sites.

```{r mean-qfqm}

dfFlux %>% group_by(Site) %>% summarise(mean(qfqm.fluxCo2.turb.qfFinl))

```

Now, lets plot the qfFinal failed percentage for all our flux data products:


```{r plot-qfqm}
dfFlux %>% 
  select(contains("qfqm")) %>% 
    pivot_longer(cols = everything(), names_to = "var") %>% 
      group_by(var) %>%  
        summarise(mean_var = mean(value) * 100) %>% 
          ggplot(aes(x = var, y = mean_var, fill = var)) + 
          geom_col() + 
          guides(x = guide_axis(angle = 90)) + 
          labs(x="Variable", y="Percent qfFinal failed") + 
          scale_fill_brewer(palette="RdYlBu")
    

```
### Removing data with failed qfFinal

We see that `qfqm.fluxCo2.turb.qfFinl` is flagged ~ 20-25% over the growing season; however,
the `qfqm.fluxCo2.stor.qfFinl` and `qfqm.fluxCo2.nsae.qfFinl` are flagged quite a 
bit more due to the complexity of the system and propagation of issues between 
sub-components to the overall system (additionally, we currently employ a conservative
quality control for the storage fluxes). Now that we have evaluated the qfFinal 
for all the `fluxCo2` data streams, lets remove flagged data from our dataframe.


```{r qfqm-remove}

dfFlux %>% select(contains("qfqm") & contains("fluxCO2"))  %>% summarise_each(sum)
dfFlux %>% select(contains("data") & contains("fluxCO2")) %>% summarise_each(funs(sum(is.na(.))))

dfFlux$data.fluxCo2.turb.flux[(which(dfFlux$qfqm.fluxCo2.turb.qfFinl== 1))] <- NaN
dfFlux$data.fluxCo2.stor.flux[(which(dfFlux$qfqm.fluxCo2.stor.qfFinl== 1))] <- NaN
dfFlux$data.fluxCo2.nsae.flux[(which(dfFlux$qfqm.fluxCo2.nsae.qfFinl== 1))] <- NaN
   
dfFlux %>% select(contains("qfqm") & contains("fluxCO2"))  %>% summarise_each(sum)
dfFlux %>% select(contains("data") & contains("fluxCO2")) %>% summarise_each(funs(sum(is.na(.))))
 

```
We see from the summary of the `fluxCo2` qfFinal and data NA's, that we have 
removed all the data from `data.fluxCo2.turb.flux`, `data.fluxCo2.turb.flux`, 
and `data.fluxCo2.turb.flux`. Now we have a clean data set to begin our diel 
flux cycle data analysis. 


## 3.  Plotting the diel cycle of CO<sub>2</sub> fluxes


Let's look at the data in a different way, the diel cycle of carbon and water 
vapor are important when evaluating carbon sequestration and ecosystem health.
We can evaluate the diel cycle by binning the data by `hour` in the day, this is
relatively straightforward using `ggplot` and the boxplot function (`geom_boxplot()`).


```{r plot-diel-cycle}

dfFlux$hour <- factor(lubridate::hour(dfFlux$timeBgn))

ggplot(dfFlux, aes(x = hour, y = data.fluxCo2.turb.flux, fill = Site)) +
  geom_boxplot() +
  stat_summary(fun = median, geom = 'line', aes(group = Site, colour = Site)) 

```

Nice! We get a glimpse of daily carbon dynamics, but something doesn't quite look right.

Note the timing of C uptake; the UTC time zone is clear here, where 
uptake occurs at times that appear to be during the night. Note that
**all** NEON data use UTC time, aka Greenwich Mean Time. 
This is true across NEON's instrumented, observational, and airborne 
measurements. When working with NEON data, it's best to keep 
everything in UTC as much as possible, otherwise it's very easy to 
end up with data in mismatched times. We may want to convert the time to Local
standard time (LST). The metadata attributes at the `site` level in the bundled 
EC HDF5 files contains fields for `ZoneTime` and `TimeDiffUtcLt`. We can use the 
`rhdf5` package `h5readAttributes()` function to grab the metadata. In this 
analysis, both the STEI and TREE are in the same timezone. Therefore, we will 
only grab metadata from one file and apply to both sites.

```{r plot-diel-lst}

fileMeta <- list.files(dirFile, pattern = paste0(".*",site[1],".*.h5"), recursive = TRUE, full.names = TRUE)[1]

siteMeta <- h5readAttributes(fileMeta, name = site[1])

siteMeta 

dfFlux$timeBgnLst <- dfFlux$timeBgn + lubridate::hours(siteMeta$TimeDiffUtcLt)
dfFlux$hourLst <- factor(lubridate::hour(dfFlux$timeBgnLst))

ggplot(dfFlux, aes(x = hourLst, y = data.fluxCo2.turb.flux, fill = Site)) +
  geom_boxplot() +
  stat_summary(fun = median, geom = 'line', aes(group = Site, colour = Site)) +
  scale_y_continuous(limits = quantile(dfFlux$data.fluxCo2.turb.flux, c(0.001, 0.999), na.rm = TRUE))

```

That looks better, now we can clearly see the diel cycle of the turbulent 
CO<sub>2</sub> flux at our STEI and TREE sites with peak carbon uptake around noon. 
The carbon uptake should align with peak solar angle and incoming 
photosynthetically active radiation (PAR). The boxplot also provides us information
about the variability of the fluxes throughout the day across the growing season
with interquartile range (IQR) represented by the box and the whiskers indicating
the threshold for outliers (black dots) as 1.5 * IQR subtracted from 1st quantile (`Q1`) 
and added to 3rd quantile (`Q3`):

Q1 - 1.5 * IQR 

Q3 + 1.5 * IQR

In the boxplots above the outlier points were greatly reduced after the qfFinal
data removal; however, some outliers remained. To focus our attention on the 
bulk of the data, we scaled the y-axis using `scale_y_continuous()` in conjunction
with the `quantile()` function to look at 99.8% of the data excluding extreme 
outliers that may have remained after our filtering. 
We know that the turbulent flux is usually the dominant flux component to NEE, 
but the storage flux can be significant at forested sites. Let's have a look at 
the storage flux and  it's impact on the NSAE flux at STEI and TREE:

```{r plot-diel-stor}

ggplot(dfFlux, aes(x = hourLst, y = data.fluxCo2.stor.flux, fill = Site)) +
  geom_boxplot() +
  stat_summary(fun = median, geom = 'line', aes(group = Site, colour = Site)) +
  scale_y_continuous(limits = quantile(dfFlux$data.fluxCo2.stor.flux, c(0.001, 0.999), na.rm = TRUE))

```

The storage flux as expected is quite smaller than the turbulent flux, and is 
dominant during the nighttime hours and evening/morning transition times when 
turbulence is generally weaker and the trees and understory vegetation
aren't photosynthesizing. It's interesting to note the stronger stronger signal
at the TREE site.Let's look how that impacts the overall NSAE flux:

```{r plot-diel-nsae}

ggplot(dfFlux, aes(x = hourLst, y = data.fluxCo2.nsae.flux, fill = Site)) +
  geom_boxplot() +
  stat_summary(fun = median, geom = 'line', aes(group = Site, colour = Site)) +
  scale_y_continuous(limits = quantile(dfFlux$data.fluxCo2.nsae.flux, c(0.001, 0.999), na.rm = TRUE))

```

 When we look at the NSAE flux we see that the STEI site appears to sequester 
 more carbon, which makes sense considering the STEI site is predominately a 
 young Aspen stand for pulpwood that is growing at a fast rate. It's difficult 
 to get a good total Carbon uptake estimate without performing some more advanced 
 gap-filling and partitioning to better understand the ecosystem carbon dynamics. 
 We can simply look at the relative cumulative flux contributions:

```{r mean-nsae}


dfFlux %>% 
  select(contains("data.fluxCo2") | "Site") %>% 
      group_by(Site) %>% 
        summarise_each(funs(mean(., na.rm = TRUE)))


```


As expected from the plots, we see that on average the STEI site takes up more 
carbon on average across the growing season when we look at `nsae` and `turb`
CO<sub>2</sub> fluxes. This obviously comes with the tradeoff that the trees are
managed for cultivation in the future; whereas,  TREE is currently managed as
an educational resource that is likely to be preserved, which has other ecosystem
health benefits.

This analysis is meant to provide a glimpse into NEON flux data products, while 
focusing primarily on CO<sub>2</sub>. This workflow can easily be adapted to look
at Sensible and Latent heat (H<sub>2</sub>O) fluxes, or be adapted to evaluate 
additional NEON sites or years of data.



