---
syncID: 457a5c01d051491297dc947939b4210f
title: "Using the NEON API in R"
description: "Tutorial for getting data from the NEON API, using R and the R package httr"
dateCreated:  2017-07-07
authors: [Claire K. Lunch]
contributors: [Christine Laney, Megan A. Jones, Donal O'Leary]
estimatedTime: 1 - 1.5 hours
packagesLibraries: [httr, jsonlite, devtools, downloader, geoNEON, neonUtilities]
topics: data-management, rep-sci
languagesTool: R, API
dataProduct:
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/NEON-general/neon-code-packages/NEON-API-How-To/NEON-API-How-To.R
tutorialSeries: 
urlTitle: neon-api-usage
---



This is a tutorial in pulling data from the NEON API or Application 
Programming Interface. The tutorial uses R and the R package httr, but the core 
information about the API is applicable to other languages and approaches.


<div id="ds-objectives" markdown="1">

## Objectives

After completing this activity, you will be able to:

* Construct API calls to query the NEON API.
* Access and understand data and metadata available via the NEON API.

## Things You’ll Need To Complete This Tutorial
To complete this tutorial you will need the most current version of R and, 
preferably, RStudio loaded on your computer.

### Install R Packages

* **httr:** `install.packages("httr")`
* **jsonlite:** `install.packages("jsonlite")`

### Additional Resources

* <a href="http://data.neonscience.org/data-api" target="_blank">Webpage for the NEON API</a>
* <a href="https://github.com/NEONScience/neon-data-api" target="_blank">GitHub repository for the NEON API</a>
* <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download & Explore NEON Data</a> tutorial

</div>

## What is an API?

If you are unfamiliar with the concept of an API, think of  an API as a 
‘middle person' that provides a communication path for a software application 
to obtain information from a digital data source. APIs are becoming a very 
common means of sharing digital information. Many of the apps that you use on 
your computer or mobile device to produce maps, charts, reports, and other 
useful forms of information pull data from multiple sources using APIs. In 
the ecological and environmental sciences, many researchers use APIs to 
programmatically pull data into their analyses. (Quoted from the NEON Observatory
Blog story: 
<a href="https://www.neonscience.org/observatory/observatory-blog/api-data-availability-viewer-now-live-neon-data-portal" target ="_blank"> API and data availability viewer now live on the NEON data portal</a>.)

### What is accessible via the NEON API?

The NEON API includes endpoints for NEON data and metadata, including 
spatial data, taxonomic data, and samples (see Endpoints below). This 
tutorial explores these sources of information using a specific data 
product as a guide. The principles and rule sets described below can 
be applied to other data products and metadata.

## Anatomy of an API call

An example API call: http://data.neonscience.org/api/v0/data/DP1.10003.001/WOOD/2015-07

This includes the base URL, endpoint, and target.

### Base URL: 
<span style="color:#A00606;font-weight:bold">http://data.neonscience.org/api/v0</span><span style="color:#A2A4A3">/data/DP1.10003.001/WOOD/2015-07</span>

Specifics are appended to this in order to get the data or metadata you're 
looking for, but all calls to an API will include the base URL. For the NEON 
API, this is http://data.neonscience.org/api/v0 --
not clickable, because the base URL by itself will take you nowhere!

### Endpoints: 
<span style="color:#A2A4A3">http://data.neonscience.org/api/v0</span><span style="color:#A00606;font-weight:bold">/data</span><span style="color:#A2A4A3">/DP1.10098.001/WOOD/2015-07</span>

What type of data or metadata are you looking for?

* **~/products**
  Information about one or all of NEON's data products

* **~/sites**
  Information about data availability at the site specified in the call

* **~/locations**
  Spatial data for the NEON locations specified in the call

* **~/data**
  Data! By product, site, and date (in monthly chunks)
  
* **~/samples**
  Information about sample relationships and sample tracking
  
* **~/taxonomy**
  Access to NEON's taxon lists, the approved scientific names for sampled taxa

### Targets:
<span style="color:#A2A4A3">http://data.neonscience.org/api/v0/data</span><span style="color:#A00606;font-weight:bold">/DP1.10098.001/WOOD/2015-07</span>

The specific data product, location, sample, etc, you want to get data for.


## Data, by way of Products

Which product do you want to get data for? Consult the <a href="http://data.neonscience.org/data-products/explore" target="_blank">Explore Data Products page</a>.

We'll pick Woody vegetation structure, DP1.10098.001

Your first thought is probably to use the **/data** endpoint. And we'll get 
there. But notice above that the API call for the **/data** endpoint includes 
the site and month of data to download. You don't want to have to guess sites 
and months at random - first, you need to see which sites and months have 
available data for the product you're interested in. That can be done either 
through the **/sites** or the **/products** endpoint; here we'll use 
**/products**.

**Note:** Checking for data availability can sometimes be skipped for the 
streaming sensor data products. In general, they are available continuously, 
and you could theoretically query a site and month of interest and expect 
there to be data by default. However, there can be interruptions to sensor 
data, in particular at aquatic sites, so checking availability first is the 
most reliable approach.

Use the products endpoint to query for Woody vegetation data. The target is 
the data product identifier, noted above, DP1.10098.001:

```{r os-avail-query}

# Load the necessary libraries
library(httr)
library(jsonlite)

# Request data using the GET function & the API call
req <- GET("http://data.neonscience.org/api/v0/products/DP1.10098.001")
req

```

The object returned from `GET()` has many layers of information. Entering the 
name of the object gives you some basic information about what you accessed. 
`Status: 200` indicates this was a successful query; the status field can be 
a useful place to look if something goes wrong. These are HTTP status codes, 
you can google them to find out what a given value indicates.

The `Content-Type` parameter tells us we've accessed a json file. The easiest 
way to translate this to something more manageable in R is to use the 
`fromJSON()` function in the `jsonlite` package. It will convert the json into 
a nested list, flattening the nesting where possible.

```{r os-query-contents}

# Make the data readable by jsonlite
req.text <- content(req, as="text")

# Flatten json into a nested list
avail <- jsonlite::fromJSON(req.text, 
                            simplifyDataFrame=T, 
                            flatten=T)

```

A lot of the content here is basic information about the data product. 
You can see all of it by running the line `print(avail)`, but
this will result in a very long printout in your console. Instead, try viewing 
list items individually. Here, we highlight a couple of interesting examples:

```{r os-query-contents-examples}

# View description of data product
avail$data$productDescription

# View data product abstract
avail$data$productAbstract

```

You may notice that some of this information is also accessible on the NEON 
data portal. The portal uses the same data sources as the API, and in many 
cases the portal is using the API on the back end, and simply adding a more 
user-friendly display to the data.

We want to find which sites and months have available data. That is in the 
`siteCodes` section. Let's look at what information is presented for each 
site:

```{r os-query-fromJSON}

# Look at the first list element for siteCode
avail$data$siteCodes$siteCode[[1]]

# And at the first list element for availableMonths
avail$data$siteCodes$availableMonths[[1]]

```

Here we can see the list of months with data for the site ABBY, which is 
the Abby Road forest in Washington state.

The section `$data$siteCodes$availableDataUrls` provides the exact API 
calls we need in order to query the data for each available site and month. 

```{r os-query-avail-data}

# Get complete list of available data URLs
wood.urls <- unlist(avail$data$siteCodes$availableDataUrls)

# Total number of URLs
length(wood.urls)

# Show first 10 URLs available
wood.urls[1:10] 

```

These URLs are the API calls we can use to find out what files are available 
for each month where there are data. They are pre-constructed calls to the 
**/data** endpoint of the NEON API.

Let's look at the woody plant data from the Rocky Mountain National Park 
(RMNP) site from October 2019. We can do this by using the `GET()` function 
on the relevant URL, which we can extract using the `grep()` function. 

Note that if you want data from more than one site/month you need to iterate 
this code, `GET()` fails if you give it more than one URL at a time. 

```{r os-query-woody-data-urls}

# Get available data for RMNP Oct 2019
woody <- GET(wood.urls[grep("RMNP/2019-10", wood.urls)])
woody.files <- jsonlite::fromJSON(content(woody, as="text"))

# See what files are available for this site and month
woody.files$data$files$name

```

If you've downloaded NEON data before via the data portal or the 
`neonUtilities` package, this should look very familiar. The format 
for most of the file names is:

**NEON.[domain number].[site code].[data product ID].[file-specific name].
[year and month of data].[basic or expanded data package].
[date of file creation]**

Some files omit the year and month, and/or the data package, since they're 
not specific to a particular measurement interval, such as the data product 
readme and variables files. The date of file creation uses the 
ISO6801 format, for example 20210114T173951Z, and can be used to determine 
whether data have been updated since the last time you downloaded.

Available files in our query for October 2019 at Rocky Mountain are all of the 
following (leaving off the initial NEON.D10.RMNP.DP1.10098.001): 

* **~.vst_perplotperyear.2019-10.basic.20210114T173951Z.csv:** data table of 
measurements conducted at the plot level every year

* **~.vst_apparentindividual.2019-10.basic.20210114T173951Z.csv:** data table 
containing measurements and observations conducted on woody individuals

* **~.vst_mappingandtagging.basic.20210114T173951Z.csv:** data table 
containing mapping data for each measured woody individual. Note year and 
month are not in file name; these data are collected once per individual 
and provided with every month of data downloaded

* **~.categoricalCodes.20210114T173951Z.csv:** definitions of the values in 
categorical variables

* **~.readme.20210123T023002Z.txt:** readme for the data product (not specific 
to dates or location)

* **~.EML.20191010-20191017.20210123T023002Z.xml:** Ecological Metadata 
Language (EML) file, describing the data product

* **~.validation.20210114T173951Z.csv:** validation file for the data product, 
lists input data and data entry validation rules
  
* **~.variables.20210114T173951Z.csv:** variables file for the data product, 
lists data fields in downloaded tables

* **~.2019-10.basic.20210114T173951Z.zip:** zip of all files in the basic 
package. Pre-packaged zips are planned to be removed; may not appear in 
response to your query

This data product doesn't have an expanded package, so we only see the 
basic package data files, and only one copy of each of the metadata files.

Let's get the data table for the mapping and tagging data. The list of files 
doesn't return in the same order every time, so we shouldn't use the position 
in the list to select the file name we want. Plus, we want code we can re-use 
when getting data from other sites and other months. So we select file urls 
based on the data table name in the file names.


```{r os-get-mapandtag-data}

vst.maptag <- read.csv(woody.files$data$files$url
                       [grep("mappingandtagging",
                             woody.files$data$files$name)])

```

Note that if there were an expanded package, the code above would return 
two URLs. In that case you would need to specify the package as well in 
selecting the URL.

Now we have the data and can access it in R. Just to show that the file we 
pulled has actual data in it, let's make a quick graphic of the species 
present and their abundances:

```{r os-plot-woody-data}

# Get counts by species 
countBySp <- table(vst.maptag$taxonID)

# Reorder so list is ordered most to least abundance
countBySp <- countBySp[order(countBySp, decreasing=T)]

# Plot abundances
barplot(countBySp, names.arg=names(countBySp), 
        ylab="Total", las=2)

```

This shows us that the two most abundant species are designated with the 
taxon codes PICOL and POTR5. We can look back at the data table, check the 
**scientificName** field corresponding to these values, and see that these 
are lodgepole pine and quaking aspen, as we might expect in the eastern 
foothills of the Rocky mountains.

Let's say we're interested in how NEON defines quaking aspen, and 
what taxon authority it uses for its definition. We can use the 
**/taxonomy** endpoint of the API to do that.


## Taxonomy

NEON maintains accepted taxonomies for many of the taxonomic identification 
data we collect. NEON taxonomies are available for query via the API; they 
are also provided via an interactive user interface, the <a href="https://data.neonscience.org/taxonomic-lists" target="_blank">Taxon Viewer</a>.

NEON taxonomy data provides the reference information for how NEON 
validates taxa; an identification must appear in the taxonomy lists 
in order to be accepted into the NEON database. Additions to the lists 
are reviewed regularly. The taxonomy lists also provide the author 
of the scientific name, and the reference text used.

The taxonomy endpoint of the API has query options that are a bit more 
complicated than what was described in the "Anatomy of an API Call" 
section above. As described above, each endpoint has a single type of 
target - a data product number, a named location name, etc. For taxonomic 
data, there are multiple query options, and some of them can be used in 
combination. Instead of entering a single target, we specify the query 
type, and then the query parameter to search for. For example, a query 
for taxa in the Pinaceae family:

<span style="color:#A2A4A3">http://data.neonscience.org/api/v0/taxonomy</span><span style="color:#A00606;font-weight:bold">/?family=Pinaceae</span>

The available types of queries are listed in the <a href="https://data.neonscience.org/data-api/endpoints/taxonomy/" target="_blank">taxonomy section</a> 
of the API web page. Briefly, they are:

* `taxonTypeCode`: Which of the taxonomies maintained by NEON are you 
looking for? BIRD, FISH, PLANT, etc. Cannot be used in combination 
with the taxonomic rank queries.
* each of the major taxonomic ranks from genus through kingdom
* `scientificname`: Genus + specific epithet (+ authority). Search is 
by exact match only, see final example below.
* `verbose`: Do you want the short (`false`) or long (`true`) response
* `offset`: Skip this number of items at the start of the list.
* `limit`: Result set will be truncated at this length.

For the first example, let's query for the loon family, Gaviidae, in the 
bird taxonomy. Note that query parameters are case-sensitive.

```{r get-loons}

loon.req <- GET("http://data.neonscience.org/api/v0/taxonomy/?family=Gaviidae")

```

Parse the results into a list using `fromJSON()`:

```{r parse-loons}

loon.list <- jsonlite::fromJSON(content(loon.req, as="text"))

```

And look at the `$data` element of the results, which contains:

* The full taxonomy of each taxon
* The short taxon code used by NEON (taxonID/acceptedTaxonID)
* The author of the scientific name (scientificNameAuthorship)
* The vernacular name, if applicable
* The reference text used (nameAccordingToID)

The terms used for each field are matched to Darwin Core (dwc) and 
the Global Biodiversity Information Facility (gbif) terms, where 
possible, and the matches are indicated in the column headers.

```{r display-loons}

loon.list$data

```

To get the entire list for a particular taxonomic type, use the 
`taxonTypeCode` query. Be cautious with this query, the PLANT taxonomic 
list has several hundred thousand entries.

For an example, let's look up the small mammal taxonomic list, which 
is one of the shorter ones, and use the `verbose=true` option to see 
a more extensive list of taxon data, including many taxon ranks that 
aren't populated for these taxa. For space here, we'll display only 
the first 10 taxa:

```{r get-mammals}

mam.req <- GET("http://data.neonscience.org/api/v0/taxonomy/?taxonTypeCode=SMALL_MAMMAL&verbose=true")
mam.list <- jsonlite::fromJSON(content(mam.req, as="text"))
mam.list$data[1:10,]

```

Now let's go back to our question about quaking aspen. To get 
information about a single taxon, use the `scientificname` 
query. This query will not do a fuzzy match, so you need to query 
the exact name of the taxon in the NEON taxonomy. Because of this, 
the query will be most useful in cases like the current one, where 
you already have NEON data in hand and are looking for more 
information about a specific taxon. Querying on `scientificname` 
is unlikely to be an efficient way to figure out if NEON recognizes 
a particular taxon.

In addition, scientific names contain spaces, which are not 
allowed in a URL. The spaces need to be replaced with the URL 
encoding replacement, %20.

Looking up the POTR5 data in the woody vegetation product, we 
see that the scientific name is Populus tremuloides Michx. 
This means we need to search for `Populus%20tremuloides%20Michx.` 
to get the exact match.

```{r get-aspen}

aspen.req <- GET("http://data.neonscience.org/api/v0/taxonomy/?scientificname=Populus%20tremuloides%20Michx.")
aspen.list <- jsonlite::fromJSON(content(aspen.req, as="text"))
aspen.list$data

```

This shows us the definition for Populus tremuloides Michx. does 
not include a subspecies or variety, and the authority for the 
taxon information (`nameAccordingToID`) is the USDA PLANTS 
database. This means NEON taxonomic definitions are aligned with 
the USDA, and is true for the large majority of plants in the 
NEON taxon system.


## Spatial data

How to get spatial data and what to do with it depends on which type of 
data you're working with.

#### Instrumentation data (both aquatic and terrestrial)
The sensor\_positions files, which are included in the list of available files, 
contain spatial coordinates for each sensor in the data. See the final section 
of the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/neon-spatial-data-basics" target="_blank">Geolocation tutorial</a> for guidance in using these files.

#### Observational data - Aquatic
Latitude, longitude, elevation, and associated uncertainties are included in 
data downloads. Most products also include an "additional coordinate uncertainty" 
that should be added to the provided uncertainty. Additional spatial data, such 
as northing and easting, can be downloaded from the API.

#### Observational data - Terrestrial
Latitude, longitude, elevation, and associated uncertainties are included in 
data downloads. These are the coordinates and uncertainty of the sampling plot; 
for many protocols it is possible to calculate a more precise location. 
Instructions for doing this are in the respective data product user guides, and 
code is in the `geoNEON` package on GitHub.

### Querying a single named location

Let's look at a specific sampling location in the woody vegetation structure 
data we downloaded above. To do this, look for the field called `namedLocation`, 
which is present in all observational data products, both aquatic and 
terrestrial. This field matches the exact name of the location in the NEON 
database.

```{r get-woody-NLs}

head(vst.maptag$namedLocation)

```

Here we see the first six entries in the `namedLocation` column, which tells us
the names of the Terrestrial Observation plots where the woody plant surveys 
were conducted. 

We can query the locations endpoint of the API for the first named location, 
`RMNP_043.basePlot.vst`. 

```{r woody-ex-NL}

req.loc <- GET("http://data.neonscience.org/api/v0/locations/RMNP_043.basePlot.vst")
vst.RMNP_043 <- jsonlite::fromJSON(content(req.loc, as="text"))
vst.RMNP_043

```

Note spatial information under `$data$[nameOfCoordinate]` and under 
`$data$locationProperties`. These coordinates represent the centroid of 
plot RMNP\_043, and should match the coordinates for the plot in the 
`vst_perplotperyear` data table. In rare cases, spatial data may be 
updated, and if this has happened more recently than the data table 
was published, there may be a small mismatch in the coordinates. In 
those cases, the data accessed via the API will be the most up to date.

Also note `$data$locationChildren`: these are the 
point and subplot locations within plot RMNP\_043. And 
`$data$locationChildrenUrls` provides pre-constructed API calls for 
querying those child locations. Let's look up the location data for 
point 31 in plot RMNP\_043.

```{r woody-child-NL}

req.child.loc <- GET(grep("31", 
                          vst.RMNP_043$data$locationChildrenUrls,
                          value=T))
vst.RMNP_043.31 <- jsonlite::fromJSON(content(req.child.loc, as="text"))
vst.RMNP_043.31

```

Looking at the easting and northing coordinates, we can see that point 
31 is about 10 meters away from the plot centroid in both directions. 
Point 31 has no child locations.

For the records with `pointID=31` in the `vst.maptag` table we 
downloaded, these coordinates are the reference location that could be 
used together with the `stemDistance` and `stemAzimuth` fields to 
calculate the precise locations of individual trees. For detailed 
instructions in how to do this, see the data product user guide.

Alternatively, the `geoNEON` package contains functions to make this 
calculation for you, including accessing the location data from the 
API. See below for details and links to tutorials.


## R Packages

NEON provides two customized R packages that can carry out many of the 
operations described above, in addition to other data transformations. 

### neonUtilities

The `neonUtilities` package contains functions that download data 
via the API, and that merge the individual files for each site and 
month in a single continuous file for each type of data table in the 
download.

For a guide to using `neonUtilities` to download and stack data files, 
check out the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore</a> tutorial.

### geoNEON

The `geoNEON` package contains functions that access spatial data 
via the API, and that calculate precise locations for terrestrial 
observational data. As in the case of woody vegetation structure, 
terrestrial observational data products are published with 
spatial data at the plot level, but more precise sampling locations 
are usually possible to calculate.

For a guide to using `geoNEON` to calculate sampling locations, 
check out the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/neon-spatial-data-basics" target="_blank">Geolocation</a> tutorial.
