---
title: Content for intro to continuous discharge tutorial
output: html_fragment
dateCreated: '2025-08-08'
editor_options: 
  markdown: 
    wrap: 72
---

This tutorial covers...

# Introduction to the NEON Continuous Discharge Data Product

::: {#ds-objectives markdown="1"}

## Objectives

After completing this activity, you will be able to:

-   Download and explore the contents of the NEON Continuous discharge data 
    product (DP4.00130.001)
-   Understand how the publication history of this data product influences the
    tables available for download
-   Create a single continuous time series of discharge data for a site via 
    data aggregation and table joining
-   Plot continuous time series of stage and discharge with associated
    uncertainties for the lifetime of a site

## Things You’ll Need To Complete This Tutorial

You can follow either the R or Python code throughout this tutorial.

-   For R users, we recommend using R version 4+ and RStudio.
-   For Python users, we recommend using Python 3.9+.

```{r setup, include=FALSE}
library(reticulate)
knitr::opts_chunk$set(dev="png")
```

## Set up: Install Packages {.tabset}

Packages only need to be installed once, you can skip this step after
the first time:

### R

-   **neonUtilities**: Basic functions for accessing NEON data
-   **tidyverse**: Collection of R packages designed for data science

```{r install, eval=FALSE}
install.packages("neonUtilities",
                 "tidyverse")
```

### Python

-   **os**: Module allowing interaction with user's operating system
-   **neonutilities**: Module for working with data frames
-   **neonutilities**: Basic functions for accessing NEON data
-   **matplotlib**: Functions for plotting

```{python p-install, eval=FALSE}
pip install os
pip install pandas
pip install neonutilities
pip install matplotlib
```

##  {.unnumbered}

### Additional Resources

-   <a href="https://www.neonscience.org/neon-utilities-python" target="_blank">Tutorial
    for using neonUtilities from a Python environment.</a>
-   <a href="https://github.com/NEONScience/NEON-Utilities/neonUtilities" target="_blank">GitHub
    repository for neonUtilities</a>
-   <a href="https://www.neonscience.org/sites/default/files/cheat-sheet-neonUtilities_0.pdf" target="_blank">neonUtilities
    cheat sheet</a>. A quick reference guide for users.
:::

## Set up: Load Packages {.tabset}

### R

```{r R-library, results="hide", message=FALSE}
library(neonUtilities)
library(tidyverse)
```

### Python

```{python p-import, eval=FALSE}
import os
import pandas as pd
import neonutilities as nu
import matplotlib.pyplot as plt
```

##  {.unnumbered}

## Set NEON Data Portal Personal Access Token (PAT) {.tabset}

It is recommended that NEON data users have a NEON Data Portal personal
access token (PAT) set as an environment variable. See instructions on
obtaining a NEON PAT at
<https://www.neonscience.org/resources/learning-hub/tutorials/neon-api-tokens-tutorial>

### R

```{r R-envvar, results="hide", message=FALSE}
Sys.setenv(NEON_PAT="YOUR_PAT")
```

### Python

```{python p-envvar, eval=FALSE}
os.environ.setdefault('NEON_PAT',"YOUR_PAT")
```

##  {.unnumbered}

## Download Data Product {.tabset}

In this tutorial, we will focus on a single site, Lower Hop Brook (HOPB)
in NEON Domain 01, but the workflow can be replicated for any site at
which continuous discharge is published by changing the siteID.

Visit the [NEON Data Portal landing page for Continuous discharge](https://data.neonscience.org/data-products/DP4.00130.001) 
for an availability chart showing all the sites for which Continuous discharge 
is published and the period of record for each site. 

For this tutorial we will download all data across the lifetime of a site 
by commenting out the start and end date in the function. If you wish to 
query a shorter period of time they can be un-commented. The NEON Data 
Portal stores data in site\*month download packages, so date range inputs 
are formatted as YYYY-MM.

For this tutorial we will also download both RELEASED and PROVISIONAL data 
by setting the include.provisinal argument to TRUE in the function. To 
download only RELEASED data this argument should be set to FALSE.
To learn more about the differences between released and provisional data, visit
<https://www.neonscience.org/resources/learning-hub/tutorials/release-provisional-tutorial>.

### R

```{r R-download, results="hide", message=FALSE}
# Download continuous discharge data across the lifetime of a site
csd <- neonUtilities::loadByProduct(dpID="DP4.00130.001",
                                    site="HOPB",
                                    # startdate="2020-10",
                                    enddate="2021-09",# **ZN** Set an end date while testing with a local 15 minute table; comment out in final tutorial
                                    release="current",# Downloads data from the most recent release
                                    include.provisional = T,# Includes provisional data
                                    package='expanded',
                                    check.size = F,
                                    token=Sys.getenv("NEON_PAT"))

# Add in the 15-minute table because it is not yet published
csd$csd_15_min <- read.csv("C:/Users/nickerson/Documents/GitHub/NEON-Data-Skills/tutorials-in-development/R/aquatics/L4-Discharge-Series/continuous-discharge-basic/csd_15_min.csv", 
                           encoding = "UTF-8",
                           header = T) # **NH** I'm confused by this step, probably because I'm just not getting it.  Where does do they get this .CSV?  Eventually users will be able to just get this through the neonUtils download, right?
```

### Python

```{python p-download, eval=FALSE}
# Download continuous discharge data across the lifetime of a site
csd = nu.load_by_product(dpid="DP4.00130.001",
                         site="HOPB",
                         # start_date="2020-10",
                         end_date="2022-09",# Set an end date while testing with a local 15 minute table; comment out in final tutorial
                         release="current",
                         include_provisional=True,
                         package="expanded",
                         check_size=False,
                         token=os.environ.get("NEON_PAT"))

# Add in the 15-minute table because it is not yet published
csd['csd_15_min'] = pd.read_csv("C:/Users/nickerson/Documents/GitHub/NEON-Data-Skills/tutorials-in-development/R/aquatics/L4-Discharge-Series/continuous-discharge-basic/csd_15_min.csv",
                                encoding="UTF-8")
csd['csd_15_min']['endDateTime'] = pd.to_datetime(csd['csd_15_min']['endDateTime'], format='mixed', utc=True)
```

##  {.unnumbered}

## Navigate Data Download

Downloads from the NEON Utilities R package and Python module contain multiple 
files, including data tables, metadata, and data product documentation. Let's
explore each set of files in turn.

### Files Associated with Downloads {.tabset}

The data we've downloaded comes as an object that is a named list/dictionary of 
objects. Let's view the contents of the download package.

#### R

```{r R-files, results="hide", message=FALSE}
# Get all file names in the download package
names(csd)
```

#### Python

```{python p-files, eval=FALSE}
# Get all file names in the download package
csd.keys()
```

In this tutorial, we downloaded the `expanded` download package. What are the 
files contained in this download package and why are they useful?

-   **csd_data_tables**: Includes the primary data tables of the continuous 
    discharge data product. We will dive deeper into data tables in the next 
    section.
-   **sensor_positions_00130**: Reports the geolocation of each sensor included 
    in the download.
-   **science_review_flags_00130**: Lists each science review flag (SRF) date 
    range, flag value, and justification applied to the data included in this 
    download.
-   **issueLog_00130**: Reports issues that may impact data quality, or changes 
    to a data product that affects all sites.
-   **variables_00130**: This file contains all the variables found in the data 
    table(s) included in this download. This includes full definitions, units, 
    and other important information.
-   **categoricalCodes_00130**: Some variables in the data tables are published 
    as strings and constrained to a standardized list of values (LOV). This file
    shows all the LOV options for variables published in this data product.
-   **readme_00130**: The readme file provides important information relevant to
    the data product and the specific instance of downloading the data.
-   **data citations**: At least one data citation file is included in each data
    download, depending on the `release` function inputs and date range queried.

###  {.unnumbered}

### Explore Data Tables {.tabset}

The basic download package for DP4.00130.001 contains two data tables that 
report continuous discharge time series data modeled from in situ pressure 
transducers and stage-discharge rating curves:

-   `csd_15_min`: Continuous discharge data averaged to 15-minute intervals. 
    This table began being published in RELEASE-2026 and is available, for a 
    given site, starting with the earliest date of corrected data.
-   `csd_continuousDischarge`: Instantaneous (1-minute) continuous discharge 
    data. Beginning with RELEASE-2026, this table is available only before the 
    earliest date of data correction for a given site.

The expanded download package for DP4.00130.001 contains multiple
additional tables that aid in understanding data relationships, model
coefficients, uncertainty propagation, and data corrections.

-   `sdrc_gaugePressureRelationship`: Reports the relationship between measured 
    gauge heights and calculated stage values derived from gauge height‐water 
    column height linear regressions.
-   `csd_gaugeWaterColumnHeightRegression`: Reports the linear regression 
    coefficients for each unique gauge height-water column height regression 
    model developed to estimate stage from surface water pressure data.
-   `csd_dataGapToFillMethodMapping`: Reports periods of gaps or erroneous data 
    in uncorrected continuous discharge data and the method used to correct the 
    period.
-   `csd_constantBiasShift`: Reports periods of continuous discharge data 
    corrected via constant bias shift and the magnitude of the shift.
-   `csd_gapFillingRegression`: Reports the linear regression coefficients for
    each unique relationship developed to correct NEON continuous discharge data
    using the method specified in the mapping table.

At the Domain 08 - Tombigbee River (TOMB) site, a stage-discharge relationship 
cannot be developed due to the influence of a downstream lock and dam system. 
Therefore, the above tables are not published for TOMB. Rather, the following 
tables are published for TOMB.

-   `csd_continuousDischargeUSGS`: USGS discharge and associated uncertainty
    based on the fit of USGS data with corresponding discharge measurements 
    collected at TOMB.
-   `csd_dischargeRegressionUSGS`: Reports the linear regression coefficients 
    for each unique relationship developed to estimate NEON discharge from USGS 
    discharge.

Below, we will explore the first few rows of each high-frequency time series 
table included in the basic download package. Add to the code below to also view
those tables included in the expanded download package.

#### R

```{r R-tables, results="hide", message=FALSE}
# Print the first 5 records in each of the time series tables to view structure
print("First 5 rows of csd_continuousDischarge")
head(csd$csd_continuousDischarge)

print("First 5 rows of csd_15_min")
head(csd$csd_15_min)
```

#### Python

```{python p-tables, eval=FALSE}
# Print the first 5 records in each of the time series tables to view structure
print("First 5 rows of csd_continuousDischarge")
print(csd['csd_continuousDischarge'].head())

print("First 5 rows of csd_15_min")
print(csd['csd_15_min'].head())
```

###  {.unnumbered}

### Explore Variables {.tabset}

The `variables_00130` file provides insight into the structure of each data 
table and associated variables included in a download package. For the 
Continuous discharge data product, it is crucial to understand the relationships
between the **csd_continuousDischarge** and **csd_15_min** tables to aggregate 
the two tables and produce a continuous timeseries. That is an exercise we will 
conduct later in this tutorial. For now, view the variables file and familiarize 
yourself with the different fields, data types, and units used in this data 
product.

#### R

```{r R-variables, results="hide", message=FALSE}
# View variables file to understand data table structure
View(csd$variables_00130)
```

#### Python

```{python p-variables, eval=FALSE}
# View variables file to understand data table structure
display(csd['variables_00130'])
```

###  {.unnumbered}

##  {.unnumbered}

## Wrangle Stage & Discharge Data

This data product was updated from uncorrected 1 minute values to corrected
15-min averages. Until these updates are fully applied backwards in time,
some data wrangling and table joining are required to 
produce a continuous time series of continuous discharge. The publication 
history of this data product is explained in detail in the NEON User Guide to 
Continuous Discharge (NEON.DP4.00130.001), which is published in the 
"Documentation" section of the [NEON Data Portal Landing Page](https://data.neonscience.org/data-products/DP4.00130.001)
for this data product.

In brief, the `csd_15_min` table was first published beginning in 2026 (data 
included in RELEASE-2026), and was published for data beginning (for most sites)
in 2021-10 (beginning of WY 2022). Prior to the first date published for a site
in `csd_15_min`, the `csd_continuousDischarge` table is published. As more WYs 
are corrected prior to WY 2022, records in `csd_continuousDischarge` will be 
removed from publication and replaced with corrected records in `csd_15_min`.

A good rule of thumb for managing NEON continuous discharge data:

| Table | Frequency | Record Description |
|:----------------------:|:----------------------:|:----------------------:|
| csd_continuousDischarge | 1 minute | uncorrected data unchanged since RELEASE-2025 |
| csd_15_min | 15 minute | if `release` = RELEASE-YYYY; corrected data first published in RELEASE-2026 |
| csd_15_min | 15 minute | if `release` = PROVISIONAL; uncorrected data processed within new pipeline | #' The last part of this line is not technically correct.  They need to look at the corection flag to see whether provisonal data is still uncorrected or not.

### Aggregate 1-minute Table to 15-minutes {.tabset}

The first step to generating a continuous period of record for a site is to 
aggregate the uncorrected 1-minute `csd_continuousDischarge` data to a 15-minute
interval. We will use this opportunity to conform the column headers of the 
aggregated table we create to fit those of `csd_15_min`.

#### R

```{r R-aggregate, results="hide", message=FALSE}
# Use the tidyverse suite to round, group, and summarize data
# Note: new column names match those published in 'csd_15_min'
csd_15_min_agg <- csd$csd_continuousDischarge%>%
  dplyr::mutate(endDateTime=lubridate::round_date(endDate,"15 min"))%>%
  dplyr::group_by(siteID,endDateTime)%>%
  dplyr::summarise(stageContinuous=mean(equivalentStage,
                                        na.rm=T),
                   stageTotalUncert=mean(stageUnc,
                                         na.rm=T),
                   dischargeContinuous=mean(maxpostDischarge,
                                            na.rm=T),
                   dischargeUpperParamUncert=mean(withParaUncQUpper2Std,
                                                  na.rm=T),
                   dischargeLowerParamUncert=mean(withParaUncQUpper1Std,
                                                  na.rm=T),
                   dischargeUpperRemnUncert=mean(withRemnUncQUpper2Std,
                                                 na.rm=T),
                   dischargeLowerRemnUncert=mean(withRemnUncQLower2Std,
                                                 na.rm=T),
                   dischargeFinalQF=sum(dischargeFinalQF==1),
                   dischargeFinalQFSciRvw=sum(dischargeFinalQFSciRvw==1,
                                              na.rm = T)
)

# Convert quality flag fields to 1 or NA
csd_15_min_agg$dischargeFinalQF[
  csd_15_min_agg$dischargeFinalQF>0
] <- 1
csd_15_min_agg$dischargeFinalQFSciRvw[
  csd_15_min_agg$dischargeFinalQFSciRvw>0
] <- 1
csd_15_min_agg$dischargeFinalQFSciRvw[
  csd_15_min_agg$dischargeFinalQFSciRvw==0
] <- NA
```

#### Python

```{python p-aggregate, eval=FALSE}
# Use pandas to round, group, and summarize data
# Note: new column names match those published in 'csd_15_min'

# Convert endDate to datetime and round to 15 minutes
csd['csd_continuousDischarge']['endDate'] = pd.to_datetime(csd['csd_continuousDischarge']['endDate'])
csd['csd_continuousDischarge']['endDateTime'] = csd['csd_continuousDischarge']['endDate'].dt.round('15min')

# Group by siteID and endDateTime, then calculate aggregations
csd_15_min_agg = csd['csd_continuousDischarge'].groupby(['siteID', 'endDateTime']).agg({
    'equivalentStage': 'mean',
    'stageUnc': 'mean',
    'maxpostDischarge': 'mean',
    'withParaUncQUpper2Std': 'mean',
    'withParaUncQUpper1Std': 'mean',
    'withRemnUncQUpper2Std': 'mean',
    'withRemnUncQLower2Std': 'mean',
    'dischargeFinalQF': lambda x: (x == 1).sum(),
    'dischargeFinalQFSciRvw': lambda x: (x == 1).sum()
}).reset_index()

# Rename columns to match csd_15_min format
csd_15_min_agg = csd_15_min_agg.rename(columns={
    'equivalentStage': 'stageContinuous',
    'stageUnc': 'stageTotalUncert',
    'maxpostDischarge': 'dischargeContinuous',
    'withParaUncQUpper2Std': 'dischargeUpperParamUncert',
    'withParaUncQUpper1Std': 'dischargeLowerParamUncert',
    'withRemnUncQUpper2Std': 'dischargeUpperRemnUncert',
    'withRemnUncQLower2Std': 'dischargeLowerRemnUncert'
})

# Convert quality flag fields to 1 or NaN
csd_15_min_agg.loc[csd_15_min_agg['dischargeFinalQF'] > 0, 'dischargeFinalQF'] = 1
csd_15_min_agg.loc[csd_15_min_agg['dischargeFinalQFSciRvw'] > 0, 'dischargeFinalQFSciRvw'] = 1
csd_15_min_agg.loc[csd_15_min_agg['dischargeFinalQFSciRvw'] == 0, 'dischargeFinalQFSciRvw'] = pd.NA
```

###  {.unnumbered}

### Join Tables {.tabset}

Because we conformed column headers in the uncorrected aggregated data, it 
should now be relatively easy to join the data to the `csd_15_min`. Let's now 
create a single data frame representing the entire period of record aggregated 
to a 15-minute interval.

#### R

```{r R-join, results="hide", message=FALSE}
# Merge aggregated table created earlier with csd_15_min from the download
csd_15_min_all <- merge(csd_15_min_agg,
                        csd$csd_15_min,
                        all = T)

# Order the timeseries data by endDateTime
csd_15_min_all <- csd_15_min_all[order(csd_15_min_all$endDateTime),]
```

#### Python

```{python p-join, eval=FALSE}
# Merge aggregated table created earlier with csd_15_min from the download
csd_15_min_all = pd.merge(csd_15_min_agg,
                          csd['csd_15_min'],
                          how='outer')

# Order the timeseries data by endDateTime
csd_15_min_all = csd_15_min_all.sort_values(by='endDateTime')
```

##  {.unnumbered}

## Plot Stage Data {.tabset}

Continuous stage is modeled by developing a linear regression between empirical
staff gauge heights (DP1.20267.001) and continuous water column height
calculated using surface water pressure data (DP1.20016.001). Uncertainty is estimated for continuous
stage by summing uncertainty associated with the calibration of the pressure 
transducer (termed: nonsystematic) and uncertainty associated with the fit of the 
linear regression (termed: systematic).

More details on the modeling algorithm for continuous stage can be found in the
ATBD for this data product (NEON.DOC.005403).

Here, we will plot the continuous stage time series for the entire period of
record, then plot a subset of data to zoom in on the time series to get a better
visual of the associated uncertainty.

#' **BH** Could we also include the staff gage measurments in these plots, liek we do for the test reports?

### R

```{r R-plot-stage, results="hide", message=FALSE, warning=FALSE}
# Use the ggplot package in tidyverse to plot continuous stage and uncertainty

# Plot 1: Entire Period of Record
csd_15_min_all%>%
  ggplot2::ggplot(aes(x=endDateTime))+
  ggplot2::geom_ribbon(aes(ymin=stageContinuous-stageTotalUncert,
                           ymax=stageContinuous+stageTotalUncert),
                       fill="grey70")+
  ggplot2::geom_line(aes(y=stageContinuous))+
  ggplot2::theme_classic()+
  ggplot2::labs(title = "Stage Series: Entire Period of Record",
                x="Date",
                y="Stage (m)")

# Plot 2: Zoomed in to March 2024
csd_15_min_all%>%
  dplyr::filter(endDateTime>="2024-03-01"
                &endDateTime<"2024-04-01")%>%
  ggplot2::ggplot(aes(x=endDateTime))+
  ggplot2::geom_ribbon(aes(ymin=stageContinuous-stageTotalUncert,
                           ymax=stageContinuous+stageTotalUncert),
                       fill="grey70")+
  ggplot2::geom_line(aes(y=stageContinuous))+
  ggplot2::theme_classic()+
  ggplot2::labs(title = "Stage Series: Zoomed In to View Uncertainty - 2024-03",
                x="Date",
                y="Stage (m)")
```

### Python

```{python p-plot-stage, eval=FALSE}
# Use matplotlib to plot continuous stage and uncertainty

# Plot 1: Entire Period of Record
fig, ax = plt.subplots(figsize=(12, 6))
ax.fill_between(csd_15_min_all['endDateTime'], 
                csd_15_min_all['stageContinuous'] - csd_15_min_all['stageTotalUncert'],
                csd_15_min_all['stageContinuous'] + csd_15_min_all['stageTotalUncert'],
                color='grey', alpha=0.7, label='Uncertainty')
ax.plot(csd_15_min_all['endDateTime'], csd_15_min_all['stageContinuous'], 
        color='black', linewidth=1)
ax.set_title('Stage Series: Entire Period of Record')
ax.set_xlabel('Date')
ax.set_ylabel('Stage (m)')
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()

# Plot 2: Zoomed in to March 2024
filtered_data = csd_15_min_all[
    (csd_15_min_all['endDateTime'] >= '2024-03-01') & 
    (csd_15_min_all['endDateTime'] < '2024-04-01')
]
fig, ax = plt.subplots(figsize=(12, 6))
ax.fill_between(filtered_data['endDateTime'], 
                filtered_data['stageContinuous'] - filtered_data['stageTotalUncert'],
                filtered_data['stageContinuous'] + filtered_data['stageTotalUncert'],
                color='grey', alpha=0.7, label='Uncertainty')
ax.plot(filtered_data['endDateTime'], filtered_data['stageContinuous'], 
        color='black', linewidth=1)
ax.set_title('Stage Series: Zoomed In to View Uncertainty - 2024-03')
ax.set_xlabel('Date')
ax.set_ylabel('Stage (m)')
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()
```

##  {.unnumbered}

## Plot Discharge Data {.tabset}

Continuous discharge is modeled within a Bayesian modeling framework using an
executable (BaM) first developed and published by [Le Coz et al. in 2014](https://doi.org/10.1016/j.jhydrol.2013.11.016).
The continuous stage series and inputs from the Stage-discharge rating curves
data product (DP4.00133.001) are used to model continuous discharge. Two types
of uncertainty are associated with continuous discharge:

- Parametric uncertainty: derived from the uncertainty of the model priors
- Remnant uncertainty: derived from how the model fits the observations

More details on the modeling algorithm for continuous discharge be found in the
ATBD for this data product (NEON.DOC.005403).

Here, we will plot the continuous discharge time series for the entire period of
record, then plot a subset of data to zoom in on the time series to get a better
visual of the associated uncertainties.

#' **BH** Similar to stage, could we also include the emperical discharge measuremnts in these plots?

### R

```{r R-plot-discharge, results="hide", message=FALSE, warning=FALSE}
# Use the ggplot package in tidyverse to plot continuous discharge and uncertainty

# Plot 1: Entire Period of Record
csd_15_min_all%>%
  ggplot2::ggplot(aes(x=endDateTime))+
  ggplot2::geom_ribbon(aes(ymin=dischargeLowerRemnUncert,
                           ymax=dischargeUpperRemnUncert),
                       fill="#D55E00")+
  ggplot2::geom_ribbon(aes(ymin=dischargeLowerParamUncert,
                           ymax=dischargeUpperParamUncert),
                       fill="#E69F00")+
  ggplot2::geom_line(aes(y=dischargeContinuous))+
  ggplot2::theme_classic()+
  ggplot2::labs(title = "Discharge Series: Entire Period of Record",
                x="Date",
                y="Discharge (L/s)")

# Plot 2: Zoomed in to March 2024
csd_15_min_all%>%
  dplyr::filter(endDateTime>="2024-03-01"
                &endDateTime<"2024-04-01")%>%
  ggplot2::ggplot(aes(x=endDateTime))+
  ggplot2::geom_ribbon(aes(ymin=dischargeLowerRemnUncert,
                           ymax=dischargeUpperRemnUncert),
                       fill="#D55E00")+
  ggplot2::geom_ribbon(aes(ymin=dischargeLowerParamUncert,
                           ymax=dischargeUpperParamUncert),
                       fill="#E69F00")+
  ggplot2::geom_line(aes(y=dischargeContinuous))+
  ggplot2::theme_classic()+
  ggplot2::labs(title = "Discharge Series: Zoomed In to View Uncertainty - 2024-03",
                x="Date",
                y="Discharge (L/s)")
```

### Python

```{python p-plot-discharge, eval=FALSE}
# Use matplotlib to plot continuous discharge and uncertainty

# Plot 1: Entire Period of Record
fig, ax = plt.subplots(figsize=(12, 6))
ax.fill_between(csd_15_min_all['endDateTime'], 
                csd_15_min_all['dischargeLowerRemnUncert'],
                csd_15_min_all['dischargeUpperRemnUncert'],
                color='#D55E00', alpha=0.7, label='Remnant Uncertainty')
ax.fill_between(csd_15_min_all['endDateTime'], 
                csd_15_min_all['dischargeLowerParamUncert'],
                csd_15_min_all['dischargeUpperParamUncert'],
                color='#E69F00', alpha=0.7, label='Parametric Uncertainty')
ax.plot(csd_15_min_all['endDateTime'], csd_15_min_all['dischargeContinuous'], 
        color='black', linewidth=1)
ax.set_title('Discharge Series: Entire Period of Record')
ax.set_xlabel('Date')
ax.set_ylabel('Discharge (L/s)')
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()

# Plot 2: Zoomed in to March 2024
filtered_data = csd_15_min_all[
    (csd_15_min_all['endDateTime'] >= '2024-03-01') & 
    (csd_15_min_all['endDateTime'] < '2024-04-01')
]
fig, ax = plt.subplots(figsize=(12, 6))
ax.fill_between(filtered_data['endDateTime'], 
                filtered_data['dischargeLowerRemnUncert'],
                filtered_data['dischargeUpperRemnUncert'],
                color='#D55E00', alpha=0.7, label='Remnant Uncertainty')
ax.fill_between(filtered_data['endDateTime'], 
                filtered_data['dischargeLowerParamUncert'],
                filtered_data['dischargeUpperParamUncert'],
                color='#E69F00', alpha=0.7, label='Parametric Uncertainty')
ax.plot(filtered_data['endDateTime'], filtered_data['dischargeContinuous'], 
        color='black', linewidth=1)
ax.set_title('Discharge Series: Zoomed In to View Uncertainty - 2024-03')
ax.set_xlabel('Date')
ax.set_ylabel('Discharge (L/s)')
plt.xticks(rotation=45)
plt.tight_layout()
plt.show()
```

##  {.unnumbered}
