---
syncID: e606b7e7bfb845b39b931bcd772903a1
title: "Plotting and Clustering Megapit Soils Data"
description: "Learn to download, analyze, and visualize NEON soils data."
dateCreated: 2020-06-30
authors: Donal O'Leary, Claire Lunch
contributors: 
estimatedTime: 45 minutes
packagesLibraries: neonUtilities
topics: data-management, rep-sci
languageTool: R, API
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/soils/megapit-data/megapit-profiles/megapit-profiles.R
tutorialSeries:
urlTitle: plot-cluster-megapit-soils-data
---

This tutorial will show you how to download NEON Megapit soils data, plot soil profiles by texture and chemical properties, and cluster the different megapit profiles according to their similarity across multiple dimensions. 

<div id="ds-objectives" markdown="1">

## Objectives

After completing this activity, you will be able to:

* Download NEON megapit data using the `neonUtilities` package.
* Join megapit data tables
* Plot profiles of megapit data by horizon
* Cluster sites into groups based on physical and chemical properties

## Things You'll Need To Complete This Tutorial
To complete this tutorial you will need R (version 4+) and, 
preferably, RStudio loaded on your computer.

### Install R Packages

* **neonUtilities**: Basic functions for accessing NEON data
* **dplyr**: Data manipulation functions
* **aqp**: Algorithms for Quantitative Pedology
* **cluster**: Clustering utilities
* **sharpshootR**: Plotting tools for clustered data
* **Ternary**: Tools for making Ternary plots

These packages are on CRAN and can be installed by 
`install.packages()`.

### Additional Resources

* <a href="https://data.neonscience.org/megapit-info" target="_blank">Images and narrative descriptions of megapit soils</a>

</div>

## Set up code environment

Before we get the data, we need to install (if not already done) and load 
the R packages needed for data load and analysis. 

As of June 2026, NEON requires an API token for data downloads, to reduce bot 
scraping and improve user support. Tokens can be generated in NEON data portal 
user accounts - log in to your account or create one, and go to the API Tokens 
section. For best practices in storing and using tokens, follow the 
instructions <a href="https://www.neonscience.org/resources/learning-hub/tutorials/api-token-setup" target="_blank">here</a>.

Install packages:

```{r set-up-env, eval=F}

install.packages("neonUtilities")
install.packages("neonOS")
install.packages("aqp")
install.packages("cluster")
install.packages("sharpshootR")
install.packages("dplyr")
install.packages("Ternary")

```

Load packages and API token.

```{r load-packages, message=FALSE, warning=FALSE}

# Load required packages
library(neonUtilities)
library(neonOS)
library(aqp)
library(cluster)
library(sharpshootR)
library(dplyr)
library(Ternary)

token <- Sys.getenv("NEON_TOKEN")

```

## Download megapit data

Documentation about the Megapit physical and chemical properties data product 
can be found here:

* <a href="https://data.neonscience.org/data-products/DP1.00096.001">Soil physical and chemical properties, Megapit DP1.00096.001</a>

If you are unfamiliar with `neonUtilities` and NEON data download workflows, we 
recommend starting with the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> tutorial.

In this exercise, we want all available data, so we won't subset by 
site or date range.

```{r download-data, results='hide', message=FALSE}

MP <- loadByProduct(dpID="DP1.00096.001", 
                    check.size = F,
                    token=token)

list2env(MP, .GlobalEnv)

```

## Join tables and convert to custom soil profile format

We'll join the horizon data to the physical and chemical characteristics data 
in order to see a depth profile of biogeochemical characteristics. The `neonOS` 
package contains the `joinTableNEON()` function, which uses metadata from the 
data product's Quick Start Guide to join tables.

```{r join-bgc}

S <- joinTableNEON(mgp_perhorizon,
                   mgp_perbiogeosample)

S <- arrange(S, siteID, horizonTopDepth)

```

There are two more things that we need to do before converting this data frame into a `SoilProfileCollection` object. First, we will make a new `siteLabel` column to use when plotting several pedons at once. 

```{r site-labels}

## combine 'domainID' and 'siteID' into a new label variable
S$siteLabel <- sprintf("%s-%s", S$domainID, S$siteID)

```

Second, we will translate the soil physical properties (sand, silt, clay percentages) into an RGB color scheme that we can use for plotting later. These steps are much easier to do now while the data are in a simple data frame.

```{r RGB-colors}

# duplicate physical property variables
S$r <- S$sandTotal # Sand is Red 'r'
S$g <- S$siltTotal # Silt is Green 'g'
S$b <- S$clayTotal # Clay is Blue 'b'

# set NA values to 100 (white)
S$r[is.na(S$r)] <- 100
S$g[is.na(S$g)] <- 100
S$b[is.na(S$b)] <- 100

# normalize values to 1 and convert to vector of colors using 'rgb()' function
S$textureColor <- rgb(red=S$r/100, 
                      green=S$g/100, 
                      blue=S$b/100, 
                      alpha=1, 
                      maxColorValue = 1)

```

We now have a data frame of biogeochemical data, organized 
by site and horizon. We can convert this to a 
`SoilProfileCollection` object using the `aqp` package.

```{r make-SPC-object}

depths(S) <- siteLabel ~ horizonTopDepth + horizonBottomDepth

```

We then move the site-level metadata to the @site attributes of `S`, our AQP object. 
This makes our `S` object easy to subset to look at a particular site. (Thanks 
to Dylan Beaudette of the USDA-NRCS for the helpful tips on how to best use 
this package!)

```{r move-site-attributes}

site(S) <- ~ siteID + nrcsDescriptionID

```

## Plot simple soil profiles

Using the plotting functions in the `aqp` package, 
let's start exploring some depth profiles. We'll start with 
a single site, the Smithsonian Environmental Research Center 
(SERC), and plot clay content by depth.

```{r plot-SERC}

# set plot margins
par(mar=c(1,0,3,15), mfrow=c(1,1), xpd=NA)

# Plot SERC clay profile
plotSPC(subset(S, siteID=="SERC"), 
        name='horizonName.x', label='siteLabel', 
        color='clayTotal', col.label='Clay Content (%)',
        col.palette=viridis::viridis(10), 
        cex.names=1, width = 0.1, 
        depth.axis = list(style="traditional",
                          line=6,
                          cex=1), 
        col.legend.cex = 1.5, n.legend=6, 
        x.idx.offset = 0, n=.88)


```

Now let's take a look at phosphorus at Wind River 
Experimental Forest (WREF).

```{r plot-WREF}

par(mar=c(1,0,3,15), mfrow=c(1,1), xpd=NA)

plotSPC(subset(S, siteID=="WREF"),  
        name='horizonName.x', label='siteLabel', 
        color='pMjelm', 
        col.label='Phosphorus (mg/Kg)',
        col.palette=viridis::viridis(10), 
        cex.names=1, width = 0.1, 
        depth.axis = list(style="traditional",
                          line=6,
                          cex=1), 
        col.legend.cex = 1.5, n.legend=4, 
        x.idx.offset = 0, n=.88)

```

## Plotting multiple sites

We can pass the plotting function multiple sites in order to compare pedons directly. 
We'll choose three very different sites with very different soils: Guanica (GUAN), 
a subtropical dry forest in Puerto Rico, Jornada (JORN), a desert in New Mexico, 
and Wind River (WREF), a temperate rainforest in Washington State.

```{r plot-three}

par(mar=c(0,2,3,2.5), mfrow=c(1,1), xpd=NA)

plotSPC(subset(S, siteID %in% c('WREF', 'JORN', 'GUAN')),
        name='horizonName.x', label='siteLabel', 
        color='sandTotal', 
        col.label='Percent Sand (%)',
        col.palette=viridis::viridis(10),
        n.legend = 5)

```


## Multivariate plotting

When analyzing soils data, we are often interested in more than one variable at a time. A classic example is texture, which in its simplest terms, is described in terms of percent Sand, Silt, and Clay. In order to plot these three variables together, we can describe each percentage as a color on the color wheel. Remember, in the section above, we assigned the color Red to Sand, Green to Silt, and Blue to Clay. We can plot these colors on the familiar three-axis plot that is often used to describe soil texture to serve as our color legend. To do this, we will use the `Ternary` package:

```{r ternary-plot}

par(mfrow=c(1, 1), mar=rep(.3, 4))

TernaryPlot(alab="% Sand \u2192", blab="% Silt \u2192", clab="\u2190 % Clay ",
            lab.col=c('red', 'green3', 'blue'),
            point='up', lab.cex=1.5, grid.minor.lines=1, axis.cex=1.5,
            grid.lty='solid', col=rgb(0.9, 0.9, 0.9), grid.col='white', 
            axis.col=rgb(0.6, 0.6, 0.6), ticks.col=rgb(0.6, 0.6, 0.6),
            padding=0.08)

cols <- TernaryPointValues(rgb)

ColourTernary(cols, spectrum = NULL, resolution=45)

```

Now, we can plot the soil texture of pedons as a color scheme:

```{r plot-physical-texture}

par(mar=c(0,2,3,2.5), mfrow=c(1,1), xpd=NA)

plotSPC(subset(S, siteID %in% c('WREF', 'JORN', 'GUAN')),
        name='horizonName.x', label='siteLabel', 
        color='textureColor')

```


## Multivariate clustering

We have 47 Megapit samples across the NEON observatory, spanning a wide range of 
soil types, textures, and chemical profiles. While it may be helpful from a 
geographic perspective to group the pedons by site, it may be more informative 
to group the samples by their inherent properties. For example, grouping soils 
by texture, or by their organic matter content. In order to make these groupings, 
we will employ a DIvisive ANAlysis (DIANA) clustering technique using the 
`cluster` package.

The clustering algorithm requires regularized data, so first we subset the data 
to records meeting the `aqp` package's quality standards, using the 
`HzDepthLogicSubset()` function.

```{r cluster-texture}

# subset to meet quality threshold
S.sub <- HzDepthLogicSubset(S)

# classify soil profiles
d <- NCSP(S.sub, vars=c('clayTotal',
                        'sandTotal', 
                        'siltTotal'), 
          k=0)

# vizualize dissimilarity matrix via divisive hierarchical clustering
d.diana <- diana(d)

# Plot the resulting dendrogram
plotProfileDendrogram(S.sub, d.diana, scaling.factor = .6, 
                      y.offset = 2, width=0.25, cex.names=.4, 
                      name='horizonName.x', label='siteLabel', 
                      color='textureColor')

```

We have a dendrogram of the soil texture profiles.

To zoom in and vizualise more clearly, we need to make a plotting area large enough to contain the full plot with labels. To do so, we will open a `PDF` graphics device, generate the plot in that device, then close and save the device using `dev.off()`. It will also be a good idea to check your current working directory, and perhaps change that to where you want to save your PDFs.

```{r plot-texture-clusters}

# Check and set working directory as needed.
getwd()
# setwd("enter path to save PDF here")

# Open 'pdf' graphic device. Define file name and large dimensions
pdf(file="NEON_Soils_Texture_Color_Clusters.pdf", width=24, height=10)

# set plot margins and generate plot
par(mar=c(12,2,10,1), mfrow=c(1,1), xpd=NA)
plotProfileDendrogram(S.sub, d.diana, scaling.factor = .6, 
                      y.offset = 2, width=0.25, cex.names=.4, 
                      name='horizonName', label='siteLabel',
                      color='textureColor')

# Close and save the device
dev.off()

```

This plot is not shown on this tutorial webpage, but you can view and download an example of the <a href="https://github.com/NEONScience/NEON-Data-Skills/blob/main/tutorials/R/soils/megapit-data/megapit-profiles/NEON_Soils_Texture_Color_Clusters.pdf">PDF here.</a>

## Clustering by Nutrients

Rather than cluster based on physical properties, we can also cluster based on nutrient contents (nitrogen, carbon, and sulfur):

```{r cluster-nutrients}

## Cluster as above, but for nutrient variables
d.nutrients <- NCSP(S.sub, 
                    vars=c('nitrogenTot',
                           'carbonTot', 
                           'sulfurTot'),
                    k=0)

# vizualize dissimilarity matrix via divisive hierarchical clustering
d.diana.nutrients <- diana(d.nutrients)

```

Let's make another PDF for our plot. However, this time it isn't as straightforward to plot the three nutrients of interest as 'rgb' colors, so we will make three separate plots for each nutrient:

```{r plot-nutrient-clusters}

# Open 'pdf' graphic device. Define file name and large dimensions
pdf(file="NEON_Soils_Nutrient_Clusters.pdf", width=24, height=14)

# Set plot margins
par(mar=c(8,2,5,1), mfrow=c(3,1), xpd=NA)

# Make plots for each nutrient of interest
plotProfileDendrogram(S.sub, d.diana.nutrients, scaling.factor = .6, 
                      y.offset = 2, width=0.25, name='horizonName', 
                      label='siteLabel', color='nitrogenTot',
                      col.label='Total Nitrogen (g/Kg)', 
                      col.legend.cex = 1.2, n.legend=6, 
                      col.palette=viridis::viridis(8))
plotProfileDendrogram(S.sub, d.diana.nutrients, scaling.factor = .6, 
                      y.offset = 2, width=0.25, name='horizonName', 
                      label='siteLabel', color='carbonTot',
                      col.label='Total Carbon (g/Kg)', 
                      col.legend.cex = 1.2, n.legend=6, 
                      col.palette=viridis::viridis(8))
plotProfileDendrogram(S.sub, d.diana.nutrients, scaling.factor = .6, 
                      y.offset = 2, width=0.25, name='horizonName', 
                      label='siteLabel', color='sulfurTot',
                      col.label='Total Sulfur (g/Kg)', 
                      col.legend.cex = 1.2, n.legend=6, 
                      col.palette=viridis::viridis(8))

# Close and save the device
dev.off()

```

This plot is not shown on this tutorial webpage, but you can view and download an example of the <a href="https://github.com/NEONScience/NEON-Data-Skills/blob/main/tutorials/R/soils/megapit-data/megapit-profiles/NEON_Soils_Nutrient_Clusters.pdf">PDF here.</a>
