---
syncID: 
title: "Methods of Exploring NEON Mosquito Data in R Part One: Data Set Up"
description: "This is part one of a series of tutorials that show users how to clean and perform an initial analysis of NEON mosquito data."
dateCreated: 2017-08-04
authors: Charlotte Roiger 
contributors: Megan A. Jones, Katie Levan, Cody Flagg
estimatedTime: 
packagesLibraries: 
topics: data-analysis, data-visualization
languagesTool: R,
dataProduct: NEON.DP1.10043.001
code1: 
tutorialSeries: neon-mosquito-data-intro
urlTitle: tba
---

This is part one of a series of tutorials that will teach you how to use NEON 
mosquito data to explore mosquito species richness and the abundance of *Culex 
tarsalis*. This series will also cover setting up data frames from existing NEON
data as well as how to obtain location information from NEON APIs. 

So that the code we are learning here makes sense within an ecological framework, 
the series that this tutorial is part of will be exploring the questions of
* what annual patterns of mosquito abundance and diversity are found? 
* how has the range of *Culex tarsalis* changed over the first three years of 
NEON data collection? 

<div id="ds-objectives" markdown="1">

## Objectives

After completing this tutorial, you will be able to:

*	Download mosquito trapping, identification, and sorting information from the 
NEON data portal
* Download precipitation and temperature data from Global Historical Climatology
Network
* Merge data frames to create one unified data frame with relevant variables to 
address research questions
* Subset data by year
* Use ggplot2 in R to create visualizations of data trends and maps  


## Things You Will Need To Complete This Tutorial

You will need the most current version of R and, preferably, RStudio loaded on
your computer to complete this tutorial.

### R Libraries to Install:

These R packages will be used in the tutorial below. Please make sure they are 
installed prior to starting the tutorial. 
 
* **dplyr:** `install.packages("dplyr")`
* **plyr:** `install.packages("plyr")`
* **tidyverse:** `install.packages("tidyverse")`
*	**mosaic:** `install.packages("mosaic")`
* **ggplot2:** `install.packages("ggplot2")`
* **devtools:**

<a href="/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

### Download The Data

This tutorial goes through how to pull data directly from the NEON data portal 
using the API. There is no additional data to download. 

****
{% include/_greyBox-wd-rscript.html %}

****

</div>

## NEON mosquito data collection

The data that we will primarily be working with in this tutorial are mosquiot abundance and diversity data collected at reguar intervals at NEON terrestrial field sites. 
The mosquitos are captured in CO<sub>2</sub> traps and identified to species. 
These data are available as NEON data product **NEON.DP1.10043.001**. More information
about this data product can be found in the 
<a href="http://data.neonscience.org/data-product-view?dpCode=DP1.10043.001" target="_blank">NEON Data Catalog</a>. 

## NEON Data API

To start off, we will be pulling NEON mosquito data from the NEON data portal 
using the NEON API. This tutorial focuses specifically on using the API for NEON
mosquito data. For more information on the API in general please see the 
<a href="/neon-api-usage" target="_blank"> *Using the NEON API in R* tutorial </a>
or the 
<a href="http://data.neonscience.org/data-api" target="_blank"> NEON API webpage</a>. 

First, let's set up our R environment to our desired working directory. 

```{r loadStuff}

# Load the necessary libraries
library(httr)
library(jsonlite)
library(dplyr, quietly=T)
library(downloader) 

# set working directory to ensure R can find the file we wish to import
# setwd("working-dir-path-here")
```

Now we can use the API call to determine if our desired mosquito data are 
available.

The API call that we will initially use looks like this 

`http://data.neonscience.org/api/v0/products/DP1.10043.001`

The initial part (in red) is the base URL that is going to the NEONScience 
data portal with the API version 0. 

`<span style="color:#A00606;font-weight:bold">http://data.neonscience.org/api/v0</span><span style="color:#A2A4A3">/products/DP1.10043.001</span>`

The next part of the call specifies exactly what data we want from the call. 

`<span style="color:#A2A4A3;font-weight:bold">http://data.neonscience.org/api/v0</span><span style="color:#A00606">/products/DP1.10043.001</span>`

Here it specifies that we want information about data `products` and specifcially 
about the data product `DP1.10043.001` --- the mosquito data. 

We will use the `GET()` function from the httr package to data we are 
interested in using the API call. 

```{r data-availability}

{
zipsByProduct(dpID="DP1.10043.001", site="all", package="basic", check.size=T)
stackByTable(paste0(getwd(), "/filesToStack-NEONDP10043"), folder=T)
}

# Request data using the GET function & the API call
req <- GET("http://data.neonscience.org/api/v0/products/DP1.10043.001")
req

```

What we get from this request is a "response" object. When we printed the response
object we got the URL, date of response, http status ("200" means it was a success, "400s" means there was some type of error), the file content type, and the size of
the file.  

<div id="ds-dataTip" markdown="1"><i class="fa fa-star"></i> **Data Tip:** Type
`?? httr` into your console to view the "Getting started with httr/httr quickstart
guide" for more on GET() and response objects. </div>

We can convert it into a JSON file to make it a bit more readable

```{r json-readable}
# make this JSON readable -> "text"
req.text <- content(req, as="text")

# Flatten data frame to see available data. 
avail <- fromJSON(req.text, simplifyDataFrame=T, flatten=T)
```

If you print out `avail` you will see a long list of important information (I'm 
not here so that we don't have to scroll through the full list --- but you should).

What we are particularly interested in is the sites that the data is available 
from and the data range represented by the data. Therefore, we can look at just 
this data. 

```{r}
# see all sites
avail$data$siteCodes

# create a list of the URL codes for next step 
mos.urls <- unlist(avail$data$siteCodes$availableDataUrls)
mos.urls

# get data availability for WOOD July 2015
mos <- GET(mos.urls)
mos.files <- fromJSON(content(mos, as="text"))

# view just the available data files 
mos.files$data$files

# Get both files
brd.count <- read.delim(brd.files$data$files$url
                        [intersect(grep("countdata", brd.files$data$files$name),
                                    grep("basic", brd.files$data$files$name))], sep=",")

brd.point <- read.delim(brd.files$data$files$url
                        [intersect(grep("perpoint", brd.files$data$files$name),
                                    grep("basic", brd.files$data$files$name))], sep=",")
```

```{r script}
# Functions to pull NEON data
source(paste(repo_location, 'get_NEON_data_by_productID.R', sep='/'))

# Pulls all data by DPID
# Requires: a known DPID, the folder location of the 'get_NEON_data_by_productID.R' script, andthe table names to pull
# Note: table names are pulled by grepl, so you just need enough of the name to get a unique table name

# Mosquito diversity data "DP1.10043.001"
# contains 6 tables: 'mos_trapping', 'mos_sorting', 'mos_expertTaxonomistIDProcessed', 'mos_expertTaxonomistIDRaw','mos_archivepooling', 'mos_barcoding'
mos = get_data_product(productID = "DP1.10043.001", 
                       df = c('mos_trapping', 'mos_sorting', 'mos_expertTaxonomistIDProcessed', 'mos_expertTaxonomistIDRaw','mos_archivepooling', 'mos_barcoding'))

# Rename table outputs for script
trap = mos$mos_trapping
sort = mos$mos_sorting
id = mos$mos_expertTaxonomistIDProcessed


#Set strings as factors equal to false thoughout

options(stringsAsFactors = FALSE) 

```


```{r example-data-portal-pull}

# Clean starting environment

# rm(list = ls())

# My working directory


# if(file.exists('~/INSERT/WORKING/DIRECTORY/HERE')){
#   repo_location = '~/INSERT/WORKING/DIRECTORY/HERE'
# }
# if(!exists('repo_location')){stop('Set the location of the neon_data repository.')}

# Functions to pull NEON data


# source(paste(repo_location, 'get_NEON_data_by_productID.R', sep='/'))

# Pulls all data by DPID
# Requires: a known DPID, the folder location of the 'get_NEON_data_by_productID.R' script, andthe table names to pull
# Note: table names are pulled by grepl, so you just need enough of the name to get a unique table name

# Mosquito diversity data "DP1.10043.001"
# contains 6 tables: 'mos_trapping', 'mos_sorting', 'mos_expertTaxonomistIDProcessed', 'mos_expertTaxonomistIDRaw','mos_archivepooling', 'mos_barcoding'


# mos = get_data_product(productID = "DP1.10043.001", 
#                        df = c('mos_trapping', 'mos_sorting', 'mos_expertTaxonomistIDProcessed', 'mos_expertTaxonomistIDRaw','mos_archivepooling', 'mos_barcoding'))
# 
# # Rename table outputs for script
# trap = mos$mos_trapping
# sort = mos$mos_sorting
# id = mos$mos_expertTaxonomistIDProcessed


#Set strings as factors equal to false thoughout

# options(stringsAsFactors = FALSE) 

```


## NEON Mosquito Data

The mosquito data on the NEON Data Portal are divided by type of information 
into six tables:
 * field collection data
 * sorting data
 * identification and pinning data
 * pathogen pooling data
 * pathogen results data
 * archiving pooling data

For this tutorial we focus on the 2014 and 2016 mosquito data for which data is 
available in those years, we will also direct most of our focus to the field
collection, sorting, and identification data. To look at all this data requires 
downloading 3 files and combining the datasheets. 

#################################################################################################
NEON provides several documents with information about the Carabid beetle protocol & 
data collection. It is highly recommended you are familiar with the
<a href="http://data.neonscience.org/data-product-view?dpCode=DP1.10022.001" 
target="_blank">data product documents </a>
prior to using NEON carabid beetle data for your research. 

#################################################################################################


We'll explore these three tables and then combine them into a couple of tables for
use with analysis. 

First, set up the R environment. 

```{r load-in-libraries-dp}

# Load packages required for script.

# move/manipulate data
library(dplyr)

# move/manipulate data
library(plyr)   

# move/manipulate data
library(tidyverse)

# good for data exploration
library(mosaic) 

# creation of plots and visualizations
library(ggplot2)   

```


### Field Collection Data Table

The field collection data table, or in this case the "trap" data frame, contains
data related to the location, time, and whether mosquitoes were present in the
sample. To get a better feel for this data we will use the `str` command to
look at the structure of the data frame. 

```{r trapping-table-dp}

#This command allows you to view the structure of the data
str(trap)

```

This table contains information related to:

* metadata about the sampling event, includes: 
   + `plotID`: Label of where a sample was collected
   + `setDate`: When traps were set
   + `collectDate`: The date of trap collection
   + `sampleID`: Unique label of sample collection events
   + `targetTaxaPresent`: An indication of whether mosquitos were found in the 
      sample 
   + `samplingProtocolVersion`: The document number and the version of the 
      sampling protocol used. These can be found in the 
<a href="http://data.neonscience.org/documents" target="_blank"> NEON Documents Library</a>. 

For sake of convenience we have only included the meta data for certain variables
that we will use to carry out our analysis. For more metadata please see:

**INSERT LINK FOR SORT METADATA**


Unique collection events have a unique `sampleID` with the format = 
`plotID.trapID.collectDate.TimeOfCollection`. 

### Sorting Data Table


```{r sort-table-dp}

#View structure of sorting data
str(sort)

```

This table contains information about weight of subsamples and weight of bycatch. 

* metadata about the subsampling event, includes:
  + `plotID`: Label of where a sample was collected
  + `setDate`: When traps were set
  + `collectDate`: The date of trap collection
  + `sampleID`: Unique label of sample collection events
  + `subsampleID`: Unique label of subsampling collection events
  + `totalWeight`: Total weight of sample
  + `subsampleWeight`: Total weight of subsample
  + `bycatchWeight`: Total weight of bycatch in the subsample

Unique records have a unique `subSampleID` (format = 
`plotID.trapID.collectDate.TimeofCollection.S.01` ). 

For sake of convenience we have only included the meta data for certain variables
that we will use to carry out our analysis. For more metadata please see:

**INSERT LINK FOR SORT METADATA**

### Identification Data Table

```{r identification-table-dp}

# view structure of the data
str(id)

```

The identification table contains information about the types of mosquitos found
in each subsample. 

* metadata about the subsampling event, includes:
  + `plotID`: Label of where a sample was collected
  + `setDate`: When traps were set
  + `collectDate`: The date of trap collection
  + `sampleID`: Unique label of sample collection events
  + `subsampleID`: Unique label of subsampling collection events
  + `individualCount`: The number of each species found in a subsample
  + `scientificName`: The Scientific name of each species found in a subsample

* metadata about the quality of the data 

The identification table contains information about all mosquitos that were 
found in each subsample. Each sample in the identification dataset contains the
target taxa and once identified is either directly archived, or first sent to an
external lab for Pathogen testing.

### Domain Data Table

```{r domain-table-dp}

#read in the data
domain.df = read.csv("NEON-mosquito-data-viz/domainid_siteid_match.csv")

str(domain.df)

```

The domain table includes information that matches site acronyms to NEON
domains. 

### Taxonomy Data Table

```{r taxonomy-table-dp}

#read in data
taxonomy = read.csv("NEON-mosquito-data-viz/mosquito_taxonomy.csv")

str(taxonomy)

```

This table contains information about mosquito taxonomy as well as the native
status of each mosquito species in the data frame.

* metadata about the subsampling event, includes:
  + `scientificName`: The Scientific name of each species found in a subsample
  + `d##NativeStatusCode`: An indicator of whether a species is native to 
     each domain
     
### Temperature Data Table

```{r temperature-table-dp}

# read in data 
temp.df = read.csv("NEON-mosquito-data-viz/temperature_maximum.csv")

str(temp.df)

```

This table contains temperature information from NOAA National Climatic Data
Center. 

* metadat about the temperature data, includes:
  + `date`: The date of when each temperature measurement was taken.
  + `station`: The station id of where each sample was taken.
  + `value`: The maximum temperature given in tenths of a degree Celsius
  + `siteID`: The closest NEON site to a NOAA weather station

## Wrangling the Data

### Obtaining location information

Before we can begin analyzing our data frames, we need to create a few new 
variables and collect all the information we need into a unified usable 
dataframe. The first step in our data wrangling process is to consolidate all of
the information stored in the trapping, identification, and sorting data frames. 
A lot of the information in the sorting data frame is very similar to what is
found in the identification data frame. But, one key difference is that the 
sorting data frame also contains the weight of the subsample, the weight of the 
bycatch, and the total weight of the sample. So we want to only select the 
columns in the sorting data frame that aren't in the id data frame.

```{r finding-unique-columns-dp}

# Merge ID with sort
cols = colnames(sort)[!colnames(sort)%in%colnames(id)]

# Merge id with subsetted sorting data frame
id <- left_join(id, sort[, c('subsampleID', cols)], 
                by = "subsampleID")

```

If we want to then merge our id data frame with the information in trap, we 
first have to simplify the trap data to lower processing times. We do that by
selecting only the rows of our trap data frame that are unique, and omitting any
rows that have repeated Plot IDs. 

```{r create-unique-trap-dp}

#Creating a dataframe with only the unique plotIDs and lat2 lon2 data for merging
uniquetrap<-trap[!duplicated(trap$plotID),c("plotID", "decimalLatitude","decimalLongitude", "elevation","nlcdClass")]

#Merging id df with latitude and longitude data
id <- merge(x = uniquetrap, y = id, by.y = "plotID", by.x = "plotID", all.y = TRUE)


```

One thing to keep in mind is that the identification and sorting data frames 
only contain samples where the mosquitoes were present. However, we might also
be interested in analyzing the samples where mosquitoes were not present. So we
need to find the rows in the trap data frame where the plot ID is not in the
id data frame and the target taxon is not present. First we create a subset of
the trap data frame where mosquitoes were not found in the sample. Since we 
want to then merge these rows with those in our id data frame we add in columns
that are present in the id dataframe but not in the trap data frame and row-bind
these two data frames together.

```{r including-trap-zero-dp}

# Get zero traps from trapping
new_trap<- trap[!trap$sampleID %in% id$sampleID & trap$targetTaxaPresent=="N",]

#Add columns in new_trap that weren't present in the ID table then add new_trap to ID table
new_trap <- new_trap[, colnames(new_trap)[colnames(new_trap)%in%colnames(id)]]

new_trap[, colnames(id)[!colnames(id)%in%colnames(new_trap)]]<-NA

id <- rbind(id,new_trap)

```

### Creating Variables and Obtaining Weather Data 

Now that we have a more complete id dataset, we want to create a couple of 
variables that could be useful in our analysis. To start, we 
will note that the individual count present in each observation of the id data
frame is only the individual count of each subsample. So to estimate the 
number of individuals in each sample we will use the sample weight,
by-catch weight, and subsample weight to generate a sample multiplier. To 
create the sample multiplier we use an if-else statement to find only the rows
in the id dataframe where by-catch weight information is present. Then we divide
the total sample weight by the by-catch weight subtracted from the subsample
weight. Next we use another if-else statement to replace all instances where
the sample multiplier is infinity with NAs. We then create a new variable called
"newindividualCount" where we multiply the individual count by the sample
multiplier.

```{r fixing-individualCount-dp}

#Convert sample weights and individual count to numeric type
id$bycatchWeight<- as.numeric(id$bycatchWeight)
id$totalWeight <- as.numeric(id$totalWeight)
id$subsampleWeight <- as.numeric(id$subsampleWeight)
id$individualCount <- as.numeric(id$individualCount)

#Creation of sample Multiplier
id$sampleMultiplier <- ifelse(is.na(id$bycatchWeight), id$totalWeight/id$subsampleWeight, id$totalWeight/(id$subsampleWeight-id$bycatchWeight))
id$sampleMultiplier <- ifelse(id$sampleMultiplier==Inf, NA, id$sampleMultiplier)
id$sampleMultiplier <- ifelse(id$subsampleWeight==0 & id$individualCount != 0, 1, id$sampleMultiplier)

#Creation of New individual Count with Multiplier
id$newindividualCount <-ifelse(is.na(id$sampleMultiplier)==F, round(id$individualCount*id$sampleMultiplier), NA)

```

Now that we have an estimate of the abundance of each species in a sample, we
also want to create a variable that takes into account the amount of time a 
trap was deployed. One issue present with creating this variable is that traps 
were either deployed overnight or collected within the space of one day. To
address this challenge, we first create a variable that returns true if the 
set date and the collect date are on the same day. Next we create two variables
that converts the minutes of the set and collect times into hours. After that, 
we use an if-else statement to find observations where set and collection dates
were on the same day, then we subtract the set hours from the collection hours
to get the number of hours that the trap was deployed. If the trap was deployed
over the period of two days, we calculate the number of hours from when the trap
was set until midnight by subtracting the set time from 24, then added the
number of hours the trap was deployed on the collect day to yield the hours of 
deployment. 

```{r trapping-hours-dp}

#Creation of a variable to test whether samples were collected on the same day or different days
id$sameDay <- ifelse(substr(id$collectDate, 9, 10) != substr(id$setDate,9,10), FALSE, TRUE)

#Creating variables that convert the time of set and collection to hours
id$setHours <-((as.numeric(substr(id$setDate,15,16))/60)+(as.numeric(substr(id$setDate,12,13))))
id$collectHours <-((as.numeric(substr(id$collectDate,15,16))/60)+(as.numeric(substr(id$collectDate,12,13))))

#variable to calculate the number of hours of trap deployment
id$HoursOfTrapping <-ifelse(id$sameDay == TRUE, id$collectHours - id$setHours, (24 - id$setHours) + id$collectHours)

#Changing hours of trapping to positive number 
id$HoursOfTrapping <- abs(as.numeric(id$HoursOfTrapping))

```

In our current id data frame, we have only the set and collect dates of each 
sample, where the collect date and time is formatted as "YYYY-MM-DDThh:mm".
However, if we want to look at the Julian date of observation for Culex 
tarsalis, we might want the date and year when the sample was collected. So we 
use the `substr` command to collect only the first four characters of the
"collectDate" variable to pull year information, then we convert year to a 
factor. However, in the id data frame there are some observations where the 
collection date was not available. For many of the observations where 
collection information is missing, the date of when the sample was recieved so
we can extract year information in a similar fashion from the "recievedDate" 
variable. 

```{r date-and-year-dp}

#Extracting year information for id
id$Year<-substr(id$collectDate,1,4)

#Extracting year information for id from both collect date and recieved date
id$receivedDate <- as.character(id$receivedDate)

id$Year<-ifelse(is.na(id$collectDate), substr(id$receivedDate,1,4), substr(id$collectDate,1,4))

#Exctracting date information for id
id$Date<-substr(id$collectDate,1,10)

```


### Finishing Touches and Filtering Data

Now that we have location and date information in a more usable format we
are almost ready to start analyzing our NEON data. However a quick exploration 
of our resulting identification data frame might reveal that the number of 
mosquito observations fluctuate greatly for the years 2012, 2013, and 2015. This
is because of changes in sampling design, making the observationsmfor these 
years less comparable to 2014 and 2016. Due to the changes in sampling
design for mosquito collection, we will continue on with our analysis and focus 
in on the data from 2014 and 2016. Another issue that we need to resolve before
moving forward with this tutorial is when mosquitoes cannot be identified. 
Since both species richness and an examination of *Culex tarsalis* both depend 
on species identification we will remove any observations in our data frame 
where samples cannot be identified down to the species or subspecies level in
our data, Please note that we have decided to include observations where 
the "taxonRank" is considered to be blank or "NA" because these observations 
could represent samples where the target taxon was not present.

```{r filtering-id-dp}

#Convert domainID to a factor.
id$domainID <- as.character(id$domainID)

#Filter by species and subspecies classification
id <- dplyr::filter(id, id$taxonRank %in% c("subspecies","species", NA, ""))

#smalle subset only containing 2014 and 2016
idsmall<-dplyr::filter(id, id$Year %in% c(2014,2016))

```

Now that we have reformatted our data frames into a more usable format we will
save "idsmall" and "uniquetrap" as csv files. These can either be read into R
again at a later time for use in the second and third part of the NEON Mosquito
Data Visualization tutorial series, or this step can be skipped if you wish to 
proceed immediately to the next tutorial in the series. 

```{r}

#Write idsmall as a csv
write.csv(idsmall, "NEON-mosquito-data-viz/idsmall.csv")

#Write uniquetrap as a csv
write.csv(uniquetrap, "NEON-mosquito-data-viz/uniquetrap.csv")

```

A link to part two of this tutorial series can be found here. In part two, we 
will use both the 'idsmall' and the 'uniquetrap' data frames to create a data
frame that captures the number of species at each site on a given date. 


## Stack NEON Data

library(devtools)
install_github("NEONScience/NEON-utilities/neonUtilities", dependencies=TRUE)
library (neonUtilities)
stackByTable("NEON_count-mosquitoes.zip")
