---
syncID: 
title: "NEON Domain and Site Shapefiles and Maps"
description: "Use files available on the NEON data portal and NEON API to create maps of NEON domains and sites."
dateCreated: 2021-04-06
authors: Claire K. Lunch
contributors: Megan Jones, Alison Dernbach, Donal O'Leary
estimatedTime: 20 minutes
packagesLibraries: sp, rgdal
topics: data-management, rep-sci
languageTool: R
dataProduct: 
code1: 
tutorialSeries:
urlTitle: neon-maps

---

This tutorial explores NEON domain- and site-level spatial data, using 
shapefiles and tabular data provided by the NEON project. The intent of 
this tutorial is to provide guidance in navigating the data and files 
provided by NEON, and in creating maps of the domain and site locations. 
We hope these data and methods will assist users in generating images and 
figures for publications or presentations related to NEON.

<div id="ds-objectives" markdown="1">

## Learning Objectives 
After completing this tutorial you will be able to: 

* Access NEON spatial data from the NEON website
* Create a simple map with NEON domains and field site locations

## Things You’ll Need To Complete This Tutorial

### R Programming Language
You will need a current version of R to complete this tutorial. We also recommend 
the RStudio IDE to work with R. 

</div>


## Setup R Environment

We'll need several R packages in this tutorial. Install the packages, if not 
already installed, and load the libraries for each. 

```{r install, eval=FALSE}

# run once to get the package, and re-run if you need to get updates
install.packages("sp")  # work with spatial data
install.packages("rgdal")  # work with spatial data

```


```{r libraries, results="hide"}

# run every time you start a script
library(sp)
library(rgdal)

options(stringsAsFactors=F)

# set working directory to ensure R can find the file we wish to import and where
# we want to save our files. 

wd <- "~/Documents/data/" # This will depend on your local environment
setwd(wd)

```

## NEON spatial data files

NEON spatial data are available in a number of different files depending on which 
spatial data you are interested in. This section covers a variety of spatial 
data files that can be directly downloaded from the 
<a href="https://www.neonscience.org" target="_blank">NEONScience.org</a> website 
instead of being delivered with a downloaded data product.

The goal of this section is to create a map of the entire Observatory that includes 
the NEON domain boundaries and differentiates between aquatic and terrestrial field 
sites. 

### Site locations & domain boundaries

Most NEON spatial data files that are not part of the data downloads 
are available on the 
<a href="https://www.neonscience.org/data/spatial-data-maps" target="_blank">Spatial Data and Maps page</a> 
as both shapefiles and kmz files. 

In addition, latitude, longitude, elevation, and some other basic metadata for each site 
are available for download from the 
<a href="https://www.neonscience.org/field-sites/explore-field-sites" target="_blank">Field Sites page</a> 
on the NEON website (linked above the image). In this summary of each field site, 
the geographic coordinates for each site correspond to the tower 
location for terrestrial sites and the center of the permitted reach 
for aquatic sites.

To continue, please download these files from the NEON website: 

* **NEON Domain Polygons:** A polygon shapefile defining NEON's domain 
boundaries. Like all NEON data the Coordinate Reference system is Geographic 
WGS 84. Available on the <a href="https://www.neonscience.org/data/spatial-data-maps" target="_blank">Spatial Data and Maps page</a>.
* **Field Site csv:** generic locations data for each NEON field site. Available on the <a href="https://www.neonscience.org/field-sites/explore-field-sites" target="_blank">Field Sites page</a>. 

The Field Site location data is also available as a shapefile and KMZ on the 
Spatial Data and Maps page. We use the file from the site list to demonstrate 
alternative ways to work with spatial data. 

## Map NEON domains 

Using the domain shapefile and the field sites list, let's make a map of NEON 
site locations.

We'll read in the spatial data using the `rgdal` and `sp` packages and plot it 
using base R. First, read in the domain shapefile. 

Be sure to move the downloaded and unzipped data files into the working directory 
you set earlier!

```{r load-domains, results="hide"}

# upload data
neonDomains <- readOGR(paste0(wd,"NEONDomains_0"), layer="NEON_Domains")

```

The data come as a Large SpatialPolygonsDataFrame. Base R has methods 
for working with this data type, and will recognize it automatically 
with the generic `plot()` function. Let's first plot the domains without 
the sites.

```{r plot-domains, fig.width=8, fig.height=6, fig.cap="Map of the United States with each NEON domain outlined"}

plot(neonDomains)

```

The data are currently in a Lat-Long projection. The map will look a little 
more familiar if we convert it to a Mercator projection. There are many, 
many map projections possible, and Mercator is distorted in very 
well-documented ways! Here, we'll use it as a demonstration, but you may 
want to use a different projection in your own work.

The `spTransform()` and `CRS()` functions in the `sp` package can be used 
to convert the projection:

```{r mercator, fig.width=8, fig.height=6}

neonMercator <- spTransform(neonDomains,
                            CRS("+proj=merc"))
plot(neonMercator)

```

## Map NEON field sites

Now that we have a map of all the NEON domains, let's plot the NEON field site 
locations on it. To do this, we need to load and explore the field sites data. 

```{r load-explore-sites}

# read in the data
neonSites <- read.delim(paste0(wd,"field-sites.csv"), sep=",", header=T)

# view data structure for each variable
str(neonSites)

```

Here there is a lot of associated data about the field sites that may be of 
interest, such as site descriptions and dominant vegetation types. For mapping 
purposes, we can see that there are both **Latitude** and **Longitude**
data, so we can plot this data onto our previous map. 

```{r plot-sites, fig.width=8, fig.height=6, fig.cap="NEON domain map with site dots added"}

plot(neonDomains)
points(neonSites$Latitude~neonSites$Longitude,
       pch=20)

```

Now we can see all the sites across the Observatory. Note that we've switched 
back to the Lat-Long projection, which makes it simple to plot the site 
locations onto the map using their latitude and longitude. To plot them on 
the Mercator projection, we would need to convert the site latitude and 
longitude values into a SpatialPoints object, and then perform the same 
conversion. See `sp` package documentation if you are interested in doing 
this.

NEON has both aquatic and terrestrial sites, with important differences 
between the two. Looking back at the variables in this data set, we can 
see that `Site.Type` designates the aquatic and terrestrial sites. 

Let's differentiate the site types by adding color to our plot, with 
terrestrial sites in green and aquatic sites in blue.

```{r sites-color, fig.width=8, fig.height=6, fig.cap="NEON domain map with site dots color-coded for aquatic and terrestrial sites"}

# create empty variable
siteCol <- character(nrow(neonSites))

# populate new variable with colors, according to Site.Type
siteCol[grep("Aquatic", neonSites$Site.Type)] <- "blue"
siteCol[grep("Terrestrial", neonSites$Site.Type)] <- "green"

# add color to points on map
plot(neonDomains)
points(neonSites$Latitude~neonSites$Longitude,
       pch=20, col=siteCol)

```

Now we can see where NEON sites are located within the domains. Note that 
a significant number of terrestrial and aquatic sites are co-located; in 
those cases both sites are plotted here, but one color may be 
superimposed over the other.

## Map locations at a specific site

Also available on the
<a href="https://www.neonscience.org/data/spatial-data-maps" target="_blank"> Spatial Data and Maps page</a> are shapefiles containing location data at the 
within-site level.


The readme file contains descriptions for each variable in the tabular files. 

```{r TOS-readme}

# load TOS plot readme
rdme <- read.delim(paste0(wd,'All_NEON_TOS_Plots_V8/readme.csv'),
                   sep=',', header=T)

# view the variables
rdme[,1]

```
