---
syncID: b152963c4883463883c3b6f0de95fd93
title: "Access and Work with NEON Geolocation Data"
description: "Use files available on the NEON data portal, NEON API, and  neonUtilities R package to access the locations of NEON sampling events and infrastructure. Calculate more precise locations for certain sampling types and reference ground sampling to airborne data."
dateCreated: 2019-09-13
authors: Claire K. Lunch
contributors: Megan Jones, Alison Dernbach, Donal O'Leary
estimatedTime: 40 minutes
packagesLibraries: sp, rgdal, maptools, broom, ggplot2, neonUtilities
topics: data-management, rep-sci
languageTool: R
dataProduct: DP1.10072.001, DP1.00024.001
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/dev-aten/tutorials/R/NEON-general/neon-code-packages/spatialData/spatialData.R
tutorialSeries:
urlTitle: neon-spatial-data-basics

---

This tutorial explores NEON geolocation data. The focus is on the locations 
of NEON observational sampling and sensors; NEON remote sensing data are 
inherently spatial and have dedicated tutorials. If you are interested in 
connecting remote sensing with ground-based measurements, the methods in 
the <a href="https://www.neonscience.org/tree-heights-veg-structure-chm" target="_blank">canopy height model tutorial</a> 
can be generalized to other data products.

<div id="ds-objectives" markdown="1">

## Learning Objectives 
After completing this tutorial you will be able to: 

* access NEON spatial data from the website and through data downloaded with the 
neonUtilities package. 
* create a simple map with NEON domains and field site locations. 
* access and plot specific sampling locations for TOS data products. 
* access and use sensor location data. 

## Things You’ll Need To Complete This Tutorial

### R Programming Language
You will need a current version of R to complete this tutorial. We also recommend 
the RStudio IDE to work with R. 

</div>


## Setup R Environment

We'll need several R packages in this tutorial. Install the packages, if not 
already installed, and load the libraries for each. 

```{r install, eval=F}

# run once to get the package, and re-run if you need to get updates
install.packages("sp")  # work with spatial data
install.packages("rgdal")  # work with spatial data
install.packages("maptools")  # work with spatial objects
install.packages("broom")  # tidy up data
install.packages("ggplot2")  # plotting
install.packages("neonUtilities")  # work with NEON data
install.packages("devtools")  # to use the install_github() function
devtools::install_github("NEONScience/NEON-geolocation/geoNEON")  # work with NEON spatial data

```


```{r libraries, results="hide"}

# run every time you start a script
library(sp)
library(rgdal)
library(maptools)
library(broom)
library(ggplot2)
library(neonUtilities)
library(geoNEON)

options(stringsAsFactors=F)

# set working directory to ensure R can find the file we wish to import and where
# we want to save our files. 

wd <- "~/Documents/data/" # This will depend on your local environment
setwd(wd)

```

## Source 1: NEON spatial data files

NEON spatial data are available in a number of different files depending on what 
spatial data you are interested in. This section covers a variety of spatial 
data files that can be directly downloaded from the 
<a href="https://www.neonscience.org" target="_blank">NEONScience.org</a> website 
instead of being delivered with a downloaded data product. Later in the tutorial 
we'll work with spatial data available with the downloaded data products.

The goal of this section is to create a map of the entire Observatory that includes 
the NEON domain boundaries and differentiates between aquatic and terrestrial field 
sites. 

### Site locations & domain boundaries

Most NEON spatial data files that are not part of the data downloads 
are available on the 
<a href="https://www.neonscience.org/data/spatial-data-maps" target="_blank">Spatial Data and Maps page</a> 
as both shapefiles and kmz files. 

In addition, latitude, longitude, elevation, and some other basic metadata for each site 
are available for download from the 
<a href="https://www.neonscience.org/field-sites/field-sites-map/list" target="_blank">Field Sites List page</a> 
on the NEON website (linked below the table). In this summary of each field site, 
the geographic coordinates for each site correspond to the tower 
location for terrestrial sites and the center of the permitted reach 
for aquatic sites.

To continue, please download these files from the NEON website: 

* **NEON Domains - Shapefile:** A polygon shapefile defining NEON's domain 
boundaries. Like all NEON data the Coordinate Reference system is Geographic 
WGS 84. Available on the <a href="https://www.neonscience.org/data/spatial-data-maps" target="_blank">Spatial Data and Maps page</a>.
* **Field Site csv:** generic locations data for each NEON field site. Available on the <a href="https://www.neonscience.org/field-sites/field-sites-map/list" target="_blank">Field Sites List page</a> (bottom of table). 

The Field Site location data is also available as a shapefile and KMZ on the 
Spatial Data and Maps page. We use the file from the site list to demonstrate 
alternative ways to work with spatial data. 

## Map NEON domains 

Using the domain shapefile and the field sites list, let's make a map of NEON 
site locations.

We'll read in the spatial data using the `rgdal` and `sp` packages and plot it 
using the `ggplot2` package. First, read in the domain shapefile. 

Be sure to move the downloaded and unzipped data files into the working directory 
you set earlier!

```{r load-domains, results="hide"}

# upload data
neonDomains <- readOGR(paste0(wd,"NEONDomains_0"), layer="NEON_Domains")

```

The data come as a Large SpatialPolygonsDataFrame, which unfortunately `ggplot`
can't use. Therefore, we'll need to make a few changes to the data structure to 
convert it to a dataframe that `ggplot` can use. 

```{r create-df, warning=FALSE}

# first, add a new column termed "id" composed of the row names of the data
neonDomains@data$id <- rownames(neonDomains@data)

# now, use tidy() to convert to a dataframe
# if you previously used fortify(), this does the same thing 
neonDomains_points <- tidy(neonDomains, region="id")

# finally, merge the new data with the data from our spatial object
neonDomainsDF <- merge(neonDomains_points, neonDomains@data, by = "id")

```

Now that the data are in a dataframe, let's check out what data are available 
for us to plot.

```{r explore-domains}

# view data structure for each variable
str(neonDomainsDF)

```

We have **long**itude and **lat**itude data, as well as names of the domains. 
With this information, and now that we have the data in a dataframe, we can plot 
it in ggplot. Let's first plot the domains without the sites. Since we are plotting 
a map, this means our longitude is plotted on the x axis and latitude is on our 
y axis. 

```{r plot-domains, fig.width=8, fig.height=6, fig.cap="Map of the United States with each NEON domain outlined"}

# plot domains
domainMap <- ggplot(neonDomainsDF, aes(x=long, y=lat)) + 
        geom_map(map = neonDomainsDF,
                aes(map_id = id),
                 fill="white", color="black", size=0.3)

domainMap

```


## Map NEON field sites
Now that we have a map of all the NEON domains, let's plot the NEON field site 
locations on it. To do this, we need to load and explore the field sites data. 

```{r load-explore-sites}

# read in the data
neonSites <- read.delim(paste0(wd,"field-sites.csv"), sep=",", header=T)

# view data structure for each variable
str(neonSites)

```

Here there is a lot of associated data about the field sites that may be of 
interest. For now, we can see that there are both **Latitude** and **Longitude**
data so we can plot this data onto our previous map. 

```{r plot-sites, fig.width=8, fig.height=6, fig.cap="Same map as above but with dots for the field site locations across the country"}

# plot the sites
neonMap <- domainMap + 
        geom_point(data=neonSites, 
                   aes(x=Longitude, y=Latitude))

neonMap 

```

Now we can see all the sites across the Observatory. However, NEON has both 
aquatic and terrestrial sites with important differences between the two. Looking 
back at the variables in this data set, we can see that `Site.Type` designates 
the aquatic and terrestrial sites. However, it goes further and includes whether 
or not they are core or relocatable sites. Let's plot that distinction as well. 

We can do this by adding color to our plot, with terrestrial sites in green and 
aquatic sites in blue. We can choose darker shades for core sites and lighter 
ones for relocatable sites. 

```{r sites-color, fig.width=8, fig.height=6, fig.cap="Same as maps above but the field site dots are now four different colors"}

# color is determined by the order that the unique values show up
# check order
unique(neonSites$Site.Type)

# add color
sitesMap <- neonMap + 
        geom_point(data=neonSites, 
                      aes(x=Longitude, y=Latitude, color=Site.Type)) + 
           scale_color_manual(values=c("lightskyblue", "forest green", 
                                       "blue4", "light green"),
                              name="",
                              breaks=unique(neonSites$Site.Type))
sitesMap

```


## Map terrestrial observation plots

The **NEON Terrestrial Observation System (TOS) sampling locations** shapefile 
is available for download from the
<a href="https://www.neonscience.org/data/spatial-data-maps" target="_blank"> Spatial Data and Maps page</a> or from the portal in the
<a href="http://data.neonscience.org/documents" target="_blank"> document library</a>. 
The download will be a zip file containing tabular files of plot centroids and 
point locations, and shapefiles of plot centroids, point locations, and polygons.

The readme file contains descriptions for each variable in the tabular files. 

```{r TOS-readme}

# load TOS plot readme
rdme <- read.delim(paste0(wd,'All_NEON_TOS_Plots_V8/readme.csv'),
                   sep=',', header=T)

# view the variables
rdme[,1]

```

You can use these files to plot the spatial layout of sampling for TOS (mosquitoes, 
beetles, plants, birds, etc.). However, instead of using these files, for this 
tutorial we'll be using the location data provided with the data downloads below, 
as well as methods in the `geoNEON` package, to explore TOS spatial data.

You can, however, modify the code above from mapping the domains and field sites to 
plot the location data if you'd like to do so. 

## Source 2: Spatial data in data downloads

The second source of data is spatial data in and accompanying data directly 
downloaded from the NEON data portal. How the data are handled are a bit different 
for the Observational Sampling data and the Instrumented Systems data, so we'll 
go through working with both types. 

## Locations for observational data

### Plot level locations
Both aquatic and terrestrial observational data downloads include spatial 
data in the downloaded files. While the specific layout varies from data type to 
data type the method for working with the data are similar. Therefore, we'll work 
with NEON Small Mammal Box Trapping data. Our end product for this section is to 
create a map of all the traps within a small mammal grid that captured a mouse in a given 
sampling bout.  

First, let's download the small mammal data from one site, Onaqui (ONAQ), in 
August 2018 to investigate. 

If downloading data using the `neonUtilities` package is new to you, check out the 
<a href="https://www.neonscience.org/neonDataStackR" target="_blank">neonUtilities tutorial</a>.

```{r get-mam-data, results="hide"}

# load mammal data
mam <- loadByProduct(dpID="DP1.10072.001", site="ONAQ",
                     startdate="2018-08", enddate="2018-08",
                     check.size=F)

```

Data downloaded this way are stored in R as a large list. For this tutorial, 
we'll work with the individual dataframes within this large list. Alternatively, 
each dataframe can be assigned as its own object. 

To find the spatial data for any given data product, view the variables files to
figure out which data table the spatial data are contained in. 

```{r find-sp-data}

View(mam$variables_10072)

```

Looking through the variables, we can see that the spatial data (`decimalLatitude` 
and `decimalLongitude`) are in the `pertrapnight` table. We can look at the first few
entries of the table. 

```{r print-mam}

head(mam$mam_pertrapnight[,1:18])

```

Did you notice that traps in plot ONAQ_020 have the same location data?
Let's check the coordinates for all traps within a single plot. 

```{r print-ONAQ020}

# view all trap locations in one plot
mam$mam_pertrapnight[which(mam$mam_pertrapnight$plotID=="ONAQ_020"),
                     c("trapCoordinate","decimalLatitude",
                       "decimalLongitude")]

```

The latitude and longitude are the same for every record! This is because the 
latitudes and longitudes provided are for the plots, not for the traps. Other 
TOS data has the same pattern: the data download contains the plot-level 
coordinates. This important information is in the **Data Product User Guide** and 
can be downloaded from the 
<a href="http://data.neonscience.org" target="_blank">NEON Data Portal</a> on a 
data product's page. Understanding nuances like this is one reason it is 
important to read the Data Product User Guide.

For many analyses, this level of spatial data is sufficient. But for other 
types of analyses, you may need more precise locations. The `geoNEON` package 
can get these data for you.

### Sampling locations 

The `getLocTOS()` function in the `geoNEON` package uses the NEON API to 
access NEON location data and then makes protocol-specific calculations 
to return precise locations for each sampling effort. This function works for a 
subset of NEON TOS data products. The list of tables and data products that can 
be entered is in the 
<a href="https://github.com/NEONScience/NEON-geolocation/tree/master/geoNEON" target="_blank">package documentation on GitHub</a>. 

For more information about the NEON API, see the 
<a href="https://www.neonscience.org/neon-api-usage" target="_blank">API tutorial</a> 
and the 
<a href="https://data.neonscience.org/data-api" target="_blank">API web page</a>. 
For more information about the location calculations used in each data product, 
see the Data Product User Guide for each product.

The `getLocTOS()` function requires two inputs:

* A data table that contains spatial data from a NEON TOS data product
* The NEON table name of that data table

For small mammal box trap locations, the function call looks like this. This 
function may take a while to download all the location data. 

```{r mam-getLocTOS, results="hide"}

# download small mam locationso
mam.loc <- getLocTOS(data=mam$mam_pertrapnight,
                           dataProd="mam_pertrapnight")

```

What additional data are now available in the data obtained by `getLocTOS()`?

```{r mam-diff}

# print variable names that are new
names(mam.loc)[which(!names(mam.loc) %in% names(mam$mam_pertrapnight))]

```

Now we have adjusted latitude, longitude, and elevation, and the 
corresponding easting and northing UTM data. We also have coordinate uncertainty 
data for these coordinates. 

We can use the easting and northing data to plot the locations of the mammal traps. 

```{r mam-grids, fig.cap="Six square points on a plot each made up of smaller dots that form either a complete square point or a partially filled in square point"}

# plot all trap locations at site
plot(mam.loc$adjEasting, mam.loc$adjNorthing, pch=".",
     xlab="Easting", ylab="Northing")

```

Each trap grid has 100 points (individual trap locations), so even with each 
trap plotted as a dot (.) we can only see a square for each grid. 

Let's zoom in on a single plot:

```{r plot-ONAQ003, fig.width=6, fig.height=6, fig.cap="dots on a plot equally spaced in 10 rows and 10 columns"}

# plot all trap locations in one grid (plot)
plot(mam.loc$adjEasting[which(mam.loc$plotID=="ONAQ_003")], 
     mam.loc$adjNorthing[which(mam.loc$plotID=="ONAQ_003")], 
     pch=".", xlab="Easting", ylab="Northing", cex=3)

```

This isn't the most interesting plot given that that each small mammal box 
trapping grid is a 10 x 10 plot of traps. 

Now, let's add a layer of data to see which of these traps caught a mammal during 
the August 2018 sampling bout. To do this we need to look at our variables file 
again and see what variable gives us information about captures. We can see that 
`trapStatus` provides the information on what happened to each trap. It has 
categorical data on the status: 

* 0 - no data
* 1 - trap not set
* 2 - trap disturbed/door closed but empty
* 3 - trap door open or closed w/ spoor left
* 4 - >1 capture in one trap
* 5 - capture
* 6 - trap set and empty

Therefore, we need to plot all trap locations in this plot (ONAQ_003) for which 
`trapStatus` is "5 - capture" (technically, we should add in a capture status of 
4 but for demonstration purposes, we'll keep it simple).

```{r plot-captures, fig.width=6, fig.height=6, fig.cap="Same plot as above with 100 equally spaced dots but 22 dots are now blue"}

# plot all captures 
plot(mam.loc$adjEasting[which(mam.loc$plotID == "ONAQ_003")], 
     mam.loc$adjNorthing[which(mam.loc$plotID == "ONAQ_003")], 
     pch=".", xlab="Easting", ylab="Northing")

points(mam.loc$adjEasting[which(mam.loc$plotID == "ONAQ_003" & 
                               mam.loc$trapStatus == "5 - capture")], 
     mam.loc$adjNorthing[which(mam.loc$plotID =="ONAQ_003" &
                              mam.loc$trapStatus == "5 - capture")],
     pch=19, col="blue")

```

In the month of data we're viewing small mammals were caught at 22 of the 100 
trap locations in this plot.

The basic techniques for working with this data can be adapted to other TOS 
location data for other data products.  

## Locations for sensor data

Downloads of instrument system (IS) data include a file called 
**sensor_positions.csv**. The sensor positions file contains information 
about the coordinates of each sensor, relative to a reference location. 

While the specifics vary, techniques are generalizable for working with sensor 
data and the sensor_positions.csv file. For this tutorial, let's look at the 
sensor locations for soil temperature (PAR; DP1.00041.001) at  
the NEON Treehaven site (TREE) in July 2018. To reduce our file size, we'll use 
the 30 minute averaging interval. Our final product from this section is to 
create a depth profile of soil temperature in one soil plot.

If downloading data using the `neonUtilties` package is new to you, check out the 
<a href="https://www.neonscience.org/neonDataStackR" target="_blank">neonUtilities tutorial</a>.

This function will download about 7 MB of data as written so we have `check.size =F` 
for ease of running the code. 

```{r soilT-load, results="hide"}
# load soil temperature data of interest 
soilT <- loadByProduct(dpID="DP1.00041.001", site="TREE",
                    startdate="2018-07", enddate="2018-07",
                    avg=30, check.size=F)

```

### Sensor positions file 
Now we can specifically look at the sensor positions file.

```{r sens-pos}

# create object for sensor positions file
pos <- soilT$sensor_positions_00041

# view names
names(pos)

# view table
View(pos)

```

The sensor locations are indexed by the `HOR.VER` variable - see the 
<a href="https://data.neonscience.org/file-naming-conventions" target="_blank">file naming conventions</a> 
page for more details. 

Using `unique()` we can view all the location indexes in this file. 

```{r pos-levs}

# view names
unique(pos$HOR.VER)

```

Soil temperature data are collected in 5 instrumented soil plots inside the 
tower footprint. We see this reflected in the data where HOR = 001 to 005. 
Within each plot, temperature is measured at 9 depths, seen in VER = 501 to 
509. At some sites, the number of depths may differ slightly.

The x, y, and z offsets in the sensor positions file are the relative distance, 
in meters, to the reference latitude, longitude, and elevation in the file. 

The HOR and VER indices in the sensor positions file correspond to the 
`verticalPosition` and `horizontalPosition` fields in `soilT$ST_30_minute`.

Note that there are two sets of position data for soil plot 001, and that 
one set has an `end` date in the file. This indicates sensors either 
moved or were relocated; in this case there was a frost heave incident. 
You can read about it in the issue log, both in the readme file and on 
the <a href="https://data.neonscience.org/data-products/DP1.00041.001" target="_blank">Data Product Details</a> page.

Since we're working with data from July 2018, and the change in 
sensor locations is dated Nov 2018, we'll use the original locations.

```{r pos-rem}
pos <- pos[-intersect(grep("001.", pos$HOR.VER),
                      which(pos$end=="")),]
```

Our goal is to plot a time series of temperature, stratified by 
depth, so let's start by joining the data file and sensor positions 
file, to bring the depth measurements into the same data frame with 
the data.

``` {r pos-join}
# paste horizontalPosition and verticalPosition together
# to match HOR.VER
soilT$ST_30_minute$HOR.VER <- paste(soilT$ST_30_minute$horizontalPosition,
                                    soilT$ST_30_minute$verticalPosition,
                                    sep=".")

# left join to keep all temperature records
soilTHV <- merge(soilT$ST_30_minute, pos, 
                 by="HOR.VER", all.x=T)
```

And now we can plot soil temperature over time for each depth. 
We'll use `ggplot` since it's well suited to this kind of 
stratification. Each soil plot is its own panel:

```{r soilT-plot, fig.cap="---UPDATE---HERE---"}

gg <- ggplot(soilTHV, 
             aes(endDateTime, soilTempMean, 
                 group=zOffset, color=zOffset)) +
             geom_line() + 
        facet_wrap(~horizontalPosition)
gg

```

We can see that as soil depth increases, temperatures 
become much more stable, while the shallowest measurement 
has a clear diurnal cycle. We can also see that 
something has gone wrong with one of the sensors in plot 
002. To remove those data, use only values where the final 
quality flag passed, i.e. `finalQF` = 0

```{r soilT-plot-noQF}
gg <- ggplot(subset(soilTHV, finalQF==0), 
             aes(endDateTime, soilTempMean, 
                 group=zOffset, color=zOffset)) +
             geom_line() + 
        facet_wrap(~horizontalPosition)
gg
```
