---
title: Content for neonUtilities tutorial
output: html_fragment
dateCreated: '2024-09-22'
---

This tutorial provides an overview of functions in the `neonUtilities` 
package in R and the `neonutilities` package in Python. These packages 
provide a toolbox of basic functionality for working with NEON data.

This tutorial is primarily an index of 
functions and their inputs; for more in-depth guidance in using these 
functions to work with NEON data, see the <a href="https://www.neonscience.org/download-explore-neon-data" target="_blank">Download and Explore</a> 
tutorial. If you are already familiar with the `neonUtilities` package, 
and need a quick reference guide to function inputs and notation, see the 
<a href="https://www.neonscience.org/sites/default/files/cheat-sheet-neonUtilities_0.pdf" target="_blank">neonUtilities cheat sheet</a>.

## Function index {.tabset}

The `neonUtilities`/`neonutilities` package contains several functions 
(use the R and Python tabs to see the syntax in each language):

### R

* `stackByTable()`: Takes zip files downloaded from the 
<a href="http://data.neonscience.org" target="_blank">Data Portal</a> or 
downloaded by `zipsByProduct()`, unzips them, and joins 
the monthly files by data table to create a single file per table.
* `zipsByProduct()`: A wrapper for the 
<a href="http://data.neonscience.org/data-api" target="_blank">NEON API</a>; 
downloads data based on data product and site criteria. Stores 
downloaded data in a format that can then be joined by 
`stackByTable()`.
* `loadByProduct()`: Combines the functionality of `zipsByProduct()`,  
`stackByTable()`, and `readTableNEON()`: Downloads the specified data, 
stacks the files, and loads the files to the R environment.
* `byFileAOP()`: A wrapper for the NEON API; downloads remote 
sensing data based on data product, site, and year criteria. 
Preserves the file structure of the original data.
* `byTileAOP()`: Downloads remote sensing data for the specified 
data product, subset to tiles that intersect a list of 
coordinates.
* `readTableNEON()`: Reads NEON data tables into R, using the 
variables file to assign R classes to each column.
* `getCitation()`: Get a BibTeX citation for a particular data 
product and release.

### Python

* `stack_by_table()`: Takes zip files downloaded from the 
<a href="http://data.neonscience.org" target="_blank">Data Portal</a> or 
downloaded by `zips_by_product()`, unzips them, and joins 
the monthly files by data table to create a single file per table.
* `zips_by_product()`: A wrapper for the 
<a href="http://data.neonscience.org/data-api" target="_blank">NEON API</a>; 
downloads data based on data product and site criteria. Stores 
downloaded data in a format that can then be joined by 
`stack_by_table()`.
* `load_by_product()`: Combines the functionality of `zips_by_product()`,  
`stack_by_table()`, and `read_table_neon()`: Downloads the specified data, 
stacks the files, and loads the files to the R environment.
* `by_file_aop()`: A wrapper for the NEON API; downloads remote 
sensing data based on data product, site, and year criteria. 
Preserves the file structure of the original data.
* `by_tile_aop()`: Downloads remote sensing data for the specified 
data product, subset to tiles that intersect a list of 
coordinates.
* `read_table_neon()`: Reads NEON data tables into R, using the 
variables file to assign R classes to each column.
* `get_citation()`: Get a BibTeX citation for a particular data 
product and release.

## {-}

<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i> If you are only interested in joining data 
files downloaded from the NEON Data Portal, you will only need to use 
`stackByTable()`. Follow the instructions in the first section of the 
<a href="https://www.neonscience.org/download-explore-neon-data" target="_blank">Download and Explore</a> 
tutorial.
</div>


## Install and load packages {.tabset}

First, install and load the package. The installation step only 
needs to be run once, and then periodically to update when new 
package versions are released. The load step needs to be run 
every time you run your code.

### R

```{r loadStuff, eval=6}

# install neonUtilities - can skip if already installed
install.packages("neonUtilities")

# load neonUtilities
library(neonUtilities)

```

### Python

```{python p-install, eval=FALSE}

# install neonutilities - can skip if already installed
# do this in the command line
pip install neonutilities

```

```{python p-loadpack}

# load neonutilities in working environment
import neonutilities as nu

```

## {-}

## Download files and load to working environment {.tabset}

The most popular function in `neonUtilities` is `loadByProduct()` 
(or `load_by_product()` in `neonutilities`). 
This function downloads data from the NEON API, merges the site-by-month 
files, and loads the resulting data tables into the programming 
environment, classifying each variable's data type appropriately. It 
combines the actions of the `zipsByProduct()`, `stackByTable()`, and 
`readTableNEON()` functions, described below.

This is a popular choice because it ensures you're always working with the 
latest data, and it ends with ready-to-use tables. However, if you 
use it in a workflow you run repeatedly, keep in mind it will re-download 
the data every time.

`loadByProduct()` works on most observational (OS) and sensor (IS) data, 
but not on surface-atmosphere exchange (SAE) data, remote sensing (AOP) 
data, and some of the data tables in the microbial data products. For 
functions that download AOP data, see the `byFileAOP()` and `byTileAOP()` 
sections in this tutorial. For functions that work with SAE data, see 
the <a href="https://www.neonscience.org/eddy-data-intro" target="_blank">NEON eddy flux data tutorial</a>. SAE functions are not yet available in Python.

The inputs to `loadByProduct()` control which data to download and how 
to manage the processing:

### R

* `dpID`: The data product ID, e.g. DP1.00002.001
* `site`: Defaults to "all", meaning all sites with available data; 
can be a vector of 4-letter NEON site codes, e.g. 
`c("HARV","CPER","ABBY")`.
* `startdate` and `enddate`: Defaults to NA, meaning all dates 
with available data; or a date in the form YYYY-MM, e.g. 
2017-06. Since NEON data are provided in month packages, finer 
scale querying is not available. Both start and end date are 
inclusive.
* `package`: Either basic or expanded data package. Expanded data 
packages generally include additional information about data 
quality, such as chemical standards and quality flags. Not every 
data product has an expanded package; if the expanded package is 
requested but there isn't one, the basic package will be 
downloaded.
* `timeIndex`: Defaults to "all", to download all data; or the 
number of minutes in the averaging interval. See example below; 
only applicable to IS data.
* `release`: Specify a particular data Release, e.g. 
`"RELEASE-2024"`. Defaults to the most recent Release. For more 
details and guidance, see the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/release-provisional-tutorial" target="_blank">
Release and Provisional</a> 
tutorial.
* `include.provisional`: T or F: Should provisional data be 
downloaded? If `release` is not specified, set to T to include 
provisional data in the download. Defaults to F.
* `savepath`: the file path you want to download to; defaults to the 
working directory.
* `check.size`: T or F: should the function pause before downloading 
data and warn you about the size of your download? Defaults to T; if 
you are using this function within a script or batch process you 
will want to set it to F.
* `token`: Optional API token for faster downloads. See the 
<a href="https://www.neonscience.org/resources/learning-hub/tutorials/neon-api-tokens-tutorial" target="_blank">
API token</a> tutorial.
* `nCores`: Number of cores to use for parallel processing. Defaults 
to 1, i.e. no parallelization.

### Python

* `dpid`: the data product ID, e.g. DP1.00002.001
* `site`: defaults to "all", meaning all sites with available data; 
can be a list of 4-letter NEON site codes, e.g. 
`["HARV","CPER","ABBY"]`.
* `startdate` and `enddate`: defaults to NA, meaning all dates 
with available data; or a date in the form YYYY-MM, e.g. 
2017-06. Since NEON data are provided in month packages, finer 
scale querying is not available. Both start and end date are 
inclusive.
* `package`: either basic or expanded data package. Expanded data 
packages generally include additional information about data 
quality, such as chemical standards and quality flags. Not every 
data product has an expanded package; if the expanded package is 
requested but there isn't one, the basic package will be 
downloaded.
* `timeindex`: defaults to "all", to download all data; or the 
number of minutes in the averaging interval. See example below; 
only applicable to IS data.
* `release`: Specify a particular data Release, e.g. 
`"RELEASE-2024"`. Defaults to the most recent Release. For more 
details and guidance, see the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/release-provisional-tutorial" target="_blank">
Release and Provisional</a> 
tutorial.
* `include_provisional`: True or False: Should provisional data be 
downloaded? If `release` is not specified, set to T to include 
provisional data in the download. Defaults to F.
* `savepath`: the file path you want to download to; defaults to the 
working directory.
* `check_size`: True or False: should the function pause before 
downloading data and warn you about the size of your download? 
Defaults to True; if you are using this function within a script or 
batch process you will want to set it to False.
* `token`: Optional API token for faster downloads. See the 
<a href="https://www.neonscience.org/resources/learning-hub/tutorials/neon-api-tokens-tutorial" target="_blank">
API token</a> tutorial.
* `cloud_mode`: Can be set to True if you are working in a cloud 
environment; provides more efficient data transfer from NEON cloud 
storage to other cloud environments.
* `progress`: Set to False to omit the progress bar during download 
and stacking.

## {-}

The `dpID` (`dpid`) is the data product identifier of the data you want 
to download. The DPID can be found on the 
<a href="http://data.neonscience.org/data-products/explore" target="_blank">
Explore Data Products page</a>.
It will be in the form DP#.#####.###

### Demo data download and read {.tabset}

Let's get triple-aspirated air temperature data (DP1.00003.001) 
from Moab and Onaqui (MOAB and ONAQ), from May--August 2018, and 
name the data object `triptemp`:

#### R

```{r loadBy, eval=F}

triptemp <- loadByProduct(dpID="DP1.00003.001", 
                          site=c("MOAB","ONAQ"),
                          startdate="2018-05", 
                          enddate="2018-08")

```
```{r loadBy-hidden, results="hide", include=FALSE}

triptemp <- loadByProduct(dpID="DP1.00003.001", 
                          site=c("MOAB","ONAQ"),
                          startdate="2018-05", 
                          enddate="2018-08",
                          check.size=F)

```

#### Python

```{python p-loadBy, eval=F}

triptemp = nu.load_by_product(dpid="DP1.00003.001", 
                              site=["MOAB","ONAQ"],
                              startdate="2018-05", 
                              enddate="2018-08")

```
```{python p-loadBy-hidden, results="hide", include=FALSE}

triptemp = nu.load_by_product(dpid="DP1.00003.001", 
                              site=["MOAB","ONAQ"],
                              startdate="2018-05", 
                              enddate="2018-08",
                              check_size=False)

```

### {-}

### View downloaded data {.tabset}

The object returned by `loadByProduct()` is a named list of data 
tables, or a dictionary of data tables in Python. To work with each 
of them, select them from the list.

#### R

```{r loadBy-list}

names(triptemp)
temp30 <- triptemp$TAAT_30min

```

If you prefer to extract each table from the list and work 
with it as an independent object, you can use the 
`list2env()` function:

``` {r env, eval=F}

list2env(trip.temp, .GlobalEnv)

```

#### Python

```{python p-loadBy-list}

triptemp.keys()
temp30 = triptemp["TAAT_30min"]

```

If you prefer to extract each table from the list and work 
with it as an independent object, you can use  
`globals().update()`:

``` {python p-env, eval=F}

globals().update(triptemp)

```

### {-}

For more details about the contents of the data tables and metadata tables, 
check out the <a href="https://www.neonscience.org/download-explore-neon-data" target="_blank">Download and Explore</a> 
tutorial.

## Join data files: stackByTable()

The function `stackByTable()` joins the month-by-site files from a data 
download. The output will yield data grouped into new files by table name. 
For example, the single aspirated air temperature data product contains 1 
minute and 30 minute interval data. The output from this function is one 
.csv with 1 minute data and one .csv with 30 minute data. 

Depending on your file size this function may run for a while. For 
example, in testing for this tutorial, 124 MB of temperature data took 
about 4 minutes to stack. A progress bar will display while the 
stacking is in progress. 

### Download the Data

To stack data from the Portal, first download the data of interest from the 
<a href="http://data.neonscience.org" target="_blank"> NEON Data Portal</a>. 
To stack data downloaded from the API, see the `zipsByProduct()` section 
below.

Your data will download from the Portal in a single zipped file. 

The stacking function will only work on zipped Comma Separated Value (.csv) 
files and not the NEON data stored in other formats (HDF5, etc). 

### Run `stackByTable()` {.tabset}

The example data below are single-aspirated air temperature. 

To run the `stackByTable()` function, input the file path to the 
downloaded and zipped file. 

#### R

```{r run-function, eval = FALSE}

# Modify the file path to the file location on your computer
stackByTable(filepath="~neon/data/NEON_temp-air-single.zip")

```

#### Python

```{python p-run-function, eval = FALSE}

# Modify the file path to the file location on your computer
nu.stack_by_table(filepath="/neon/data/NEON_temp-air-single.zip")

```

### {-}

In the same directory as the zipped file, you should now have an unzipped 
directory of the same name. When you open this you will see a new directory 
called **stackedFiles**. This directory contains one or more .csv files 
(depends on the data product you are working with) with all the data from 
the months & sites you downloaded. There will also be a single copy of the 
associated variables, validation, and sensor_positions files, if applicable 
(validation files are only available for observational data products, and 
sensor position files are only available for instrument data products).

These .csv files are now ready for use with the program of your choice.

To read the data tables, we recommend using `readTableNEON()`, 
which will assign each column to the appropriate data type, based on the 
metadata in the variables file. This ensures time stamps and missing 
data are interpreted correctly.

## Load data to environment {.tabset}

### R

```{r readtable, eval=F}

SAAT30 <- readTableNEON(
  dataFile='~/stackedFiles/SAAT_30min.csv',
  varFile='~/stackedFiles/variables_00002.csv'
)

```

### Python

```{python p-readtable, eval=F}

SAAT30 = read_table_neon(
  dataFile='/stackedFiles/SAAT_30min.csv',
  varFile='/stackedFiles/variables_00002.csv'
)

```

## {-}

### Other function inputs {.tabset}

Other input options in `stackByTable()` are:

* `savepath` : allows you to specify the file path 
where you want the stacked files to go, overriding the default. 
Set to `"envt"` to load the files to the working environment.
* `saveUnzippedFiles` : allows you to keep the unzipped, unstacked 
files from an intermediate stage of the process; by default they 
are discarded.

Example usage:

#### R

```{r run-options, eval = FALSE}

stackByTable(filepath="~neon/data/NEON_temp-air-single.zip", 
             savepath="~data/allTemperature", saveUnzippedFiles=T)

tempsing <- stackByTable(filepath="~neon/data/NEON_temp-air-single.zip", 
                         savepath="envt", saveUnzippedFiles=F)

```

#### Python

```{python p-run-options, eval = FALSE}

nu.stack_by_table(filepath="/neon/data/NEON_temp-air-single.zip", 
                  savepath="/data/allTemperature", 
                  saveUnzippedFiles=True)

tempsing <- nu.stack_by_table(filepath="/neon/data/NEON_temp-air-single.zip", 
                              savepath="envt", 
                              saveUnzippedFiles=False)

```

### {-}

## Download files to be stacked: zipsByProduct() {.tabset}

The function `zipsByProduct()` is a wrapper for the NEON API, it 
downloads zip files for the data product specified and stores them in 
a format that can then be passed on to `stackByTable()`.

Input options for `zipsByProduct()` are the same as those for 
`loadByProduct()` described above.

Here, we'll download single-aspirated air temperature (DP1.00002.001) 
data from Wind River Experimental Forest (WREF) for April and May of 
2019.

### R

```{r run-zipsByProduct, eval = FALSE}

zipsByProduct(dpID="DP1.00002.001", site="WREF", 
              startdate="2019-04", enddate="2019-05",
              package="basic", check.size=T)

```

Downloaded files can now be passed to `stackByTable()` to be 
stacked.

```{r zips-to-stack, eval = FALSE}

stackByTable(filepath=paste(getwd(), 
                            "/filesToStack00002", 
                            sep=""))

```

### Python

```{python p-run-zipsByProduct, eval = FALSE}

nu.zips_by_product(dpid="DP1.00002.001", site="WREF", 
                   startdate="2019-04", enddate="2019-05",
                   package="basic", check_size=True)

```

Downloaded files can now be passed to `stackByTable()` to be 
stacked.

```{python p-zips-to-stack, eval = FALSE}

nu.stack_by_table(filepath=os.getcwd()+
                  "/filesToStack00002")

```

## {-}

For many sensor data products, download sizes can get 
very large, and `stackByTable()` takes a long time. The 
1-minute or 2-minute files are much larger than the 
longer averaging intervals, so if you don't need high-
frequency data, the `timeIndex` input option lets you choose 
which averaging interval to download.

This option is only applicable to sensor (IS) data, 
since OS data are not averaged.

### Download by averaging interval {.tabset}

Download only the 30-minute data for single-aspirated 
air temperature at WREF:

#### R

```{r run-zipsByProduct-avg, eval = FALSE}

zipsByProduct(dpID="DP1.00002.001", site="WREF", 
              startdate="2019-04", enddate="2019-05",
              package="basic", timeIndex=30, 
              check.size=T)

```

#### Python

```{python p-run-zipsByProduct-avg, eval = FALSE}

nu.zips_by_product(dpid="DP1.00002.001", site="WREF", 
                   startdate="2019-04", 
                   enddate="2019-05", package="basic", 
                   timeindex=30, check_size=True)

```

### {-}

The 30-minute files can be stacked and loaded as usual.

## Download remote sensing files {.tabset}

Remote sensing data files can be very large, and NEON remote sensing 
(AOP) data are stored in a directory structure that makes them easier 
to navigate. `byFileAOP()` downloads AOP files from the API while 
preserving their directory structure. This provides a convenient way 
to access AOP data programmatically.

Be aware that downloads from `byFileAOP()` can take a VERY long time, 
depending on the data you request and your connection speed. You 
may need to run the function and then leave your machine on and 
downloading for an extended period of time.

Here the example download is the Ecosystem Structure data product at 
Hop Brook (HOPB) in 2017; we use this as the example because it's a 
relatively small year-site-product combination.

### R

```{r aop-files, eval = FALSE}

byFileAOP("DP3.30015.001", site="HOPB", 
          year=2017, check.size=T)

```

### Python

```{python p-aop-files, eval = FALSE}

nu.by_file_aop(dpid="DP3.30015.001", 
               site="HOPB", year=2017, 
               check_size=True)

```

## {-}

The files should now be downloaded to a new folder in your 
working directory.

## Download remote sensing files for specific coordinates {.tabset}

Often when using remote sensing data, we only want data covering a
certain area - usually the area where we have coordinated ground 
sampling. `byTileAOP()` queries for data tiles containing a 
specified list of coordinates. It only works for the tiled, AKA 
mosaicked, versions of the remote sensing data, i.e. the ones with 
data product IDs beginning with "DP3".

Here, we'll download tiles of vegetation indices data (DP3.30026.001) 
corresponding to select observational sampling plots. For more information 
about accessing NEON spatial data, see the 
<a href="https://www.neonscience.org/neon-api-usage" target="_blank">
API tutorial</a> and the in-development <a href="https://github.com/NEONScience/NEON-geolocation/tree/master/geoNEON" 
target="_blank">
geoNEON package</a>.

For now, assume we've used the API to look up the plot centroids of 
plots SOAP_009 and SOAP_011 at the Soaproot Saddle site. You can 
also look these up in the Spatial Data folder of the 
<a href="https://data.neonscience.org/documents" target="_blank">
document library</a>. 
The coordinates of the two plots in UTMs are 298755,4101405 and 
299296,4101461. These are 40x40m plots, so in looking for tiles 
that contain the plots, we want to include a 20m buffer. The 
"buffer" is actually a square, it's a delta applied equally to 
both the easting and northing coordinates.

### R

```{r byTile, eval=F}

byTileAOP(dpID="DP3.30026.001", site="SOAP", 
          year=2018, easting=c(298755,299296),
          northing=c(4101405,4101461),
          buffer=20)

```

### Python

```{python p-byTile, eval=F}

nu.by_tile_aop(dpid="DP3.30026.001", 
               site="SOAP", year=2018, 
               easting=[298755,299296],
               northing=[4101405,4101461],
               buffer=20)

```

## {-}

The 2 tiles covering the SOAP_009 and SOAP_011 plots have  
been downloaded.


