---
syncID: 994e7c10f35b4243bbaa31243d6c2355
title: "Using the neonOS Package to Check for Duplicates and Join Tables"
description: "Use the functions in the neonOS package to process NEON observational data: check for duplicate data records and join data tables."
dateCreated: 2022-03-25
authors: [Claire K. Lunch]
contributors: 
estimatedTime: 40 minutes
packagesLibraries: neonOS, neonUtilities
topics: data-management, rep-sci
languageTool: R
dataProduct: 
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/NEON-general/neon-code-packages/neonOS/neonOS-duplicates-joins.R
tutorialSeries:
urlTitle: neonOS-duplicates-joins

---

NEON observational data are diverse in both content and in data structure. 
Generally, data are published in a set of data tables, each of which 
corresponds to a particular activity: for example, several data 
products include a field collection data table, a sample processing 
data table, and a table of laboratory analytical results.

### Joining data tables

Depending on the analysis you want to carry out, there may be data in 
multiple tables that you want to bring together. For example, it is 
very common that species identifications are in a different table from 
chemical composition or pathogen status. For species-specific analyses, 
you would need to draw on multiple tables. There are a variety of ways 
to do this, but one of the simplest is to join the tables and create a 
single flat table.

The Quick Start Guides and Data Product User Guides provide information 
about the relationships between different data tables, and we 
recommend you consult these documents for the data products you are 
working with to gain the full picture. Quick Start Guides (QSGs) are 
included in the download package when you download data via the Data 
Portal, and can also be viewed on the Data Product Details page for 
each data product; for example, see the page for <a href="https://data.neonscience.org/data-products/DP1.10026.001" target="_blank">Plant foliar traits</a>. 
Scroll down to the Documentation section to see the QSG. The Data 
Product User Guide is also available in the Documentation menu.

To join related tables, the `neonOS` package provides the 
`joinTableNEON()` function, which checks the Table joining section of 
the QSG, and uses the information there to join the tables if 
possible. If the join isn't possible, or if it requires customized code, 
the function will return an informative error, and usually refer you 
to the QSG for more details.

### Duplicates

One of the most common data entry errors that occurs in NEON OS data is 
duplicate entry. NEON data entry applications and ingest validation rules 
are designed to prevent duplicate entry where possible, but errors can't 
be avoided completely. Consequently, NEON metadata for each OS data product 
(the `variables` file) includes an indicator of which data fields, taken 
together, should define a unique record. This combination of fields is 
called the "primary key" for the data table. The `neonOS` function 
`removeDups()` uses these metadata to identify duplicate records. 
Depending on the content of the duplicate records, they may be resolved to 
a single record or marked as unresolvable - see below for details.

In this tutorial, we will de-duplicate and then join two data tables in the 
Aquatic plant bryophyte chemical properties (DP1.20063.001) data product, 
using it as an example to demonstrate the operation of these two functions. 
Then we will take a look at Mosquitoes sampled from CO2 traps (DP1.10043.001), 
which contains more complicated table relationships, and see how to modify 
the code to work with those tables.


<div id="ds-objectives" markdown="1">

## Objectives
After completing this activity, you will be able to:

 * Identify and resolve duplicate data using the `neonOS` package 
 * Join NEON data tables using the `neonOS` package 

## Things You’ll Need To Complete This Tutorial
You will need a version of R (4.0 or higher. This code may work with 
earlier versions but it hasn't been tested) and, preferably, `RStudio` 
loaded on your computer to complete this tutorial.

### Install R Packages

* **neonUtilities:** `install.packages("neonUtilities")`
* **neonOS**: `install.packages("neonOS")`

## Additional Resources

* NEON <a href="http://data.neonscience.org" target="_blank"> Data Portal </a>
* <a href="https://www.neonscience.org/resources/code-hub" target="_blank">NEON Code Hub</a>

</div>


## Set Up R Environment and Download Data

First install and load the necessary packages.

```{r setup, eval=c(9,10,11), results="hide"}

# install packages. you can skip this step if 
# the packages are already installed
install.packages("neonUtilities")
install.packages("neonOS")
install.packages("ggplot2")

# load packages
library(neonUtilities)
library(neonOS)
library(ggplot2)

```

We'll use aquatic plant chemistry (DP1.20063.001) as the example dataset. Use 
the `neonUtilities` function `loadByProduct()` to download and read in the 
data. If you're not familiar with the `neonUtilities` package and how to use 
it to access NEON data, we recommend you follow the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> 
tutorial before proceeding with this one.

Here, we'll use the same subset of aquatic plant chemistry data as in the 
Download and Explore tutorial. Get the data from the Prairie Lake, Suggs Lake, 
and Toolik Lake sites, specifying the 2022 data release.

``` {r load-data, results="hide"}

apchem <- loadByProduct(dpID="DP1.20063.001", 
                        site=c("PRLA","SUGG","TOOK"), 
                        package="expanded",
                        release="RELEASE-2022",
                        check.size=F)

```

Copy each of the downloaded tables into the R environment.

``` {r env, results="hide"}

list2env(apchem, .GlobalEnv)

```


## Identify and Resolve Duplicates

As noted above, duplicate data entry is a common error in human 
data entry. `removeDups()` uses the definitional metadata in the 
variables file to identify duplicates. It requires two inputs: 
the data table, and the corresponding variables file.

``` {r remove-dups-biomass}

apl_biomass <- removeDups(data=apl_biomass, 
                          variables=variables_20063)

```

There were no duplicates found in the `apl_biomass` data table. 
A `duplicateRecordQF` quality flag has been added to the table, 
and you can confirm that there were no duplicates by checking 
the values of the flag.

``` {r view-flag-values}

unique(apl_biomass$duplicateRecordQF)

```

All data have flag value = 0, indicating they are not duplicated. 
Let's check the `apl_plantExternalLabDataPerSample` table.

``` {r remove-dups-analytes, results="hide"}

apl_plantExternalLabDataPerSample <- removeDups(
  data=apl_plantExternalLabDataPerSample, 
  variables=variables_20063)

```

```
10 duplicated key values found, representing 20 non-unique records. Attempting to resolve.
  |==========================================================================================| 100%
2 resolveable duplicates merged into matching records
2 resolved records flagged with duplicateRecordQF=1
16 unresolveable duplicates flagged with duplicateRecordQF=2
```

The function output tells you there were 20 duplicates (out of 26108 
total data records). Four of those duplicates were resolved, leaving two 
records flagged as resolved duplicates, with `duplicateRecordQF=1`. The 
remaining 16 couldn't be resolved, and were flagged with 
`duplicateRecordQF=2`.

What does it mean for duplicates to be resolved? Some duplicates represent 
identical data that have been entered multiple times, whereas some duplicates 
have the same values in the primary key, but differ in data values. 
`removeDups()` has a fairly narrow set of criteria for resolving to a single 
record:

1. If one data record has empty fields that are populated in the other record, the non-empty values are retained.
2. If records are identical except for uid, remarks, and/or personnel (`identifiedBy`, `recordedBy`, etc) fields, unique values are concatenated within the non-matching fields, separated by `|` (pipes).

Records that can be merged to a single record by these criteria are flagged 
with `duplicateRecordQF=1`. Records with mis-matched data that can't be merged 
are retained as-is and flagged with `duplicateRecordQF=2`.

Note that even in fully identical duplicates, the `uid` field (universal 
identifier) will always be unique. Thus the `uid` field in merged records 
will always contain the pipe-delimited set of uids of the original records.

What does this look like in practice? Let's look at the two resolved 
duplicates:

``` {r view-resolved}

apl_plantExternalLabDataPerSample[which(
  apl_plantExternalLabDataPerSample$duplicateRecordQF==1),]

```

You can see that both records have two pipe-delimited uids, and are 
flagged.

Let's look at the unresolveable duplicates:

``` {r view-unresolved}

apl_plantExternalLabDataPerSample[which(
  apl_plantExternalLabDataPerSample$duplicateRecordQF==2),]

```

The key for this data table is the sample identifier and analyte, and 
here there are multiple records with the same sample identifier, both for 
carbon and nitrogen values. The most likely scenario is that these are 
unlabeled replicate analyses, i.e., the lab ran multiple analyses on the 
same samples for quality control purposes, but forgot to label them 
accordingly.

Now, how should you proceed with your analysis? Of course, that is up to 
you, and depends on your analysis. Because these appear to be unlabeled 
analytical replicates, I would probably average the analyte values, but 
a decision like this can't be made automatically - `removeDups()` can 
identify the records of concern, but what to do with them is a 
judgement call.

Of course, NEON scientists also review NEON data and identify 
duplicates as part of quality assurance and quality control procedures, 
and resolve them if possible. In the data download step above, we 
accessed RELEASE-2022. The data release is stable and reproducible over 
time, but duplicates you find in one release may be resolved in future 
releases.

## Join Data Tables

If you are using `neonOS` to check for duplicates and also to join data 
tables, the duplicate step should always come first. Because duplicate 
identification uses the variables file to determine uniqueness of data 
records, the duplicate check step requires that the data match the 
variables file exactly, which they won't after being joined.

To join the `apl_biomass` and `apl_plantExternalLabDataPerSample` tables, 
we input both tables to the `joinTableNEON()` function. It uses the 
information provided in NEON quick start guides to determine whether the 
join is possible, and if it is, which fields to use to perform the join.

```{r join, results="hide"}

aqbc <- joinTableNEON(apl_biomass,
                      apl_plantExternalLabDataPerSample)

```

After joining tables, always take a look at the resulting table and 
make sure it makes sense. Errors in joining can easily result in 
completely nonsensical data. If you're not familiar with table 
joining operations, check out a lesson on the basics. The chapter on 
relational data in <a href="https://r4ds.had.co.nz/relational-data.html" target="_blank">R for Data Science</a> is a good one.

When checking your results, keep in mind that the default behavior of 
`joinTableNEON()` is a full join, i.e., all records from both original 
tables are retained, even if they don't match. For a small number of 
table pairs, the Quick Start Guide specifies a left join, and in those 
cases `joinTableNEON()` performs a left join.

Let's take a look at the aquatic plant table join:

```{r aqbc-check}

nrow(apl_biomass)
nrow(apl_plantExternalLabDataPerSample)
nrow(aqbc)

```

The number of rows in the joined table is larger than both of the 
original tables, but smaller than the sum of the original two. This 
suggests that most of the records in each of the original tables had 
a matching record in the other table, but some didn't.

View the full table:

```{r aqbc-view, results="hide"}

View(aqbc)

```

(Table not displayed here due to size)

Here we can see we have several rows per `chemSubsampleID` for most 
`chemSubsampleID`s. Each row corresponds to one of the chemical 
analytes, and the biomass data are repeated on each row. At the bottom 
of the table are a number of biomass records with no corresponding 
chemistry data; these explain why the merged table is larger than the 
original chemistry table.

This table structure is consistent with the original tables and with 
the intended join, so we're satisfied all is well. If you're working 
with a different data product and encounter something unexpected or 
undesirable in a joined table, contact NEON using the 
<a href="https://www.neonscience.org/about/contact-us" target="_blank">Contact Us</a> page.

We can now connect chemical content to taxon, as in the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> 
tutorial. Let's look at nitrogen content by species and site:

```{r aq-fig}

gg <- ggplot(subset(aqbc, analyte=="nitrogen"),
             aes(scientificName, analyteConcentration, 
                 group=scientificName, 
                 color=scientificName)) +
             geom_boxplot() + 
             facet_wrap(~siteID) +
        theme(axis.text.x=element_text(angle=90,
                                       hjust=1,
                                       size=4)) +
        theme(legend.position="none") +
        ylab("Nitrogen (%)") + 
        xlab("Scientific name")
gg

```


## Other Input Options

After downloading the data above, we ran `list2env()` to make each table 
an independent object in the R environment, and then provided the tables 
to the two functions as-is. This worked because the names of the objects 
were identical to the table names, so the functions were able to figure 
out which tables they were. If the object names are not exactly equal to 
the table names, you will need to input the table names separately. If 
we hadn't used `list2env()`, this is how we would proceed:

```{r with-table-names, results="hide"}

bio.dup <- removeDups(data=apchem$apl_biomass,
                      variables=apchem$variables_20063,
                      table="apl_biomass")
chem.dup <- removeDups(data=apchem$apl_plantExternalLabDataPerSample,
                      variables=apchem$variables_20063,
                      table="apl_plantExternalLabDataPerSample")
aq.join <- joinTableNEON(table1=bio.dup, 
                         table2=chem.dup,
                         name1="apl_biomass",
                         name2="apl_plantExternalLabDataPerSample")

```

## More Complicated Table Joins

In the aquatic plant chemistry example, we were able to join two tables 
in a single step. In some cases, the relationship between tables is 
more complicated, and joining is more difficult. In these cases, 
`joinTableNEON()` will provide an error message, and usually direct 
you to the Quick Start Guide for more information. Let's walk through 
how you can use this information, using Mosquitoes sampled from CO2 traps 
(DP1.10043.001) from Toolik Lake as an example.

```{r load-mos, results="hide"}

mos <- loadByProduct(dpID="DP1.10043.001",
                     site="TOOL", 
                     release="RELEASE-2022",
                     check.size=F)
list2env(mos, .GlobalEnv)

```

Let's say we're interested in evaluating which mosquito species are found in 
which vegetation types. The species identifications are in the 
`mos_expertTaxonomistIDProcessed` table, and the trapping conditions are in 
the `mos_trapping` table. So we'll attempt to join those two tables.

```{r try-join-mos, eval=FALSE}

mos.sp <- joinTableNEON(mos_trapping,
                        mos_expertTaxonomistIDProcessed)

```

```
Error in joinTableNEON(mos_trapping, mos_expertTaxonomistIDProcessed) : 
  Tables mos_trapping and mos_expertTaxonomistIDProcessed can't be joined directly, but can each be joined to a common table(s). Consult quick start guide for details.
```

The function returns an error, telling us it can't perform a simple join 
on the two tables, but there is a table they can each join to, which can be 
used to join them indirectly. As directed, we refer to the Quick Start Guide 
for DP1.10043.001, on the <a href="https://data.neonscience.org/data-products/DP1.10043.001" target="_blank">Data Product Details</a> page.

From the QSG, we learn that `mos_sorting` is the intermediate table 
between `mos_trapping` and `mos_expertTaxonomistIDProcessed`.

First, let's join `mos_trapping` and `mos_sorting`:

```{r join-trap-sort, results="hide"}

mos.trap <- joinTableNEON(mos_trapping,
                          mos_sorting)

```

Now, this next step is a bit odd. We've created a merged table of 
`mos_trapping` and `mos_sorting`, but we know 
`mos_expertTaxonomistIDProcessed` can only join to `mos_sorting`. So 
we pass the merged table to `joinTableNEON()`, telling the function to 
use the join instructions for `mos_sorting` and 
`mos_expertTaxonomistIDProcessed`.

```{r join-sort-tax, results="hide"}

mos.tax <- joinTableNEON(mos.trap,
                         mos_expertTaxonomistIDProcessed,
                         name1="mos_sorting")

```

When you join data in this way, check carefully that the resulting 
table is structured logically and contains the data you expect it to. 
Looking at the merged table, we now have multiple records for each 
trapping event, with one record for each species captured in that event, 
plus a set of records for traps that either caught no mosquitoes or 
couldn't be deployed, and thus have no species identifications. This 
is consistent with the table join we performed, so everything appears 
to be correct.

Let's take a look at species occurrence by habitat, as we set out to 
do:

```{r mos-fig}

gg <- ggplot(mos.tax, 
             aes(scientificName, individualCount, 
                 group=scientificName, 
                 color=scientificName)) +
             geom_boxplot() + 
        facet_wrap(~nlcdClass) +
        theme(axis.text.x=element_blank()) +
        ylab("Count") +
        xlab("Scientific name")
gg

```




