## What is a function?

```
input --> function does something --> output
```

One of the main benefits of writing functions is to avoid repetition of your
code. When you write code you should aim for laziness. If you are about to copy
a piece of code 5 times and just change a variable in each instance, you are
better off converting it into a function. In general, you should never repeat
yourself when you write code, it's called the DRY (Don't Repeat Yourself)
principle.

Another advantage is that your code is self contained and any variable that you
create inside a function will not be exported into your global environment.

## How to write functions in R?

```{r, purl=FALSE}
## Fahrenheit to Celsius
(70 - 32) * 5/9
(65 - 32) * 5/9
(85 - 32) * 5/9
```

Let's convert this into a function:

```{r, purl=FALSE}
fahr_to_celsius <- function(temp) {
    (temp - 32) * 5/9
}
```

Don't forget to re-evaluate your function, after modifying it.

A little better:

```{r, purl=FALSE}
convert_fahr <- function(temp, to) {
    res <- (temp - 32) * 5/9
    if (to == "kelvin") {
        res <- res + 273.15
    }
    res
}
```


With functions you can easily control the format of the input and avoid the
chances for typos or other small mistakes.

```{r, purl=FALSE}
convert_fahr <- function(temp, to = c("celsius", "kelvin")) {
    to <- tolower(to)
    to <- match.arg(to)
    res <- (temp - 32) * 5/9
    if (to == "kelvin") {
        res <- res + 273.15
    }
    res
}
```

Let's refactor this function into something even better that will allow you to
easily expend the `convert_fahr` function to other units:

```{r, purl=FALSE}
fahr_to_celsius <- function(temp) {
    (temp - 32) * 5/9
}

celsius_to_kelvin <- function(temp) {
    temp + 273.15
}

convert_fahr <- function(temp, to = c("celsius", "kelvin")) {
    to <- tolower(to)
    to <- match.arg(to)
    res <- fahr_to_celsius(temp)
    if (to == "kelvin") {
        res <- celsius_to_kelvin(res)
    }
    res
}
```

Your intentions are clear, everything is self contained, you can easily debug,
test and document each of the steps involved.

### Challenge

Write a function that converts pounds in kilograms (divides by 2.2).

Stretch goal: and in grams.
