---
title: "Functions for data"
output:
  ioslides_presentation:
    highlight: pygments
    widescreen: yes
---

# Basics

## Using functions to...

Now that we know how to write functions, we can use this concept for 

- preparing our data sets for analysis, 
- generating the intermediary data files, 
- and exporting them to CSV so that we can, for instance, share them with our collaborators.

## Functionalizing a chunk

Let's start with the chunk from the manuscript:

```{r, purl=TRUE}
## Gathering all the data files
split_gdp_files <- list.files(path = "data-raw", pattern = "gdp-percapita\\.csv$", 
                              full.names = TRUE)
split_gdp_list <- lapply(split_gdp_files, read.csv)
gdp <- do.call("rbind", split_gdp_list)
```

- Enclose these lines of code inside curly brackets 
- Don't foget to return the `gdp` variable on the last line

## Functionalizing a chunk (cont.)

- Enclose these lines of code inside curly brackets 
- Don't foget to return the `gdp` variable on the last line

```{r, purl=FALSE}
gather_gdp_data <- function() {
    split_gdp_files <- list.files(path = "data-raw", pattern = "gdp-percapita\\.csv$", 
                                  full.names = TRUE)
    split_gdp_list <- lapply(split_gdp_files, read.csv)
    gdp <- do.call("rbind", split_gdp_list)
    gdp
}
```

<br>

**Demo 1:** Function returns a data.frame with all countries combined

## Making functions more versatile

- Use the folder where the files are stored and the pattern as arguments: `path` and `pattern`,
respectively

- Allows for re-usability for another project where a similar operation (combining many CSV files 
into a single `data.frame`) would be needed

```{r, purl=FALSE}
gather_data <- function(path = "data-raw", pattern = "gdp-percapita\\.csv$") {
  split_files <- list.files(path = path, pattern = pattern, full.names = TRUE)
  split_list <- lapply(split_files, read.csv)
  res <- do.call("rbind", split_list)
  res
}
```

**Demo 2:** Revised function with inputs

## Caveats

- The code here is pretty simple because we know that all datasets have
exactly the same columns, but in a real life example, we might way to add
additional checks to ensure that we won't be introducing any issues.

- This also illustrates how general you need to be when writing your
functions. We could spend a lot of time optimizing and writing a function
that would work on all cases. Sometimes it's worth your time, sometimes it
might distract from your primary goal: writing the manuscript.

## Towards automation

- We can create a `make_csv` function to automatically generate CSV files from our
data sets. 

- This might come handy if you want to send your intermediate datasets
to your collaborators or if you want to inspect more closely that everything is
working as it should.

- This function should take a dataframe and make a CSV file out of it.

## Writing the function {.build}

- Start by just typing the `write.csv`
- Use `row.names = FALSE` because we don't want them in the output
- `verbose` is useful for keeping track of progress
- `dir.create` creates the directory in the path, `recursive = TRUE` only
does it if directory dosn't exist

```{r, purl=FALSE}
make_csv <- function(obj, path, file, verbose = TRUE) {
  if (verbose) {
    message("Creating csv file: ", file)
  }
  dir.create(path, showWarnings = FALSE, recursive = TRUE)  
  write.csv(obj, file = paste0(path, "/", file), row.names = FALSE)
}
```

## Putting it all togehter

Combine the two functions we just wrote (`make_csv` and `gather_data`) to generate 
a CSV file that contains the data from all countries:

```{r, eval=FALSE}
gdp_data <- gather_data()
make_csv(gdp_data, path = "data-output", file = "gdp.csv", verbose = TRUE)
```

<br>

**Demo 3:** Run the functions.

## Your turn

Transform these two pieces of code into functions:

```{r, eval=FALSE}
## Turn this into a function called get_mean_lifeExp
mean_lifeExp_by_cont <- gdp %>%
    group_by(continent, year) %>%
    summarize(mean_lifeExp = mean(lifeExp)) %>%
    as.data.frame
```

```{r, eval=FALSE}
## Turn this into a function called get_latest_lifeExp
latest_lifeExp <- gdp %>%
    filter(year == max(gdp$year)) %>%
    group_by(continent) %>%
    summarize(latest_lifeExp = mean(lifeExp)) %>%
    as.data.frame
```

<!--

# Long computations

## Caching

- Caching is available in R Markdown but it can be pretty fragile. 

- For instance, caching is only based on whether the code in your chunk changes and 
doesn't check if your data on your hard drive is changing.

## Saving R objects

If the output of your R code can't be represented in a CSV file,
so you need to save it directly into an R object

```{r, purl=TRUE}
## If you need to save an R object to avoid the repetition of long computations
make_rds <- function(obj, file, ..., verbose = TRUE) {
    if (verbose) {
        message("Creating rds file: ", file)
    }
    saveRDS(obj, file = file)
    invisible(file.exists(file))
}
```

## Use in R Markdown documents

In your R Markdown document, you can do:

`gdp <- readRDS(file="data-output/gdp.rds")`

or maybe even:

```{r, eval=FALSE}
## An example of the kind of code you could use to work with time-consuming
## computations in R.
if ( !file.exists("data-output/gdp.rds")) {
    gdp <- gather_gdp_data() ## long computation...
    make_rds(gdp, file="data-output/gdp.rds")
}
gdp <- readRDS(file="data-output/gdp.rds")
```

-->