---
syncID: 774cad9fb9214d159157581471423312
title: "Create A Square Buffer Around a Plot Centroid in R"
description: "This tutorial walks you through creating square polygons from a plot centroid (x,y format) in R."
dateCreated: 2014-10-29
authors: Leah A. Wasser, Natalie Robinson, Sarah Elmendorf, Megan A. Jones
contributors:
estimatedTime: 1.0 - 1.5 Hours
packagesLibraries: sp, rgdal
topics: spatial-data-gis, raster
languagesTool: R
dataProduct:
code1: /R/primer-raster-data/Creating-Square-Plot-Boundaries-From-Centroids-in-R.R
tutorialSeries: primer-raster-data-R
urlTitle: field-data-polygons-centroids
---

Want to use plot centroid values (marking the center of a plot) in x,y format 
to get the plot boundaries of a certain size around the centroid? This tutorial
is for you! 

If the plot is a circle, we can generate the plot boundary using a buffer 
function in R or a GIS program. However, creating a **square** 
boundary around a centroid requires an alternate approach. This tutorial 
presents a way to create square polygons of a given radius (referring to half 
of the plot's width) for each plot centroid location in a dataset.

**Special thanks** to 
<a href="http://stackoverflow.com/users/489704/jbaums" target="_blank"> jbaums</a>
from StackOverflow for helping with the SpatialPolygons code!

<div id="ds-objectives" markdown="1">

## Learning Objectives
After completing this activity, you will be able to: 

* Create square polygons around a centroid point. 
* Export shapefiles from R using the `writeOGR()` function.

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and, preferably, `RStudio` loaded
on your computer to complete this tutorial.

### Install R Packages

* **rgdal:** `install.packages("rgdal")`
* **sp:** `install.packages("sp")`

<a href="https://www.neonscience.org/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

## Download Data

<h3> <a href="https://ndownloader.figshare.com/files/7907590"> NEON Teaching Data Subset: Field Site Spatial Data</a></h3>

These remote sensing data files provide information on the vegetation at the 
<a href="https://www.neonscience.org/" target="_blank"> National Ecological Observatory Network's</a> 
<a href="https://www.neonscience.org/field-sites/field-sites-map/SJER" target="_blank"> San Joaquin Experimental Range</a> 
and 
<a href="https://www.neonscience.org/field-sites/field-sites-map/SOAP" target="_blank"> Soaproot Saddle</a> 
field sites. The entire dataset can be accessed by request from the 
<a href="http://data.neonscience.org" target="_blank"> NEON Data Portal</a>.

<a href="https://ndownloader.figshare.com/files/7907590" class="link--button link--arrow">
Download Dataset</a>




This data download contains several files. You will only need the `SJERPlotCentroids.csv`
file for this tutorial. The path to this file is: NEON-DS-Field-Site-Spatial-Data/SJER/PlotCentroids/SJERPlotCentroids.csv . 
The other data files in the downloaded data directory are used for related tutorials. 
You should set your working directory to the parent directory of the downloaded 
data to follow the code exactly. 

****

**Set Working Directory:** This lesson assumes that you have set your working 
directory to the location of the downloaded and unzipped data subsets. 

<a href="https://www.neonscience.org/set-working-directory-r" target="_blank"> An overview
of setting the working directory in R can be found here.</a>

**R Script & Challenge Code:** NEON data lessons often contain challenges that reinforce 
learned skills. If available, the code for challenge solutions is found in the
downloadable R script of the entire lesson, available in the footer of each lesson page.


</div>

Our x,y coordinate centroids come in a ".csv" (Comma Separated Value) file with 
the plot ID that goes with the data. The data we are using today were collected 
at the National Ecological Observatory Network field site at the 
San Joaquin Experimental Range (SJER) in California. 

## Load .csv, Setup Plots 

To work with our spatial data in R, we can use the `rgdal` package and the 
`sp` package. Once we've loaded these packages and set the working directory to
the where our .csv file with the data is located, we can load our data. 

```{r load-data}
# load the sp and rgdal packages

library(sp)
library(rgdal)

# set working directory to data folder
#setwd("pathToDirHere")


# read in the NEON plot centroid data 
# `stringsAsFactors=F` ensures character strings don't import as factors
centroids <- read.csv("C:/Users/mccahill/Documents/GitHub/NEON-DS-Field-Site-Spatial-Data/SJER/PlotCentroids/SJERPlotCentroids.csv", stringsAsFactors=FALSE)
```

Let's look at our data. This can be done several ways but one way is to view 
the structure (`str()`) of the data. 

``` {r view-data}
# view data structure
str(centroids)
```

We can see that our data consists of five distinct types of data:

* Plot_ID: denotes the plot
* Point: denotes where the point is taken -- all are centroids
* northing: northing coordinate for point
* easting: easting coordinate for point
* Remarks: any other remarks from those collecting the data

It would be nice to have a metadata file with this .csv to confirm the coordinate
reference system (CRS) that the points are in, however, without one, based on 
the numbers, we can assume it is in Universal Transverse Mercator (UTM). And 
since we know the data are from the San Joaquin Experimental Range, that is in
UTM zone 11N. 

## Part 1: Create Plot Boundary

Now that we understand our centroid data file, we need to set how large our plots
are going to be. The next piece of code sets the "radius"" for the plots. 
This radius will later be used to calculate vertex locations that define the plot 
perimeter. 

In this case, let's use a **radius of 20m**. This means that the edge of each plot 
(not the corner) is 20m from the centroid. Overall this will create a 40 m x 40 m 
square plot. 

Units: Radius is in meters, matching the UTM CRS. If you're coordinates were in 
lat/long or some other CRS than you'd need to modify the code. 

Plot Orientation: Our code is based on simple geometry and assumes that plots 
are oriented North-South. If you wanted a different orientation, 
adjust the math accordingly to find the corners. 

``` {r set-radius}
# set the radius for the plots
radius <- 20 # radius in meters

# define the plot edges based upon the plot radius. 

yPlus <- centroids$northing+radius
xPlus <- centroids$easting+radius
yMinus <- centroids$northing-radius
xMinus <- centroids$easting-radius

```

When combining the coordinates for the vertices, it is important to close the 
polygon. This means that a square will have 5 instead of 4 vertices. The fifth 
vertex is identical to the first vertex. Thus, by repeating the first vertex 
coordinate (xMinus,yPlus) the polygon will be closed. 

The `cbind()` function allows use to combine or bind together data by column. Make
sure to create the vertices in an order that makes sense. We recommend starting 
at the NE and proceeding clockwise. 

``` {r create-perimeter}
# calculate polygon coordinates for each plot centroid. 
square=cbind(xMinus,yPlus,  # NW corner
	xPlus, yPlus,  # NE corner
	xPlus,yMinus,  # SE corner
	xMinus,yMinus, # SW corner
	xMinus,yPlus)  # NW corner again - close ploygon

```

Next, we will associate the centroid plot ID, from the .csv file, with the plot 
perimeter polygon that we create below. First, we extract the Plot_ID from our 
data. Note that because we set `stringsAsFactor` to false when importing, we can 
extract the Plot_IDs using the code below. If we hadn't do that, our IDs would 
come in as factors and we'd thus have to use the code
`ID=as.character(centroids$Plot_ID)`. 

``` {r get-id}

# Extract the plot ID information
ID=centroids$Plot_ID

```

We are now left with two key "groups" of data:

* a dataframe `square` which has the points for our new 40x40m plots
* a list`ID` with the `Plot_ID`s for each new 40x40m plot

If all we wanted to do was get these points, we'd be done. But no, we want to 
be able to create maps with our new plots as polygons and have them as spatial 
data objects for later analyses.

## Part 2: Create Spatial Polygons

Now we need to convert our dataframe `square` into a SpatialPolygon object. This
particular step is somewhat confusing. Please consider reading up on the 
SpatialPolygon object in R
<a href="https://cran.r-project.org/web/packages/sp/sp.pdf" target="_blank"> in 
the `sp` package documentation (pg 86) </a> 
or check out this 
<a href="http://stackoverflow.com/questions/26620373/spatialpolygons-creating-a-set-of-polygons-in-r-from-coordinates" target="_blank"> StackOverflow thread</a>. 

Two general consideration: 

First, spatial polygons require a list of lists. Each list contains the xy 
coordinates of each vertex in the polygon - in order. It is always important 
to include the closing vertex as we discussed above -- you'll have to repeat the 
first vertex coordinate.

Second, we need to specify the CRS string for our new polygon. We will do this
with a `proj4string`. We can either type in the `proj4string` (as we do below) or
we can grab the string from another file that has CRS information.
To do this, we'd use the syntax: 

`proj4string =CRS(as.character(FILE-NAME@crs))` 

For example, if we imported a GeoTIFF file called "canopy" that was in a 
UTM coordinate system, we could type `proj4string-CRS(as.character(canopy@crs))`. 

### Method 1: mapply function
We'll do this in two different ways. The first, using the `mapply()` function
is far more efficient. However, the function hides a bit of what is going on so
next we'll show how it is done without the function so you understand it.

``` {r mapply}
# create spatial polygons from coordinates
polys <- SpatialPolygons(mapply(function(poly, id) 
		{
	  xy <- matrix(poly, ncol=2, byrow=TRUE)
	  Polygons(list(Polygon(xy)), ID=id)
	  }, 
	split(square, row(square)), ID),
	proj4string=CRS(as.character("+proj=utm +zone=11 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")))
```

Let's create a simple plot to see our new SpatialPolygon data. 

``` {r polys-plot}
# plot the new polygons
plot(polys)

```

Yay! We created polygons for all of our plots! 

### Method 2: Using loops

Let's do the process again with simpler R code so that we understand how the 
process works. Keep in mind that loops are less efficient to process your data 
but don't hide as much under the box. Once you understand how this works, we 
recommend the `mapply()` function for your actual data processing. 

``` {r looping-it}
# First, initialize a list that will later be populated
# a, as a placeholder, since this is temporary
a <- vector('list', length(2))

# loop through each centroid value and create a polygon
# this is where we match the ID to the new plot coordinates
for (i in 1:nrow(centroids)) {  # for each for in object centroids
	  a[[i]]<-Polygons(list(Polygon(matrix(square[i, ], ncol=2, byrow=TRUE))), ID[i]) 
	  # make it an Polygon object with the Plot_ID from object ID
	}

# convert a to SpatialPolygon and assign CRS
polysB<-SpatialPolygons(a,proj4string=CRS(as.character("+proj=utm +zone=11 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")))

```

Let's see if it worked with another simple plot. 

``` {r polysB-plot}
# plot the new polygons
plot(polysB)

```

Good. The two methods return the same plots. We now have our new plots saved as 
a SpatialPolygon but how do we share that with our colleagues? One way is to turn
them into shapefiles, which can be read into R, Python, QGIS, ArcGIS, and many
other programs. 

## Part 3: Export to Shapefile

Before you can export a shapefile, you need to convert the `SpatialPolygons` to a 
`SpatialPolygonDataFrame`. Note that in this step you could add additional 
attribute data if you wanted to! 

``` {r create-spdf}

# Create SpatialPolygonDataFrame -- this step is required to output multiple polygons.
polys.df <- SpatialPolygonsDataFrame(polys, data.frame(id=ID, row.names=ID))
```

Let's check out the results before we export. And we can add color this time. 

```{r polysdf-plot}
plot(polys.df, col=rainbow(50, alpha=0.5))
```

When we want to export a spatial object from R as a shapefile, `writeOGR()` is a 
nice function. It writes not only the shapefile, but also the associated 
Coordinate Reference System (CRS) information as long as it is associated with 
the spatial object (e.g., if it was identified when creating the SpatialPolygons 
object). 

To do this we need the following arguments:

1. the name of the spatial object (`polys.df`)
2. file path from the current working directory for the directory where we want 
to save our shapefile. If we want it in our current directory we can simply use `'.'`.
3.the name of the new shapefile (`2014Plots_SJER`)
4. the driver which specifies the file format (`ESRI Shapefile`)

We can now export the spatial object as a shapefile.

``` {r write-ogr}

# write the shapefiles 
writeOGR(polys.df, '.', '2014Plots_SJER', 'ESRI Shapefile')
```

And there you have it -- a shapefile with a square plot boundary around your
centroids. Bring this shapefile into QGIS or whatever GIS package you prefer 
and have a look! 

For more on working with shapefiles in R, check out our 
<a href="https://www.neonscience.org/vector-data-series" target="_blank"> *Working with Vector Data in R* series </a>.
