---
syncID: FILL IN
title: "Using Footprints and Reflectance Indices to Interpret Flux Data"
description: "Explore the relationship between NDVI in the flux footprint and NEE. This tutorial serves as a guide to bringing together the flux footprints and remote sensing data more generally."
dateCreated: 2022-06-07
authors: [Claire K. Lunch, Chris Florian]
contributors: 
estimatedTime: 40 minutes
packagesLibraries: neonUtilities, raster
topics: 
languageTool: R
dataProduct: DP4.00200.001, DP3.30026.001
code1: FILL IN
tutorialSeries:
urlTitle: flux-footprint-NDVI

---

Flux footprints represent an estimate of the region of land surface 
contributing to the fluxes estimated by eddy covariance. The exact 
footprint varies from moment to moment, due to changes in wind 
speed and direction, so footprints are typically presented as 
proportional estimates of the contribution from each point over the 
relevant time period. See 
<a href="https://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.922.4124&rep=rep1&type=pdf" target="_blank">Amiro 1998</a> for details and considerations about 
footprint calculations.

Because they represent regions of land around the flux tower, flux 
footprints are ideally suited to be used in combination with remotely 
sensed data. In this tutorial, we attempt to use flux footprints 
together with NDVI to estimate net ecosystem exchange (NEE), but the 
general principles used here can be applied to many data types in the 
remote sensing and eddy covariance suites (for example, canopy water 
content and latent heat fluxes, canopy height increment and 
cumulative carbon uptake, etc).

### Background: NDVI in the flux footprint

Let's say we want to use reflectance data to scale up NEE, by 
extrapolating flux data across a larger landscape. A first step might 
be to understand how reflectance relates to NEE from year to year in 
the same location.

NEON remote sensing flights follow a detailed schedule, as described 
on the <a href="https://www.neonscience.org/data-collection/flight-schedules-coverage" target="_blank">flight schedule page</a>. Although flights are generally 
carried out at a similar time of year at each site over time, weather 
conditions can shift the exact date, and local climate and phenology 
may result in flights capturing slightly different conditions on the 
ground in different years. And of course, local conditions each year 
may result in different vegetation and soil conditions even at the same 
time of year and same phenological stage.

The flux footprints enable us to greatly improve the accuracy of the 
relationship between reflectance and fluxes - instead of guessing at 
the appropriate region around the tower to use for the comparison, we 
can use the region identified by the footprint, and weighted by the 
relative contribution of each pixel to the footprint.

To begin to bring together these data sources, in this tutorial, we will 
examine the relationship between NDVI within the flux footprint and NEE 
on the day the remote sensing flight took place. In order to give 
ourselves the best odds of detecting differences, we'll focus on sites 
where there can be large differences in NDVI from year to year, and at 
different times of year: grasslands and deciduous forests.


<div id="ds-objectives" markdown="1">

## Objectives
After completing this activity, you will be able to:

 * Create a raster of NEON flux footprint data
 * Get the NDVI data corresponding to the footprint area
 * Calculate the weighted average NDVI within the footprint

## Things You’ll Need To Complete This Tutorial
You will need a version of R (4.0 or higher. This code may work with 
earlier versions but it hasn't been tested) and, preferably, `RStudio` 
loaded on your computer to complete this tutorial.

### Install R Packages

* **neonUtilities:** `install.packages("neonUtilities")`
* **raster**: `install.packages("raster")`
* **rhdf5**: `install.packages("BiocManager")`, then `BiocManager::install("rhdf5")`

## Additional Resources

* NEON <a href="http://data.neonscience.org" target="_blank"> Data Portal </a>
* <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore tutorial</a>
* <a href="https://www.neonscience.org/resources/learning-hub/tutorials/eddy-data-intro" target="_blank">Introduction to NEON flux data</a>

</div>


## Set Up R Environment and Download Data

First install and load the necessary packages.

```{r setup, eval=c(10,11), results="hide"}

# install packages. can skip this step if already installed
# update neonOS installation after CRAN release
install.packages("neonUtilities")
install.packages("raster")
install.packages("BiocManager")
BiocManager::install("rhdf5")

# load packages
library(neonUtilities)
library(raster)

```

For this tutorial, we will need three data sets from two NEON 
data products: NDVI is published in Vegetation indices (DP3.30026.001), 
and NEE and footprints are published in Eddy covariance (DP4.00200.001). 
If you're unfamiliar with downloading and wrangling NEON data, start with 
the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore</a> tutorial, and consider the 
<a href="https://www.neonscience.org/resources/learning-hub/tutorials/eddy-data-intro" target="_blank">flux data</a> tutorial as well.

INSERT DESCRIPTION OF SITES USED
BLAN, UKFS deciduous
CPER, DCFS, NOGP grassland
KONZ, CLBJ deciduous/grassland


``` {r load-data, results="hide"}

apchem <- loadByProduct(dpID="DP1.20063.001", 
                        site=c("PRLA","SUGG","TOOK"), 
                        package="expanded",
                        release="RELEASE-2022",
                        check.size=F)

```

Copy each of the downloaded tables into the R environment.

``` {r env, results="hide"}

list2env(apchem, .GlobalEnv)

```




