---
syncID: FILL IN
title: "Using the neonOS Package to Check for Duplicates and Join Tables"
description: "Use the functions in the neonOS package to process NEON observational data: check for duplicate data records, and join data tables."
dateCreated: 2022-03-25
authors: [Claire K. Lunch]
contributors: 
estimatedTime: 40 minutes
packagesLibraries: neonOS, neonUtilities
topics: data-management, rep-sci
languageTool: R
dataProduct: 
code1: FILL IN
tutorialSeries:
urlTitle: neonOS-duplicates-joins-tutorial

---

NEON observational data are diverse in both content and in data structure. 
Generally, data are published in a set of data tables, each of which 
corresponds to a particular activity: for example, several data 
products includes a field collection data table, a sample processing 
data table, and a table of laboratory analytical results.

### Joining data tables

Depending on the analysis you want to carry out, there may be data in 
multiple tables that you want to bring together. For example, it is 
very common that species identifications are in a different table from 
chemical composition or pathogen status. For species-specific analyses, 
you would need to draw on multiple tables. There are a variety of ways 
to do this, but the simplest is to join the tables and create a single 
flat table.

The Quick Start Guides and Data Product User Guides provide information 
about the relationships between different data tables, and we 
recommend you consult these documents for the data products you are 
working with to gain the full picture. For the step of joining related 
tables, the `neonOS` package provides the `joinTableNEON()` function, 
which will join tables if their relationship allows this to be done 
automatically, and will provide an informative message if this is not 
possible.

### Duplicates

One of the most common data entry errors that occurs in NEON OS data is 
duplicate entry. NEON data entry applications and ingest validation rules 
are designed to prevent duplicate entry where possible, but errors can't 
be avoided completely. Consequently, NEON metadata for each data product 
(the variables file) includes an indicator of which data fields, taken 
together, should define a unique record. The `neonOS` function `removeDups()` 
uses these metadata to identify duplicate records. Depending on the content 
of the duplicate records, they may be resolved to a single record or 
marked as unresolvable - see below for details.

In this tutorial, we will de-duplicate and then join two data tables in the 
Aquatic plant bryophyte chemical properties (DP1.20063.001) data product, 
using it as an example to demonstrate the operation of these two functions.


<div id="ds-objectives" markdown="1">

## Objectives
After completing this activity, you will be able to:

 * Identify and resolve duplicate data using the `neonOS` package 
 * Join NEON data tables using the `neonOS` package 

## Things You’ll Need To Complete This Tutorial
You will need a version of R (4.0 or higher. This code may work with 
earlier versions but it hasn't been tested) and, preferably, `RStudio` 
loaded on your computer to complete this tutorial.

### Install R Packages

* **neonUtilities:** `install.packages("neonUtilities")`
* **neonOS**: `devtools::install_github("NEONScience/NEON-OS-data-processing/neonOS")`

## Additional Resources

* NEON <a href="http://data.neonscience.org" target="_blank"> Data Portal </a>
* <a href="https://www.neonscience.org/resources/code-hub" target="_blank">NEON Code Hub</a>

</div>


## Set Up R Environment and Download Data

First install and load the necessary packages.

```{r setup, eval=c(8,9), results="hide"}

# install packages. can skip this step if already installed
# update neonOS installation after CRAN release
install.packages("neonUtilities")
devtools::install_github("NEONScience/NEON-OS-data-processing/neonOS")

# load packages
library(neonUtilities)
library(neonOS)

```

We'll use aquatic plant chemistry (DP1.20063.001) as the example dataset. Use 
the `neonUtilities` function `loadByProduct()` to download and read in the 
data. If you're not familiar with the `neonUtilities` package and how to use 
it to access NEON data, we recommend you follow the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> 
tutorial before proceeding with this one.

Here, we'll use the same subset of aquatic plant chemistry data as in the 
Download and Explore tutorial. Get the data from the Prairie Lake, Suggs Lake, 
and Toolik Lake sites, specifying the 2022 data release.

``` {r load-data, results="hide"}

apchem <- loadByProduct(dpID="DP1.20063.001", 
                        site=c("PRLA","SUGG","TOOK"), 
                        package="expanded",
                        release="RELEASE-2022",
                        check.size=F)

```

Copy each of the downloaded tables into the R environment.

``` {r env, results="hide"}

list2env(apchem, .GlobalEnv)

```


## Identify and Resolve Duplicates

As noted above, duplicate data entry is a common error in human 
data entry. `removeDups()` uses the definitional metadata in the 
variables file to identify duplicates. It requires two inputs: 
the data table, and the corresponding variables file.

``` {r remove-dups-biomass}

apl_biomass <- removeDups(data=apl_biomass, 
                          variables=variables_20063)

```

There were no duplicates found in the `apl_biomass` data table. 
A `duplicateRecordQF` quality flag has been added to the table, 
and you can confirm that there were no duplicates by checking 
the values of the flag.

``` {r remove-dups-biomass}

unique(apl_biomass$duplicateRecordQF)

```

All data have flag value = 0, indicating they are not duplicated. 
Let's check the `apl_plantExternalLabDataPerSample` table.

``` {r remove-dups-analytes, results="hide"}

apl_plantExternalLabDataPerSample <- removeDups(
  data=apl_plantExternalLabDataPerSample, 
  variables=variables_20063)

```

```
10 duplicated key values found, representing 20 non-unique records. Attempting to resolve.
  |==========================================================================================| 100%
2 resolveable duplicates merged into matching records
2 resolved records flagged with duplicateRecordQF=1
16 unresolveable duplicates flagged with duplicateRecordQF=2
```

The function output tells you there were 20 duplicates (out of 26108 
total data records). Four of those duplicates were resolved, leaving two 
records flagged as resolved duplicates, with `duplicateRecordQF=1`. The 
remaining 16 couldn't be resolved, and were flagged with 
`duplicateRecordQF=2`.

What does it mean for duplicates to be resolved? Some duplicates represent 
identical data that have been entered multiple times, whereas some duplicates 
have the same key values, but differ in data values. `removeDups()` has a 
fairly narrow set of criteria for resolving:

1. If one data record has empty fields that are populated in the other record, and the records otherwise match, the non-empty values are retained.
2. If records are identical except for uid, remarks, and/or personnel (`identifiedBy`, `recordedBy`, etc) fields, unique values are concatenated within the non-matching fields, separated by `|` (pipes).

Records that can be merged to a single record by those criteria are flagged 
with `duplicateRecordQF=1`. Records with mis-matched data that can't be merged 
are retained and flagged with `duplicateRecordQF=2`.

Note that even in fully identical duplicates, the uid field will always be 
unique. Thus the uid field in merged records will always contain the 
pipe-delimted set of uids of the original records.

What does this look like in practice? Let's look at the two resolved 
duplicates:

``` {r view-merged}

apl_plantExternalLabDataPerSample[which(
  apl_plantExternalLabDataPerSample$duplicateRecordQF==1),]

```

You can see that both records have two pipe-delimited uids, and are 
flagged.

Let's look at the unresolveable duplicates:

``` {r view-merged}

apl_plantExternalLabDataPerSample[which(
  apl_plantExternalLabDataPerSample$duplicateRecordQF==2),]

```

The key for this data table is the sample identifier and analyte, and 
here there are multiple records with the same sample identifier, both for 
carbon and nitrogen values. The most likely scenario is that these are 
unlabeled replicate analyses.

Now, how should you proceed with your analysis? Of course, that is up to 
you, and depends on your analysis. Because these appear to be unlabeled 
analytical replicates, I would probably average the analyte values, but 
a decision like this can't be made automatically - `removeDups()` can 
identify the records of concern, but what to do with them is a 
judgement call.

Of course, NEON scientists also evaluate NEON data to identify 
duplicates as part of quality assurance and quality control procedures, 
and to resolve them if possible. In the data download step above, we 
accessed RELEASE-2022. The data release is stable and reproducible over 
time, but these duplicates may be resolved in future releases of these 
data.

## Join Data Tables

If you are using `neonOS` to check for duplicates and also to join data 
tables, the duplicate step should always come first. Because duplicate 
identification uses the variables file to determine uniqueness of data, 
the duplicate check step requires that the data match the variables 
file exactly, which they won't after being joined.


