---
syncID: FILL IN
title: "Using the neonOS Package to Find Samples"
description: "Use the functions in the neonOS package to explore the relationships between samples in the NEON database."
dateCreated: 2022-03-29
authors: [Claire K. Lunch]
contributors: 
estimatedTime: 40 minutes
packagesLibraries: neonOS, neonUtilities
topics: data-management, rep-sci
languageTool: R
dataProduct: 
code1: FILL IN
tutorialSeries:
urlTitle: neonOS-samples-tutorial

---


## Background About Samples

The definition of a sample in the NEON data system is very broad. For 
example, at sites with woody vegetation, each tagged tree is a "sample", 
even though it never leaves the field. Leaves and other samples taken from 
a given tree are considered subsamples of the tree, and may have their 
own subsamples.

The sample system at NEON is hierarchical, meaning that any given sample 
can have both parent and child samples. These relationships are not one to 
one, each sample may have multiple parents and/or multiple children. Samples 
with multiple parents are the rarest type, and are referred to as mixtures.

Samples may be measured repeatedly over time, as in the case of trees, or they 
may be consumed or destroyed in the process of analysis, such as some water 
chemistry samples. Often, they are subsampled several times, with different 
subsamples analyzed for different purposes. A significant portion of NEON 
samples are completely or partially preserved at the NEON Biorepository, where 
they are available to the community by request for further analyses.

In this tutorial, we will explore the sample relationships of a tagged tree 
and a mosquito sampling event, to help us understand the NEON sample system, 
and how to use the tools available for learning about the samples.

<div id="ds-objectives" markdown="1">

## Objectives
After completing this activity, you will be able to:

 * Find all samples related to a particular NEON sample, using the `neonOS` package 
 * Understand the hierarchical relationships of NEON samples
 * Find the archived samples associated with a particular sample collection

## Things You’ll Need To Complete This Tutorial
You will need a version of R (4.0 or higher. This code may work with 
earlier versions but it hasn't been tested) and, preferably, `RStudio` 
loaded on your computer to complete this tutorial.

### Install R Packages

* **neonUtilities:** `install.packages("neonUtilities")`
* **neonOS**: `devtools::install_github("NEONScience/NEON-OS-data-processing/neonOS")`

## Additional Resources

* NEON <a href="http://data.neonscience.org" target="_blank"> Data Portal </a>
* <a href="https://www.neonscience.org/resources/code-hub" target="_blank">NEON Code Hub</a>
* NEON <a href="https://data.neonscience.org/sample-explorer" target="_blank">Sample Explorer</a>
* NEON <a href="https://biorepo.neonscience.org/portal/" target="_blank">Biorepository Portal</a>

</div>

## Set Up R Environment and Download Data

First install and load the necessary packages.

```{r setup, eval=c(8,9), results="hide"}

# install packages. can skip this step if already installed
# update neonOS installation after CRAN release
install.packages("neonUtilities")
devtools::install_github("NEONScience/NEON-OS-data-processing/neonOS")

# load packages
library(neonUtilities)
library(neonOS)

```

We'll download Vegetation structure (DP1.10098.001) data from 
Rocky Mountain National Park (RMNP), and Mosquitoes sampled from 
CO2 traps (DP1.10043.001) from Toolik Lake (TOOL).

If you're not familiar with the `neonUtilities` package and how to use 
it to access NEON data, we recommend you follow the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> 
tutorial before proceeding with this one.

``` {r load-data, results="hide"}

veg <- loadByProduct(dpID="DP1.10098.001", 
                     site="RMNP", 
                     release="RELEASE-2022",
                     check.size=F)

mos <- loadByProduct(dpID="DP1.10043.001",
                     site="TOOL",
                     release="RELEASE-2022",
                     check.size=F)

```

## Find Related Samples

Of course both of these datasets contain many samples. Let's check out 
the tagged plant with `individualID = NEON.PLA.D10.RMNP.03623`.

``` {r view-vst}

veg$vst_mappingandtagging[which(veg$vst_mappingandtagging$individualID==
                                  "NEON.PLA.D10.RMNP.03623"),]

```

We see from the data that this is a quaking aspen (Populus tremuloides). 
Now use the `getSampleTree()` function from the `neonOS` package to find 
all samples related to this particular sample.

``` {r query-taggedPlant}

potr.samp <- getSampleTree("NEON.PLA.D10.RMNP.03623")

```

SOME STUFF ABOUT HOW TO UNDERSTAND WHAT YOU'RE LOOKING AT?

``` {r plant-network}

potr.samp$id <- potr.samp$sampleClass

edges <- potr.samp[,c("sampleClass","parentSampleClass")]
colnames(edges) <- c("to", "from")

visNetwork(nodes=potr.samp,
           edges=edges) %>%
  visEdges(arrows="to")

```

Poking around with mosquito data here, figuring out what makes sense for tutorial

``` {r mosquito-explore}

mos.samp <- getSampleTree("TOOL_087.20190607.1236")

# this runs forever
mos.samp <- getSampleTree("TOOL_087.20190607.1236.S.01")

mos.samp$id <- mos.samp$sampleUuid
nodes <- mos.samp[,"id"]
nodes <- unique(nodes)

edges <- mos.samp[,c("sampleUuid","parentSampleUuid")]
colnames(edges) <- c("to", "from")

visNetwork(nodes=nodes,
           edges=edges) %>%
  visEdges(arrows="to") #%>%
  visNodes(label=mos.samp$sampleTag,
           shape="dot",
           size=10)


```

