---
syncID: FILL IN
title: "Using the neonOS Package to Find Samples"
description: "Use the functions in the neonOS package to explore the relationships between samples in the NEON database."
dateCreated: 2022-03-29
authors: [Claire K. Lunch]
contributors: 
estimatedTime: 40 minutes
packagesLibraries: neonOS, neonUtilities
topics: data-management, rep-sci
languageTool: R
dataProduct: 
code1: FILL IN
tutorialSeries:
urlTitle: neonOS-samples-tutorial

---


## Background About Samples

The definition of a sample in the NEON data system is very broad. For 
example, at sites with woody vegetation, each tagged tree is a "sample", 
even though it never leaves the field. Leaves and other samples taken from 
a given tree are considered subsamples of the tree, and may have their 
own subsamples.

The sample system at NEON is hierarchical, meaning that any given sample 
can have both parent and child samples. These relationships are not one to 
one, each sample may have multiple parents and/or multiple children. Samples 
with multiple parents are the rarest type, and are referred to as mixtures.

Samples may be measured repeatedly over time, as in the case of trees, or they 
may be consumed or destroyed in the process of analysis, such as some water 
chemistry samples. Often, they are subsampled several times, with different 
subsamples analyzed for different purposes. A significant portion of NEON 
samples are completely or partially preserved at the NEON Biorepository, where 
they are available to the community by request for further analyses.

In this tutorial, we will explore the sample relationships of a tagged tree 
and a mosquito sampling event, to help us understand the NEON sample system, 
and how to use the tools available for learning about the samples.

<div id="ds-objectives" markdown="1">

## Objectives
After completing this activity, you will be able to:

 * Find all samples related to a particular NEON sample, using the `neonOS` package 
 * Understand the hierarchical relationships of NEON samples
 * Find the archived samples associated with a particular sample collection

## Things You’ll Need To Complete This Tutorial
You will need a version of R (4.0 or higher. This code may work with 
earlier versions but it hasn't been tested) and, preferably, `RStudio` 
loaded on your computer to complete this tutorial.

### Install R Packages

* **neonUtilities:** `install.packages("neonUtilities")`
* **neonOS**: `install.packages("neonOS")`

## Additional Resources

* NEON <a href="http://data.neonscience.org" target="_blank"> Data Portal </a>
* <a href="https://www.neonscience.org/resources/code-hub" target="_blank">NEON Code Hub</a>
* NEON <a href="https://data.neonscience.org/sample-explorer" target="_blank">Sample Explorer</a>
* NEON <a href="https://biorepo.neonscience.org/portal/" target="_blank">Biorepository Portal</a>

</div>

## Set Up R Environment and Download Data

First install and load the necessary packages.

```{r setup, eval=FALSE}

# install packages. can skip this step if 
# the packages are already installed
install.packages("neonUtilities")
install.packages("visNetwork")
install.packages("neonOS")

# load packages
library(visNetwork)
library(neonUtilities)
library(neonOS)

```

```{r libraries, include=FALSE}
library(neonUtilities)
library(neonOS)
library(visNetwork)
```

We'll download Vegetation structure (DP1.10098.001) data from 
Rocky Mountain National Park (RMNP), and Rodent pathogen status, 
tick-borne (DP1.10064.002) from all available sites.

If you're not familiar with the `neonUtilities` package and how to use 
it to access NEON data, we recommend you follow the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore NEON Data</a> 
tutorial before proceeding with this one.

``` {r load-data, results="hide"}

veg <- loadByProduct(dpID="DP1.10098.001", 
                     site="RMNP", 
                     release="RELEASE-2022",
                     check.size=F)

rpt <- loadByProduct(dpID="DP1.10064.002",
                     release="RELEASE-2022",
                     check.size=F)

```

## Find Related Samples

Of course both of these datasets contain many samples. Let's check out 
the tagged plant with `individualID = NEON.PLA.D10.RMNP.03623`.

``` {r view-vst}

veg$vst_mappingandtagging[which(veg$vst_mappingandtagging$individualID==
                                  "NEON.PLA.D10.RMNP.03623"),]

```

We see from the data that this is a quaking aspen (Populus tremuloides). 
Now use the `getSampleTree()` function from the `neonOS` package to find 
all samples related to this particular sample.

``` {r query-taggedPlant}

potr.samp <- getSampleTree("NEON.PLA.D10.RMNP.03623")

```

Let's take a look at the table this function returns:

``` {r view-potr}

potr.samp

```

There are 10 columns. The first 5 are:

* sampleUuid: Unique identifier for the sample in the NEON database
* sampleTag: Identifier for the sample. Frequently human-readable, may be non-unique across sampleClasses.
* barcode: Identifier in scannable barcode on the sample. Unique within NEON.
* archiveGuid: Globally unique identifier for an archived sample.
* sampleClass: Category or type of sample. Used within NEON to classify samples at different levels in the hierarchy, and define expectations for sample handling.

For each row, we describe the sample identified in these 5 columns as the 
focal sample. The next 5 columns contain the same identifiers and 
descriptors, but for the parent sample of the focal sample on each row. Note 
that each parent sample identified in the table will also be a focal sample 
in at least one row.

Because samples can have multiple parents, and parents can have multiple 
children, there may be rows with the sample focal sample but different 
parents, and rows with different focal samples but the same parents. In 
the case of the sample hierarchy for `NEON.PLA.D10.RMNP.03623`, most of 
the samples are children of the sample with barcode A00000125509. This is 
a foliar sample, which is taken from the tree and then split into 
subsamples for various analyses.

It's usually much easier to understand these relationships visually than in 
a table. The parent-child relationships create a 
<a href="https://en.wikipedia.org/wiki/Directed_acyclic_graph" target="_blank">directed acyclic graph</a>, and there are many tools available for viewing such graphs. Here, 
we'll use the `visNetwork` package.

To be understood by the `visNetwork` package, we need to define the "nodes" 
and "edges" of the graph - that is, the samples and their relationships.

``` {r plant-network}

potr.samp$id <- potr.samp$sampleUuid

edges <- potr.samp[,c("sampleUuid","parentSampleUuid")]
colnames(edges) <- c("to", "from")

visNetwork(nodes=potr.samp,
           edges=edges) %>%
  visEdges(arrows="to")

```

This graph displays the relationships, but we don't know which sample 
is which. It helps to add the sampleClass as a label:

``` {r plant-network-with-labels}

potr.samp$label <- potr.samp$sampleClass

visNetwork(nodes=potr.samp,
           edges=edges) %>%
  visNodes(shape="ellipse", font=list(size=10)) %>%
  visEdges(arrows="to")

```

Now we have a clearer picture. There is one sample with no parents, in 
the class `taggedPlant`. Referring back to the table, we can see this is 
the tree itself, that we initially queried, `NEON.PLA.D10.RMNP.03623`. It 
has two children, one of which is the foliar sample that has many 
subsamples (children) of its own. If you're familiar with the Foliar traits 
data product (DP1.10026.001), you'll be able to identify these subsamples 
easily, as the samples for analysis of lignin, carbon and nitrogen, etc.

There is also one sample with a class label indicating archive, and we can 
see in the table that it has an `archiveGuid` identifier. You can look up 
this sample at the NEON Biorepository, where it is stored: 
<a href="https://biorepo.neonscience.org/portal/collections/individual/index.php?occid=737499&clid=0" target="_blank">barcode A00000134072</a>

TO DO: Example from rodent pathogens? How to conclude? How should people use this info?

