## Samples

Now, we may be interested in what other data have been collected on 
some of the trees in the RMNP woody vegetation dataset. The 
**/samples** endpoint lets you explore the sample relationships and 
sample histories of samples in the NEON database.

The way NEON samples are defined, each tagged tree is a "sample", even 
though it never leaves the field. Leaves and other samples taken from 
a given tree are considered subsamples of the tree, and may have their 
own subsamples. These hierarchical relationships are accessible via the 
**/samples** endpoint.

Like the **/taxonomy** endpoint, the **/samples** endpoint has 
multiple query options.

Going back to the `vst.maptag` data table we downloaded, the data field 
containing the sample identifiers for the individual trees is 
`individualID`. Let's look up a specific tree, an aspen, 
NEON.PLA.D10.RMNP.03861.

In order to query a sample, both a sample tag (= ID) and a sample class are 
needed. So we will first need to look up the sample tag and see which 
classes it exists in (this step is necessary because NEON sample IDs 
may be repeated in different classes, i.e. different types of samples).

```{r get-sample-class}

samp.class.req <- GET("https://data.neonscience.org/api/v0/samples/classes?sampleTag=NEON.PLA.D10.RMNP.03861")
samp.class.list <- jsonlite::fromJSON(content(samp.class.req, as="text"))
samp.class.list$data

```

There is only one possible sample class for the sample tag 
NEON.PLA.D10.RMNP.03861, `taggedPlant`. So we can proceed to query for 
sample information using the tag and class together.

```{r get-sample-view}

samp.view.req <- GET("https://data.neonscience.org/api/v0/samples/view?sampleTag=NEON.PLA.D10.RMNP.03861&sampleClass=taggedPlant")
samp.view.list <- jsonlite::fromJSON(content(samp.view.req, as="text"))
samp.view.list$data

```
