---
layout: post
title: "Plotting stream discharge data for a NEON field site in R"
date:   2015-07-15
lastModified: 2015-09-30
createddate:   2015-07-15
estimatedTime: 1.0 Hour
packagesLibraries: 
authors: Leah A. Wasser
categories: [coding-and-informatics]
category: coding-and-informatics
tags: [R,time-series]
mainTag: time-series
description: "Open up and explore hyperspectral imagery in HDF format R. Combine multiple bands to create a raster stack. Use these steps to create various band combinations such as RGB, Color-Infrared and False color images."
code1: 
image:
  feature: hierarchy_folder_purple.png
  credit: Colin Williams NEON, Inc.
  creditlink: http://www.neonscience.org
permalink: /R/StreamDischarge-TimeSeries-R/
code1: R/2015-09-15-StreamDischarge-TimeSeries-R.Rmd
comments: true
---

<section id="table-of-contents" class="toc">
  <header>
    <h3>Contents</h3>
  </header>
<div id="drawer" markdown="1">
*  Auto generated table of contents
{:toc}
</div>
</section><!-- /#table-of-contents -->

<div id="objectives">
<strong>R Skill Level:</strong> Beginner / Intermediate

<h3>Goals / Objectives</h3>
After completing this activity, you will:
<ol>
<li>Import a time series formated csv file.</li>
<li>Modify a date field that contains non-date characters and turn it into a proper
date formated field.</li>
<li>Plot the data.</li>
<li>Convert time to a local time zone.</li>
</ol>

<h3>R Libraries to Install</h3>
<ul>

</ul>

<h3>Data to Download</h3>
<a href="{{ site_baseurl }}/data/streamDischarge.zip" class="btn btn-success"> 
DOWNLOAD NEON Stream Discharge Data for Posey Creek</a>. 
<p>The data in this csv file were collected at 
<a href="http://www.neonscience.org/science-design/field-sites/posey-creek" target="_blank"> 
Posey Creek field site located in Virginia (NEON Domain 02)</a> and processed at NEON headquarters. The 
entire dataset and data for other field sites can be accessed from the <a href="http://neonscience.orgdata.neonscience.org" target="_blank">NEON data portal.</a>
</p>  

<h3>Pre-reqs</h3>
<p>.</p>
</div>


```{r load-libraries}

library("ggplot2")
```

##Import Our Data
First, let's import the csv that contains the time series data.

```{r import-data}

# import csv making sure factors are set to false

options(stringsAsFactors = FALSE)
#import stream discharge data
disArik <- read.csv("streamDischarge/NEON.D02.POSE.DP1.20048.001.dsc_fieldData.csv")
```

```{r fix-dates }
#fix the dates
# 2014-08-19T00:43:22Z
#remove the funny T
disArik$collectDate <- gsub("T", " ", disArik$collectDate)

#remove the z
disArik$collectDate <- gsub("Z", " ",  disArik$collectDate)

#set the field to be a date field
disArik$collectDate <- as.POSIXct(disArik$collectDate,format = "%Y-%m-%d %H:%M:%S", 
                        tz = "GMT")

#list of time zones
#https://en.wikipedia.org/wiki/List_of_tz_database_time_zones
#convert to local time for pretty plotting
attributes(disArik$collectDate)$tzone <- "America/Denver"
```

```{r plot-data}
qplot(x=disArik$collectDate,y=disArik$streamStage,
      xlab = "Collection Date", ylab="Stream Stage",
      main="Stream Stage for Arikaree - NEON Site")
```
