---
layout: post
title: "GREPL, Filter and Piping Using R DPLYR - An Intro"
date:   2015-1-28 20:49:52
dateCreated:   2015-05-27 14:49:52
lastModified:   2015-05-28 20:49:52
estimatedTime: 1.0 - 1.5 Hours
packagesLibraries: dplyr
authors: Natalie Robinson, Kate Thibault
contributors: 
categories: [coding-and-informatics]
category: coding-and-informatics
tags: [R]
mainTag: R
description: "Learn how to use the GREPL, Summarise and piping functions in R dplyr."
code1: 2015-05-27-R-dplyr-GREPL-Summarise-Piping.R
image:
  feature:  
  credit:  
  creditlink:  
permalink: /R/GREPL-Filter-Piping-in-DPLYR-Using-R/
comments: true
---

#Intro to dplyr
When working with data frames in R, it is often useful to manipulate and 
summarize data. The `dplyr` library in `R` offers one of the most comprehensive 
library of functions to perform common manipulative tasks. In addition, the dplyr 
functions are often of a simpler syntax than most other data manipulation 
functions in R.


#Elements of dplyr
There are several elements of `dplyr` that are unique to the library, and that
do very cool things!

##Functions for manipulating data
_excerpted from <http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html>_
Dplyr aims to provide a function for each basic verb of data manipulating, like:

 * filter() (and slice())
      * filter rows based on values in specified columns
 * arrange()
      * sort data by values in specified columns 
 * select() (and rename())
      * view and work with data from only specified columns
 * distinct()
      * view and work with only unique values from specified columns
 * mutate() (and transmute())
      * add new data to the data frame
 * summarise()
      * calculate specified summary statistics on data
 * sample_n() and sample_frac()
      * return a random sample of rows
 
##Format of function calls
The single table verb functions share these features:

  * The first argument is a data frame (or a dplyr special class tbl_df).
      * dplyr can work with data frames as is, but if you'??re dealing with large
        data it'??s worthwhile to convert them to a tbl_df, to avoid printing 
        a lot of data to the screen.
  * The subsequent arguments describe how to manipulate the data (e.g., based on
    which columns, using which summary statistics), and you can refer to columns
    directly (without using $).
  * The result is a new data frame
  * Function calls do not generate 'side-effects'; you always have to assign the
    results to an object
  
##Grouped operations
Certain functions (e.g., group_by, summarise, and other 'aggregate functions')
allow you to get information for groups of data, in one fell swoop. This is like 
performing database functions with knowing SQL or any other db specific code. 
Powerful stuff!

##Piping
We often need to get a subset of data using one function, and then use 
another function to do something with that subset (and we may do this multiple 
times). This leads to nesting functions, which can get messy and hard to keep 
track of. Enter 'piping', dplyr's way of feeding the output of one function into 
another, and so on, without the hassleof parentheses and brackets. Piping looks like:

  * data frame %>%
         function to apply first %>%
               function to apply second %>%
                     function to apply third


For example, if we want to find the mean body weight of male mice, we'd do this
(read as 'for data frame myMammalData, select only males and return the mean
weight):

  * myMammalData %>%
         filter(sex=='m') %>%
               summarise (mean_weight = mean(weight))
                      

#Goals/Objectives of this lesson
After completing this activity, you be able to use the following dplry functions:
 1. filter, alone and combined with simple pattern matching (grepl)
 2. group_by
 3. Summarise
 
You will also know how to pipe functions.

```{r load-dplyr-library-import-data}

#install dplyr library
#install.packages('dplyr')

#load library
library('dplyr')

#setwd('insert path to data files here')
#download example data - NEON small mammal capture data from D01 Harvard Forest - a site located in the heart of the Lyme disease epidemic.
myData <- read.csv('NEON.D01.HARV.DP1.10072.001.mam_capturedata.csv', header = T, 
                       stringsAsFactors = FALSE, strip.white = TRUE, 
                       na.strings = '')

```

#filter
 * function to extract only a subset of rows from a data frame according to specified conditions
 * similar to the base function subset, but with simpler syntax
  * inputs: data object, any number of conditional statements on the named columns of the data object
  * outputs: a data object of the same class as the input object (e.g., data.frame in, data.frame out) with only those rows that meet the conditions

```{r dplyr filter function}

#for example, let's create a new data frame that contains only female Peromyscus mainculatus, 
# one of the key small mammal players in the life cycle of Lyme disease-causing bacterium.

data_PeroManicFemales <- filter(myData, scientificName == 'Peromyscus maniculatus', sex == 'F')

#Note that we were able to put multiple conditions into the filter statement, pretty cool!

#how many female P. maniculatus are in the dataset
print(paste('In 2014, NEON technicians captured', nrow(data_PeroManicFemales), 'female Peromyscus maniculatus at Harvard Forest.', sep = ' '))

```

#grepl
This is a function in the base package that is part of the suite of Regular Expressions functions. grepl uses regular expressions to match patterns in character strings. Regular expressions offer very powerful and useful tricks for data manipulation. They can be complicated and therefore are a challenge to learn, but well worth it! Here, we present a very simple example.
* inputs: pattern to match, character vector to search for a match
* outputs: a logical vector indicating whether the pattern was found within each element of the input character vector

```{r using grepl with the dplyr filter function}

#In reality, all species of Peromyscus are viable players in Lyme disease transmission, so we really should be looking at all species of Peromyscus. Since we don't have genera split out as a separate field, we have to search within the scientific name string for the genus -- this is a simple example of pattern matching - 

#we will use the dplyr function filter in combination with the base function grepl to accomplish this

data_PeroFemales <- filter(myData, grepl('Peromyscus', scientificName), sex == 'F')

#how many female Peromyscus are in the dataset
print(paste('In 2014, NEON technicians captured', nrow(data_PeroFemales), 'female Peromyscus spp. at Harvard Forest.', sep = ' '))

```

#group_by + summarise
An alternative to using the filter function to subset the data (and make a new data object in the process),
is to calculate summary statistics based on some grouping factor. We'll use group_by, which does basically the same thing as SQL or other tools for interacting with relational databases. For those unfamiliar with SQL, no worries - dplyr provides lots of additional functionality for working with databases (local and remote) that does not require knowledge of SQL. Handy! 

The group_by function in dplyr allows you to perform functions on a subset of a dataset without having to create multiple new objects or construct for loops. The combination of group_by and summarise are great for generating simple summaries (counts, sums) of grouped data. NOTE - be careful about using summarise for other summary functions - e.g.,  you need to think about weighting for means and variances, and summary doesn't work precisely for medians if there is any missing (e.g., there was no value recorded, maybe for good reasons!) data.

```{r dplyr group_by & summarise}

#Since the diversity of the entire small mammal community has been shown to impact disease dynamics among the key reservoir species, we really want to know more about the demographics of the whole community.

#We can quickly generate counts by species and sex in 2 lines of code, using group_by and summarise.

dataBySpSex <- group_by(myData, scientificName, sex)
#Note - the output does not look any different than the original data.frame, but the application of subsequent functions (e.g., summarise) to this new data frame will produce distinctly different results than if you tried the same operations on the original

countsBySpSex <- summarise(dataBySpSex, n_individuals = n())

```

#use piping to put functions together
We created multiple new data objects during our explorations of dplyr functions, above. Now let's put it all together, by chaining functions together in order to make one new data object that encapsulates all of the previously sought information: filter on only females, grepl to get only Peromyscus spp., group_by individual species, and summarise the numbers of individuals.

```{r dplyr piping to combine functions}

#Let's combine several dplyr functions in order to get a summary of the numbers of individuals of female Peromyscus species in our dataset.

dataBySpFem <- myData %>%
                 filter(grepl('Peromyscus', scientificName), sex == 'F') %>%
                   group_by(scientificName) %>%
                      summarise(n_individuals = n())



#For reference, let's see what we would have needed to do for the same output but using R's base functions

#First, subset the data to only female Peromyscus
dataFemPero  <- myData[myData$sex=='F' & grepl('Peromyscus',myData$scientificName),]

#Option 1) Use aggregate and then rename columns--------------------------------
dataBySpFem_agg <-aggregate(dataFemPero$sex ~ dataFemPero$scientificName, data = dataFemPero, FUN = length)
names(dataBySpFem_agg) <- c('scientificName','n_individuals')

#Option 2) Do it by hand--------------------------------------------------------
#Get the unique scientificNames in the subset
sppInDF <- unique(dataFemPero$scientificName[!is.na(dataFemPero$scientificName)])

#Calculate the numbers of individuals of each species
sciName <- vector(); numInd <- vector()
for (i in sppInDF){
  sciName <- c(sciName,i)
  numInd <- c(numInd, length(which(dataFemPero$scientificName==i)))
}

#Create the desired output data frame
dataBySpFem_byHand <- data.frame('scientificName'=sciName, 'n_individuals'=numInd)

```