---
title: "Intro-Create-Read-HDF5"
output: html_document
---

First, let's get R setup. We will use the RHDF5 library.

```{r}

#Call the R HDF5 Library
#source("http://bioconductor.org/biocLite.R")
#biocLite("rhdf5")
library("rhdf5")
```

We'll start by creating a new HDF5 file. Let's call it vegData.

```{r}

# Create hdf5 file
h5createFile("vegData.h5")

#create a group called testData within the H5 file
h5createGroup("vegData.h5", "aNEONSite")

#view the structure of the h5 we've created
h5ls("vegData.h5")

# create some sample, numeric data 
a <- rnorm(n=40, m=1, sd=1) 
someData <- matrix(a,nrow=20,ncol=2)

# add some sample data to the H5 file located in the location1 group
# we'll call the dataset "precip_data" and store it 
# in the location1 group within the H5 file
h5write(someData, file = "vegData.h5", name="aNEONSite/temperature")

# let's check out the H5 structure again
h5ls("vegData.h5")

# we can look at everything too 
# but be cautious using this command!
h5dump("vegData.h5")


```


## Next let's add some attributes to our data

```{r}

#open the file, create a class
fid <- H5Fopen("vegData.h5")
#open up the dataset to add attributes to, as a class
did <- H5Dopen(fid, "aNEONSite/temperature")

# Provide the NAME and the ATTR (what the attribute says) 
# for the attribute.
h5writeAttribute(did, attr="Here is a description of the data",
                 name="Description")
h5writeAttribute(did, attr="Meters",
                 name="Units")

#let's add some attributes to the group
did2 <- H5Gopen(fid, "aNEONSite/")
h5writeAttribute(did2, attr="San Joachin Experimental Range",
                 name="SiteName")
h5writeAttribute(did2, attr="Southern California",
                 name="Location")

#close the files, groups and the dataset when you're done writing to them!
H5Dclose(did)
H5Gclose(did2)
H5Fclose(fid)

#look at the attributes of the precip_data dataset
h5readAttributes(file = "vegData.h5", 
                 name = "aNEONSite/temperature")

#look at the attributes of the aNEONsite group
h5readAttributes(file = "vegData.h5", 
                 name = "aNEONSite")

# let's grab some data from the H5 file
testSubset <- h5read(file = "vegData.h5", 
                 name = "aNEONSite/temperature")

testSubset2 <- h5read(file = "vegData.h5", 
                 name = "aNEONSite/temperature",
                 index=list(NULL,1))
H5close()


#create a quick plot of the data
hist(testSubset2[])
```

## Challenge -- 
Open up the D17_2013_vegStr.csv in R. 
Create a new H5 file called vegStructure.

Add a group in your h5 file called SJER. Add the veg structure data to that folder!
Add some attributes the data. Do the same with the SOAP csv.
NOTE: R is not good at subsetting compound datasets. 

```{r}
options(stringsAsFactors = FALSE)
newData <- read.csv("D17_2013_vegStr.csv")

```


