---
layout: post
title: "Use the neonDataStackR package to access NEON data"
date:   2017-08-01
authors: [Megan A. Jones ]
contributors: [ ] 
dateCreated: 2017-08-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [neonDataStackR]
category: [self-paced-tutorial]
tags: [ ]
mainTag: R
tutorialSeries: [ ]
description: "Convert data downloaded from the NEON Data Portal in zipped month-by-site files into a table with all data of interest. Temperature data is used as an example. "
code1: neonDataStackR-temp.R
image:
  feature: laptopCode.png
  credit: National Ecological Observatory Network (NEON)
  creditlink: 
permalink: /R/neonDataStackR-temp
comments: true
---

This tutorial goes over how to convert data downloaded from the NEON Data Portal 
in zipped month-by-site files into individual files with all data from the 
given site(s) and months. Temperature data is used as an example.

## Download the Data
To start, you must have your data of interest downloaded from the 
<a href="data.neonscience.org" target="_blank"> NEON Data Portal</a>.  

The stacking function will only work on Comma Seperated Value (.csv) files and 
not the NEON data stored in other formats (HDF5, etc). 

Your data will download in a single zipped file. 

The example data is any single-aspirated air temperature available from 
1 January 2015 to 31 December 2016. 

## neonDataStackR package

This package was written to stack data downloaded in month-by-site files into a 
full table with all the data of interest from all sites in the downloaded date
range.  

More information on the package see the README in the associated GitHub repo 
<a href="https://github.com/NEONScience/NEON-utilities/tree/master/neonDataStackR" target="_blank"> NEONScience/NEON-utilities</a>. 

First, install the package from the GitHub repo. You must have the **devtools** 
package installed to do this. Then load the package. 

```{r loadStuff}

library(devtools)
install_github("NEONScience/NEON-utilities/neonDataStackR", dependencies=TRUE)
library (neonDataStackR)

```

Now there is a single function to run in this package `stackByTable()`. The output
will yield data grouped into new files by table name.  For example the single 
aspirated air temperature data product contains 1 minute and 30 minute interval
data. The output from this function is one .csv with 1 minute data and one .csv 
with 30 minute data. 

Depending on your file size this function may run for a while. The 2015 and 2016
single aspirated air temperature from two sites that I used for the example took
about 25 minutes to complete.  


```{r run-function, eval = FALSE}
stackByTable("data/NEON_temp-air-single.zip")

```

```{r sample-output, eval=FALSE}

Unpacked  2016-02-SERC-DP1.00002.001-basic-20160708035158.zip
Unpacked  2016-03-SERC-DP1.00002.001-basic-20160708035642.zip
Joining, by = c("domainID", "siteID", "horizontalPosition", "verticalPosition", "startDateTime", "endDateTime", "tempSingleMean", "tempSingleMinimum", "tempSingleMaximum", "tempSingleVariance", "tempSingleNumPts", "tempSingleExpUncert", "tempSingleStdErMean", "finalQF")
Joining, by = c("domainID", "siteID", "horizontalPosition", "verticalPosition", "startDateTime", "endDateTime", "tempSingleMean", "tempSingleMinimum", "tempSingleMaximum", "tempSingleVariance", "tempSingleNumPts", "tempSingleExpUncert", "tempSingleStdErMean", "finalQF")
Stacked  SAAT_1min
Joining, by = c("domainID", "siteID", "horizontalPosition", "verticalPosition", "startDateTime", "endDateTime", "tempSingleMean", "tempSingleMinimum", "tempSingleMaximum", "tempSingleVariance", "tempSingleNumPts", "tempSingleExpUncert", "tempSingleStdErMean", "finalQF")
Joining, by = c("domainID", "siteID", "horizontalPosition", "verticalPosition", "startDateTime", "endDateTime", "tempSingleMean", "tempSingleMinimum", "tempSingleMaximum", "tempSingleVariance", "tempSingleNumPts", "tempSingleExpUncert", "tempSingleStdErMean", "finalQF")
Stacked  SAAT_30min
Finished: All of the data are stacked into  2  tables!

```

From the single-aspirated air temperature data we are given two final tables. 
One with 1 minute intervals: **SAAT_1min** and one for 30 minute intervals: **SAAT_30min**.  

These .csv files are now ready for use.  

