---
layout: post
title: "Plotting stream discharge data for a NEON field site in R"
date:   2015-07-15
createddate:   2015-07-15
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
estimatedTime: 1.0 Hour
packagesLibraries: 
authors: Leah A. Wasser
contributors: [ ]
categories: [coding-and-informatics]
mainTag: 
tags: [R,time-series]
tutorialSeries: [] 
description: " "
code1: R/StreamDischarge-TimeSeries-R.Rmd
image:
  feature: hierarchy_folder_purple.png
  credit: National Ecological Observatory Network
  creditlink: http://www.neonscience.org
permalink: /R/StreamDischarge-TimeSeries-R/
comments: true
---

{% include _toc.html %}


**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

# Objectives
After completing this tutorial, you will be able to:

* Import a .csv file with time series data.
* Modify a date field that contains non-date characters and turn it into a proper
date formated field.
* Plot the data.
* Convert time to a local time zone.

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and, preferably, RStudio loaded on
your computer to complete this tutorial.

## R Libraries to Install

[More on Packages in R - Adapted from Software Carpentry.]({{ site.baseurl }}/R/Packages-In-R/)

## Data to Download

<a href="{{ site_baseurl }}/data/streamDischarge.zip" class="btn btn-success"> 
DOWNLOAD NEON Stream Discharge Data for Posey Creek</a>. 

The data in this Teaching Data Subset (csv file) were collected at 
<a href="http://www.neonscience.org/science-design/field-sites/posey-creek" target="_blank"> Posey Creek field site located in Virginia (NEON Domain 02)</a>
and processed at NEON headquarters. The entire dataset and data for other field 
sites can be accessed from the 
<a href="http://data.neonscience.org" target="_blank">NEON data portal.</a>
  
</div>

We'll start by loading the `ggplot2` package which enables plotting functions
that we will use into our R enviornment.  

### Load Data

```{r load-libraries}
# load packages
library("ggplot2")

# set working directory to ensure R can find the file we wish to import
# setwd("working-dir-path-here")

```

Now we can import the .csv file that contains the stream discharge data.

```{r import-data}

# import stream discharge data
# we don't want strings to automatically be factors

dis_ARIK <- read.csv("streamDischarge/NEON.D02.POSE.DP1.20048.001.dsc_fieldData.csv",
										stringsAsFactors = FALSE)

# view the structure of the data
str(dis_ARIK)
```

## Work with Dates
One of the first s

2014-08-19T00:43:22Z

```{r fix-dates }
# fix the dates
# 2014-08-19T00:43:22Z
#remove the funny T
disArik$collectDate <- gsub("T", " ", disArik$collectDate)

#remove the z
disArik$collectDate <- gsub("Z", " ",  disArik$collectDate)

#set the field to be a date field
disArik$collectDate <- as.POSIXct(disArik$collectDate,format = "%Y-%m-%d %H:%M:%S", 
                        tz = "GMT")

#list of time zones
#https://en.wikipedia.org/wiki/List_of_tz_database_time_zones
#convert to local time for pretty plotting
attributes(disArik$collectDate)$tzone <- "America/Denver"
```

```{r plot-data}
qplot(x=disArik$collectDate,y=disArik$streamStage,
      xlab = "Collection Date", ylab="Stream Stage",
      main="Stream Stage for Arikaree - NEON Site")
```
