---
syncID: 
title: "Methods of Exploring NEON Mosquito Data in R Part Three: "
description: "This is part one of a series of tutorials that show users how to clean and perform an initial analysis of NEON mosquito data."
dateCreated: 2017-08-04
authors: Charlotte Roiger 
contributors: Megan A. Jones, Katie Levan, Cody Flagg, Maya R. Stahl
estimatedTime: 
packagesLibraries: dplyr, ggplot2
topics: data-analysis, data-visualization
languagesTool: R
dataProduct: DP1.10043.001
code1: 
tutorialSeries: neon-mosquito-data-intro
urlTitle: tba
---

This is part three of a series of tutorials will teach you how to use NEON 
mosquito data to explore mosquito species richness and the abundance of *Culex 
tarsalis*. For this tutorial we will focus in on how to calculate mosquito
species richness, how to add a variable that takes into account temperature,
and how to visualize some of the trends we might find in species richness.

**R Skill Level:** Intermediate - you've got the basics of `R` down

<div id="objectives" markdown="1">

# Goals & Objectives

After completing this tutorial series, you will be able to:

* Use the `ddply` function to calculate species richness
* Create and apply a temperature lag function to the species richness data frame
* Use `ggplot2` in `R` to create visualizations of data trends and maps  

## Things You Will Need To Complete This Tutorial

You will need the most current version of `R` and, preferably, RStudio loaded on
your computer to complete this tutorial. You will also need to download the
NEON mosquito data covered in part one of this tutorial series. Please click
here to access part one of this tutorial series.

# 6/20: Link needs to be updated.

### R Libraries to Install:

These R packages will be used in the tutorial below. Please make sure they are 
installed prior to starting the tutorial. 
 
* **dplyr:** `install.packages("dplyr")`
* **tidyverse:** `install.packages("tidyverse")`
*	**plyr:** `install.packages("plyr")`
*	**mosaic:** `install.packages("mosaic")`
* **ggplot2:** `install.pcakages("ggplot2")`


</div>

## Tutorial Set Up

If you are starting this tutorial immediately after the completion of part one
of the NEON Mosquito Data Visualization series, feel free to skip this portion 
of the tutorial. If not, our first step is to set up our working directory, 
load in the packages and libraries necessary, and download the mosquito data.

```{r load-libraries}

# Load packages required for entire script. 
library(dplyr)     # move/manipulate data
library(plyr)      # move/manipulate data
library(tidyverse) # move/manipulate data
library(mosaic)    # good for data exploration
library(ggplot2)   # creation of plots and visualizations

#Set strings as factors equal to false thoughout
options(stringsAsFactors = FALSE) 

# set working directory to ensure R can find the file we wish to import

#setwd("working-dir-path-here")

#Read in the data from part one of this tutorial series

idsmall = read.csv("NEON-mosquito-data-viz/idsmall.csv")

uniquetrap = read.csv("NEON-mosquito-data-viz/uniquetrap.csv")

```

## NEON Mosquito Data Visualization Part Two: Species Richness


###Calculating Species Richness and  Obtaining Location Information

Now that we have our data frames in a more usable format, we want to explore 
mosquito species richness to see if there is a pattern in relation to 
sample latitude. The first step we need to take to explore this topic is to
calculate species richness at each site. However if we calculate species 
richness for each site, we lose some complexity in our data since sites were 
sampled multiple times. So we want to calculate the species richness per site
but also take into account the date at which each sample was taken. To tackle 
this problem we use the `ddply` command from the `plyr` package to count the 
number of unique scientific names for each plot and date. We also choose to 
include certain variables in the idsmall data frame that could be relevant to
our exploration of species richness.


```{r create-specrich}

#Creating a new data frame called specrich from idsmall

specrich <- ddply(idsmall, ~ siteID + domainID + Date, summarize, num_species = length(unique(scientificName)))

```

Next we merge our newly formed species ricness data frame with location
information and convert latitude and longitude information into a numeric. To 
start exploring species richness we will make a scatter plot of species richness
over latitude using the package 'ggplot2'. Another factor that we might want to
take into account is the year in which the sample was taken to see if patterns
in species richness differ by year. We will incorporate this variable into our 
scatterplot of species richness over latitude by coloring the dots in our plot 
with their corresponding year.

```{r plot-richness}

#Omit all instances where date is unknown
specrich <- specrich[complete.cases(specrich$Date),]

#Merging to get lat2 lon2 data

#First extract site id for the uniquetrap data frame
uniquetrap$siteID <- substr(uniquetrap$plotID,1,4)

#Omit any instances where lat and lon is not known for a site
uniquetrap <- uniquetrap[uniquetrap$decimalLatitude!="",]

#taking only the observations that are not duplicated
uniquetrap <- uniquetrap[!duplicated(uniquetrap$siteID),]

#merge with lat lon data excluding plotID
specrich <- merge(x = uniquetrap[,c("plotID", "decimalLatitude", "decimalLongitude", "elevation.y", "nlcdClass.y", "siteID")], y = specrich, by.y = "siteID", by.x = "siteID")

#Changing latitude to a numeric and date to date class
specrich$decimalLatitude<-as.numeric(specrich$decimalLatitude)

#Creating a Year variable
specrich$Year <- substr(specrich$Date, 1,4)

# Plotting Species Richness over Latitude
ggplot(specrich,aes(decimalLatitude, num_species))+
  geom_point(aes(colour = Year), size = 2)+
  labs(x = "Latitude", y = "Number of Species")+
  ggtitle("Species Richness by Latitude")

```

What we can see from the plot of species richness by latitude is that there 
appears to be a large amount of clustering between 28 and about 47 degrees 
latitude. This result makes sense since most NEON sampling sites are located on 
the continental United States. Another thing we might notice is that there
does not appear to be a recognizable pattern between species richness and 
latitude for both 2014 and 2016. However, latitude is a proxy variable for other
environmental factors such as temperature, which plays a crucial role in 
mosquito development. So we will develop a metric that captures the
maximum temperature of the days prior to collection. 

###Creating a Temperature Lag Function

To examine the maximum temperature of days before the collection date, we will 
develop a function that takes the average maximum temperature of the 14 days
prior to the sample collection date. To start, we define a function that takes 
the date and site identification as inputs, and use the filter command to create
a data frame that contains temperature information for each observation. We then
take an average of the maximum temperature for the 14 days prior to collection
at each site and also count the number of days where the maximum temperature was
greater than 16 degrees Celsius.  

```{r templag-function}

Templag <- function(siteID, date){ 
  date <- as.Date(date) #Converting all date inputs into date format
  filter1 <- temp.df[grepl(siteID, temp.df$siteID),] #subset our temp data by site id
  filter2 <- filter1[filter1$date >= date - 14 & filter1$date < date + 1,] # subset by date
  TwoWeekAvgT <- mean(filter2$value)/10 #standardize temperature values
  filter3 <- filter2[filter2$value > 16,] #select rows with max temp greater than 16 degrees C
  GDD <- length(unique(filter3$date)) # count the number of unique dates in data frame
  return(list(TwoWeekAvgT=TwoWeekAvgT, GDD=GDD))
}

```

Now that our temperature lag function is complete, we will use the `mapply`
command to apply it to our species richness dataframe and create a matrix
output. We then create two new variables that capture the average maximum
temperature and the number of degreedays for the two weeks before collection. 
Once the two temperature variables are created, we then create a scatterplot of 
species richness over the average maximum temperature. 

```{r apply-templag}

#Applying temp lag fucntion to species richness data frame
SPTWAMT.mat<-mapply(Templag, specrich$siteID, specrich$Date)

#Creating variables that capture average max temp and degree days 
specrich$TwoWeekAvgMT<- c(unlist(SPTWAMT.mat[1,]))
specrich$DegreeDays <- c(unlist(SPTWAMT.mat[2,]))

#Scatterplot of species richness and two week average max temp
ggplot(specrich, aes(TwoWeekAvgMT, num_species))+
  geom_point(color = "green")+
  labs(x = "Two Week Average Maximum Temperature in Celsius", y = "Number of Species")+
  ggtitle("Species Richness by Two Week Average Maximum Temperature")

```

Looking at the scatterplot of species richness by two week average maximum 
temperature, we can see a sort of distribution where species richness is highest
where the average maximum temperature two weeks prior to collection was between 
25 to 32 degrees Celsius. This sort of pattern suggests that there is some sort
of relationship between species richness and the two week average maximum
temperature prior to collection. 

<div id = "challenge" markdown = "1">

###Challenge: Creating a Precipitation Lag Function

Now that we've created and applied a function that can calculate the average 
maximum temperature over a two week period before collection, try creating a
function that gauges the amount of precipitation at a sampling site before
collection and creating your own visualizations using the function outputs. 

</div>


