---
syncID: 
title: "Methods of Exploring NEON Mosquito Data in R Part Four: Culex tarsalis"
description: "This is part one of a series of tutorials that show users how to clean and perform an initial analysis of NEON mosquito data."
dateCreated: 2017-08-04
authors: Charlotte Roiger 
contributors: Megan A. Jones, Katie Levan, Cody Flagg, Maya R. Stahl
estimatedTime: 
packagesLibraries: dplyr, ggplot2
topics: data-analysis, data-visualization
languagesTool: R
dataProduct: DP1.10043.001
code1: 
tutorialSeries: neon-mosquito-data-intro
urlTitle: tba
---

This is part four of a series of tutorials that will teach you how to use NEON 
mosquito data to explore mosquito species richness and the abundance of *Culex 
tarsalis*. This tutorial will focus on examining trends in abundance and 
mapping the range of *Culex tarsalis*.

**R Skill Level:** Intermediate - you've got the basics of `R` down

<div id="objectives" markdown="1">

# Goals & Objectives

After completing this tutorial series, you will be able to:

*	Use the `gather` function in the `tidyr` package to transform data from a wide
  to long format
* Use ggplot2 in R to create visualizations of data trends and maps 
* Map NEON domains and sites from shapefiles

## Things You Will Need To Complete This Tutorial
You will need the most current version of R and, preferably, RStudio loaded on
your computer to complete this tutorial. You will also need to download the
NEON mosquito data covered in part one of this tutorial series. Please click
here to access part one of this tutorial series

### R Libraries to Install:

These R packages will be used in the tutorial below. Please make sure they are 
installed prior to starting the tutorial. 
 
* **dplyr:** `install.packages("dplyr")`
* **plyr:** `install.packages("plyr")`
* **tidyverse:** `install.packages("tidyverse")`
*	**mosaic:** `install.packages("mosaic")`
* **ggplot2:** `install.packages("ggplot2")`
* **gridExtra:** `install.packages("gridExtra")`
* **scales:** `install.packages("scales")`
* **foreign:** `install.packages("foreign")`
* **rgeos:** `install.packages("rgeos")`
* **maptools:** `install.packages("maptools")`
* **raster:** `install.packages("raster")`
* **rbokeh:** `install.packages("rbokeh")`
* **rgdal:** `install.packages("rgdal")`

</div>

## Tutorial Set Up

If you are starting this tutorial immediately after the completion of part one, part two, and part three of the NEON Mosquito Data Visualization series, feel free to skip 
this portion of the tutorial, make sure to load in the correct libraries. 
If not, our first step is to set up our working directory, load in the packages 
and libraries necessary, and download the mosquito data.

```{r load-libraries}

# Load packages required for script. 
library(dplyr)     # move/manipulate data
library(plyr)      # move/manipulate data
library(tidyverse) # move/manipulate data
library(mosaic)    # good for data exploration
library(ggplot2)   # creation of plots and visualizations
library(gridExtra) # arrange plots side by side
library(scales)    # used to set axes values manually
library(foreign)   # read in data stred in R
library(rgeos)     # used for creating maps of NEON field sites
library(maptools)  # used for creating maps of NEON field sites
library(raster)    # manipulate spatial data
library(rbokeh)    # used for creating maps of NEON field sites
library(rgdal)     # manipulate and read spatial data requires sp, will use proj.4 if installed

#Set strings as factors equal to false thoughout
options(stringsAsFactors = FALSE) 

# set working directory to ensure R can find the file we wish to import

#setwd("working-dir-path-here")

#Read in the data from part one of this tutorial series

idsmall = read.csv("NEON-mosquito-data-viz/idsmall.csv")

uniquetrap = read.csv("NEON-mosquito-data-viz/uniquetrap.csv")

#First extract site id for the uniquetrap data frame
uniquetrap$siteID <- substr(uniquetrap$plotID,1,4)

#Omit any instances where lat and lon is not known for a site
uniquetrap <- uniquetrap[uniquetrap$decimalLatitude!="",]

#taking only the observations that are not duplicated
uniquetrap <- uniquetrap[!duplicated(uniquetrap$siteID),]

#Re-run the templag function covered in part two of this tutorial series if not already  in your global environment

Templag <- function(siteID, date){
  date <- as.Date(date)
  filter1 <- temp.df[grepl(siteID, temp.df$siteID),]
  filter2 <- filter1[filter1$date >= date - 14 & filter1$date < date + 1,]
  TwoWeekAvgT <- mean(filter2$value)/10
  filter3 <- filter2[filter2$value > 16,]
  GDD <- length(unique(filter3$date))
  # return(GDD)
  # how to return multiple objects from a single function call
  return(list(TwoWeekAvgT=TwoWeekAvgT, GDD=GDD))
}


```


## NEON Mosquito Data Visualization Part Three: Abundance and Range of Culex tarsalis


### Creating Culex tarsalis Data Frames

To begin our investigation of spatial patterns and abundance of the species
*Culex tarsalis*, we need to create data frames with only the information
relating to *Culx tarsalis* is present. But first we take a look at the number
times *Culex tarsalis* was present in our data.

```{r first-look-at-Cut}

#Instances of Cu. tarsalis by Year
table(id$Year, id$scientificName=="Culex tarsalis")

```

We can see from the table that *Culex tarsalis* was sampled most fequently in 
the years 2014 and 2016. So very similar to our examination of species ricness,
we will be focusing on these two years. Now we will subset our id data frame
to only include rows of data where *Culex tarsalis* is in the 'scientificName'
variable column. We next create an even smaller subset of our data by selecting
rows whose year was either 2014 or 2016.

```{r create-Cut-Dfs}

#Tracking number of sites where Cu. tarsalis has been identified
tars<- id[grepl("Culex tarsalis", id$scientificName),]
tarssmall <- tars[tars$Year %in% c(2014,2016), ]

#Site Level variable
tarsSiteLevel <- ddply(tarssmall,~siteID + domainID + Date + Year, summarize, siteAbundance = sum(newindividualCount))

#merge with location information
tarsSiteLevel <- merge(x = uniquetrap[,c("siteID", "decimalLatitude", "decimalLongitude", "elevation.y", "nlcdClass.y")], y = tarsSiteLevel, by.y = "siteID", by.x = "siteID")

tarsSiteLevel$decimalLatitude<- as.numeric(tarsSiteLevel$decimalLatitude)
tarsSiteLevel$decimalLongitude<- as.numeric(tarsSiteLevel$decimalLongitude)

```

Next we create variables that captures the two week average maximum temperature
and the number of days greater than 16 degrees Celsius before collection by 
applying the temperature lag function to the new *Culex tarsalis* data frame
('tarssmall'). 


```{r apply-templag-again}
#Applying temperature lag function to tarssmall
CTTWAMT.mat<- mapply(Templag, tarsSiteLevel$siteID, tarsSiteLevel$Date)

#Create variables in tarsSiteLevel for two week temp lag and degree days
tarsSiteLevel$TwoWeekAvgMT <-c(unlist(CTTWAMT.mat[1,]))
tarsSiteLevel$DegreeDays <- c(unlist(CTTWAMT.mat[2,]))


```

We want to explore the range of *Culex tarsalis* for the years 2014 and 
2016, however, it might be helpful to know the regions in which *Culex tarsalis* 
is considered to be native. One problem with obtaining the native status of 
*Culex tarsalis* is that the native status information is contained in the 
taxonomy data frame as a single row of data, as demonstrated below.

```{r obtain-taxonomy}

#A look at the row of data in the taxonomy data frame 
taxonomy[grepl("Culex tarsalis", taxonomy$scientificName),]

```

Notice how the native status information in the taxonomy data is separated by
domain in separate columns in a 'wide' format. We need the native status
information to be in the form of a column or vector to be added to our current
'tarssmall' data frame, so we use the `gather` command from the `tidyr` package
to find the columns in the taxonomy data frame that contain native status
information and transformed that row of information into a new data frame.

```{r taxonomy-dataframe}

#Selecting certain columns of the tarstax data frame to find native status information
tarstax <- tidyr:: gather(taxonomy[grepl("Culex tarsalis", taxonomy$scientificName), grepl("NativeStatusCode", colnames(taxonomy))], "Domain ID")

#Omitting rows of data where no information is available
tarstax<-tarstax[c(5:15,17:23),]

#Creation of a dataframe that includes domain identification and native status
tarstax.df <-data.frame("domainID"= unique(domain.df$domainid), "DomainNativeStatus"= tarstax)

```

Once the new data frame containing native status information is created, we 
merge it with our existing "tarsSiteLevel" data frame and then subset by year. 
The reason that we choose to subset by year is so we can separate 

# 6/20: Comment is unfinished. 

```{r finishing-tarsSiteLevel}

#Merge Cu. tarsalis taxonomy df with tars df
tarsSiteLevel <- merge(x = tarstax.df[,c("domainID", "DomainNativeStatus.value")], y = tarsSiteLevel, by.x = "domainID", by.y = "domainID")

#Creation of subsets by Year
tars2014 <- tarsSiteLevel[tarsSiteLevel$Year == 2014, ]
tars2016 <- tarsSiteLevel[tarsSiteLevel$Year == 2016, ]

```


## Visualizing Culex tarsalis Abundance

To start exploring the abundance of *Culex tarsalis*, we will first be examining
abundance in relation to time for the years 2014 and 2016. To make a more clear
comparison between the years 2014 and 2016 we will be plotting these graphs
side by side rather than trying to plot both sets of data on one graph.
To plot two graphs side by side we will use the `grid.arrange` command from the
`gridExtra` package. 

```{r Cut-abundance-plot}

# Abundance of Culex tarsalis for 2014 and 2016

plot2016 <-ggplot(tars2016, aes(as.Date(Date), as.numeric(siteAbundance)))+
  geom_point(color = "darkblue", size = 2)+ 
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "1 month")+ #X-axis breaks set at 1 month
  labs(x = "Date", y = "Abundance of Culex tarsalis")+
  ggtitle("Date versus Abundance of Culex tarsalis for 2016")


plot2014 <-ggplot(tars2014, aes(as.Date(Date), as.numeric(siteAbundance)))+
  geom_point(color = "dark green", size = 2)+
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "1 month", limits = as.Date(c("2014-05-01","2014-11-01")))+ #X-axis breaks set at 1 month and limits set to match 2016 date limits
  labs(x = "Date", y = "Abundance of Culex tarsalis")+
  ggtitle("Date versus Abundance of Culex tarsalis for 2014")

#Command that arranges both 'plot2014' and 'plot2016' site by side
grid.arrange(plot2014,plot2016)

```

One thing to take away from this first visualization is that the abundance in 
*Culex tarsalis* across all NEON sites, reaches a maximum right around August 
for both 2014 and 2016. Another interesting take away from this visualization is
that *Culex tarsalis* is present in NEON samples as early as April in 2016 
however in 2014 this species wasn't present until late May. This suggests that 
there is some sort of seasonal fluctuation in abundance that differs not only 
by month but also by year. Similar to our exploration of species richness in
part two of this data tutorial we will explore whether the two week average 
maximum temperature is related to the *Culex tarsalis*. First however we will 
look at how the two week average maximum temperature fluctuates throughout the 
years 2014 and 2016.

```{r Cut-temp-plot}

# Maximum temperature for 2014 and 2016
tempplot2016 <-ggplot(tars2016, aes(as.Date(Date), as.numeric(TwoWeekAvgMT)))+
  geom_point(color = "red", size = 2)+
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "1 month")+ #X-axis breaks set at 1 month
  labs(x = "Date", y = "Average Max Temperature in C")+
  ggtitle("Date versus Two Week Average Maximum Temperature for 2016")
tempplot2014 <-ggplot(tars2014, aes(as.Date(Date), as.numeric(TwoWeekAvgMT)))+
  geom_point(color = "purple", size = 2)+
  scale_x_date(labels = date_format("%m-%Y"), date_breaks = "1 month", limits = as.Date(c("2014-05-01","2014-11-01")))+ #X-axis breaks set at 1 month and limits set to match 2016
  labs(x = "Date", y = "Average Max Temperature in C")+
  ggtitle("Date versus Two Week Average Maximum Temperature for 2014")

grid.arrange(tempplot2014, tempplot2016)

```

Looking at the above plots of temperature over time we can see that for these
two years, temperature reaches a maximum around the month of August. August
is also when we might expect to see a spike in abundance for *Culex tarsalis*
so it may be reasonable to conclude that these two variables are related.
Below we have a couple of plots that compare the temperature to abundance of 
*Culex tarsalis* over time for the years 2014 and 2016.

```{r Cut-temp-abundance-comp}

#Plot comparison between temperature and abundance
grid.arrange(tempplot2014, plot2014)

grid.arrange(tempplot2016, plot2016)

```


### Visualizing Range of Culex tarsalis 

Now we will turn our attention to mapping the range of *Culex tarsalis* using
`ggplot2`. This portion of the lesson is important not only because it is 
interesting to see the range of this particular species but it will also
teach you how to map NEON domains using shape files. But before we start the 
mapping process we will first create two data frames that capture the
observations where *Culex tarsalis* was first sampled in each domain and then
at each site. To accomplish this goal, we first order our "tarsSiteLevel"
by date. Then we subset the "tarsSiteLevel" data frame by taking a list of 
unique domain or site IDs and match it to where they first occur in our ordered 
data frame. 

```{r Cut-order-df}

#Order "tarsSiteLevel" by date 
tarsSiteLevel <- tarsSiteLevel[order(as.Date(tarsSiteLevel$Date)),]

#Creation of a dataframe where Culex tarsalis first found in each domain
tars.firstdomain <- tarsSiteLevel[match(unique(tarsSiteLevel$domainID), tarsSiteLevel$domainID),]

#Creation of a dataframe where Culex tarsalis first found at each site
tars.firstsite <- tarsSiteLevel[match(unique(tarsSiteLevel$siteID), tarsSiteLevel$siteID),]

```

Next we will `require` a couple of the `R` packages necessary to map shape files.
At this point make sure to check your working directory so that `R` can find 
where you've saved the shapefiles to your computer. 

```{r require-packages}

#Requiring packages
require("rgdal") # requires sp, will use proj.4 if installed
require("maptools")
require("ggplot2")
require("plyr")

```

Now we will set up our working directory so that `R` can read in the necessary
files for mapping NEON domains.  Please input the file path to where you have 
stored the shape files in the below code example and before running please 
remove the pound signs from the beginning of each line of code.

```{r Author-wd, echo = FALSE}

if(file.exists('~/GitHub/mosquito-intern')){ # This is the author's working directory
  repo_location = '~/GitHub/mosquito-intern'
}

```

```{r Cut-wd-path-example}

# if(file.exists('~/INSERT/YOUR/WD/PATH/HERE')){ 
#   repo_location = '~/GitHub/mosquito-intern'
# }

```

Once our working directory is set up we will now read in the shape files and
start to create our map. One thing to note is that the line of code that creates
"NEONmap.points" can be tricky because `fortify` is a part of both `maptools` 
and `ggplot2`. Since we want to look at the range of *Culex tarsalis*, we will also
create a variable that indicates whether *Culex tarsalis* is native to each 
NEON domain.

```{r read-in-shape-files}

#Read in file  
NEONmap = readShapePoly(paste(repo_location, 'resources/spatial files/NEON-domain-map/NEON_Domains.shp', sep='/'))

# Manipulate for plotting
NEONmap@data$id = rownames(NEONmap@data)

NEONmap.points = ggplot2::fortify(NEONmap)

NEONmap.df = join(NEONmap.points, NEONmap@data, by="id")

NEONmap.df$DomainID<-as.character(NEONmap.df$DomainID)

#Creation of a variable that indicates whether Culex tarsalis is native to a domain
NEONmap.df$nativestat<-ifelse(NEONmap.df$DomainID %in% c(1,2,3,5:17), "Native", "Absent")

```

# 6/20: readShapePoly can be replaced with readOGR
# NEONmap <- readOGR(dsn=path.expand("working dw here"), layer="layer here")

# 6/20: Line 389 throws error 'Error in fortify.SpatialPolygonsDataFrame(NEONmap, region = "id") : `region' not supported'. I removed 'region = ID' and the code ran. Can add region color codes manually later if needed.

Here is our first visualization of where *Culex tarsalis* is considered to be 
native. 

```{r CuT-map-viz}

# Map visualization of the range of Culex tarsalis
ggplot()+
  #geom_polygon maps the shapefiles into a polygon that looks like the united states and color by native status
  geom_polygon(data=NEONmap.df, aes(long, lat, group=group, fill= as.factor(nativestat)))+
  #geom_path creates the domain lines on the map
  geom_path(data=NEONmap.df, aes(long,lat, group=group), color = 'black')+
  #manual color native status 
  scale_fill_manual(values = c("lightyellow2", "gold")) +
  labs( x = "Longitude", y ="Latitude")+
  ggtitle("Map of Culex tarsalis Observation and Native Status")+
  #Increase size of text for title
  theme(plot.title = element_text(size = 16, face = "bold"))

```

# 6/20: After line 415 throws error code 'Error: 'data' must be uniquely named but has duplicated columns', I added code 'colnames(NEONmap.df) <- make.unique(names(NEONmap.df))' in order to stop the error. 

When we look at the map of NEON domains, it would seem as if *Culex tarsalis* is 
native to almost all of the continental United States. However this result is
somewhat misleading since native status is filled based on the entirety of 
domains so *Culex tarsalis* could be native to one part of a domain but absent
from another. The best example of this is in Alaska where *Culex tarsalis* is not 
known to exist, however the lower portion of alaska is colored to indicate 
native status because it is a part of a different domain that is located on the 
continental U.S where this species is found. So to add a bit more complexity to 
our map of the range of  *Culex tarsalis* we will create a series of points to
represent sites where this species is present or absent. To create this sort of 
variable we will have to take a sort of indirect route, manipulating our 
"domain.df" data frame so we can include all sites that were sampled in 2014 and
2016 but also single out the sites where *Culex tarsalis* was present. 

```{r Cut-map-viz-presence}

#merge domain and unique trap to obtain latitude and longitude information
domain.df <- merge(x = uniquetrap[,c("siteID","decimalLatitude", "decimalLongitude")], y = domain.df, by.x = "siteID", by.y = "siteid")

#merge with "tars.firstsite" to obtain where Culex tarsalis present 
domain.df <- merge(x = tars.firstsite[,c("siteID","DomainNativeStatus.value")], y = domain.df, by.x = "siteID", by.y = "siteID", all.y = TRUE)

#create variable to indicate whether Culex tarsalis is present
domain.df$tarsPresent <- ifelse(is.na(domain.df$DomainNativeStatus.value), "Not Present", "Present" )

#subset domain.df to include only sites sampled in 2014 and 2016
domain.df<- domain.df[domain.df$siteID %in% idsmall$siteID,]

#convert variables to proper type
domain.df$tarsPresent<- as.factor(domain.df$tarsPresent)
domain.df$decimalLatitude<- as.numeric(domain.df$decimalLatitude)
domain.df$decimalLongitude<- as.numeric(domain.df$decimalLongitude)

#mapping NEON domains along with presence/absence of Culex tarsalis
ggplot()+
  geom_polygon(data = NEONmap.df, aes(long, lat, group=group, fill= as.factor(nativestat)))+
  geom_path(data = NEONmap.df, aes(long,lat, group=group), color = 'black')+
  scale_fill_manual(values = c("lightyellow2", "gold")) +
  geom_point(data = domain.df, aes(decimalLongitude,decimalLatitude, color = tarsPresent),size = 2)+
  labs( x = "Longitude", y ="Latitude")+
  ggtitle("Map of Culex tarsalis Observation and Native Status")+
  theme(plot.title = element_text(size = 16, face = "bold"))

```

# 6/20: The final map needed the decimalLatitude and decimalLongitude to be switched in the geom_point call.

A second visualization of NEON domains using our newly created indicator
variable reveals that there are a good deal of sites where *Culex tarsalis* is 
considered to be native but doesn't appear to be present in samples. A reason
for this could be because *Culex tarsalis* might not be native to that portion
of the domain that our map is not displaying correctly, or this species just 
was not sampled in these areas during the years 2014 or 2016 and could be found
in future samples. Regardless of these discrepancies one interesting take away
from this visualization is that *Culex tarsalis* has been sampled in Alaska 
where it has previously been absent, suggesting that this species has undergone
a range expansion. 

<div id = "challenge" markdown = "1">

###Challenge: Creating a Range Map for a Different Species

Now that we've created a range and native status map for the species 
*Culex tarsalis* try applying the same techniques to a different species in the 
NEON data frame. A good candidate for this challenge could be *Aedes Albopictus*
however feel free to choose whatever your favorite mosquito species might be. 

</div>

