# Functions for data

Now that we know how to write functions, we can use this concept for preparing our
data sets for analysis, generating the intermediary data files, and exporting
them to CSV so that we can, for instance, share them with our collaborators.

Let's start with the chunk from the manuscript:

```{r, purl=TRUE}
## Gathering all the data files
split_gdp_files <- list.files(path = "data-raw", pattern = "gdp-percapita\\.csv$", full.names = TRUE)
split_gdp_list <- lapply(split_gdp_files, read.csv)
gdp <- do.call("rbind", split_gdp_list)
```

The simplest function we can write from this chunk is simply to enclose these
lines of code inside curly brackets and not forgetting to return the `gdp`
variable on the last line:

```{r, purl=FALSE}
gather_gdp_data <- function() {
    split_gdp_files <- list.files(path = "data-raw", pattern = "gdp-percapita\\.csv$", full.names = TRUE)
    split_gdp_list <- lapply(split_gdp_files, read.csv)
    gdp <- do.call("rbind", split_gdp_list)
    gdp
}
```

We can make this function more general by using the folder where the files are
stored and the pattern we use as arguments (`path` and `pattern`
respectively). This way, we could re-use this function for another project where
a similar operation (combining many CSV files into a single `data.frame`) would be
needed.

```{r, purl=FALSE}
gather_data <- function(path = "data-raw", pattern = "gdp-percapita\\.csv$") {
    split_files <- list.files(path = path, pattern = pattern, full.names = TRUE)
    split_list <- lapply(split_gdp_files, read.csv)
    gdp <- do.call("rbind", split_gdp_list)
    gdp
}
```

> ## A word of warning
>
> - the code here is pretty simple because we know that all datasets have
>   exactly the same columns, but in a real life example, we might way to add
>   additional checks to ensure that we won't be introducing any issues.
> - this also illustrates how general you need to be when writing your
>   functions. We could spend a lot of time optimizing and writing a function
>   that would work on all cases. Sometimes it's worth your time, sometimes it
>   might distract from your primary goal: writing the manuscript.

## Towards automation

We can create a `make_csv` function to automatically generate CSV files from our
data sets. This might come handy if you want to send your intermediate datasets
to your collaborators or if you want to inspect more closely that everything is
working as it should.

This function takes a data frame and make a CSV file out of it.


```{r, purl=FALSE}
make_csv <- function(obj, file, ...,  verbose = TRUE) {
    if (verbose) {
        message("Creating csv file: ", file)
    }
    write.csv(obj, file = file, row.names = FALSE, ...)
}
```

Now, we can combine the two functions we just wrote (`make_csv` and
`gather_data`) to generate a CSV file that contains the data from all countries:

```{r, purl=FALSE}
gdp_data <- gather_data()
make_csv(gdp_data, file = "data-output/gdp.csv")
```

### Your turn

Transform into functions these two pieces of code.

```{r, purl=TRUE}
## Turn this into a function called get_mean_lifeExp
mean_lifeExp_by_cont <- gdp %>%
    group_by(continent, year) %>%
    summarize(mean_lifeExp = mean(lifeExp)) %>%
    as.data.frame
```

```{r, purl=TRUE}
## Turn this into a function called get_latest_lifeExp
latest_lifeExp <- gdp %>%
    filter(year == max(gdp$year)) %>%
    group_by(continent) %>%
    summarize(latest_lifeExp = mean(lifeExp)) %>%
    as.data.frame
```

## Long computations

[aside: talk about it if time permits]

Caching is available in knitr but it can be pretty fragile. For instance, the
caching is only based on whether the code in your chunk changes and doesn't
check if your data on your hard drive is changing.

In other cases, the output of your R code can't be represented into a CSV files,
so you need to save it directly into an R object.

```{r, purl=TRUE}
## If you need to save an R object to avoid the repetition of long computations
make_rds <- function(obj, file, ..., verbose = TRUE) {
    if (verbose) {
        message("Creating rds file: ", file)
    }
    saveRDS(obj, file = file)
    invisible(file.exists(file))
}
```

Then in your knitr document, you can do:

`gdp <- readRDS(file="data-output/gdp.rds")`

or maybe even:

```{r, eval=FALSE}
## An example of the kind of code you could use to work with time-consuming
## computations in R.
if ( !file.exists("data-output/gdp.rds")) {
    gdp <- gather_gdp_data() ## long computation...
    make_rds(gdp, file="data-output/gdp.rds")
}
gdp <- readRDS(file="data-output/gdp.rds")
```
