---
syncID: 63716f3d5a5745f7b0ce7f3bf9de982d
title: "Classify a Raster using Threshold Values in R"
description: "Classify rasters in R using threshold values."
dateCreated:  2016-05-01
authors: Leah A. Wasser, Kyla Dahlin
contributors: Megan A. Jones
estimatedTime:
packagesLibraries: [raster]
topics: HDF5, raster
languagesTool: R
dataProduct:
code1: institute-materials/day2_tuesday/classify-raster-threshold-R.R
tutorialSeries: institute-day2
---

## About

In this tutorial, we will walk through how to classify a raster file using
defined value ranges in R.

First, let's load the required libraries.

```{r load-libraries, warning=FALSE, results='hide', message=FALSE}
# load libraries
library(raster)
library(rhdf5)
library(rgdal)

# be sure to set your working directory
# setwd("~/Documents/data/NEONDI-2016") # Mac
# setwd("~/data/NEONDI-2016")  # Windows

#source("/Users/lwasser/Documents/GitHub/neon-aop-package/neonAOP/R/aop-data.R")

```

## Import LiDAR data

To begin, we will open the NEON LiDAR Digital Surface and Digital Terrain Models
(DSM and DTM) which are in Geotiff format.

```{r import-lidar }

# read LiDAR canopy height model
chm <- raster("NEONdata/D17-California/TEAK/2013/lidar/TEAK_lidarCHM.tif")

```

## View CHM

``` {r explore-chm}
# assign chm values of 0 to NA
chm[chm==0] <- NA

# do the values in the data look reasonable?
plot(chm,
     main="Canopy Height \n LowerTeakettle, California")

hist(chm,
     main="Distribution of Canopy Height  \n Lower Teakettle, California",
     xlab="Tree Height (m)",
     col="springgreen")


```

## Import Aspect Data

Next, we'll import an aspect dataset - one of the NEON data products.

<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i>**Data Tip:** You can calculate aspect in R
if you have a surface or elevation model as follows:  
`terrain(your.raster.data, opt = "aspect", unit = "degrees", neighbors = 8)` .
</div>

```{r import-aspect-data }

# calculate aspect of cropped DTM
# aspect <- terrain(your-dem-or-terrain-data.tif, opt = "aspect", unit = "degrees", neighbors = 8)
# example code:
# aspect <- terrain(all.data[[3]], opt = "aspect", unit = "degrees", neighbors = 8)
# read in aspect .tif
aspect <- raster("NEONdata/D17-California/TEAK/2013/lidar/TEAK_lidarAspect.tif")

plot(aspect,
     main="Aspect for Lower Teakettle Field Site",
     axes=F)

```

## Threshold Based Raster Classification

Next, we will create a classified raster object. To do this we need to:

1. Create a matrix containing the threshold ranges and the associated "class"
2. Use the `raster::reclassify` function to create a new raster.

Our range of values are as follows:
We will assign all north facing slopes "1" and south facing "2".

**North Facing Slopes:** 0-45 degrees, 315-360 degrees; class=1

**South Facing Slopes:** 135-225 degrees; class=2


```{r classify-raster }

# first create a matrix of values that represent the classification ranges
# North facing = 1
# South facing = 2
class.m <- c(0, 45, 1,
             45, 135, NA,
             135, 225, 2,  
             225 , 315, NA,
             315, 360, 1)
class.m

# reshape the object into a matrix with columns and rows
rcl.m <- matrix(class.m, 
                ncol=3, 
                byrow=TRUE)
rcl.m

# reclassify the raster using the reclass object - rcl.m
asp.ns <- reclassify(aspect, 
                     rcl.m)

# plot outside of the plot region

# make room for a legend
par(xpd = FALSE, mar=c(5.1, 4.1, 4.1, 4.5))
# plot
plot(asp.ns,
     col=c("white","blue","green"), # hard code colors, unclassified (0)=white,
		 #N (1) =blue, S(2)=green
     main="North and South Facing Slopes \nLower Teakettle",
     legend=F)
# allow legend to plot outside of bounds
par(xpd=TRUE)
# create the legend
legend((par()$usr[2] + 20), 4103300,  # set x,y legend location
       legend = c("North", "South"),  # make sure the order matches the colors, next
       fill = c("blue", "green"),
       bty="n") # turn off border

```

## Export Classified Raster

```{r export-geotiff, eval=FALSE}

# export geotiff
writeRaster(asp.ns,
            filename="outputs/TEAK/Teak_nsAspect.tif",
            format="GTiff",
            options="COMPRESS=LZW",
            overwrite = TRUE,
            NAflag = -9999)

```

<div id="ds-challenge" markdown="1">
## Challenge: Classify a Raster using Threshold Values

1. Have a look at the code that you created for this lesson. Now imagine that you are
your future self. Document your script so that your methods and process  is clear.
2. In documenting your script, synthesize the outputs.Do they tell you anything
about vegetation structure at the field site?
3. Create the following threshold classified outputs:

* A raster where NDVI values are classified into categories 0 (low greenness,
NDVI <.3), 1 (medium greenness, NDVI =.3-.6) and 2 (high greenness, NDVI>.6).
* A raster where canopy height is classified. Explore the CHM data and chose
threshold values that make sense given the distribution of values in the data.

Be sure to document your workflow as you go using Markdown within your R Markdown
document.

When you are done, knit your outputs to HTML. Save the file as
LastName_Tues_classifyThreshold.html. Add these to the Tuesday directory in your
**DI-NEON-participants** Git direcotry and push them to your fork in GitHub.
Merge with the central repository using a pull request.
</div>
