This code snippet provides instruction on working with two different NEON data products to estimate tree height: 

* **DP3.30015.001, Ecosystem structure**, aka Canopy Height Model (CHM) 
* **DP1.10098.001, Vegetation structure**

The <a href="https://data.neonscience.org/data-products/DP3.30015.001" target="_blank">CHM data</a> are derived from the Lidar point cloud data collected by the remote sensing platform. 
The <a href="https://data.neonscience.org/data-products/DP1.10098.001" target="_blank">vegetation structure data</a> are collected by by field staff on the ground. We will be using data from the Wind River Experimental Forest NEON field site located in Washington state. The 
predominant vegetation there are tall evergreen conifers. 

<div id="ds-objectives" markdown="1">

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R loaded on your computer to complete this tutorial.

</div>

## 1. Setup

Start by installing and loading packages (if necessary) and setting 
options. One of the packages we'll be using, `geoNEON`, is only 
available via GitHub, so it's installed using the `devtools` package. 
The other packages can be installed directly from CRAN.

Installation can be run once, then periodically to get package updates.

```{r install_packages, eval=FALSE}

install.packages("neonUtilities")
install.packages("neonOS")
install.packages("sp")
install.packages("raster")
install.packages("devtools")
devtools::install_github("NEONScience/NEON-geolocation/geoNEON")

```

Now load packages. This needs to be done every time you run code. 
We'll also set a working directory for data downloads.

```{r load-packages, results="hide"}

library(sp)
library(raster)
library(neonUtilities)
library(neonOS)
library(geoNEON)

options(stringsAsFactors=F)

# set working directory
# adapt directory path for your system
wd <- "~/data"
setwd(wd)

```

## 2. Vegetation structure data

Download the vegetation structure data using the `loadByProduct()` function in
the `neonUtilities` package. Inputs needed to the function are:

* `dpID`: data product ID; woody vegetation structure = DP1.10098.001
* `site`: (vector of) 4-letter site codes; Wind River = WREF
* `package`: basic or expanded; we'll download basic here
* `check.size`: should this function prompt the user with an estimated download size? Set to `FALSE` here for ease of processing as a script, but good to leave as default `TRUE` when downloading a dataset for the first time.

Refer to the <a href="https://www.neonscience.org/sites/default/files/cheat-sheet-neonUtilities.pdf" target="_blank">cheat sheet</a> 
for the `neonUtilities` package for more details if desired.

```{r veglist, results="hide"}

veglist <- loadByProduct(dpID="DP1.10098.001", 
                         site="WREF", 
                         package="basic", 
                         check.size = FALSE)

```

Use the `getLocTOS()` function in the `geoNEON` package to get 
precise locations for the tagged plants. Refer to the package 
documentation for more details.

```{r vegmap, results="hide"}

vegmap <- getLocTOS(veglist$vst_mappingandtagging, 
                          "vst_mappingandtagging")

```

Now we have the mapped locations of individuals in the `vst_mappingandtagging` 
table, and the annual measurements of tree dimensions such as height and 
diameter in the `vst_apparentindividual` table. To bring these measurements 
together, join the two tables, using the `joinTableNEON()` function from the 
`neonOS` package. Refer to the <a href="https://data.neonscience.org/data-products/DP1.10098.001" target="_blank">Quick Start Guide</a> 
for Vegetation structure for more information about the data tables and the 
joining instructions `joinTableNEON()` is using.

```{r veg_merge}

veg <- joinTableNEON(veglist$vst_apparentindividual, 
                     vegmap, 
                     name1="vst_apparentindividual",
                     name2="vst_mappingandtagging")

```

Let's see what the data look like! Make a stem map of the plants in 
plot WREF_075. Note that the `circles` argument of the `symbols()` function expects a radius, but 
`stemDiameter` is just that, a diameter, so we will need to divide by two. 
And `stemDiameter` is in centimeters, but the mapping scale is in meters, 
so we also need to divide by 100 to get the scale right.

```{r plot-1}

symbols(veg$adjEasting[which(veg$plotID=="WREF_075")], 
        veg$adjNorthing[which(veg$plotID=="WREF_075")], 
        circles=veg$stemDiameter[which(veg$plotID=="WREF_075")]/100/2, 
        inches=F, xlab="Easting", ylab="Northing")

```

And now overlay the estimated uncertainty in the location of each stem, 
in blue:

```{r plot-2}

symbols(veg$adjEasting[which(veg$plotID=="WREF_075")], 
        veg$adjNorthing[which(veg$plotID=="WREF_075")], 
        circles=veg$stemDiameter[which(veg$plotID=="WREF_075")]/100/2, 
        inches=F, xlab="Easting", ylab="Northing")
symbols(veg$adjEasting[which(veg$plotID=="WREF_075")], 
        veg$adjNorthing[which(veg$plotID=="WREF_075")], 
        circles=veg$adjCoordinateUncertainty[which(veg$plotID=="WREF_075")], 
        inches=F, add=T, fg="lightblue")

```

## 3. Canopy height model data

Now we'll download the CHM tile covering plot WREF_075. Several 
other plots are also covered by this tile. We could download all tiles 
that contain vegetation structure plots, but in this exercise we're 
sticking to one tile to limit download size and processing time.

The `tileByAOP()` function in the `neonUtilities` package allows for 
download of remote sensing tiles based on easting and northing 
coordinates, so we'll give it the coordinates of all the trees in 
plot WREF\_075 and the data product ID, DP3.30015.001 (note that if 
WREF\_075 crossed tile boundaries, this code would download all 
relevant tiles).

The download will include several metadata files as well as the data 
tile. Load the data tile into the environment using the `raster` package.

```{r get-chm, results="hide"}

byTileAOP(dpID="DP3.30015.001", site="WREF", year="2017", 
          easting=veg$adjEasting[which(veg$plotID=="WREF_075")], 
          northing=veg$adjNorthing[which(veg$plotID=="WREF_075")],
          check.size=FALSE, savepath=wd)

chm <- raster(paste0(wd, "/DP3.30015.001/neon-aop-products/2017/FullSite/D16/2017_WREF_1/L3/DiscreteLidar/CanopyHeightModelGtif/NEON_D16_WREF_DP3_580000_5075000_CHM.tif"))

```

Let's view the tile.

```{r plot-chm}

plot(chm, col=topo.colors(5))

```
