---
syncID: 68d6be70d17d4ab9807cd46081e6ee63
title: "Introduction to HDF5 Files in R"
description: "Learn how to build a HDF5 file in R from scratch! Add groups, datasets and attributes. Read data out from the file."
dateCreated: 2015-05-20 15:22:52
authors: Leah A. Wasser
contributors:
estimatedTime: 1.0 - 1.5 Hours
packagesLibraries: rhdf5
topics: HDF5
languagesTool: R
dataProduct:
code1: /HDF5/Intro-HDF5-R.R
tutorialSeries: intro-hdf5-r-series
---

<div id="ds-objectives" markdown="1">

## Learning Objectives
After completing this tutorial, you will be able to:

* Understand how HDF5 files can be created and structured in R using the 
rhdf5 libraries.
* Understand the three key HDF5 elements: 
		* the HDF5 file itself,
		* groups,and 
		* datasets.
* Understand how to add and read attributes from an HDF5 file.


## Things You’ll Need To Complete This Tutorial
To complete this tutorial you will need the most current version of R and, 
preferably, RStudio loaded on your computer.

### R Libraries to Install:

* **rhdf5**

[ {{ site.baseurl }} R/Packages-In-R/](More on Packages in R - Adapted from Software Carpentry.)

### Data to Download
We will use the file below in the optional challenge activity at the end of this 
tutorial.
{% include/dataSubsets/_data_Field-Site-Spatial-Data.html %}

***
{% include/_greyBox-wd-rscript.html %}
***

### Additional Resources

Consider reviewing the documentation for the <a href="http://www.bioconductor.org/packages/release/bioc/manuals/rhdf5/man/rhdf5.pdf" target="_blank">RHDF5 package</a>.

</div>


## About HDF5
The HDF5 file can store large, heterogeneous datasets that include metadata. It 
also supports efficient `data slicing`, or extraction of particular subsets of a 
dataset which means that you don't have to read  large files read into the 
computers memory / RAM in their entirety in order work with them. 

[Read more about HDF5 here.]({{ site.baseurl }}/HDF5/About/)

## HDF5 in R

To access HDF5 files in `R`, we will use the `rhdf5` library which is part of 
the <a href="http://www.bioconductor.org" target="_blank">Bioconductor</a> 
suite of `R` libraries. It might also be useful to install 
<a href="http://www.hdfgroup.org/products/java/hdfview/" target="_blank">the 
free HDF5 viewer</a> which will allow you to explore the contents of an HDF5 file 
using a graphic interface.

[More about working with HDFview and a hands-on activity here]({{ site.baseurl }}/HDF5/Exploring-Data-HDFView/).


First, let's get R setup. We will use the rhdf5 library. To access HDF5 files in 
R, we will use the rhdf5 library which is part of the Bioconductor suite of R 
packages. As of Aug. 2017 this package was not yet on CRAN. 

```{r load libraries}

# Install rhdf5 package (only need to run if not already installed)
#source("http://bioconductor.org/biocLite.R")
#biocLite("rhdf5")

# Call the R HDF5 Library
library("rhdf5")

# set working directory to ensure R can find the file we wish to import and where
# we want to save our files
#setwd("working-dir-path-here")

```

Read more about the 
<a href="http://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.pdf" target="_blank"> `rhdf5` package here.</a>

## Create an HDF5 File in R

Now, let's create a basic H5 file with one group and one dataset in it. 

```{r create_new_hdf5_file}

# Create hdf5 file
h5createFile("vegData.h5")
h5createFile()

# create a group called aNEONSite within the H5 file
h5createGroup("vegData.h5", "aNEONSite")

# view the structure of the h5 we've created
h5ls("vegData.h5")
```

Next, let's create some dummy data to add to our H5 file.

```{r create_sample_data }
# create some sample, numeric data 
a <- rnorm(n=40, m=1, sd=1) 
someData <- matrix(a,nrow=20,ncol=2)
```

Write the sample data to the H5 file.

```{r add_data_to_file }

# add some sample data to the H5 file located in the aNEONSite group
# we'll call the dataset "temperature"
h5write(someData, file = "vegData.h5", name="aNEONSite/temperature")

# let's check out the H5 structure again
h5ls("vegData.h5")

```

View a "dump" of the entire HDF5 file. NOTE: use this command with CAUTION if you
are working with larger datasets!

``` {r view_h5_contents }

# we can look at everything too 
# but be cautious using this command!
h5dump("vegData.h5")

# Close the file. This is good practice.
H5close()

```


## Add Metadata (attributes)

Let's add some metadata (called attributes in HDF5 land) to our dummy temperature
data. First, open up the file.

```{r add-attributes-to-file}

# open the file, create a class
fid <- H5Fopen("vegData.h5")
# open up the dataset to add attributes to, as a class
did <- H5Dopen(fid, "aNEONSite/temperature")

# Provide the NAME and the ATTR (what the attribute says) for the attribute.
h5writeAttribute(did, attr="Here is a description of the data",
                 name="Description")
h5writeAttribute(did, attr="Meters",
                 name="Units")

```

Now we can add some attributes to the file. 

``` {r group-attributes }
# let's add some attributes to the group
did2 <- H5Gopen(fid, "aNEONSite/")

h5writeAttribute(did2, attr="San Joaquin Experimental Range",
                 name="SiteName")

h5writeAttribute(did2, attr="Southern California",
                 name="Location")

# close the files, groups and the dataset when you're done writing to them!
H5Dclose(did)

H5Gclose(did2)

H5Fclose(fid)

```

## Working with an HDF5 File in R

Now that we've created our H5 file, let's use it! First, let's have a look at 
the attributes of the dataset and group in the file.

```{r read & review attributes}

# look at the attributes of the precip_data dataset
h5readAttributes(file = "vegData.h5", 
                 name = "aNEONSite/temperature")

# look at the attributes of the aNEONsite group
h5readAttributes(file = "vegData.h5", 
                 name = "aNEONSite")

# let's grab some data from the H5 file
testSubset <- h5read(file = "vegData.h5", 
                 name = "aNEONSite/temperature")

testSubset2 <- h5read(file = "vegData.h5", 
                 name = "aNEONSite/temperature",
                 index=list(NULL,1))
H5close()


```

Once we've extracted data from our H5 file, we can work with it
in R. 

``` {r access-plot-data}

# create a quick plot of the data
hist(testSubset2)

```

 <div id="ds-challenge" markdown="1">
## Challenge: Work with HDF5 Files

Time to practice the skills you've learned. Open up the D17_2013_SJER_vegStr.csv 
in R. 

* Create a new HDF5 file called `vegStructure`.
* Add a group in your HDF5 file called `SJER`. 
* Add the veg structure data to that folder.
* Add some attributes the SJER group and to the data. 
* Now, repeat the above with the D17_2013_SOAP_vegStr csv.
* Name your second group SOAP

Hint: `read.csv()` is a good way to read in .csv files. 

</div>

``` {r challenge-code-HDF5-file, echo=FALSE}

# options(stringsAsFactors = FALSE)
# newData <- read.csv("D17_2013_SJER_vegStr.csv")

```

