---
syncID: e0778998a7ea4b7c93da38a364d12a65
title: "Explore and work with NEON biodiversity data from aquatic ecosystems"
description: "Download and explore NEON macroinvertebrate data. This includes instruction on how to convert to long and wide tables, as well as an exploration of alpha, beta, and gamma diversity from Jost (2007)."
dateCreated: 2020-06-22
authors: Eric R. Sokol
contributors: Donal O'Leary, Felipe Sanchez
estimatedTime: 1 Hour
packagesLibraries: tidyverse, neonUtilities, vegan
topics: organisms, data-viz
languagesTool: R
dataProduct: DP1.20120.001
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/biodiversity/aquatic-biodiversity/01_working_with_NEON_macroinverts/01_working_with_NEON_macroinverts.R
tutorialSeries: 
urlTitle: aquatic-diversity-macroinvertebrates
---

<div id="ds-objectives" markdown="1">

## Learning Objectives 
After completing this tutorial you will be able to: 

* Download NEON macroinvertebrate data.
* Organize those data into long and wide tables.
* Calculate alpha, beta, and gamma diversity following Jost (2007).

## Things You’ll Need To Complete This Tutorial

### R Programming Language
You will need a current version of R to complete this tutorial. We also recommend 
the RStudio IDE to work with R. 

### R Packages to Install
Prior to starting the tutorial ensure that the following packages are installed. 

* **tidyverse:** `install.packages("tidyverse")`
* **neonUtilities:** `install.packages("neonUtilities")`
* **vegan:** `install.packages("vegan")`

<a href="https://www.neonscience.org/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

</div>

## Introduction
Biodiversity is a popular topic within ecology, but quantifying and describing biodiversity precisely can be elusive. In this tutorial, we will describe many of the aspects of biodiversity using NEON's <a href="https://data.neonscience.org/data-products/DP1.20120.001">Macroinvertebrate Collection data</a>.

## Load Libraries and Prepare Workspace
First, we will load all necessary libraries into our R environment. If you have not already installed these libraries, please see the 'R Packages to Install' section above.

There are also two optional sections in this code chunk: clearing your environment, and loading your NEON API token. Clearing out your environment will erase _all_ of the variables and data that are currently loaded in your R session. This is a good practice for many reasons, but only do this if you are completely sure that you won't be losing any important information! Secondly, your NEON API token will allow you increased download speeds, and helps NEON __anonymously__ track data usage statistics, which helps us optimize our data delivery platforms, and informs our monthly and annual reporting to our funding agency, the National Science Foundation. Please consider signing up for a NEON data user account and using your token <a href="https://www.neonscience.org/neon-api-tokens-tutorial">as described in this tutorial here</a>.

```{r load libraries, warning=FALSE}

# clean out workspace

#rm(list = ls()) # OPTIONAL - clear out your environment
#gc()            # Uncomment these lines if desired

# load libraries 
library(tidyverse)
library(neonUtilities)
library(vegan)


# source .r file with my NEON_TOKEN
# source("my_neon_token.R") # OPTIONAL - load NEON token
# See: https://www.neonscience.org/neon-api-tokens-tutorial

```

## Download NEON Macroinvertebrate Data
Now that the workspace is prepared, we will download NEON macroinvertebrate data using the neonUtilities function `loadByProduct()`.

```{r download-data, message=FALSE, warning=FALSE, results='hide'}

# Macroinvert dpid
my_dpid <- 'DP1.20120.001'

# list of sites
my_site_list <- c('ARIK', 'POSE', 'MAYF')

# get all tables for these sites from the API -- takes < 1 minute
all_tabs_inv <- neonUtilities::loadByProduct(
  dpID = my_dpid,
  site = my_site_list,
  #token = NEON_TOKEN, #Uncomment to use your token
  check.size = F)

```

## Macroinvertebrate Data Munging
Now that we have the data downloaded, we will need to do some 'data munging' to reorganize our data into a more useful format for this analysis. First, let's take a look at some of the tables that were generated by `loadByProduct()`:

```{r download-overview, message=FALSE, warning=FALSE}


# what tables do you get with macroinvertebrate 
# data product
names(all_tabs_inv)

# extract items from list and put in R env. 
all_tabs_inv %>% list2env(.GlobalEnv)

# readme has the same informaiton as what you 
# will find on the landing page on the data portal

# The variables file describes each field in 
# the returned data tables
View(variables_20120)

# The validation file provides the rules that 
# constrain data upon ingest into the NEON database:
View(validation_20120)

# the categoricalCodes file provides controlled 
# lists used in the data
View(categoricalCodes_20120)

```

Next, we will perform several operations in a row to re-organize our data. Each step is described by a code comment.

```{r munging-and-organizing, message=FALSE, warning=FALSE}

# It is good to check for duplicate records. This had occurred in the past in 
# data published in the inv_fieldData table in 2021. Those duplicates were 
# fixed in the 2022 data release. 
# Here we use sampleID as primary key and if we find duplicate records, we
# keep the first uid associated with any sampleID that has multiple uids

de_duped_uids <- inv_fieldData %>% 
  
  # remove records where no sample was collected
  filter(!is.na(sampleID)) %>%  
  group_by(sampleID) %>%
  summarise(n_recs = length(uid),
                   n_unique_uids = length(unique(uid)),
                   uid_to_keep = dplyr::first(uid)) 





# Are there any records that have more than one unique uid?
max_dups <- max(de_duped_uids$n_unique_uids %>% unique())





# filter data using de-duped uids if they exist
if(max_dups > 1){
  inv_fieldData <- inv_fieldData %>%
  dplyr::filter(uid %in% de_duped_uids$uid_to_keep)
}





# extract year from date, add it as a new column
inv_fieldData <- inv_fieldData %>%
  mutate(
    year = collectDate %>% 
      lubridate::as_date() %>% 
      lubridate::year())




# extract location data into a separate table
table_location <- inv_fieldData %>%

  # keep only the columns listed below
  select(siteID, 
         domainID,
         namedLocation, 
         decimalLatitude, 
         decimalLongitude, 
         elevation) %>%
  
  # keep rows with unique combinations of values, 
  # i.e., no duplicate records
  distinct()




# create a taxon table, which describes each 
# taxonID that appears in the data set
# start with inv_taxonomyProcessed
table_taxon <- inv_taxonomyProcessed %>%

  # keep only the coluns listed below
  select(acceptedTaxonID, taxonRank, scientificName,
         order, family, genus, 
         identificationQualifier,
         identificationReferences) %>%

  # remove rows with duplicate information
  distinct()



# taxon table information for all taxa in 
# our database can be downloaded here:
# takes 1-2 minutes
# full_taxon_table_from_api <- neonUtilities::getTaxonTable("MACROINVERTEBRATE", token = NEON_TOKEN)




# Make the observation table.
# start with inv_taxonomyProcessed

# check for repeated taxa within a sampleID that need to be added together
inv_taxonomyProcessed_summed <- inv_taxonomyProcessed %>% 
  select(sampleID,
         acceptedTaxonID,
         individualCount,
         estimatedTotalCount) %>%
  group_by(sampleID, acceptedTaxonID) %>%
  summarize(
    across(c(individualCount, estimatedTotalCount), ~sum(.x, na.rm = TRUE)))
  



# join summed taxon counts back with sample and field data
table_observation <- inv_taxonomyProcessed_summed %>%
  
  # Join relevant sample info back in by sampleID
  left_join(inv_taxonomyProcessed %>% 
              select(sampleID,
                     domainID,
                     siteID,
                     namedLocation,
                     collectDate,
                     acceptedTaxonID,
                     order, family, genus, 
                     scientificName,
                     taxonRank) %>%
              distinct()) %>%
  
  # Join the columns selected above with two 
  # columns from inv_fieldData (the two columns 
  # are sampleID and benthicArea)
  left_join(inv_fieldData %>% 
              select(sampleID, eventID, year, 
                     habitatType, samplerType,
                     benthicArea)) %>%
  
  # some new columns called 'variable_name', 
  # 'value', and 'unit', and assign values for 
  # all rows in the table.
  # variable_name and unit are both assigned the 
  # same text strint for all rows. 
  mutate(inv_dens = estimatedTotalCount / benthicArea,
         inv_dens_unit = 'count per square meter')





# check for duplicate records, should return a table with 0 rows
table_observation %>% 
  group_by(sampleID, acceptedTaxonID) %>% 
  summarize(n_obs = length(sampleID)) %>%
  filter(n_obs > 1)




# extract sample info
table_sample_info <- table_observation %>%
  select(sampleID, domainID, siteID, namedLocation, 
         collectDate, eventID, year, 
         habitatType, samplerType, benthicArea, 
         inv_dens_unit) %>%
  distinct()




# remove singletons and doubletons
# create an occurrence summary table
taxa_occurrence_summary <- table_observation %>%
  select(sampleID, acceptedTaxonID) %>%
  distinct() %>%
  group_by(acceptedTaxonID) %>%
  summarize(occurrences = n())





# filter out taxa that are only observed 1 or 2 times
taxa_list_cleaned <- taxa_occurrence_summary %>%
  filter(occurrences > 2)





# filter observation table based on taxon list above
table_observation_cleaned <- table_observation %>%
  filter(acceptedTaxonID %in%
             taxa_list_cleaned$acceptedTaxonID,
         !sampleID %in% c("MAYF.20190729.CORE.1",
                          "MAYF.20200713.CORE.1",
                          "POSE.20160718.HESS.1")) 
                      #this is an outlier sampleID






# some summary data
sampling_effort_summary <- table_sample_info %>%
  
  # group by siteID, year
  group_by(siteID, year, samplerType) %>%
  
  # count samples and habitat types within each event
  summarise(
    event_count = eventID %>% unique() %>% length(),
    sample_count = sampleID %>% unique() %>% length(),
    habitat_count = habitatType %>% 
        unique() %>% length())




# check out the summary table
sampling_effort_summary %>% as.data.frame() %>% 
  head() %>% print()


```

## Working with 'Long' data
'Reshaping' your data to use as an input to a particular fuction may require you to consider: do I want 'long' or 'wide' data? Here's a link to <a href="https://www.theanalysisfactor.com/wide-and-long-data/">a great article from 'the analysis factor' that describes the differences</a>.

For this first step, we will use data in a 'long' table:

```{r long-data, fig.cap= "Horizontal bar graph showing the number of taxa for each taxonomic rank at the D02:POSE, D08:MAYF, and D10:ARIK sites. Including facet_wrap to the ggplot call creates a seperate plot for each of the faceting arguments, which in this case are domainID and siteID.", message=FALSE, warning=FALSE}

# no. taxa by rank by site
table_observation_cleaned %>% 
  group_by(domainID, siteID, taxonRank) %>%
  summarize(
    n_taxa = acceptedTaxonID %>% 
        unique() %>% length()) %>%
  ggplot(aes(n_taxa, taxonRank)) +
  facet_wrap(~ domainID + siteID) +
  geom_col()
```
```{r long-data-2,  fig.cap= "Bar graph of the occurence of each taxonomic order at the D02:POSE, D08:MAYF, and D10:ARIK sites. Occurence data at each site is depicted as stacked bars for each order, where a red bar represents D10:ARIK, a green bar represents D08:MAYF, and a blue bar represents the D02:POSE site. The data has also been reordered to show the greatest to least occuring taxonomic order from left to right.", message=FALSE, warning=FALSE}
# library(scales)
# sum densities by order for each sampleID
table_observation_by_order <- 
    table_observation_cleaned %>% 
    filter(!is.na(order)) %>%
    group_by(domainID, siteID, year, 
             eventID, sampleID, habitatType, order) %>%
    summarize(order_dens = sum(inv_dens, na.rm = TRUE))
  
  
# rank occurrence by order
table_observation_by_order %>% head()


# stacked rank occurrence plot
table_observation_by_order %>%
  group_by(order, siteID) %>%
  summarize(
    occurrence = (order_dens > 0) %>% sum()) %>%
    ggplot(aes(
        x = reorder(order, -occurrence), 
        y = occurrence,
        color = siteID,
        fill = siteID)) +
    geom_col() +
    theme(axis.text.x = 
              element_text(angle = 45, hjust = 1))

```
```{r long-data-3, fig.cap= "Box plots of the log density of each taxonomic order per site. This graph consists of three box plots, organized vertically in one column, that correspond to log density data for each site. This is achieved through the use of the Facet_grid function in the ggplot call.", message=FALSE, warning=FALSE}
# faceted densities plot
table_observation_by_order %>%
  ggplot(aes(
    x = reorder(order, -order_dens), 
    y = log10(order_dens),
    color = siteID,
    fill = siteID)) +
  geom_boxplot(alpha = .5) +
  facet_grid(siteID ~ .) +
  theme(axis.text.x = 
            element_text(angle = 45, hjust = 1))



```

### Making Data 'wide'
For the next process, we will need to make our data table in the 'wide' format.

```{r make-wide, message=FALSE, warning=FALSE}


# select only site by species density info and remove duplicate records
table_sample_by_taxon_density_long <- table_observation_cleaned %>%
  select(sampleID, acceptedTaxonID, inv_dens) %>%
  distinct() %>%
  filter(!is.na(inv_dens))

# table_sample_by_taxon_density_long %>% nrow()
# table_sample_by_taxon_density_long %>% distinct() %>% nrow()



# pivot to wide format, sum multiple counts per sampleID
table_sample_by_taxon_density_wide <- table_sample_by_taxon_density_long %>%
  tidyr::pivot_wider(id_cols = sampleID, 
                     names_from = acceptedTaxonID,
                     values_from = inv_dens,
                     values_fill = list(inv_dens = 0),
                     values_fn = list(inv_dens = sum)) %>%
  column_to_rownames(var = "sampleID") 

# check col and row sums -- mins should all be > 0
colSums(table_sample_by_taxon_density_wide) %>% min()
rowSums(table_sample_by_taxon_density_wide) %>% min()

```

## Multiscale Biodiversity


Reference:
Jost, L. 2007. Partitioning diversity into independent alpha and beta components. Ecology 88:2427–2439. <a href="https://doi.org/10.1890/06-1736.1">https://doi.org/10.1890/06-1736.1</a>.

These metrics are based on Robert Whittaker's multiplicative diversity where
* gamma is regional biodiversity
* alpha is local biodiversity (e.g., the mean diversity at a patch)
* and beta diversity is a measure of among-patch variability in community composition. 

Beta could be interpreted as the number of "distinct" communities present within the region.

The relationship among alpha, beta, and gamma diversity is:
   __beta = gamma / alpha__ 

The influence of relative abundances over the calculation of alpha, beta, and gamma diversity metrics is determined by the coefficient q. The coefficient "q" determines the "order" of the diversity metric, where q = 0 provides diversity measures based on richness, and higher orders of q give more weight to taxa that have higher abundances in the data. Order q = 1 is related to Shannon diveristy metrics, and order q = 2 is related to Simpson diversity metrics.

#### Alpha diversity is average local richness.
Order q = 0 alpha diversity calculated for our dataset returns a mean local richness (i.e., species counts) of ~30 taxa per sample across the entire data set.

```{r calc-alpha, message=FALSE, warning=FALSE}

# Here we use vegan::renyi to calculate Hill numbers
# If hill = FALSE, the function returns an entropy
# If hill = TRUE, the function returns the exponentiated
# entropy. In other words:
# exp(renyi entropy) = Hill number = "species equivalent"

# Note that for this function, the "scales" argument 
# determines the order of q used in the calculation

table_sample_by_taxon_density_wide %>%
  vegan::renyi(scales = 0, hill = TRUE) %>%
  mean()

```

#### Comparing alpha diversity calculated using different orders:

Order q = 1 alpha diversity returns mean number of "species equivalents" per sample in the data set. This approach incorporates evenness because when abundances are more even across taxa, taxa are weighted more equally toward counting as a "species equivalent". For example, if you have a sample with 100 individuals, spread across 10 species, and each species is represented by 10 individuals, the number of order q = 1 species equivalents will equal the richness (10).

Alternatively, if 90 of the 100 individuals in the sample are one species, and the other 10 individuals are spread across the other 9 species, there will only be 1.72 order q = 1 species equivalents, whereas, there are still 10 species in the sample.

```{r simulated-abg, message=FALSE, warning=FALSE}
# even distribution, orders q = 0 and q = 1 for 10 taxa
vegan::renyi(
  c(spp.a = 10, spp.b = 10, spp.c = 10, 
    spp.d = 10, spp.e = 10, spp.f = 10, 
    spp.g = 10, spp.h = 10, spp.i = 10, 
    spp.j = 10),
  hill = TRUE,
  scales = c(0, 1))

# uneven distribution, orders q = 0 and q = 1 for 10 taxa
vegan::renyi(
  c(spp.a = 90, spp.b = 2, spp.c = 1, 
    spp.d = 1, spp.e = 1, spp.f = 1, 
    spp.g = 1, spp.h = 1, spp.i = 1, 
    spp.j = 1),
  hill = TRUE,
  scales = c(0, 1)) 

```

## Comparing orders of q for NEON data

Let's compare the different orders q = 0, 1, and 2 measures of alpha diversity across the samples collected from ARIK, POSE, and MAYF.

```{r compare-q-NEON, message=FALSE, warning=FALSE}

# Nest data by siteID
data_nested_by_siteID <- table_sample_by_taxon_density_wide %>%
  tibble::rownames_to_column("sampleID") %>%
  left_join(table_sample_info %>% 
                select(sampleID, siteID)) %>%
  tibble::column_to_rownames("sampleID") %>%
  nest(data = -siteID)

data_nested_by_siteID$data[[1]] %>%
  vegan::renyi(scales = 0, hill = TRUE) %>%
  mean()

# apply the calculation by site for alpha diversity
# for each order of q
data_nested_by_siteID %>% mutate(
  alpha_q0 = purrr::map_dbl(
    .x = data,
    .f = ~ vegan::renyi(x = .,
                        hill = TRUE, 
                        scales = 0) %>% mean()),
  alpha_q1 = purrr::map_dbl(
    .x = data,
    .f = ~ vegan::renyi(x = .,
                        hill = TRUE, 
                        scales = 1) %>% mean()),
  alpha_q2 = purrr::map_dbl(
    .x = data,
    .f = ~ vegan::renyi(x = .,
                        hill = TRUE, 
                        scales = 2) %>% mean())
  )

# Note that POSE has the highest mean alpha diversity



# To calculate gamma diversity at the site scale,
# calculate the column means and then calculate 
# the renyi entropy and Hill number
# Here we are only calcuating order 
# q = 0 gamma diversity
data_nested_by_siteID %>% mutate(
  gamma_q0 = purrr::map_dbl(
    .x = data,
    .f = ~ vegan::renyi(x = colMeans(.),
                        hill = TRUE, 
                        scales = 0)))

# Note that POSE has the highest gamma diversity



# Now calculate alpha, beta, and gamma using orders 0 and 1 
# for each siteID
diversity_partitioning_results <- 
  data_nested_by_siteID %>% 
  mutate(
    n_samples = purrr::map_int(data, ~ nrow(.)),
    alpha_q0 = purrr::map_dbl(
      .x = data,
      .f = ~ vegan::renyi(x = .,
                          hill = TRUE, 
                          scales = 0) %>% mean()),
    alpha_q1 = purrr::map_dbl(
      .x = data,
      .f = ~ vegan::renyi(x = .,
                          hill = TRUE, 
                          scales = 1) %>% mean()),
    gamma_q0 = purrr::map_dbl(
      .x = data,
      .f = ~ vegan::renyi(x = colMeans(.),
                          hill = TRUE, 
                          scales = 0)),
    gamma_q1 = purrr::map_dbl(
      .x = data,
      .f = ~ vegan::renyi(x = colMeans(.),
                          hill = TRUE, 
                          scales = 1)),
    beta_q0 = gamma_q0 / alpha_q0,
    beta_q1 = gamma_q1 / alpha_q1)


diversity_partitioning_results %>% 
  select(-data) %>% as.data.frame() %>% print()

```

```{r local-regional-var, echo=F}

##########################################################
# local and regional scale variability
#########################################

# This doesn't make sense to do for macroinverts unless we can get 
# finer spatial resolution. 

# devtools::install_github("sokole/ltermetacommunities/ltmc")
# library(ltmc)
# 
# # Variability in aggregate biomass following
# # Wang and Loreau (2014)
# metacommunity_variability_results_POSE <- table_observation %>%
#   filter(!is.na(inv_dens), siteID == "POSE") %>%
#   ltmc::metacommunity_variability(
#     data_long = .,
#     time_step_col_name = "year",
#     site_id_col_name = "namedLocation",
#     taxon_id_col_name = "order",
#     biomass_col_name = "inv_dens",
#     variability_type = "agg")

```

## Using NMDS to ordinate samples

Finally, we will use Nonmetric Multidimensional Scaling (NMDS) to ordinate samples as shown below:

```{r NMDS, fig.cap=c("Two-dimension ordination plot of NMDS results. NMDS procedure resulted in a stress value of 0.21. Plot contains sampleIDs depicted in circles, and species, which have been labeled using the ordilabel function.","Ordination plots of community composition faceted by year. These plots were acheived by merging NMDS scores with sampleID information in order to plot samples by sampler type(shape) and siteID(color).","Ordination plots in community composition space faceted by siteID and habitat type. Points are colored to represent different years, as well as different shapes for sampler type. "), message=FALSE, warning=FALSE}


# create ordination using NMDS
my_nmds_result <- table_sample_by_taxon_density_wide %>% vegan::metaMDS()

# plot stress
my_nmds_result$stress
p1 <- vegan::ordiplot(my_nmds_result)
vegan::ordilabel(p1, "species")

# merge NMDS scores with sampleID information for plotting
nmds_scores <- my_nmds_result %>% 
  vegan::scores() %>%
  .[["sites"]] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("sampleID") %>%
  left_join(table_sample_info)


# # How I determined the outlier(s)
nmds_scores %>% arrange(desc(NMDS1)) %>% head()
nmds_scores %>% arrange(desc(NMDS1)) %>% tail()


# Plot samples in community composition space by year
nmds_scores %>%
  ggplot(aes(NMDS1, NMDS2, color = siteID, 
             shape = samplerType)) +
  geom_point() +
  facet_wrap(~ as.factor(year))

# Plot samples in community composition space
# facet by siteID and habitat type
# color by year
nmds_scores %>%
  ggplot(aes(NMDS1, NMDS2, color = as.factor(year), 
             shape = samplerType)) +
  geom_point() +
  facet_grid(habitatType ~ siteID, scales = "free")
  



```
