---
syncID: 457a5c01d051491297dc947939b4210f
title: "Using the NEON API in R"
description: "Tutorial for getting data from the NEON API, using R and the R package httr"
dateCreated:  2017-07-07
authors: [Claire K. Lunch]
contributors: [Christine Laney, Megan A. Jones]
estimatedTime: 1 - 1.5 hours
packagesLibraries: httr, jsonlite, devtools, geoNEON, neonDataStackR
topics: data-management, rep-sci
languagesTool: R, API
dataProduct:
code1: R/NEON-API/NEON-API-How-To
tutorialSeries: 
urlTitle: neon-api-usage
---


This is a tutorial in pulling data from the NEON API. The tutorial uses R 
and the R package httr, but the core information about the API is applicable 
to other languages and approaches.

There are 3 basic categories of NEON data:

1. Observational - Data collected by a human in the field, or in an analytical laboratory, e.g. beetle identification, foliar isotopes
1. Instrumentation - Data collected by an automated, streaming sensor, e.g. net radiation, soil carbon dioxide
1. Remote sensing - Data collected by the airborne observation platform, e.g. LIDAR, surface reflectance

This lesson covers the first two types of data. NEON remote sensing data are not currently available through the API.


<div id="ds-objectives" markdown="1">

# Objectives
After completing this activity, you will:

* Be able to pull observational, instrumentation, and geolocation data from the NEON API.
* Be able to transform API-accessed data from JSON to tabular format for analyses.

## Things You’ll Need To Complete This Tutorial
To complete this tutorial you will need the most current version of R and, 
preferably, RStudio loaded on your computer.

### Install R Packages

* **httr:** `install.packages("httr")`
* **jsonlite:** `install.packages("jsonlite")`
* **dplyr:** `install.packages("dplyr")`
* **devtools:** `install.packages("devtools")`
* **geoNEON:** `devtools::install_github("NEONScience/NEON-geolocation/geoNEON")`
* **neonDataStackR:** `devtools::install_github("NEONScience/NEON-utilities/neonDataStackR")`

### Additional Resources

* [Website for the NEON API](http://data.neonscience.org/data-api)
* [GitHub repository for the NEON API](https://github.com/NEONScience/neon-data-api)
* [ROpenSci wrapper for the NEON API](https://github.com/ropenscilabs/nneo) (not covered in this tutorial)

</div>

## Wondering what an API is?

If you are unfamiliar with the concept of an API, think of  an API as a 
‘middleperson' that provides a communication path for a software application 
to obtain information from a digital data source. APIs are becoming a very 
common means of sharing digital information. Many of the apps that you use on 
your computer or mobile device to produce maps, charts, reports, and other 
useful forms of information pull data from multiple sources using APIs. In 
the ecological and environmental sciences, many researchers use APIs to 
programmatically pull data into their analyses. (Quoted from the NEON Observatory
Blog story: 
<a href="http://www.neonscience.org/observatory/observatory-blog/api-data-availability-viewer-now-live-neon-data-portal" target ="_blank"> API and data availability viewer now live on the NEON data portal</a>.)

## Anatomy of an API call

An example API call: http://data.neonscience.org/api/v0/data/DP1.10003.001/WOOD/2015-07

This includes the base URL, endpoint, and target.

### Base URL: 
<span style="color:#A00606;font-weight:bold">http://data.neonscience.org/api/v0</span><span style="color:#C6C5C5">/data/DP1.10003.001/WOOD/2015-07</span>

Specifics are appended to this in order to get the data or metadata you're 
looking for, but all calls to this API will include the base URL. For the NEON 
API, this is http://data.neonscience.org/api/v0 --
not clickable, because the base URL by itself will take you nowhere!

### Endpoints: 
<span style="color:#C6C5C5">http://data.neonscience.org/api/v0</span><span style="color:#A00606;font-weight:bold">/data</span><span style="color:#C6C5C5">/DP1.10003.001/WOOD/2015-07</span>

What type of data or metadata are you looking for?

* **~/products**
  Information about one or all of NEON's data products

* **~/sites**
  Information about data availability at the site specified in the call

* **~/locations**
  Spatial data for the NEON locations specified in the call

* **~/data**
  Data! By product, site, and date (in monthly chunks).

### Targets:
<span style="color:#C6C5C5">http://data.neonscience.org/api/v0/data</span><span style="color:#A00606;font-weight:bold">/DP1.10003.001/WOOD/2015-07</span>

The specific data product, site, or location you want to get data for.

## Observational data
Which product do you want to get data for? Consult the <a href="http://data.neonscience.org/data-product-catalog" target="_blank">data product catalog</a>.

We'll pick Breeding landbird point counts, DP1.10003.001

First query the products endpoint of the API to find out which sites and dates 
have data available. In the products endpoint, the target is the numbered 
identifier for the data product:

``` {r os-avail-query}

# Load the necessary libaries
library(httr)
library(jsonlite)
library(dplyr, quietly=T)

# Request data using the GET function & the API call
req <- GET("http://data.neonscience.org/api/v0/products/DP1.10003.001")
req
```

The object returned from `GET()` has many layers of information. Entering the 
name of the object gives you some basic information about what you downloaded. 

The `content()` function returns the contents in the form of a highly nested 
list. This is typical of JSON-formatted data returned by APIs. 

``` {r os-query-contents}

# View requested data
req.content <- content(req, as="parsed")
req.content

```

To get a more accessible view of which sites have data for which months, you'll 
need to extract data from the nested list. There are a variety of ways to do this, 
in this tutorial we'll explore a couple of them. Here we'll use `fromJSON()`, in 
the jsonlite package, which doesn't fully flatten the nested list, but gets us 
the part we need. To use it, we need a text version of the content. Data this 
way is not as human readable but is readable by the `fromJSON()` function. 

``` {r os-query-fromJSON}
# make this JSON readable -> "text"
req.text <- content(req, as="text")

# Flatten data frame to see available data. 
avail <- fromJSON(req.text, simplifyDataFrame=T, flatten=T)
avail

```

The object contains a lot of information about the data product, including: 

* keywords under `$data$keywords`, 
* references for documentation under `$data$specs`, 
* data availability by site and month under `$data$siteCodes`, and 
* specific URLs for the API calls for each site and month under 
`$data$siteCodes$availableDataUrls`.

We need `$data$siteCodes` to tell us what we can download.
`$data$siteCodes$availableDataUrls` allows us to avoid writing the API 
calls ourselves in the next steps.

``` {r os-query-avail-data}

# get data availability list for the product
bird.urls <- unlist(avail$data$siteCodes$availableDataUrls)
bird.urls

```

These are the URLs showing us what files are available for each month where 
there are data. 

Let's look at the bird data from Woodworth (WOOD) site from July 2015. We can do 
this by using the above code but now specifying which site/date we want using 
the `grep()` function. 

Note, that if there was only one month of data from a site, you can leave off the
date in the function. If you want date from more than one site/month you need 
to iterate this code, GET fails if you give it more than one URL. 

``` {r os-query-bird-data-urls}
# get data availability for WOOD July 2015
brd <- GET(bird.urls[grep("WOOD/2015-07", bird.urls)])
brd.files <- fromJSON(content(brd, as="text"))

# view just the available data files 
brd.files$data$files

```

In this output, `name` and `url` are key fields. It provides us with the names 
of the files available for this site and month, and URLs where we can get the 
files. We'll use the file names to pick which ones we want.

The available files include both **data** and **metadata**, and both the **basic**
and **expanded** data packages. Typically the expanded package includes additional 
quality or uncertainty data, either in additional files or additional fields 
than in the basic files. Basic and expanded data packages are available for 
most NEON data products (some only have basic). Metadata are described by file 
name below.

The format for most of the file names are:

**NEON.[domain number].[site code].[data product ID].[file-specific name].
[date of file creation] **

Some files omit the domain and site, since they're not specific to a 
location, like the data product readme. The date of file creation uses the 
ISO6801 format, in this case 20170720T182547Z, and can be used to determine 
whether data have been updated since the last time you downloaded.

Available files in our query for July 2015 at Woodworth are all of the following
(leaving off the initial NEON.D09.WOOD.10003.001): 

* **~.2015-07.expanded.20170720T182547Z.zip:** zip of all files in the expanded 
package

* **~.brd_countdata.2015-07.expanded.20170720T182547Z.csv:** count data table, 
expanded package version: counts of birds at each point

* **~.brd_perpoint.2015-07.expanded.20170720T182547Z.csv:** point data table, 
expanded package version: metadata at each observation point

* **NEON.Bird Conservancy of the Rockies.brd_personnel.csv:** personnel data 
table, accuracy scores for bird observers

* **~.2015-07.basic.20170720T182547Z.zip:** zip of all files in the basic package

* **~.brd_countdata.2015-07.basic.20170720T182547Z.csv:** count data table, 
basic package version: counts of birds at each point

* **~.brd_perpoint.2015-07.basic.20170720T182547Z.csv:** point data table, 
basic package version: metadata at each observation point

* **NEON.DP1.10003.001_readme.txt:** readme for the data product (not specific 
to dates or location). Appears twice in the list, since it's in both the basic 
and expanded package

* **~.20150101-20160613.xml:** Ecological Metadata Language (EML) file. Appears 
twice in the list, since it's in both the basic and expanded package
  
* **~.validation.20170720T182547Z.csv:** validation file for the data product, 
lists input data and data entry rules. Appears twice in the list, since it's in 
both the basic and expanded package
  
* **~.variables.20170720T182547Z.csv:** variables file for the data product, 
lists data fields in downloaded tables. Appears twice in the list, since it's 
in both the basic and expanded package


We'll get the data tables for the point data and count data in the basic 
package. The list of files doesn't return in the same order every time, so we 
won't use position in the list to select. Plus, we want code we can re-use 
when getting data from other sites and other months. So we select files 
based on the data table name and the package name.


``` {r os-get-bird-data}

# Get both files
brd.count <- read.delim(brd.files$data$files$url
                        [intersect(grep("countdata", brd.files$data$files$name),
                                    grep("basic", brd.files$data$files$name))], sep=",")

brd.point <- read.delim(brd.files$data$files$url
                        [intersect(grep("perpoint", brd.files$data$files$name),
                                    grep("basic", brd.files$data$files$name))], sep=",")

```

Now we have the data and can access it in R. Just to show that the files we 
pulled have actual data in them, let's make a quick graphic showing 

``` {r os-plot-bird-data}
# Cluster by species
clusterBySp <- brd.count %>% 
	group_by(scientificName) %>% 
  summarize(total=sum(clusterSize))

# Reorder so list is ordered most to least abundance
clusterBySp <- clusterBySp[order(clusterBySp$total, decreasing=T),]

# Plot
barplot(clusterBySp$total, names.arg=clusterBySp$scientificName, 
        ylab="Total", cex.names=0.5, las=2)

```

Wow! There are lots of *Agelaius phoeniceus* (Red-winged Blackbirds) at WOOD in July. 


## Instrumentation data

The process is essentially the same for sensor data. We'll do the same series of 
queries for Soil Temperature, DP1.00041.001. Let's use data from Moab in March 
2017 this time.

```{r soil-data}
# Request soil temperature data
req.soil <- GET("http://data.neonscience.org/api/v0/products/DP1.00041.001")

# make this JSON readable
# Note how we've change this from two commands into one here
avail.soil <- fromJSON(content(req.soil, as="text"), simplifyDataFrame=T, flatten=T)

# get data availability list for the product
temp.urls <- unlist(avail.soil$data$siteCodes$availableDataUrls)

# get data availability from location/date of interest
tmp <- GET(temp.urls[grep("MOAB/2017-03", temp.urls)])
tmp.files <- fromJSON(content(tmp, as="text"))
tmp.files$data$files$name

```

These file names start and end the same way as the observational files, but the 
middle is a little more cryptic. The structure from beginning to end is: 

**NEON.[domain number].[site code].[data product ID].00000.
[soil plot number].[depth].[averaging interval].[data table name].
[year]-[month].[data package].[date of file creation]**

So **NEON.D13.MOAB.DP1.00041.001.00000.002.504.030.ST_30_minute.
2017-03.basic.20170804T063725Z.csv** is the:
* NEON (`NEON.`)
* Domain 13 (`.D13.`)
* Moab field site (`.MOAB.`) 
* soil temperature data (`.DP1.00041.001.`)
* (internal NEON identifier, always 00000 in published data) (`.00000.`)
* collected in Soil Plot 2, (`.002.`)
* at the 4th depth below the surface (`.504.`)
* and reported as a 30-minute mean of (`.030.` and `.ST_30_minute.`)
* only for the period of March 2017 (`.2017-03.`)
* and provided in a basic data package (`.basic.`)
* published on Aug 4, 2017 at 06:37:25 GMT (`.20170804T063725Z.`).

More information about interpreting file names can be found in the readme that 
accompanies each download. 

Note: NEON instrumentation data products are currently (Fall/Winter 2017) being 
reprocessed on a rolling basis, and as each product is re-deployed, the files will 
begin to appear in the format described here. Products that haven't yet 
been reprocessed will return file names like the above, but will be missing the 
year, month, package, and date of creation specifications.

Let's get data (and the URL) for only the plot and depth described above by selecting 
`002.504.030` and the word `basic` in the file name.

Go get it:

``` {r os-get-soil-data}

soil.temp <- read.delim(tmp.files$data$files$url
                        [intersect(grep("002.504.030", tmp.files$data$files$name),
                                   grep("basic", tmp.files$data$files$name))], sep=",")

```

Now we have the data and can use it to conduct our analyses. To show that we do 
in fact have the data let's plot the mean soil temperature by date. 


``` {r os-plot-soil-data}
# plot temp ~ date
plot(soil.temp$soilTempMean~soil.temp$startDateTime, pch=".", xlab="Date", ylab="T")

```

As, we'd expect we see daily fluctuation in soil temperature. 

## Geolocation data

You may have noticed some of the spatial data referenced above are a bit vague, 
e.g. "soil plot 2, 4th depth below the surface."

How to get spatial data and what to do with it depends on which type of 
data you're working with.

#### Instrumentation data (both aquatic and terrestrial)
Stay tuned - spatial data for instruments are in the process of entry into 
the NEON database.

#### Observational data - Aquatic
Latitude, longitude, elevation, and associated uncertainties are included in 
data downloads. Most products also include an "additional coordinate uncertainty" 
that should be added to the provided uncertainty. Additional spatial data, such 
as northing and easting, can be downloaded from the API.

#### Observational data - Terrestrial
Latitude, longitude, elevation, and associated uncertainties are included in 
data downloads. These are the coordinates and uncertainty of the sampling plot; 
for many protocols it is possible to calculate a more precise location. 
Instructions for doing this are in the respective data product user guides, and 
code is in the `geoNEON` package on GitHub.

### Querying a single named location
Let's look at the named locations in the bird data we downloaded above. To do this, 
look for the field called `namedLocation`, which is present in all observational 
data products, both aquatic and terrestrial.

```{r get-bird-NLs}
# view named location
head(brd.point$namedLocation)

```

Here we see the first six entries in the `namedLocation` column which tells us
the names of the Terrestrial Observation plots where the bird surveys were 
conducted. 

We can query the locations endpoint of the API for the first named location, 
`WOOD_013.birdGrid.brd`. 

```{r brd-ex-NL}
# location data 
req.loc <- GET("http://data.neonscience.org/api/v0/locations/WOOD_013.birdGrid.brd")

# make this JSON readable
brd.WOOD_013 <- fromJSON(content(req.loc, as="text"))
brd.WOOD_013

```

Note spatial information under `$data$[nameOfCorrdinate]` and under 
`$data$locationProperties`. Also note `$data$locationChildren`: these are the 
finer scale locations that can be used to calculate precise spatial data for 
bird observations.

For convenience, we'll use the `geoNEON` package to make the calculations. 
First we'll use `def.extr.geo.os()` to get the additional spatial information 
available through the API, and look at the spatial resolution available in the 
initial download:

```{r brd-extr-NL}

# load the geoNEON package
library(geoNEON)

# extract the spatial data
brd.point.loc <- def.extr.geo.os(brd.point)

# plot bird point locations 
# note that decimal degrees is also an option in the data
symbols(brd.point.loc$api.easting, brd.point.loc$api.northing, 
        circles=brd.point.loc$coordinateUncertainty, 
        xlab="Easting", ylab="Northing", tck=0.01, inches=F)

```

And use `def.calc.geo.os()` to calculate the point locations of observations.

```{r brd-calc-NL}

brd.point.pt <- def.calc.geo.os(brd.point, "brd_perpoint")


# plot bird point locations 
# note that decimal degrees is also an option in the data
symbols(brd.point.pt$api.easting, brd.point.pt$api.northing, 
        circles=brd.point.pt$adjCoordinateUncertainty, 
        xlab="Easting", ylab="Northing", tck=0.01, inches=F)

```

Now you can see the individual points where the respective point counts were 
located. 

## Coming soon

Remote sensing data were added to the API recently (late September 2017). This 
tutorial will be expanded to include a section on remote sensing data. In the 
meantime, we can give you the guidance that all the query structure is the same 
for remote sensing data, but once you get to the files themselves, different 
handling will be needed, since they are not tabular data.

At the top of this tutorial, we installed the `neonDataStackR` package. 
This is a custom R package that stacks the monthly files provided by 
the NEON data portal into a single continuous file for each type of 
data table in the download. It currently handles files downloaded from 
the data portal, but not files pulled from the API. That functionality 
will be added soon!

For a guide to using `neonDataStackR` on data downloaded from the portal, 
look <a href="{{ site.baseurl }}/neonDataStackR" target="_blank">here</a>.
