---
syncID: 03868e86e4d34d1fa88caacaddf95a57
title: "Plot Continuous & Discrete Data Together"
description: "This tutorial discusses ways to plot plant phenology (discrete time series) and single-aspirated temperature ((near-)continuous time series) together."
dateCreated: 2017-08-01
authors: Lee Stanish, Megan A. Jones, Natalie Robinson
contributors: Katie Jones, Cody Flagg, Felipe Sanchez
estimatedTime:
packagesLibraries: dplyr, ggplot2, lubridate, gridExtra, scales
topics: timeseries, meteorology, phenology, organisms
languagesTool: R
dataProduct: NEON.DP1.10055, 
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/biodiversity/neon-phenology-temp/03-plot-discrete-continuous-data-pheno-temp/03-plot-discrete-continuous-data-pheno-temp.R
tutorialSeries:  neon-pheno-temp-series
urlTitle: neon-pheno-temp-plots-r
---

This tutorial discusses ways to plot plant phenology (discrete time
series) and single-aspirated temperature (continuous time series) together.
It uses data frames created in the first two parts of this series,
<a href = "https://www.neonscience.org/osis-pheno-temp-series" target = "_blank">*Work with NEON OS & IS Data - Plant Phenology & Temperature*</a>. 
If you have not completed these tutorials, please download the dataset below.

<div id="ds-objectives" markdown="1">

## Objectives
After completing this tutorial, you will be able to:

 * plot multiple figures together with grid.arrange()
 * plot only a subset of dates

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and, preferably, `RStudio` loaded
on your computer to complete this tutorial.

### Install R Packages

* **neonUtilities:** `install.packages("neonUtilities")`
* **ggplot2:** `install.packages("ggplot2")`
* **dplyr:** `install.packages("dplyr")`
* **gridExtra:** `install.packages("gridExtra")`

<a href="https://www.neonscience.org/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

### Download Data 

This tutorial is designed to have you download data directly from the NEON
portal API using the neonUtilities package. However, you can also directly 
download this data, prepackaged, from FigShare. This data set includes all the 
files needed for the *Work with NEON OS & IS Data - Plant Phenology & Temperature* 
tutorial series. The data are in the format you would receive if downloading them
using the `zipsByProduct()` function in the neonUtilities package. 

<a href = "https://ndownloader.figshare.com/files/22775042">Direct Download: **NEON Phenology & Temp Time Series Teaching Data Subset (v2 - 2017-2019 data)** (12 MB)</a>

</div>


To start, we need to set up our R environment. If you're continuing from the 
previous tutorial in this series, you'll only need to load the new packages.

```{r setup-env}
# Install needed package (only uncomment & run if not already installed)
#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("scales")

# Load required libraries
library(ggplot2)
library(dplyr)
library(gridExtra)
library(scales)

options(stringsAsFactors=F) #keep strings as character type not factors

# set working directory to ensure R can find the file we wish to import and where
# we want to save our files. Be sure to move the download into your working directory!
wd <- "~/Documents/data/" # Change this to match your local environment
setwd(wd)
```

If you don't already have the R objects, `temp_day` and `phe_1sp_2018`, loaded
you'll need to load and format those data. If you do, you can skip this code. 

```{r load-data}
# Read in data -> if in series this is unnecessary
temp_day <- read.csv(paste0(wd,'NEON-pheno-temp-timeseries/NEONsaat_daily_SCBI_2018.csv'))

phe_1sp_2018 <- read.csv(paste0(wd,'NEON-pheno-temp-timeseries/NEONpheno_LITU_Leaves_SCBI_2018.csv'))

# Convert dates
temp_day$Date <- as.Date(temp_day$Date)
# use dateStat - the date the phenophase status was recorded
phe_1sp_2018$dateStat <- as.Date(phe_1sp_2018$dateStat)

```

## Separate Plots, Same Panel

In this dataset, we have phenology and temperature data from the Smithsonian
Conservation Biology Institute (SCBI) NEON field site. There are a variety of ways 
we may want to look at this data, including aggregated at the site level, by
a single plot, or viewing all plots at the same time but in separate plots. In 
the *Work With NEON's Plant Phenology Data* and the 
*Work with NEON's Single-Aspirated Air Temperature Data* tutorials, we created 
separate plots of the number of individuals who had leaves at different times 
of the year and the temperature in 2018.

However, plot the data next to each other to aid comparisons. The `grid.arrange()` 
function from the gridExtra package can help us do this. 

```{r stacked-plots, fig.cap="One graphic showing two plots arranged vertically by using the grid.arrange function form the gridExtra package. The top plot shows a bar plot of the counts of Liriodendrum tulipifera (LITU) individuals at the Smithsonian Conservation Biology Institute (SCBI) for the year 2018. The bottom plot shows a scatter plot of daily maximum temperatures(of 30 minute interval means) for the year 2018 at the Smithsonian Conservation Biology Institute (SCBI)." }
# first, create one plot 
phenoPlot <- ggplot(phe_1sp_2018, aes(dateStat, countYes)) +
    geom_bar(stat="identity", na.rm = TRUE) +
    ggtitle("Total Individuals in Leaf") +
    xlab("") + ylab("Number of Individuals")

# create second plot of interest
tempPlot_dayMax <- ggplot(temp_day, aes(Date, dayMax)) +
    geom_point() +
    ggtitle("Daily Max Air Temperature") +
    xlab("Date") + ylab("Temp (C)")

# Then arrange the plots - this can be done with >2 plots as well.
grid.arrange(phenoPlot, tempPlot_dayMax) 

```

Now, we can see both plots in the same window.  But, hmmm... the x-axis on both 
plots is kinda wonky. We want the same spacing in the scale across the year (e.g.,
July in one should line up with July in the other) plus we want the dates to 
display in the same format(e.g. 2016-07 vs. Jul vs Jul 2018).

## Format Dates in Axis Labels
The date format parameter can be adjusted with `scale_x_date`. Let's format the x-axis
ticks so they read "month" (`%b`) in both graphs. We will use the syntax:

`scale_x_date(labels=date_format("%b"")`

Rather than re-coding the entire plot, we can add the `scale_x_date` element
to the plot object `phenoPlot` we just created. 


<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i> **Data Tip:** 

1) You can type `?strptime` into the R 
console to find a list of date format conversion specifications (e.g. %b = month).
Type `scale_x_date` for a list of parameters that allow you to format dates 
on the x-axis.

2) If you are working with a date & time
class (e.g. POSIXct), you can use `scale_x_datetime` instead of `scale_x_date`.
</div>


```{r format-x-axis-labels, fig.cap="Graphic showing the arranged plots created in the previous step, with the x-axis formatted to only read 'month' in both plots. However, it is important to note that this step only partially fixes the problem. The plots still have different ranges on the x-axis, which makes it harder to see trends. The top plot shows a bar plot of the counts of Liriodendrum tulipifera (LITU) individuals at the Smithsonian Conservation Biology Institute (SCBI) for the year 2018. The bottom plot shows a scatter plot of daily maximum temperatures(of 30 minute interval means) for the year 2018 at the Smithsonian Conservation Biology Institute (SCBI)." }
# format x-axis: dates
phenoPlot <- phenoPlot + 
  (scale_x_date(breaks = date_breaks("1 month"), labels = date_format("%b")))

tempPlot_dayMax <- tempPlot_dayMax +
  (scale_x_date(breaks = date_breaks("1 month"), labels = date_format("%b")))

# New plot. 
grid.arrange(phenoPlot, tempPlot_dayMax) 

```

But this only solves one of the problems, we still have a different range on the
x-axis which makes it harder to see trends. 

## Align data sets with different start dates  

Now let's work to align the values on the x-axis. We can do this in two ways, 
1) setting the x-axis to have the same date range or 2) by filtering the dataset 
itself to only include the overlapping data. Depending on what you are trying to 
demonstrate and if you're doing additional analyses and want only the overlapping 
data, you may prefer one over the other. Let's try both. 

### Set range of x-axis
Alternatively, we can set the x-axis range for both plots by adding the `limits` 
parameter to the `scale_x_date()` function. 

```{r set-x-axis, fig.cap="Graphic showing the arranged plots created in the previous step, with the x-axis formatted to only read 'month', and scaled so they align with each other. This is achieved by adding the limits parameter to the scale_x_date function in the ggplot call. The top plot shows a bar plot of the counts of Liriodendrum tulipifera (LITU) individuals at the Smithsonian Conservation Biology Institute (SCBI) for the year 2018. The bottom plot shows a scatter plot of daily maximum temperatures(of 30 minute interval means) for the year 2018 at the Smithsonian Conservation Biology Institute (SCBI)." }
# first, lets recreate the full plot and add in the 
phenoPlot_setX <- ggplot(phe_1sp_2018, aes(dateStat, countYes)) +
    geom_bar(stat="identity", na.rm = TRUE) +
    ggtitle("Total Individuals in Leaf") +
    xlab("") + ylab("Number of Individuals") +
    scale_x_date(breaks = date_breaks("1 month"), 
                  labels = date_format("%b"),
                  limits = as.Date(c('2018-01-01','2018-12-31')))

# create second plot of interest
tempPlot_dayMax_setX <- ggplot(temp_day, aes(Date, dayMax)) +
    geom_point() +
    ggtitle("Daily Max Air Temperature") +
    xlab("Date") + ylab("Temp (C)") +
    scale_x_date(date_breaks = "1 month", 
                 labels=date_format("%b"),
                  limits = as.Date(c('2018-01-01','2018-12-31')))

# Plot
grid.arrange(phenoPlot_setX, tempPlot_dayMax_setX) 

```

Now we can really see the pattern over the full year. This emphasizes the point
that during much of the late fall, winter, and early spring none of the trees 
have leaves on them (or that data were not collected - this plot would not 
distinguish between the two). 

### Subset one data set to match other
Alternatively, we can simply filter the dataset with the larger date range so 
the we only plot the data from the overlapping dates. 

```{r align-datasets-replot, fig.cap="Graphic of the arranged plots created in the previous steps with only the data that overlap. This was achieved by filtering the daily max temperature data by the observation date in the total individuals in Leaf dataset. The top plot shows a bar plot of the counts of Liriodendrum tulipifera (LITU) individuals at the Smithsonian Conservation Biology Institute (SCBI) for the year 2018. The bottom plot shows a scatter plot of daily maximum temperatures(of 30 minute interval means) for the year 2018 at the Smithsonian Conservation Biology Institute (SCBI)." }
# filter to only having overlapping data
temp_day_filt <- filter(temp_day, Date >= min(phe_1sp_2018$date) & 
                         Date <= max(phe_1sp_2018$date))

# Check 
range(phe_1sp_2018$date)
range(temp_day_filt$Date)

#plot again
tempPlot_dayMaxFiltered <- ggplot(temp_day_filt, aes(Date, dayMax)) +
    geom_point() +
    scale_x_date(breaks = date_breaks("months"), labels = date_format("%b")) +
    ggtitle("Daily Max Air Temperature") +
    xlab("Date") + ylab("Temp (C)")


grid.arrange(phenoPlot, tempPlot_dayMaxFiltered)

```

With this plot, we really look at the area of overlap in the plotted data (but 
this does cut out the time where the data are collected but not plotted). 

## Same plot with two Y-axes

What about layering these plots and having two y-axes (right and left) that have
the different scale bars?

Some argue that you should not do this as it can distort what is actually going 
on with the data. The author of the ggplot2 package is one of these individuals. 
Therefore, you cannot use `ggplot()` to create a single plot with multiple y-axis 
scales. You can read his own discussion of the topic on this 
<a href="https://stackoverflow.com/questions/3099219/plot-with-2-y-axes-one-y-axis-on-the-left-and-another-y-axis-on-the-right/3101876#3101876" target="_blank" > StackOverflow post</a>.

---

However, individuals have found work arounds for these plots. The code below
is provided as a demonstration of this capability. Note, by showing this code 
here, we don't necessarily endorse having plots with two y-axes.

This code is adapted from code by <a href="heareresearch.blogspot.com/2014/10/10-30-2014-dual-y-axis-graph-ggplot2_30.html" target="_blank">Jake Heare</a>. 

```{r two-y-axes-ggplot}

# Source: http://heareresearch.blogspot.com/2014/10/10-30-2014-dual-y-axis-graph-ggplot2_30.html

# Additional packages needed
library(gtable)
library(grid)


# Plot 1: Pheno data as bars, temp as scatter
grid.newpage()
phenoPlot_2 <- ggplot(phe_1sp_2018, aes(dateStat, countYes)) +
  geom_bar(stat="identity", na.rm = TRUE) +
  scale_x_date(breaks = date_breaks("1 month"), labels = date_format("%b")) +
  ggtitle("Total Individuals in Leaf vs. Temp (C)") +
  xlab(" ") + ylab("Number of Individuals") +
  theme_bw()+
  theme(legend.justification=c(0,1),
        legend.position=c(0,1),
        plot.title=element_text(size=25,vjust=1),
        axis.text.x=element_text(size=20),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=20),
        axis.title.y=element_text(size=20))


tempPlot_dayMax_corr_2 <- ggplot() +
  geom_point(data = temp_day_filt, aes(Date, dayMax),color="red") +
  scale_x_date(breaks = date_breaks("months"), labels = date_format("%b")) +
  xlab("") + ylab("Temp (C)") +
  theme_bw() %+replace% 
  theme(panel.background = element_rect(fill = NA),
        panel.grid.major.x=element_blank(),
        panel.grid.minor.x=element_blank(),
        panel.grid.major.y=element_blank(),
        panel.grid.minor.y=element_blank(),
        axis.text.y=element_text(size=20,color="red"),
        axis.title.y=element_text(size=20))

g1<-ggplot_gtable(ggplot_build(phenoPlot_2))
g2<-ggplot_gtable(ggplot_build(tempPlot_dayMax_corr_2))

pp<-c(subset(g1$layout,name=="panel",se=t:r))
g<-gtable_add_grob(g1, g2$grobs[[which(g2$layout$name=="panel")]],pp$t,pp$l,pp$b,pp$l)

ia<-which(g2$layout$name=="axis-l")
ga <- g2$grobs[[ia]]
ax <- ga$children[[2]]
ax$widths <- rev(ax$widths)
ax$grobs <- rev(ax$grobs)
ax$grobs[[1]]$x <- ax$grobs[[1]]$x - unit(1, "npc") + unit(0.15, "cm")
g <- gtable_add_cols(g, g2$widths[g2$layout[ia, ]$l], length(g$widths) - 1)
g <- gtable_add_grob(g, ax, pp$t, length(g$widths) - 1, pp$b)

grid.draw(g)


# Plot 2: Both pheno data and temp data as line graphs
grid.newpage()
phenoPlot_3 <- ggplot(phe_1sp_2018, aes(dateStat, countYes)) +
  geom_line(na.rm = TRUE) +
  scale_x_date(breaks = date_breaks("months"), labels = date_format("%b")) +
  ggtitle("Total Individuals in Leaf vs. Temp (C)") +
  xlab("Date") + ylab("Number of Individuals") +
  theme_bw()+
  theme(legend.justification=c(0,1),
        legend.position=c(0,1),
        plot.title=element_text(size=25,vjust=1),
        axis.text.x=element_text(size=20),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=20),
        axis.title.y=element_text(size=20))

tempPlot_dayMax_corr_3 <- ggplot() +
  geom_line(data = temp_day_filt, aes(Date, dayMax),color="red") +
  scale_x_date(breaks = date_breaks("months"), labels = date_format("%b")) +
  xlab("") + ylab("Temp (C)") +
  theme_bw() %+replace% 
  theme(panel.background = element_rect(fill = NA),
        panel.grid.major.x=element_blank(),
        panel.grid.minor.x=element_blank(),
        panel.grid.major.y=element_blank(),
        panel.grid.minor.y=element_blank(),
        axis.text.y=element_text(size=20,color="red"),
        axis.title.y=element_text(size=20))

g1<-ggplot_gtable(ggplot_build(phenoPlot_3))
g2<-ggplot_gtable(ggplot_build(tempPlot_dayMax_corr_3))

pp<-c(subset(g1$layout,name=="panel",se=t:r))
g<-gtable_add_grob(g1, g2$grobs[[which(g2$layout$name=="panel")]],pp$t,pp$l,pp$b,pp$l)

ia<-which(g2$layout$name=="axis-l")
ga <- g2$grobs[[ia]]
ax <- ga$children[[2]]
ax$widths <- rev(ax$widths)
ax$grobs <- rev(ax$grobs)
ax$grobs[[1]]$x <- ax$grobs[[1]]$x - unit(1, "npc") + unit(0.15, "cm")
g <- gtable_add_cols(g, g2$widths[g2$layout[ia, ]$l], length(g$widths) - 1)
g <- gtable_add_grob(g, ax, pp$t, length(g$widths) - 1, pp$b)

grid.draw(g)

```

