---
layout: post
title: "Clean & Combine Carabid Beetle Data from the NEON Data Portal"
date:  2016-12-16
authors: [Katie LeVan]
contributors: [Megan A. Jones] 
dateCreated: 2016-12-16
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
description: "This tutorial explains how to clean and combine the three related NEON carabid beetle datatables."
image:
  feature: TeachingModules.jpg
  credit: A National Ecological Observatory Network (NEON) - Teaching Module
  creditlink: http://www.neonscience.org
permalink: /R/carabid-clean-data
code1: carabid-beetle-data/Beetle-Data-Clean-Portal-Data.R
code2: carabid-beetle-data/carabid-NEON-data-cleanup.R
comments: false
---

{% include _toc.html %}

This tutorial focuses on cleaning and combining the three tables from the 
NEON Data Portal related to Carabid Beetle data. This tutorial accompanies the 
<a href="{{ site.basurl }}/R/carabid-explore-data" target="_blank"> *Work with NEON Carabid Beetle Data* tutorial</a> 
which explains the structure of the data and explains how to work with the 
combined data. 

**R Skill Level:** Introduction - you've got the basics of `R` down and 
understand the general structure of tabular data.

<div id="objectives" markdown="1">

# Objectives
After completing this tutorial, you will:

* ADD

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and, preferably, RStudio loaded on
your computer to complete this tutorial.

### Install R Packages

These R packages will be used in the tutorial below. Please make sure they are 
installed prior to starting the tutorial. 
 
* **dplyr:** `install.packages("dplyr")`
* **plyr:** `install.packages("plyr")`


### Download The Data
**NOTE: eventually turn these into teaching data subsets with others, then change to download buttons**
You can download cleaned data files [here](//github.com/klevan/carabid-workshop/blob/master/data/zip%20files/cleaned-Data.zip), 
NOAA weather data for each site [here](//github.com/klevan/carabid-workshop/blob/master/data/NOAA%20weather%20data%20for%202014.csv), 
NEON map shp files [here](//github.com/klevan/carabid-workshop/blob/master/data/zip%20files/map%20data.zip) 
and the script we will be modifying [here](//github.com/klevan/carabid-workshop/blob/master/code/data-analysis.R). 
</div>

## NEON Carabid Beetle Data

The carabid beetle data on the NEON Data Portal are divided by site and into 
three table per site. 

NEON carabid data are stored in three tables: 
 * field collection data, 
 * sorting data and 
 * identification and pinning data. 

For this tutorial we focus on the 2014 data and the 13 sites
for which data is available in that year. To look at all this data requires 
downloading 39 files (one field, sorting and pinning data for each site) and 
combining the datasheets across all sites for each type of data (i.e., field, 
sorting and pinning). 

NEON provides several documents with information about the Carabid beetle protocal & 
data collection. It is highly recommended you are familiar with the
<a href="http://data.neonscience.org/data-products/DP1.10022.001" 
target="_blank">data product documents </a>
prior to using NEON carabid beetle data for your research. 

We'll explore these three tables and then combine them into a single clean
table for use with analysis. 

First, set up the R environment. 

``` {r load-data}
# Load packages required for entire script. 
library(plyr)      # move/manipulate data
library(dplyr)     # move/manipulate data

# set working directory to ensure R can find the file we wish to import
# set to the `carabid-2014-NEON` directory
#setwd("working-dir-path-here")

```


### Field Collection Data Table

Read in the field collection data table. 

``` {r field-table}
# read in the data
car.field.HARV <- read.csv(
      file="rawData_portalDownloads/NEON.D01.HARV.DP1.10022.001.bet_fielddata.csv",
      stringsAsFactors = FALSE
      )

# view structure of the data
str(car.field.HARV)

```

This table contains information related to:

* the location of each trap in the first 12 fields (i.e., latitude, longitude, 
NLCD class)
* metadata about the sampling event, including
   + `setDate`: when traps were set
   + `collectDate`: the date of trap collection
   + `daysOfTrapping`: the number of days a given trap was in the field
   + `samplingProtocol`: the document number of the sampling protocol used. These
   can be found in the 
<a href="http://data.neonscience.org/documents" target="_blank"> NEON Documents Library</a>. 
* metadata about the quality of the data 

Unique collection events have a unique `sampleID` with the format = 
`plotID.trapID.collectDate`. 

### Sorting Data Table


``` {r sort-table}
# read in sorting data
car.sort.HARV <- read.csv(
      file="rawData_portalDownloads/NEON.D01.HARV.DP1.10022.001.bet_sorting.csv",
      stringsAsFactors = FALSE
      )

# view structure of the data
str(car.sort.HARV)

```

This table contains information about what specimens were found in traps. Sample 
types include:

* vertebrate bycatch (herps = `vert bycatch herp`; mammals = `vert bycatch mam`), 
* non-Carabidae invertebrate bycatch (`invert bycatch`), 
* and all information about carabids. 

Older data (2014 and prior) divide carabid data into two categories: 

* `common carabid`, for which all taxonomic and abundance data is listed in the 
sorting dataset, and 
* `other carabid`, a catch-all category indicating that the relevant taxonomic 
and abundance information for specimens with that `sampleID` is provided in the 
pinning dataset. 

Data after 2014 do not separate between these two group. **Katie, is this correct. My notes from when I asked you weren't very clear**

Unique records have a unique `associatedSampleID` (format = 
`plotID.trapID.collectDate.taxonID.tubeID` ). For records where taxonomic 
information is unknown the following shorthand codes are used in place of the 
`taxonID` based on `sampleType`:

* `ib` for invertebrate bycatch, 
* `vb` for vertebrate bycatch, 
* `cc` for common carabid, 
* `oc` for other carabid.

### Pinning Data Table

``` {r field-table}
# read in data
car.pin.HARV <- read.csv(
      file="rawData_portalDownloads/NEON.D01.HARV.DP1.10022.001.bet_IDandpinning.csv",
      stringsAsFactors = FALSE
      )

# view structure of the data
str(car.pin.HARV)

```

The pinning table contains information about all carabids that were pinned or 
pointed. Each sample in the pinning dataset has a unique identifier 
(`individualID`). The data in this table include location of the sample (but 
fewer details than the field table) as well as the information about when and how
it was identified and pinned/pointed.


### IDs to Connect Tables

To link the Field table to the sorting table you would use XXXX? 
The information in the pinning table are traceable to the 
sorting data via the `sampleID`. Every `sampleID` in the pinning data matches an
`associatedSampleID` in the sorting dataset. 

## Combine the Data Tables

``` {r combine-function}
# set strings as factors as false throughout
options(stringsAsFactors = FALSE) 

# this function allows for XXXX.
multipleCombine <- function(input, ply = llply){
  require(plyr)
  require(dplyr)
  ply(input, function(x){
    t <- read.table(x, header=TRUE, sep=",",stringsAsFactors = FALSE) # read the csv
    t1 <- rbind(t) # rbind it to a temporary variable
    return(t1) # return the full variable
  }
  )
}

```

We want to set paths that direct to all the data. 

**Katie, I'm thinking we could pull the USGS weather data from this lesson. Thoughts?**

``` {r auto-paths}

# setting paths
pathToData <- paste(path,'rawData_portalDownloads',sep='/')
pathToWeatherData <- paste(path,'weatherData',sep='/')
setwd(pathToData)

```

Next we want to create objects that will be used as suffixes for each table: 

* `field`: Data that is recorded in the field when samples are recovered from 
pitfall traps
* `sort`: Data recorded in the lab during an initial sort; data on vertebrate 
and invertebrate bycatch is recorded in this table; data on carabids that were 
not pinned is recorded here
* `pin`:  Data on identified carabids that were pinned; many of these Carabids 
will eventually be available for loan from archival facilities
* `weather`: the 

**Katie, what is "bet", is it short for beetle? If so, I might change to `car` as
it is consistent w/ the Explore lesson?**.

``` {r table-suffix}

# csv files, by type, where the combined data will get added.

field <- 'bet_fielddata.csv' 
sort <- 'bet_sorting.csv' 
pin <- 'bet_IDandpinning.csv'
weather <- 'NOAA weather data for 2014.csv'
```

Then we create a list of the files in each directory. 

**Katie -- what does "module" mean in the notes here? 

``` {r file-function-neon}
# This function will grab the file paths of all the data in the individual directory

fileList <- list.files(pathToData, full.names=TRUE) # list all the files, full.names=TRUE is necessary for ldplay/lapply to work below
field <- fileList[grep(field,fileList)] # subset to just the ones in your module, using prefix, if needed
sort <- fileList[grep(sort,fileList)] # subset to just the ones in your module, using prefix, if needed
pin <- fileList[grep(pin,fileList)] # subset to just the ones in your module, using prefix, if needed
```

Now we can do the same thing but with the weather data. Here we don't want to 
bring in all the weather data field only column 8 (`PRCP`, precipitation), and 12 to 14 (`TMAX`, `TMIN`,`TOBS`, temperature measures). 

``` {r file-function-weather}
fileList <- list.files(pathToWeatherData, full.names=TRUE) # list all the files, full.names=TRUE is necessary for ldplay/lapply to work below
# use grep to create
weather <- fileList[grep(weather,fileList)]
weather <- read.table(weather, header=TRUE, sep=",",stringsAsFactors = FALSE); weather[,c(8,12:14)] <- weather[,c(8,12:14)]/10;weather$DATE <- paste(substr(weather$DATE,1,4),substr(weather$DATE,5,6),substr(weather$DATE,7,8),sep = '-') ;weather$DATE <- as.Date(weather$DATE,format="%Y-%m-%d")
```

Now we need to pull it together and create three data frames 

``` {r data-frames}
# Three dataframes compiling all the NEON data are created below
bet_field = multipleCombine(field, ply = ldply) # The field data from all sites
bet_sort = multipleCombine(sort, ply = ldply) # The sorting data from all sites
bet_pin = multipleCombine(pin, ply = ldply) # The data on pinned Carabidae
```

It might be nice to have a single object with all the spatial data in it. We can
call this `gisData` 

``` {r spatial-data}
gisData <- unique.data.frame(bet_field[c("domainID","siteID","plotID","trapID","nlcdClass",
                                         "decimalLatitude","decimalLongitude","geodeticDatum",
                                         "coordinateUncertainty","elevation","elevationUncertainty")])
```

Katie - ? not sure what this is for or what the note means. 
``` {r not-sure-!}
bet_field %>% 
	filter(missingRecordsPerBoutQF==0) %>% 
	select(-missingRecordsPerBoutQF) ->
	bet_field # These are misleading. If FOPS didn't set a trap, it isn't in any table

# clean up the objects no longer needed. 
rm(field,sort,pin,fileList)

````


## Fixing Plausible Errors in Data

With any data it is a good idea to review it to make sure that 
### Errors in bet_field data

#### Resolving duplicates


**Katie - this seems very specific to a sample, do we need ot teach? If so, how
would they know to do this?**

``` {r resolve-duplicates}
# resolving duplicates

for (i in 1:dim(bet_field)[1]){
  if(bet_field$plotID[i]=="JERC_030" & bet_field$boutNumber[i]==3) {
    bet_field$setDate[i] <- "2014-07-02"
    bet_field$collectDate[i] <- "2014-07-16"
  }
}
```

``` {r standard-ID}

# Standardizing sampleID
for (i in 1:dim(bet_field)[1]){
  bet_field$sampleID[i] <- paste(bet_field$plotID[i],bet_field$trapID[i],
                                 paste0(unlist(strsplit(as.character(bet_field$collectDate[i]),split="-"))[1],
                                        unlist(strsplit(as.character(bet_field$collectDate[i]),split="-"))[2],
                                        unlist(strsplit(as.character(bet_field$collectDate[i]),split="-"))[3]),
                                 sep = '.')
}

```


``` {r remove-dups}
# Removing uid and duplicate records
# use dplyr to filter for duplicates and then select them by sampleID
bet_field %>% 
  filter(duplicateCollectionEventQF==2) %>% 
  select(sampleID) -> bet_dups

# create a vector we will fill with those sampleIDs to remove.
recordsToRemove <- vector()

# fill `recordsToRemove` based on 
for (i in unique(bet_dups$sampleID)){
  bet_field %>% 
    filter(sampleID==i) %>% 
    select(uid)-> dup_uids
  recordsToRemove <- c(recordsToRemove,dup_uids[-1,])
}

# 
for (i in recordsToRemove){
  bet_field <- bet_field[-match(i,bet_field$uid),]
}

#
bet_field <- bet_field[,2:26]; rm(recordsToRemove,dup_uids,bet_dups)

```

### Errors in bet_sort 

``` {r }
# Make sampleIDs all Caps
bet_sort$associatedSampleID <- toupper(bet_sort$associatedSampleID)

# Known uids that are complete duplicates for bycatch
uidToRemove <- c("ED7318A8BAEB438684E8F09EAABE185F",
								 "F6C95A226CE747B4A628C9D11EB07730",
								 "2C716CDD2DAC4173AA853257EC522B1E", 
								 "3867CCADDB344AA28A72E8188D1CA6A8",
								 "6650FC14AD64406BACA54D6E57862C7F",
								 "EDD0884075574497BCFC281B7B58E50C")

# Finding additional records that are duplicated in invert bycatch
bet_sort %>% 
  filter(duplicateSampleIDQF==2,sampleType=="invert bycatch") -> a
for(i in unique(a$sampleID)){
  a %>% 
    filter(sampleID==i)-> a1
  uidToRemove <- append(uidToRemove,a1$uid[2])
}

# Finding records that are duplicates in the common carabid group
bet_sort %>% 
  filter(duplicateSampleIDQF==2,sampleType=="common carabid") -> a
a <- a[11:42,]
for(i in unique(a$associatedSampleID)){
  a %>% 
    filter(associatedSampleID==i) -> a1
  a1 %>% filter(individualCount==a1$individualCount) -> a1
  uidToRemove <- append(uidToRemove,a1$uid)
}
rm(a,a1)

# Remove duplicates
bet_sort <- bet_sort[-match(uidToRemove,bet_sort$uid),2:dim(bet_sort)[2]]

# Adding individual count numbers for records missing that data; fixing dates
for (i in 1:dim(bet_sort)[1]){
  if(bet_sort$sampleType[i]=='vert bycatch mam' & is.na(bet_sort$individualCount[i])==TRUE){
    bet_sort$individualCount[i] <- 1
  }
  if(bet_sort$sampleType[i]=='vert bycatch herp' & is.na(bet_sort$individualCount[i])==TRUE){
    bet_sort$individualCount[i] <- 1
  }
  if(length(unlist(strsplit(bet_sort$collectDate[i],split = '/')))==3){
    bet_sort$collectDate[i] <- paste(unlist(strsplit(bet_sort$collectDate[i],split = '/'))[3],
                                     ifelse(nchar(unlist(strsplit(bet_sort$collectDate[i],split = '/'))[1])==2,
                                            unlist(strsplit(bet_sort$collectDate[i],split = '/'))[1],
                                            paste0(0,unlist(strsplit(bet_sort$collectDate[i],split = '/'))[1])),
                                     ifelse(nchar(unlist(strsplit(bet_sort$collectDate[i],split = '/'))[2])==2,
                                            unlist(strsplit(bet_sort$collectDate[i],split = '/'))[2],
                                            paste0(0,unlist(strsplit(bet_sort$collectDate[i],split = '/'))[2])),
                                     sep = '-')
  }
  if(length(unlist(strsplit(bet_sort$etOHChangeDate[i],split = '/')))==3){
    bet_sort$etOHChangeDate[i] <- paste(unlist(strsplit(bet_sort$etOHChangeDate[i],split = '/'))[3],
                                        ifelse(nchar(unlist(strsplit(bet_sort$etOHChangeDate[i],split = '/'))[1])==2,
                                               unlist(strsplit(bet_sort$etOHChangeDate[i],split = '/'))[1],
                                               paste0(0,unlist(strsplit(bet_sort$etOHChangeDate[i],split = '/'))[1])),
                                        ifelse(nchar(unlist(strsplit(bet_sort$etOHChangeDate[i],split = '/'))[2])==2,
                                               unlist(strsplit(bet_sort$etOHChangeDate[i],split = '/'))[2],
                                               paste0(0,unlist(strsplit(bet_sort$etOHChangeDate[i],split = '/'))[2])),
                                        sep = '-')
  }
  if(length(unlist(strsplit(bet_sort$processingDate[i],split = '/')))==3){
    bet_sort$processingDate[i] <- paste(unlist(strsplit(bet_sort$processingDate[i],split = '/'))[3],
                                        ifelse(nchar(unlist(strsplit(bet_sort$processingDate[i],split = '/'))[1])==2,
                                               unlist(strsplit(bet_sort$processingDate[i],split = '/'))[1],
                                               paste0(0,unlist(strsplit(bet_sort$processingDate[i],split = '/'))[1])),
                                        ifelse(nchar(unlist(strsplit(bet_sort$processingDate[i],split = '/'))[2])==2,
                                               unlist(strsplit(bet_sort$processingDate[i],split = '/'))[2],
                                               paste0(0,unlist(strsplit(bet_sort$processingDate[i],split = '/'))[2])),
                                        sep = '-')
  }
  if(length(unlist(strsplit(bet_sort$identifiedDate[i],split = '/')))==3){
    bet_sort$identifiedDate[i] <- paste(unlist(strsplit(bet_sort$identifiedDate[i],split = '/'))[3],
                                        ifelse(nchar(unlist(strsplit(bet_sort$identifiedDate[i],split = '/'))[1])==2,
                                               unlist(strsplit(bet_sort$identifiedDate[i],split = '/'))[1],
                                               paste0(0,unlist(strsplit(bet_sort$identifiedDate[i],split = '/'))[1])),
                                        ifelse(nchar(unlist(strsplit(bet_sort$identifiedDate[i],split = '/'))[2])==2,
                                               unlist(strsplit(bet_sort$identifiedDate[i],split = '/'))[2],
                                               paste0(0,unlist(strsplit(bet_sort$identifiedDate[i],split = '/'))[2])),
                                        sep = '-')
  }
}


bet_sort$sampleIDPLUStaxa <- substr(bet_sort$sampleID,1,nchar(bet_sort$sampleID)-3)

```


### Errors in the bet_pin

``` {r errors-bet_pin}
# Add 
bet_pin$individualCount <-  1 # The pinning table represents single instances of individuals

# Make sampleIDs all Caps
bet_pin$sampleID <- toupper(bet_pin$sampleID)

# Fixing dates
for (i in 1:dim(bet_pin)[1]){
  if(length(unlist(strsplit(bet_pin$collectDate[i],split = '/')))==3){
    bet_pin$collectDate[i] <- paste(unlist(strsplit(bet_pin$collectDate[i],split = '/'))[3],
                                    ifelse(nchar(unlist(strsplit(bet_pin$collectDate[i],split = '/'))[1])==2,
                                           unlist(strsplit(bet_pin$collectDate[i],split = '/'))[1],
                                           paste0(0,unlist(strsplit(bet_pin$collectDate[i],split = '/'))[1])),
                                    ifelse(nchar(unlist(strsplit(bet_pin$collectDate[i],split = '/'))[2])==2,
                                           unlist(strsplit(bet_pin$collectDate[i],split = '/'))[2],
                                           paste0(0,unlist(strsplit(bet_pin$collectDate[i],split = '/'))[2])),
                                    sep = '-')
  }
  if(length(unlist(strsplit(bet_pin$processingDate[i],split = '/')))==3){
    bet_pin$processingDate[i] <- paste(unlist(strsplit(bet_pin$processingDate[i],split = '/'))[3],
                                       ifelse(nchar(unlist(strsplit(bet_pin$processingDate[i],split = '/'))[1])==2,
                                              unlist(strsplit(bet_pin$processingDate[i],split = '/'))[1],
                                              paste0(0,unlist(strsplit(bet_pin$processingDate[i],split = '/'))[1])),
                                       ifelse(nchar(unlist(strsplit(bet_pin$processingDate[i],split = '/'))[2])==2,
                                              unlist(strsplit(bet_pin$processingDate[i],split = '/'))[2],
                                              paste0(0,unlist(strsplit(bet_pin$processingDate[i],split = '/'))[2])),
                                       sep = '-')
  }
  if(length(unlist(strsplit(bet_pin$identifiedDate[i],split = '/')))==3){
    bet_pin$identifiedDate[i] <- paste(unlist(strsplit(bet_pin$identifiedDate[i],split = '/'))[3],
                                       ifelse(nchar(unlist(strsplit(bet_pin$identifiedDate[i],split = '/'))[1])==2,
                                              unlist(strsplit(bet_pin$identifiedDate[i],split = '/'))[1],
                                              paste0(0,unlist(strsplit(bet_pin$identifiedDate[i],split = '/'))[1])),
                                       ifelse(nchar(unlist(strsplit(bet_pin$identifiedDate[i],split = '/'))[2])==2,
                                              unlist(strsplit(bet_pin$identifiedDate[i],split = '/'))[2],
                                              paste0(0,unlist(strsplit(bet_pin$identifiedDate[i],split = '/'))[2])),
                                       sep = '-')
  }
}

# Correcting sampleIDs
# Set aside things that are completely accurate
matches <- (bet_pin$sampleID %in% bet_sort$associatedSampleID); bet_pin <- cbind(bet_pin,matches)

bet_pin %>% 
	filter(matches==TRUE) -> 
	bet_pin1; bet_pin1 <- 
	bet_pin1[,-match('matches',colnames(bet_pin1))] # 3778 correct records

bet_pin %>% 
	filter(matches==FALSE) -> 
	bet_pin; bet_pin <- 
	bet_pin[,-match('matches',colnames(bet_pin))] # 2168 incorrect records

# Adding a tube number '01' shows it exists
bet_pin$sampleID2 <- paste(bet_pin$sampleID,'01',sep=".")
matches <- (bet_pin$sampleID2 %in% bet_sort$associatedSampleID); bet_pin <- cbind(bet_pin,matches)
bet_pin %>% filter(matches==TRUE) -> bet_pin2; bet_pin2$sampleID <- bet_pin2$sampleID2 # 998 correct records
bet_pin2 <- bet_pin2[,-match('matches',colnames(bet_pin2))]; bet_pin2 <- bet_pin2[,-match('sampleID2',colnames(bet_pin2))]; bet_pin1 <- rbind(bet_pin1,bet_pin2); rm(bet_pin2)
bet_pin %>% filter(matches==FALSE) -> bet_pin; bet_pin <- bet_pin[,-match('matches',colnames(bet_pin))]; bet_pin <- bet_pin[,-match('sampleID2',colnames(bet_pin))] # 1170 incorrect records left to fix


# Of the sampleIDs that don't have a match in the sort table
for (i in bet_pin$individualID){
  bet_sort %>% 
    filter(plotID==bet_pin$plotID[match(i,bet_pin$individualID)],
           trapID==bet_pin$trapID[match(i,bet_pin$individualID)],
           collectDate==bet_pin$collectDate[match(i,bet_pin$individualID)],
           sampleType!='vert bycatch mam',
           sampleType!='vert bycatch herp',
           sampleType!='invert bycatch') -> b
  # Nothing in the sort table matches the plot/trap/date combo
  if(dim(b)[1]==0){
    bet_pin$sampleID2[match(i,bet_pin$individualID)] <- ''
  }
  
  # Only one option in the sort table
  if(dim(b)[1]==1){
    bet_pin$sampleID2[match(i,bet_pin$individualID)] <- b$associatedSampleID[1]
  }
  
  # Many options in the sort table
  if(dim(b)[1]>1){
    # And the pinned specimen was a 'common carabid' with a taxonID in the sample ID
    bet_sort %>% 
      filter(sampleIDPLUStaxa==paste(bet_pin$plotID[match(i,bet_pin$individualID)],
                                     bet_pin$trapID[match(i,bet_pin$individualID)],
                                     substr(bet_pin$sampleID[match(i,bet_pin$individualID)],12,19),
                                     bet_pin$taxonID[match(i,bet_pin$individualID)],
                                     sep='.')) -> b1
    if(dim(b1)[1]>0){
      bet_pin$sampleID2[match(i,bet_pin$individualID)] <- b1$associatedSampleID[1]
    } else{
      # And the pinned specimen was a 'common carabid' given a morphospecies assignment 
      bet_sort %>% 
        filter(substr(sampleID,1,19)==paste(bet_pin$plotID[match(i,bet_pin$individualID)],
                                            bet_pin$trapID[match(i,bet_pin$individualID)],
                                            substr(bet_pin$sampleID[match(i,bet_pin$individualID)],12,19),
                                            sep='.'),
               morphospeciesID==bet_pin$morphospeciesID[match(i,bet_pin$individualID)],nchar(morphospeciesID)>0) -> b1
      if(dim(b1)[1]>0){
        bet_pin$sampleID2[match(i,bet_pin$individualID)] <- b1$associatedSampleID[1]
      } else{
        # An 'other carabid' exists that fits the bill
        bet_sort %>% 
          filter(sampleIDPLUStaxa==paste(bet_pin$plotID[match(i,bet_pin$individualID)],
                                         bet_pin$trapID[match(i,bet_pin$individualID)],
                                         substr(bet_pin$sampleID[match(i,bet_pin$individualID)],12,19),
                                         'OC',
                                         sep='.')) -> b1
        if(dim(b1)[1]>0){
          bet_pin$sampleID2[match(i,bet_pin$individualID)] <- b1$associatedSampleID[1]
        }
      }
    }
  }   
  bet_pin$numRecords[match(i,bet_pin$individualID)] <- dim(b)[1]
  b=0
}
bet_pin %>% filter(numRecords==0)-> bet_pin2 # 213 samples with no provenance in the sort table
bet_pin %>% filter(numRecords>0)-> bet_pin # 957 records that might exist
for (i in 1:dim(bet_pin)[1]){
  if (substr(bet_pin$sampleID[i],1,19)==substr(bet_pin$sampleID2[i],1,19)){
    bet_pin$plausible[i] <- TRUE
  } else {
    bet_pin$plausible[i] <- FALSE    
  }
}
bet_pin %>% filter(plausible==TRUE) -> bet_pin3; bet_pin3$sampleID <- bet_pin3$sampleID2; bet_pin3 <- bet_pin3[,1:23]; bet_pin1 <- rbind(bet_pin1,bet_pin3); rm(bet_pin3) # 561 plausible fixes
bet_pin %>% filter(plausible==FALSE) -> bet_pin # 396 with incorrect sampleIDs

# Are the last incorrect IDs other carabids?
bet_pin$sampleID2 <- paste(substr(bet_pin$sampleID,1,19),'OC.01',sep='.'); bet_pin <- bet_pin[,1:24]
plausible <- bet_pin$sampleID2 %in% bet_sort$associatedSampleID; bet_pin <- cbind(bet_pin,plausible) 
bet_pin %>% filter(plausible==TRUE) -> bet_pin3; bet_pin3$sampleID <- bet_pin3$sampleID2; bet_pin3 <- bet_pin3[,1:23]; bet_pin1 <- rbind(bet_pin1,bet_pin3); rm(bet_pin3) # 313 could have been other carabids
bet_pin %>% filter(plausible==FALSE) -> bet_pin; bet_pin <- bet_pin[,1:23]; bet_pin2 <- bet_pin2[,1:23] # 83 with incorrect sampleIDs

# Some records in the pin table are for 'other carabids'; but were never entered in the sort table
bet_pin <- rbind(bet_pin,bet_pin2); rm(bet_pin2) # All records without sort info
bet_pin$sampleID <- paste(substr(bet_pin$sampleID,1,19),'OC.01',sep=".") 

# Adding missing pinning records into the sort table
bet_sort <- bet_sort[,1:24]
bet_sortExtraRecords <- unique.data.frame(cbind(bet_pin[,c(2:9,12:21,23)],
                                                bet_sort[1:dim(bet_pin)[1],c('associatedSampleID','etOHChangeDate','targetTaxaPresent','sampleType','duplicateSampleIDQF')]))
bet_sortExtraRecords$associatedSampleID <- bet_sortExtraRecords$sampleID; bet_sortExtraRecords$sampleType <- 'other carabid'; bet_sortExtraRecords$targetTaxaPresent <- 'Y'
bet_sortExtraRecords <- bet_sortExtraRecords[,colnames(bet_sort)]; bet_sortExtraRecords[,c(6,8,10,13:24)] <- ''; bet_sortExtraRecords$remarks <- 'record not originally found in sorting data; presumed to exist based on pinning data'
bet_sort <- rbind(bet_sort,unique.data.frame(bet_sortExtraRecords)); rm(bet_sortExtraRecords) # Now sorting table is complete

# Appending fixed pinning records to majority of pinning samples
bet_pin <- rbind(bet_pin,bet_pin1); rm(bet_pin1) # Now pinning data is complete

# Adding missing records
# Records in the pin and sort without corresponding field data
matches <- substr(bet_sort$associatedSampleID,1,19)%in%bet_field$sampleID
bet_sortA <- cbind(bet_sort,matches); bet_sortA %>% filter(matches==FALSE,sampleType!='invert bycatch')->bet_sortA; bet_sortA$sampleID <- substr(bet_sortA$associatedSampleID,1,19)
missingFieldRecords <- unique.data.frame(bet_sortA[c('domainID','siteID',"plotID","trapID","collectDate","sampleID")]); rm(bet_sortA)

fieldRecords <- bet_field[1:dim(missingFieldRecords)[1],]
fieldRecords[,c('uid','setDate',"boutNumber","eventID","daysOfTrapping",
                "cupStatus","lidStatus","fluidLevel","trapReset","remarks",'recordedBy',
                "duplicateCollectionEventQF", "compareSetCollectDateQF")] <- ''
missingFieldRecords <- cbind(missingFieldRecords,
                             fieldRecords[1:dim(missingFieldRecords)[1],c("setDate","boutNumber","eventID","daysOfTrapping","cupStatus","lidStatus","fluidLevel",                
                                                                                              "trapReset","samplingProtocol","recordedBy","remarks",
                                                                                              "duplicateCollectionEventQF")])
for (i in 1:dim(missingFieldRecords)[1]){
  gisData %>% 
    filter(plotID==missingFieldRecords$plotID[i],trapID==missingFieldRecords$trapID[i])->a
  missingFieldRecords$nlcdClass[i] <- a$nlcdClass[1]
  missingFieldRecords$decimalLatitude[i] <- a$decimalLatitude[1]
  missingFieldRecords$decimalLongitude[i]<- a$decimalLongitude[1]
  missingFieldRecords$geodeticDatum[i]<- a$geodeticDatum[1]
  missingFieldRecords$coordinateUncertainty[i]<- a$coordinateUncertainty[1]
  missingFieldRecords$elevation[i]<- a$elevation[1]
  missingFieldRecords$elevationUncertainty[i]<- a$elevationUncertainty[1]
}

missingFieldRecords <- missingFieldRecords[colnames(bet_field)]
missingFieldRecords$remarks <- 'This record is assumed to exist based on the presence of sorting or pinning data'
bet_field <- rbind(bet_field,missingFieldRecords);rm(missingFieldRecords,plausible,uidToRemove,gisData,matches,fieldRecords)
for (i in 1:dim(bet_field)[1]){
  if(bet_field$boutNumber[i]!='10' &
     bet_field$boutNumber[i]!='11' & 
     bet_field$boutNumber[i]!=''){
    bet_field$eventID[i] <- paste('BET',bet_field$plotID[i],'2014',paste0(0,bet_field$boutNumber[i]),sep='.')    
  } 
  if(bet_field$boutNumber[i]!='10'& 
     bet_field$boutNumber[i]!='11'){
    bet_field$eventID[i] <- paste('BET',bet_field$plotID[i],'2014',bet_field$boutNumber[i],sep='.')    
  }
}
for(i in 1:dim(bet_pin)[1]){
  if(unlist(strsplit(bet_pin$sampleID[i],split = '\\.'))[4]=='OC'){
    bet_pin$sampleType[i] <- 'other carabid'
  } else{
    bet_pin$sampleType[i] <- 'common carabid'
  }
}

# Updating 'OTHE' code for carabids
bet_pin %>% filter(taxonID=='OTHE') %>% select(remarks)->a; a <- sort(unique(a$remarks))
for (i in 1:dim(bet_pin)[1]){
  if(is.na(bet_pin$taxonID[i])==FALSE){
  if (bet_pin$taxonID[i]=='OTHE'){
    for (j in a){
      if (bet_pin$remarks[i]==j & match(j,a)<4){
        bet_pin$taxonID[i] <- 'CYCINC'
        bet_pin$scientificName[i] <- 'Cyclotrachelus incisus'
        bet_pin$taxonRank[i] <- 'species'
        bet_pin$scientificNameAuthorship[i] <- 'LeConte'
      } 
      if (bet_pin$remarks[i]==j & match(j,a)>3){
        bet_pin$taxonID[i] <- 'HARRUB'
        bet_pin$scientificName[i] <- 'Harpalus rubripes'
        bet_pin$taxonRank[i] <- 'species'
        bet_pin$scientificNameAuthorship[i] <- 'Duftschmid'
      } 
    }
  }
  }
}
bet_pin <- bet_pin[,2:24]

```


## Compiling Carabid Abundance & Diversity 


``` {r dates}
# Converting collectDate into a date format
bet_field$collectDate <- as.Date(bet_field$collectDate,format="%Y-%m-%d")
bet_sort$individualCount <- as.numeric(bet_sort$individualCount)
bet_field$daysOfTrapping <- as.numeric(bet_field$daysOfTrapping)

```

### Abundance for different sampleTypes
In the beetle data in 2014 and prior, beetles were separated into two `sampleTypes`:
common carabid and other carabid. The way abundance is calculated differs. We 
need to calculate it in two ways. 

1. If the sampleType=='common carabid', then the individualCount reflects the 
true number of Carabids of a given species that were in the trap. This means that 
functionally the 'individualCount' of any 'common carabid' beetle with a sample 
ID that is not in the pinning table at all can be processed for diversity/abundance 
as though none were pinned. Pinning records just tell you that a pinned individual 
from that trap is available

2. If the sampleType=='other carabid' in the bet_sort, then there is at least 
one beetle present from the trap indicated in the record,and to calculate 
abundance we need to count the number of rows associated with that sampleID

``` {r sampleType-abund}
# Getting the 'common carabid' data
bet_sort %>% 
  filter(sampleType=='common carabid')-> bet_cc

for (i in 1:dim(bet_cc)[1]){
  # For sorting data missing counts, check the pinning table for records.
  if(is.na(bet_cc$individualCount[i])==TRUE){
    bet_pin %>% 
      filter(plotID==bet_cc$plotID[i],trapID==bet_cc$trapID[i],
             collectDate==bet_cc$collectDate[i],taxonID==bet_cc$taxonID[i])-> a
    if(dim(a)[1]>0){
      bet_cc$individualCount[i] <- dim(a)[1]
    }
  }
  # For sorting data missing counts in the pinning table entirely, assign a minimum count of 1
  if(is.na(bet_cc$individualCount[i])==TRUE){ 
    bet_cc$individualCount[i] <- 1
  }
}

# Filling in other carabid data
# sorting records for which pinning data doesn't exist
bet_sort %>% 
  filter(sampleType=='other carabid') %>% 
  anti_join(bet_pin,by=c('associatedSampleID'= 'sampleID')) -> bet_oc1
bet_oc1$individualCount <- 1 # Can't find the corresponding records in the pinning table, but at least one must have been pinned/exist

# 'other carabid' sorting records that DO have pinning records
bet_sort %>% 
  filter(sampleType=='other carabid') %>% 
  semi_join(bet_pin,by=c('associatedSampleID'='sampleID')) -> bet_oc

for (i in unique(bet_oc$associatedSampleID)){
  bet_pin %>% 
    filter(sampleType=='other carabid',sampleID==i) -> a 
  a <- unique.data.frame(a[c("domainID","siteID" ,"plotID","trapID","collectDate",
                             "sampleID","taxonID","scientificName","taxonRank",
                             "identificationQualifier","scientificNameAuthorship","morphospeciesID",'sampleType')])
  
  a$targetTaxaPresent <- "Y"
  a$etOHChangeDate <- ''
  a$processingDate <- ''
  a$associatedSampleID <- a$sampleID
  a$sampleID <- ''
  a$individualCount <- -999
  a$duplicateSampleIDQF <- -1 
  a$identificationReferences <- ''
  a$identifiedBy <- '' 
  a$identifiedDate <- ''
  a$recordedBy <- ''
  a$remarks <- ''
  
  a <- a[colnames(bet_sort)]
  bet_oc1 <- rbind(bet_oc1,a)
}

rm(bet_oc)


for (i in 1:dim(bet_oc1)[1]){
  if (bet_oc1$individualCount[i]==-999){
    if(nchar(bet_oc1$taxonID[i])>4){
      bet_pin %>% 
        filter(sampleID==bet_oc1$associatedSampleID[i],taxonID==bet_oc1$taxonID[i],
               identificationQualifier==bet_oc1$identificationQualifier[i]) -> a
      bet_oc1$individualCount[i] <- dim(a)[1]
    }
    if(nchar(bet_oc1$morphospeciesID[i])>10){
      bet_pin %>% 
        filter(sampleID==bet_oc1$associatedSampleID[i],morphospeciesID==bet_oc1$morphospeciesID[i]) -> a
      bet_oc1$individualCount[i] <- dim(a)[1]
    }
  }
  bet_sort %>% 
    filter(associatedSampleID==bet_oc1$associatedSampleID[i]) -> a
  bet_pin %>% 
    filter(sampleID==bet_oc1$associatedSampleID[i]) -> a1
  if(bet_oc1$recordedBy[i]==''){
  bet_oc1$etOHChangeDate[i] <- max(a$etOHChangeDate)
  bet_oc1$processingDate[i] <- max(a$processingDate)
  bet_oc1$identifiedDate[i] <- max(a1$identifiedDate)
  bet_oc1$identifiedBy[i] <- max(a1$identifiedBy)
  bet_oc1$recordedBy[i] <- max(a1$recordedBy)
}
}

# If the 'bet_pin' list an identification for a beetle with a sampleType=='other carabid' 
# We have to assume all the beetles of that type were pinned and are listed in the 'bet_pin' table
# As per the protocol.

# This is the table that will be used to build the abundance statistics
bet_div <- as.data.frame(bind_rows(bet_cc,bet_oc1))
```


## Adding Beetle Info To the Field Data

``` {r pin-to-field-data}
# Adding beetle abundance and richness; bycatch info; weather
for (i in 1:dim(bet_field)[1]){
  # adding abundance
  bet_div %>% 
    filter(plotID==bet_field$plotID[i],trapID==bet_field$trapID[i],collectDate==bet_field$collectDate[i])-> x
  bet_field$beetleAbundance[i] <- sum(x$individualCount,na.rm = TRUE)
  # adding richness (including morphospecies)
  unique.data.frame(x[c('taxonID','morphospeciesID')]) -> x1
  bet_field$beetleRichness[i] <-  dim(x1)[1]
  # richness excluding morphospecies
  x1 %>% 
    filter(nchar(taxonID)>1) -> x1
  bet_field$beetleRichnessNoMorphospecies[i] <- dim(x1)[1]
  
  # bycatch
  bet_sort %>% 
    filter(sampleType!='other carabid',sampleType!='common carabid',
           plotID==bet_field$plotID[i],trapID==bet_field$trapID[i],
           collectDate==bet_field$collectDate[i])-> a
  # mammal bycatch
  a %>% 
    filter(sampleType=='vert bycatch mam') -> a1
  bet_field$numMammalsCaught[i] <- sum(a1$individualCount,na.rm = TRUE)
  # amphibian/reptile bycatch
  a %>% 
    filter(sampleType=='vert bycatch herp') -> a1  
  bet_field$numHerpsCaught[i] <- sum(a1$individualCount,na.rm = TRUE)
  # invertebrate bycatch
  a %>% 
    filter(sampleType=='invert bycatch') -> a1
  bet_field$invertBycatchPresent[i] <- dim(a1)[1]  

  print(i)
}
rm(a,a1,b1,x,x1,b,bet_cc,bet_oc1)

```


