---
title: "Soil temperature and moisture controls on soil microbial biomass"
syncID: b5183dec9d8844d28e5b892971517ff8
description: Explore the relationships between microbial biomass and soil temperature and moisture
authors: Edward Ayres
contributors: 
estimatedTime: 1 hour
packagesLibraries: neonUtilities
topics: soil temperature, soil moisture, soil microbial biomass
languagesTool: R
dataProduct: DP1.10104.001, DP1.10086.001
code1: "https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/soils/soil-microbe-biomass/soil-microbe-biomass-drivers.R"
tutorialSeries: 
urlTitle: "soil-biomass-drivers"
dateCreated: "2024-3-26"
---

Soil microbial biomass is the mass of bacteria and fungi in the soil and although it is a relatively small component of soil organic matter, these microbes play a major role in several ecosystem processes, including decomposition, nutrient cycling, and soil carbon stabilization. There are many ways to measure microbial biomass and NEON uses the amount of phospholipid fatty acids (PLFAs) in the soil because they turn over quickly making them a good proxy for the amount of living microbial mass. Many factors are known to influence soil microbial biomass, such as soil organic carbon content, vegetation type and litter inputs, soil type, and abiotic conditions, such as temperature and water availability. In this tutorial we'll investigate the relationship between soil microbial biomass and soil temperature and moisture at one NEON site. This will involve downloading NEON data, combining data from different NEON data products, filtering out suspect data, and exploring the microbial biomass relationship with soil temperature and moisture. We'll be using two different NEON data products: 

* **DP1.10104.001, Soil microbe biomass**
* **DP1.10086.001, Soil physical and chemical properties, periodic**

<a href="https://data.neonscience.org/data-products/DP1.10104.001" target="_blank">Soil microbial biomass</a> and corresponding <a href="https://data.neonscience.org/data-products/DP1.10086.001" target="_blank">soil sample temperature and moisture</a> are measured in ten plots at each NEON terrestrial site one-to-three times per year, with each site always sampled during the historic peak in vegetation greenness and sometimes sampled during other seasons or seasonal transition. 

We will be using data from the <a href="https://www.neonscience.org/field-sites/guan" target="_blank">Guanica Forest</a> (GUAN) site in Puerto Rico. The site is in a subtropical dry zone along the southern coast of Puerto Rico characterized by low annual rainfall averaging 840 mm and high evapotranspiration. Given that temperatures are warm and fairly stable at this subtropical site and precipitation is relatively low, we might expect to see a stronger relationship between microbial biomass and soil moisture than with soil temperature, so let's see what the data show.

<div id="ds-objectives" markdown="1">

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R loaded on your computer to complete this tutorial.

</div>

## 1. Setup

Start by installing (if necessary) and loading the `neonUtilities` package.

Installation can be run once, then periodically to get package updates.

```{r install_packages, eval=FALSE}

install.packages("neonUtilities")

```

Now load the package. This needs to be done every time you run code. 

```{r load-packages, results="hide"}

library(neonUtilities)

```


## 2. Download the data

Download the soil microbial biomass and soil properties 
data using the `loadByProduct()` function in the `neonUtilities` package. Inputs 
needed for the function are:

* `dpID`: data product ID; soil microbe biomass = DP1.10104.001 (or soil properties, periodic = DP1.10086.001)
* `site`: (vector of) 4-letter site codes; Guanica Forest = GUAN
* `release`: data release to download; we'll use the current latest release: RELEASE-2024
* `check.size`: should this function prompt the user with an estimated download size? Set to `FALSE` here for ease of processing as a script, but good to leave as default `TRUE` when downloading a dataset for the first time.

Since microbial biomass is only measured a few times per year in ten plots the amounts of data is relatively small so we'll download all available data by not specifying start and end dates. Refer to the <a href="https://www.neonscience.org/sites/default/files/cheat-sheet-neonUtilities_0.pdf" target="_blank">cheat sheet</a> 
for the `neonUtilities` package for more details if desired.

```{r soildata, results="hide"}

microb <- loadByProduct(dpID = "DP1.10104.001", 
                        site = "GUAN", 
                        release = "RELEASE-2024", 
                        check.size = F)

soil <- loadByProduct(dpID = "DP1.10086.001", 
                      site = "GUAN", 
                      release = "RELEASE-2024", 
                      check.size = F)

```

## 3. Exploring the data

First let's take a look at the microbial data that we downloaded, which is a list of seven tables but we're just going to use two of them: `microb$sme_scaledMicrobialBiomass` and `microb$variables_10104`. We can use the `head()` function to get a sense of what the data look like. There are a lot of columns corresponding to a range of different lipids that might be detected in the phospholipid fatty acid (PLFA) analysis, but we're primarily interested in total microbial biomass as indicated by total lipid concentration (`totalLipidConcentration`). We can also see that at least some of the extractions might not have work correctly as indicated by the `analysisResultsQF` column, so we'll want to filter those out later.

```{r explore microb}

# View the first few rows of the microbial biomass table
head(microb$sme_scaledMicrobialBiomass)

```

There is known contamination of lipid c18To0 from the Buyer and Sasser high-throughput PLFA method that comes from the solid phase extraction 96 well plates used. Prior to November 2021 this lipid was excluded from the results and no correction to the total lipid concentration data is necessary, however, since then c18To0 has been reported in the data and needs to be subtracted from the total lipid concentration to get a reliable estimate of microbial biomass. Note that an alternative approach to correcting the total lipid concentration data is to perform a blank correction based on lipid concentrations reported in blanks in the `microb$sme_batchResults` table.

```{r total lipid conc correction}

# Identify pre- (i.e., c18To0ScaledConcentration = NA) and post-November 2021 data
preNov2021 <- is.na(microb$sme_scaledMicrobialBiomass$c18To0ScaledConcentration)

# Add the pre-November 2021 total lipid concentration data to a new column (no correction needed)
microb$sme_scaledMicrobialBiomass$correctedTotLipidConc[preNov2021] <- 
  microb$sme_scaledMicrobialBiomass$totalLipidScaledConcentration[preNov2021]

# Add the corrected (totalLipidScaledConcentration - c18To0ScaledConcentration) total lipid concentration data to the new column
microb$sme_scaledMicrobialBiomass$correctedTotLipidConc[!preNov2021] <- 
  microb$sme_scaledMicrobialBiomass$totalLipidScaledConcentration[!preNov2021] - 
  microb$sme_scaledMicrobialBiomass$c18To0ScaledConcentration[!preNov2021]

```

To find the units of the total lipid concentration data we can use the `microb$variables_10104` table, which provides information about each column in the data product. The units are nanomoles of lipids per gram of soil.

```{r microb units}

# Identify the units of the totalLipidConcentration data
microb$variables_10104[grep("totalLipidConcentration", microb$variables_10104$fieldName), c("description", "units")]

```


Next let's look at the soil properties data, which contains 14 tables but we'll just use three of them: `soil$sls_soilCoreCollection` which includes soil temperature at the time of sampling, `soil$sls_soilMoisture`, and lastly `soil$variables_10086` to identify the measurement units.

```{r explore soil}

# View the first few rows of the soil core table, which includes soil temperature
head(soil$sls_soilCoreCollection)

# View the first few rows of the soil moisture table
head(soil$sls_soilMoisture)

```

We can find the units of the soil temperature and soil moisture data in `soil$variables_10086`, which are °C and g water/g soil, respectively.

```{r soil units}

# Identify the units of the soilTemp data
soil$variables_10086[grep("soilTemp", soil$variables_10086$fieldName), c("description", "units")]

# Identify the units of the soilTemp data
soil$variables_10086[grep("soilMoisture", soil$variables_10086$fieldName), c("description", "units")]

```


## 4. Combining the data

We have all the data we need but they are currently in several different tables that need to be combined. We'll use the `merge()` function and combine the tables using the `sampleID` column.

```{r temp-rows}

# Merge microbial biomass and soil core data tables
coreMicrob <- merge(soil$sls_soilCoreCollection, microb$sme_scaledMicrobialBiomass, by="sampleID")

# Merge soil moisture table with microbial and core table
swcMicrob <- merge(soil$sls_soilMoisture, coreMicrob, by="sampleID")

```


## 5. Microbial biomass relationship with soil temperature

Next let's identify the sampling plots so we can show the relationship between microbial biomass and soil temperature for each plot. Then we'll create colors for each sampling plot to use in the graph, identify the rows that passed the sample QA/QC tests, and plot the temperature-microbial biomass relationship.

```{r mb-temp plot}

# Identify the different soil plots represented in the table
soilPlots <- unique(swcMicrob$plotID)

# Create a color vector with a different color for each soil plot
colors <- rainbow(length(soilPlots))

# Identify rows that passed the QA/QC tests
goodRows <- grep("OK", swcMicrob$analysisResultsQF)

# Create an empty plotting window for the soil temperature and microbial biomass relationship
plot(swcMicrob$soilTemp, 
     swcMicrob$correctedTotLipidConc, 
     pch=NA, 
     xlab="Soil temperature (degrees C)", 
     ylab="Microbial biomass (nmol lipids/g soil)")

# Loop through the soil plots and add the temperature-microbial biomass relationship
for(i in 1:length(soilPlots)){
  points(swcMicrob$soilTemp[intersect(goodRows, grep(soilPlots[i], swcMicrob$plotID))], 
         swcMicrob$correctedTotLipidConc[intersect(goodRows, grep(soilPlots[i], swcMicrob$plotID))], 
         col=colors[i])
}

# Add a plot legend
legend("topleft", legend=soilPlots, col=colors, pch=1, bty="n")

```

There isn't an obvious relationship between soil temperature and microbial biomass and the range in soil temperature is relatively narrow due to the subtropical location of the site.


## 6. Microbial biomass relationship with soil moisture

Now let's make a similar plot using soil moisture instead of soil temperature.

```{r mb-swc plot}

# Create an empty plotting window for the soil moisture and microbial biomass relationship
plot(swcMicrob$soilMoisture, 
     swcMicrob$correctedTotLipidConc, 
     pch=NA, 
     xlab="Soil moisture (g water / g soil)", 
     ylab="Microbial biomass (nmol lipids/g soil)")

# Loop through the soil plots and add the moisture-microbial biomass relationship
for(i in 1:length(soilPlots)){
  points(swcMicrob$soilMoisture[intersect(goodRows, grep(soilPlots[i], swcMicrob$plotID))], 
         swcMicrob$correctedTotLipidConc[intersect(goodRows, grep(soilPlots[i], swcMicrob$plotID))], 
         col=colors[i])
}

# Add a plot legend
legend("topleft", legend=soilPlots, col=colors, pch=1, bty="n")

```

There's a lot of noise, but generally there appears to be a positive relationship between soil moisture and microbial biomass. Given the warm but relatively dry climate at GUAN it is perhaps not surprising that microbial biomass increases when more water is available in the soil.

It is likely that many other factors besides soil temperature and moisture also influence microbial biomass. Some of these relationships could be further explored using these data products (e.g., soil pH or organic carbon content) as well as other NEON data (e.g., root biomass). In addition, NEON data from across the US could be used to identify the dominant drivers of microbial biomass in different climates.