```{r setup, echo = FALSE, results='hide', message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
```

```{r author-info}
first_name <- "your first name"
last_name <- "your last name"
affiliation <- "your affiliation"
address <- "your address"
email <- "your email"
```

```{r manuscript_options}
break_year <- 1985
```

# Increase in life expectancy through time: a continental analysis

> By `r last_name`, `r first_name`
>
> Affiliation: `r affiliation`, `r address`
>
> Email: `r email`

In this paper, we will examine the change in life expectancy through time across
the five continents.


```{r prepare-data}
split_gdp_files <- list.files(path = "data-raw", pattern = "gdp-percapita\\.csv$", full.names = TRUE)
split_gdp_list <- lapply(split_gdp_files, read.csv)
gdp <- do.call("rbind", split_gdp_list)
```

## The situation

```{r mean-lifeExp-per-continent}
mean_lifeExp_by_cont <- gdp %>% group_by(continent, year) %>%
  summarize(mean_lifeExp = mean(lifeExp)) %>% as.data.frame
```

```{r summary-statistics}
latest_lifeExp <- gdp %>% filter(year == max(gdp$year)) %>%
  group_by(continent) %>%
  summarize(latest_lifeExp = mean(lifeExp)) %>%
  as.data.frame
```

There are important disparities in life expectancy across the five
continents. It ranges from `r min(latest_lifeExp[["latest_lifeExp"]])`
in `r latest_lifeExp[["continent"]][which.min(latest_lifeExp$latest_lifeExp)]`,
to `r max(latest_lifeExp[["latest_lifeExp"]])`
in `r latest_lifeExp[["continent"]][which.max(latest_lifeExp$latest_lifeExp)]`.

These disparities are also reflected in the increases in life expectancy between
`r min(gdp$year)` and `r max(gdp$year)`.

```{r lifeExp-trends-per-continent}
ggplot(mean_lifeExp_by_cont, aes(x = year, y = mean_lifeExp, colour = continent)) +
  geom_line() + facet_wrap(~ continent) + theme(legend.position = "top")
```

To investigate these differences in more detail, we compared the rate of increase
for the periods `r min(gdp$year)`-`r break_year` and `r break_year`-`r max(gdp$year)`.

```{r before-after-1985}
mdl_before_afr <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                     subset = (continent == "Africa" & year <= break_year))
mdl_after_afr <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                    subset = (continent == "Africa" & year > break_year))
coef_before_after_afr <- rbind(c("Africa", "before", coef(mdl_before_afr)[2]),
                               c("Africa", "after", coef(mdl_after_afr)[2]))

mdl_before_asi <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                     subset = (continent == "Asia" & year <= break_year))
mdl_after_asi <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                    subset = (continent == "Asia" & year > break_year))
coef_before_after_asi <- rbind(c("Asia", "before", coef(mdl_before_asi)[2]),
                               c("Asia", "after", coef(mdl_after_asi)[2]))

mdl_before_ame <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                     subset = (continent == "Americas" & year <= break_year))
mdl_after_ame <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                    subset = (continent == "Americas" & year > break_year))
coef_before_after_ame <- rbind(c("Americas", "before", coef(mdl_before_ame)[2]),
                               c("Americas", "after", coef(mdl_after_ame)[2]))

mdl_before_eur <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                     subset = (continent == "Europe" & year <= break_year))
mdl_after_eur <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                    subset = (continent == "Europe" & year > break_year))
coef_before_after_eur <- rbind(c("Europe", "before", coef(mdl_before_eur)[2]),
                               c("Europe", "after", coef(mdl_after_eur)[2]))

mdl_before_oce <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                     subset = (continent == "Oceania" & year <= break_year))
mdl_after_oce <- lm(mean_lifeExp ~ year, data = mean_lifeExp_by_cont,
                    subset = (continent == "Oceania" & year > break_year))
coef_before_after_oce <- rbind(c("Oceania", "before", coef(mdl_before_oce)[2]),
                               c("Oceania", "after", coef(mdl_after_oce)[2]))

coef_before_after <- rbind(
    coef_before_after_afr,
    coef_before_after_ame,
    coef_before_after_asi,
    coef_before_after_eur,
    coef_before_after_oce)
coef_before_after <- as.data.frame(coef_before_after)
names(coef_before_after) <- c("continent", "period", "trend")
coef_before_after$trend <- as.numeric(levels(coef_before_after$trend)[coef_before_after$trend])
coef_before_after$period <- factor(coef_before_after$period, levels = c("before", "after"))
```

```{r change-trend-plot}
ggplot(coef_before_after, aes(x = period, y = trend, colour = continent, group = continent)) +
  geom_point() + geom_path()
```

```{r change-trend}
change_trend <- coef_before_after %>% group_by(continent) %>% summarize(diff = trend[2] - trend[1])
largest_neg_change_index <- which.min(change_trend$diff)
largest_pos_change_index <- which.max(change_trend$diff)
least_change_index <- which.min(abs(change_trend$diff))
```

The largest negative change is observed for `r change_trend$continent[largest_neg_change_index]` with a
decrease of `r change_trend$diff[largest_neg_change_index]`.year<sup>-1</sup>
between the two time periods; while `r change_trend$continent[largest_pos_change_index]`
experienced the largest increase with a difference of
`r change_trend$diff[largest_pos_change_index]`.year<sup>-1</sup> during the
same interval.
`r change_trend$continent[least_change_index]` stayed the most constant between the
two time periods.
