---
layout: post
title: "NEON's Plant Phenology Data"
date:   2017-08-01
authors: [Megan A. Jones, Natalie Robinson, Lee Stanish]
contributors: [ ] 
dateCreated: 2017-08-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [ dplyr, ggplot2]
category: [self-paced-tutorial]
tags: [time-series, phenology, organisms]
mainTag:  neon-pheno-temp-series
tutorialSeries:  [neon-pheno-temp-series]
description: "Learn to work with NEON plant phenology observation data."
code1: 
image:
  feature: codedFieldJournal.png
  credit: National Ecological Observatory Network (NEON)
  creditlink: 
permalink: /R/plant-pheno-data
comments: true
---

{% include _toc.html %}

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

# Objectives
After completing this activity, you will:

 * 

## Things You’ll Need To Complete This Tutorial
You will need the most current version of `R` and, preferably, `RStudio` loaded
on your computer to complete this tutorial.

### Install R Packages
* **NAME:** `install.packages("NAME")`


[More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

### Download Data 

{% include/dataSubsets/_data_NEON-pheno-temp-timeseries.html %}

****
{% include/_greyBox-wd-rscript.html %}

****

## Additional Resources

* Data wrangling cheatsheet

</div>

Big Question: Plants change throughout the year - these are phenophases. 
Why do they change? 

Explore Phenology Data 

Download data  
- user guide

Briefly: 

* where is the data from (the topic of site selection)
* how is it collected 
* Which plants/growth form should we choose? 
* growth-form vs. species discussion

Final choice: - LITU


But how do we choose what aspect of phenophase to look at: 

Phenophases that are recorded (deciduous broad-leaf)/ all with Y/N & % of canopy/plant:

* breaking leaf buds 
* increasing leaf size 
* leaves 
* open flowers 
* colored leaves
* falling leaves 

Criteria for choosing 

* something with progress
* comparable across sites or growth forms? 

Final choice: leaves  (main teaching), colored leaves  (challenges)


Status codes
Status intensity - every time they go out. 
PerIndividualPerYear - once a year, "metadata" about the plants 

```{r loadStuff}

library(dplyr)
library(ggplot2)
library(lubridate)


# set working directory 
setwd('/Users/lstanish/Downloads/')


# Read in data
ind <- read.csv('NEON-pheno-temp-timeseries/pheno/phe_perindividual.csv', stringsAsFactors = FALSE )

status <- read.csv('NEON-pheno-temp-timeseries/pheno/phe_statusintensity.csv', stringsAsFactors = FALSE)

```

Explore the data, let's get to know what the dataframe looks like.

```{r look-ind}
#What are the fieldnames in this dataset?
names(ind)

#how many rows are in the data?
nrow(ind)

#look at the first six rows of data.
head(ind)

# look at the structure of the dataframe.
str(ind)

```

Add protocal transect map 

Link GitHub repo - https://github.com/NEONScience/NEON-geolocation

Dates - opened first in excel (06/14/2014) vs not (2014-06-14)

```{r look-status}

# What variables are included in this dataset?
names(status)
nrow(status)
head(status)
str(status)

# min date
min(status$date)
max(status$date)

```

The `uid` is no important to understanding the data so we are going to remove uid. 
However, if you are every reporting an error in the data you should include this
with your report. 

```{r remove-uid}

ind <- select(ind,-uid)
status <- select (status, -uid)

``` 

## Clean up 

* remove duplicates (full rows)
* convert date
* retain only the latest `editedDate` in the perIndividual table.

### Remove duplicates


```{r remove-duplicates}

ind_noD <- distinct(ind)
nrow(ind_noD)

status_noD<-distinct(status)
nrow(status_noD)

```

## Figure out which names overlap 

```{r same-fieldnames}

sameName <- intersect(names(status_noD), names(ind_noD))
sameName

```

These fields have different values we want to keep. Rename common fields before joining: 
* editedDate
* measuredBy
* recordedBy
* samplingProtocolVersion
* remarks
* dataQF

```{r rename-column}

# rename status editedDate
status_noD <- rename(status_noD, editedDateStat=editedDate, measuredByStat=measuredBy, recordedByStat=recordedBy, samplingProtocolVersionStat=samplingProtocolVersion, remarksStat=remarks, dataQFStat=dataQF)


```


## Convert to Date

Our `addDate` and `date` columns are stored as a `character` class. We need to convert it to 
date-time class. 

```{r as-date-only }

# convert column to date class
ind_noD$editedDate <- as.Date(ind_noD$editedDate)
str(ind_noD$editedDate)

status_noD$date <- as.Date(status_noD$date)
str(status_noD$date)

```



## Retain only the latest indivdual record

Only the latest `editedDate` on ind


```{r filter-edit-date}
# retain only the max of the date for each individualID
ind_last <- ind_noD %>%
	group_by(individualID) %>%
	filter(editedDate==max(editedDate))

# oh wait, duplicate dates, retain only one
ind_lastnoD <- ind_last %>%
	group_by(editedDate, individualID) %>%
	filter(row_number()==1)

```

## Join dataframes

```{r join-dfs-error}

# Create a new dataframe "phe_ind" with all the data from status and some from ind_lastnoD
phe_ind <- left_join(status_noD, ind_lastnoD)


```

Ack!  Two different data types.  Why?  NA in taxonID is a logicial, but all the 
names are character.  

Try it again.  

`taxonID` and `scientificName` are provided for convenience in Status table, but
most up to date data is always in the `phe_perindividual.csv` files. Therefore, 
we'll remove from ...


```{r join-dfs}
# drop taxonID, scientificName
status_noD <- select (status_noD, -taxonID, -scientificName)

# Create a new dataframe "phe_ind" with all the data from status and some from ind_lastnoD
phe_ind <- left_join(status_noD, ind_lastnoD)

```


### What do we do with the data?  


Build a DF of interest with single site, species, and phenophase called `phe_1sp`.


## Select Site(s) of Interest

```{r filter-site}

# set site of interest
siteOfInterest <- "SCBI"

# use filter to select only the site of Interest 
# using %in% allows one to add a vector if you want more than one site. 
phe_1sp <- filter(phe_ind, siteID %in% siteOfInterest)

```

## Select Species of Interest


```{r filter-species}

# see which species are present
unique(phe_1sp$taxonID)

speciesOfInterest <- "LITU"

#subset to just "LITU"
# here just use == but could also use %in%
phe_1sp <- filter(phe_1sp, taxonID==speciesOfInterest)

# check that it worked
unique(phe_1sp$taxonID)

```


## Select Species of Interest

```{r filter-phonophase}

# see which species are present
unique(phe_1sp$phenophaseName)

phenophaseOfInterest <- "Leaves"

#subset to just the phenosphase of Interest 
phe_1sp <- filter(phe_1sp, phenophaseName %in% phenophaseOfInterest)

# check that it worked
unique(phe_1sp$phenophaseName)

```

## Total in Phenophase of Interest

Calculate the phenophaseStatus total Yes of total Individuals

```{r calc-total-yes}

# Total in status by day
sampSize <- count(phe_1sp, date)
inStat <- phe_1sp %>%
	group_by(date) %>%
  count(phenophaseStatus)
inStat <- full_join(sampSize, inStat, by="date")

# Retain only Yes
inStat_T <- filter(inStat, phenophaseStatus %in% "yes")

```

## Plot with ggplot
The `ggplot()` function within the `ggplot2` package gives us more control
over plot appearance. However, to use `ggplot` we need to learn a slightly 
different syntax. Three basic elements are needed for `ggplot()` to work:

 1. The **data_frame:** containing the variables that we wish to plot,
 2. **`aes` (aesthetics):** which denotes which variables will map to the x-, y-
 (and other) axes,  
 3. **`geom_XXXX` (geometry):** which defines the data's graphical representation
 (e.g. points (`geom_point`), bars (`geom_bar`), lines (`geom_line`), etc).
 
The syntax begins with the base statement that includes the `data_frame`
(`inStat_T`) and associated x (`date`) and y (`n`) variables to be
plotted:

`ggplot(inStat_T, aes(date, n))`

To successfully plot, the last piece that is needed is the `geom`etry type. For example, if we want to create a scatterplot, we can add `+ geom_point()`.

## Bar Plots with ggplot
We can use ggplot to create bar plots too. Let's create a bar plot of total 
number of individuals with a phenophase of "leaves". A bar plot might be a better way to represent a total value. To create a bar plot, we change the `geom` element from
`geom_point()` to `geom_bar()`.  

The default setting for a ggplot bar plot -  `geom_bar()` - is a histogram
designated by `stat="bin"`. However, in this case, we want to plot count values. 
We can use `geom_bar(stat="identity")` to force ggplot to plot actual values.


```{r plot-leaves-total}

# plot number of individuals in leaf

phenoPlot <- ggplot(inStat_T, aes(date, n.y)) +
    geom_bar(stat="identity", na.rm = TRUE) +
    ggtitle("Total Individuals in Leaf") +
    xlab("Date") + ylab("Number of Individuals") +
    theme(plot.title = element_text(lineheight=.8, face="bold", size = 20)) +
    theme(text = element_text(size=18))

phenoPlot

```


```{r plot-leaves-percentage}


inStat_T$percent<- ((inStat_T$n.y)/inStat_T$n.x)*100

# plot percent of leaves

phenoPlot_P <- ggplot(inStat_T, aes(date, percent)) +
    geom_bar(stat="identity", na.rm = TRUE) +
    ggtitle("Proportion in Leaf") +
    xlab("Date") + ylab("% of Individuals") +
    theme(plot.title = element_text(lineheight=.8, face="bold", size = 20)) +
    theme(text = element_text(size=18))

phenoPlot_P

```

Why don't we have the nice drop-off in % leaves at the end of the year like we do when we plot individual counts? This is when it's important to understand the protocol, because once a plant falls below 5% of green or colored leaves, it stops being monitored. This makes it look like the percentages stay high, when in reality it's just because non-leafy plants aren't being counted. 

## Re-evaluate dataset
The plots demonstrate that, while the 2016 data show the nice expected pattern 
of increasing leaf-out, peak, and drop-off, we seem to be missing the increase 
in leaf-out in 2015. That may create problems with downstream analyses. Let's 
filter the dataset to include just 2016.


## Select 2016 SCBI data

```{r filter-to-2016}

# use filter to select only the site of Interest 
# using %in% allows one to add a vector if you want more than one site. 
phe_1sp_2016 <- filter(phe_1sp, date >= "2016-01-01")

```



```{r write-csv, echo = FALSE}
# Write .csv (this will be read in new in subsuquent lessons)
write.csv(phe_1sp_2016, file="NEONpheno_LITU_Leaves_SCBI_2016.csv")

```


