Challenge (at end) was originally at end of Image-Raster-Data-In-R tutorial. 
However, band 90 won't stack with the others and despite the text saying "band 90, 
that you fixed in step 1", the tutorial doens't teach how to fix it.  I can't 
find any of our tutorials that fix it.  
I started to add a section When Raster's Won't Stack as this does seem like great
info (extent & resolution don't match), but I can't figure out how to do it at 
this point. 

Removing from lesson so a working lesson can be published, but want to retain and 
potentially add it back in later!  -- MJ 16 March 17

---

## When Rasters Won't Stack


``` {r band-90}
# import band 90 as a raster
band90raster <- raster("band90.tif")

# view attributes
band90raster
rgbRaster


```

Notice that the resolution and extent are different. We won't be able to create
a raster stack or raser brick (coming up next) with the other bands if they don't
line up. 

Let's see what it looks like.
``` {r plot-90}

# plot band90
plot(band90raster)

```

Hmm, band90 certainly appears to be of the same location as bands 19, 34, and 58. 
The tags in the GeoTIFF must have been set incorrectly to start. 

First, lets fix the extent

``` {r set-extent}
# define known extent from other bands


```

---- START CHALLENGE -----

 <div id="ds-challenge" markdown="1">
## Challenge: Create & Plot Image

Apply what you've learned to a new problem! A color infrared image is the 
combination of the near-infrared, red, and green bands. For our data

* green: band 34
* red: band 58 and 
* near-infrared: band 90. 

Using the  band90.tiff raster in the RGB directory with the other rasters, and 

1. Create a color infrared image (as a false color RGB image). 
1. Crop the image to the same dimensions as the previous RGB images (or 
experiment with your own cropping dimensions). 
1. Export your results as a GeoTiff. 

Hint: NIR->R->G = R->G->B

</div>

``` {r challenge-code-plot-crop-rgb, echo=FALSE}


# create color infrared stack
nrgRaster <- stack(band90raster,band58,band34)

# plot an RGB version of the cropped stack
plotRGB(rgbRaster_crop,r=3,g=2,b=1, stretch = "lin")

```

