---
title: "NEON Spatial Data"
contributors: Collin Storlie
dataProduct: DP3.30024.001
description: "Downloading and mapping NEON raster data"
estimatedTime: NULL
languagesTool: R
packagesLibraries: raster, neonUtilities, rgdal, ggplot2
authors: Matthew Helmnus, Donal O'Leary
topics: spatial, raster, shape
---

This tutorial will walk through downloading and mapping NEON spatial data

<div id="ds-objectives" markdown="1">

## Learning Objectives
After completing this activity, you will be able to:

* Import NEON raster data from aerial images using the neonUtilities package
* Plot raster data using ggplot2
* Overlay shape files onto your raster maps.

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and, preferably, `RStudio` loaded
on your computer to complete this tutorial.

### Install R Packages

* **raster:** `install.packages("raster")`
* **neonUtilities:** `install.packages("neonUtilities")`
* **rgdal:** `install.packages("rgdal")`

<a href="https://www.neonscience.org/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

### Download Data

The entire dataset can be accessed by request from the 
<a href="http://data.neonscience.org" target="_blank"> NEON Data Portal</a>.

AOP Flight Data for a single site and year, SoapRoot Saddle 2019, will be downloaded via the neonUtilities R package.  There are 1462 files totalling ~ 1.8 Gigabytes.  Download may take up to ~30 minutes.  Took me 23 minutes

## Recommended Reading

You may benefit from reviewing these related resources prior to this tutorial: 

* <a href="https://www.neonscience.org/raster-res-extent-pixels-r" target="_blank"> *The Relationship Between Raster Resolution, Spatial Extent & Number of Pixels - in R* tutorial.</a>
* <a href="https://www.neonscience.org/raster-data-r" target="_blank" > Please read through *Raster Data in R - The Basics* tutorial.</a>
* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank"> The `raster` R package documentation.</a>
* <a href="https://www.neonscience.org/field-sites/field-sites-map/SOAP" target="_blank"> Soaproot Saddle site description.</a> 
* <a href="https://www.neonscience.org/soaproot-saddle-overview-soap-drone-video" target="_blank"> Short overview video.</a>
* <a href="https://www.neonscience.org/intro-lidar-r-series" target="_blank"> See the NEON lidar tutorial series. </a>
* <a href="https://www.neonscience.org/chm-dsm-dtm-gridded-lidar-data" target="_blank"> What is a CHM, DSM, and DTM? </a>
</div>


# Prompt:
# Do you think it easy to have code that downloads lidar height and 
# tree cover for the full AOP footprint, do arithmetic, and then shapefile intersections?

```{r set-up}

# Load the required packages for this tutorial

library(neonUtilities)
library(raster)
library(rgdal)

# Create an object which is the file path to your working directory
# This will be used to read / write files with

wd = "/Users/collin/Documents/GitProjects/NEON-Data-Skills/tutorials-in-development/R/NEON_Raster/data/"

```

We will download the lidar-derived 'Elevation' dataset, which includes the 
digital terrain model (DTM) and digital surface model (DSM)
         
Can download multiple sites e.g., c("SOAP","HARV") and years
check.size=T is helpful to avoid accidentally downloading many Gbs!

```{r NEON-data-download}
byFileAOP("DP3.30024.001", site="SOAP", 
          year="2019", check.size=T) # 1.8 Gb's, 1462 files, started at 08:47, finished at 09:10.  Total 23 minutes.
```

Let's look at the digital terraim model (DTM) which is a model of the
Earth's surface with all of the vegetation removed (ground returns only).

List out the files
The directory structure is provided by the NEON Portal when you download the data
Full names are provided
REGEX expression ends in $, means the pattern is searched at the end
I.e. the files must END in .tif, not just have .tif present within them


```{r tif-filepath-vector}
f <- list.files(path = paste0(wd,"DP3.30024.001/2019/FullSite/D17/2019_SOAP_4/L3/DiscreteLidar/DTMGtif/"), 
                pattern = ".tif$", full.names = TRUE)
```

lapply takes a vector a filepaths
And applies the 'raster' function to them
Reading in each filepath as a raster and storing it as an element of the list
Each filepath is 'tile', represents a different location

```{r raster-list}
rl <- lapply(f, raster)
```

Merge combines tiles to form a larger map

```{r raster-merge}
DTM=do.call(merge, c(rl, tolerance = 1)) # Took about an hour with SOAP, need a note about this
```

Let's save the mosaic to avoid having to combine all the tiles again.

```{r raster-write}
writeRaster(DTM, filename = paste0(wd,"/DP3.30024.001/2019/DTM.tif"))
```

if you want to read this in again later:

```{r raster-read}
DTM=raster(paste0(wd,"/DP3.30024.001/2019/DTM.tif"))
```

plot the full DTM

```{r raster-plot}
plot(DTM)
```

### Download Shape Data

Now let's start working with some NEON shape data.  Click the links to visit the NEON spatial data page, and then click the 'Download button'.

* <a href="https://neon.maps.arcgis.com/home/item.html?id=f27616de7f9f401b8732cdf8902ab1d8" target="_blank"> Flight Boxes </a>
* <a href="https://neon.maps.arcgis.com/home/item.html?id=4a381f124a73490aa9ad7b1df914d6d8" target="_blank"> Site Sampling Boundaries </a>
* <a href="https://neon.maps.arcgis.com/home/item.html?id=3af642ac5b5b422fbc8c09132d0e13cb" target="_blank"> Tower Point Locations </a>
* <a href="https://neon.maps.arcgis.com/home/item.html?id=e05760bfb8224e37bf918ca1d775b20c" target="_blank"> TOS Plot  Points </a>
* <a href="https://neon.maps.arcgis.com/home/item.html?id=d87cd176dd6a468294fc0ac70918c631" target="_blank"> TIS Airshed Footprint </a>

Unzip the files to a location, preferably a folder within your work directory
Make a note about how to do this properly, command line and explanation of point'n'click too

##Add flightbox shapedata to the raster

AOP_fb=readOGR(paste0(wd,"/AOP_Flightboxes/AOP_flightboxesAllSites.shp"))
# What is the projection for this file?
crs(AOP_fb)
# Select just SOAP
SOAP_AOP_fb=AOP_fb[AOP_fb$siteID=="SOAP",]
# re-project this shapefile into UTM11 to pair with the DTM
SOAP_AOP_fb=spTransform(SOAP_AOP_fb, crs(DTM))
plot(SOAP_AOP_fb, add=T)

##Add field site outlines.

field_sampling_boundaries=readOGR(paste0(wd,"Field_Sampling_Boundaries/terrestrialSamplingBoundaries.shp"))
# What is the projection for this file?
crs(field_sampling_boundaries)
# Select just SOAP
SOAP_field_boundary=field_sampling_boundaries[field_sampling_boundaries$siteID=="SOAP",]
SOAP_field_boundary=spTransform(SOAP_field_boundary, crs(DTM))
plot(SOAP_field_boundary, add=T, border="green3")







