---
layout: post
title: "Dealing with Spatial Extents when working with Heterogeneous Data"
date:   2016-06-10
authors: [Leah A. Wasser, Kyla Dahlin]
instructors: [Leah, Naupaka]
time:
contributors: [Megan A. Jones]
dateCreated:  2016-05-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rhdf5]
categories: [self-paced-tutorial]
mainTag: institute-day2
tags: [R, HDF5]
tutorialSeries: [institute-day2]
description: "This tutorials covers one way to manage data with varying spatial 
extent coverage."
code1: institute-materials/day2_tuesday/mask-raster-lidar-hyperspec-fusion-R.R
image:
  feature: 
  credit: 
  creditlink:
permalink: /R/masks-data-fusion-R/
comments: false
---

## About

In this tutorial, we will learn how to process data that cover different spatial
extents in `R`. Note that this assumes the data are on the same raster grid and
same resolution. 

First, let's load the required libraries.

```{r load-libraries, warning=FALSE, results='hide', message=FALSE}
# load libraries
library(raster)
library(rhdf5)
library(rgdal)

# be sure to set your working directory
# setwd("~/Documents/data/NEONDI-2016") # Mac
# setwd("~/data/NEONDI-2016")  # Windows

## import functions
# install devtools (only if you have not previously intalled it)
#install.packages("devtools")
# call devtools library
#library(devtools)

# install from github
#install_github("lwasser/neon-aop-package/neonAOP")
# call library
library(neonAOP)

# source("/Users/lwasser/Documents/GitHub/neon-aop-package/neonAOP/R/aop-data.R")

```

## Import LiDAR data

To begin, let's open the NEON LiDAR Digital Surface and Digital Terrain Models
(DSM and DTM) which are in GeoTIFF format.

```{r import-lidar }

# read aspect data from previous lesson
TEAK_nsAspect <- raster("outputs/TEAK/TEAK_nsAspect.tif")

```

## Mask Data

Once we have created a threhold classified raster, we can use it for different things.
One application is to use it as an analysis mask for another dataset. 

Let's try to find all pixels that have an NDVI value >.6 and are north facing. 

```{r mask-data }
# Define the file name to be opened
f <- "NEONdata/D17-California/TEAK/2013/spectrometer/reflectance/Subset3NIS1_20130614_100459_atmcor.h5"

# define the CRS in EPGS format for the file
epsg <- 32611
# create a list of bands
bands <- c(60,83)

ndvi.stack <- create_stack(f, 
                           bands, 
                           epsg=32611)

# calculate NDVI
ndvi <- (ndvi.stack[[2]]-ndvi.stack[[1]]) / (ndvi.stack[[2]]+ndvi.stack[[1]])
names(ndvi) <- "TEAK_hsiNDVI"

# let's test this out
plot(ndvi)

# let's create a mask
ndvi[ndvi<.6] <- NA
plot(ndvi)

# force the two to have the same CRS
crs(ndvi) <- crs(TEAK_nsAspect)
```

# Create Raster Stack

Next, let's create a raster stack of ndvi and aspect.
```{r create-stack}

new.stack <- stack(TEAK_nsAspect, ndvi)

```

Notice we get an error. Why?

Let's compare the extents of the two objects. 

```{r compare-extents }
# view extents of both objects
extent(ndvi)
extent(TEAK_nsAspect)

# are the extents the same?
extent(ndvi) == extent(TEAK_nsAspect)

```

The extents are slightly different. They are one pixel apart in ymin and xmax.
Thus, when we try to create a stack, we get an error. All layers in a stack
need to be of the same extent.

We can create an if statement that checks the extent and crops both rasters to 
the overlap. Let's try it. 


```{r check-extents }

# check the extents of the two layers -- if they are different
# crop both datasets
if (extent(ndvi) == extent(TEAK_nsAspect)){
  print("Extents are the same, no need to crop")
  } else {
  # calculate overlap between the two datasets
  overlap <- intersect(extent(ndvi), extent(TEAK_nsAspect))
  # now let's crop both datasets to the overlap region
  ndvi <- crop(ndvi, overlap)
  asp.ns <- crop(TEAK_nsAspect, overlap)
  print("Extents are different, cropping data")
  }

# let's try to create a stack again.
new.stack <- stack(TEAK_nsAspect, ndvi)

```


```{r create-mask}

# mask out only pixels that are north facing and NDVI >.6
nsFacing.ndvi <- mask(new.stack[[1]], new.stack[[2]])
nsFacing.ndvi[nsFacing.ndvi==0] <- NA

```

## Create Final Plot

```{r plot-data}
# plot extent
plot.extent <- extent(nsFacing.ndvi)

# plot 
plot(nsFacing.ndvi,
     main="North & South Facing pixels, NDVI > .6",
     col=c("blue","green"),
     legend=F)

# allow legend to plot outside of bounds
par(xpd=TRUE)

legend((par()$usr[2] + 20), plot.extent@ymax-200, # set x,y legend location
       legend = c("North", "South"),
       fill = c("blue", "green"), 
       bty="n") # turn off border

```

## Export Classified Raster

```{r export-geotiff, eval=FALSE}

# export geotiff 
writeRaster(nsFacing.ndvi,
            filename="outputs/TEAK/TEAK_nsAspect_hsiNDVI.tif",
            format="GTiff",
            options="COMPRESS=LZW",
            overwrite = TRUE,
            NAflag = -9999)


```

