---
syncID: 1df47c5596574e6c8a9e03ba49d004df
title: "Calculate NDVI from NEON Hyperspectral Remote Sensing Data in R"
description: "Intro to HDF5"
dateCreated:  2016-05-01
authors: Leah A. Wasser, Kyla Dahlin
contributors: Edmund Hart, Megan A. Jones
estimatedTime:
packagesLibraries: rhdf5
topics: HDF5
languagesTool:
dataProduct:
code1: R/calculate_ndvi_h5.R
tutorialSeries:
urlTitle: 
---

First, let's load the required libraries.

```{r load-libraries, warning=FALSE, results='hide', message=FALSE}

# load libraries
library(raster)
library(rhdf5)
library(rgdal)

```

## Load Functions

Once we have done the work to build our functions, we can perform routine tasks
over and over using those functions.

Note: if you have not yet loaded the `neonAOP` package, please see directions 
for how to do this at the top of the 
<a href="{{ site.basurl }}/R/plot-spectral-signature/" target="_blank"> *Plot a Spectral Signature from Hyperspectral Remote Sensing data in R - HDF5* tutorial</a>. 

```{r load-functions }

## call function library
library(neonAOP)

```

## Calculate NDVI

Next, we can use the `create_stack` function to create a raster stack of the
red and near-infrared bands that we need to calculate NDVI.

```{r create-NDVI }

# set working directory
# setwd("~/Documents/data/NEONDI-2016")
# setwd("~/data/NEONDI-2016")  # Windows

# Define the file name to be opened
f <- "NEONdata/D17-California/TEAK/2013/spectrometer/reflectance/Subset3NIS1_20130614_100459_atmcor.h5"

# define CRS
epsg=32611

# Calculate NDVI
# select bands to use in calculation (red, NIR)
ndvi_bands <- c(58, 90)

#create raster list and then a stack using those two bands
ndvi_stack <-  create_stack(ndvi_bands,
                            file = f,
                            epsg = epsg)

# calculate NDVI
NDVI <- function(x) {
	  (x[,2]-x[,1])/(x[,2]+x[,1])
}

ndvi_rast <- calc(ndvi_stack, NDVI)

# clear out plots
# dev.off(dev.list()["RStudioGD"])

plot(ndvi_rast,
     main="NDVI for the NEON TEAK Field Site")


```

## Export to GeoTIFF

```{r export-ndvi, eval=FALSE}

# export as a GeoTIFF
writeRaster(ndvi_rast,
            file="outputs/TEAK/ndvi_2013.tif",
            format="GTiff",
            overwrite=TRUE)

```

## Plot NDVI

```{r import-lidar}

DSM <- raster("NEONdata/D17-California/TEAK/2013/lidar/TEAK_lidarDSM.tif")  

slope <- terrain(DSM, opt='slope')
aspect <- terrain(DSM, opt='aspect')

# create hillshade
hill <- hillShade(slope, aspect, 40, 270)

plot(hill,
     col=grey(1:100/100),
     main="NDVI for the Lower Teakettle Field site",
     legend=FALSE)

plot(ndvi_rast,
     add=TRUE,
     alpha=.3
     )
```
