---
syncID: 3b5f4a2c871844069c94c3bb04c15389
title: "NEON Hyperspectral Remote Sensing Data in R - Efficient Processing Using Functions"
description: "Accessing NEON "
dateCreated:  2016-05-01
authors: Leah A. Wasser
contributors: Edmund Hart
estimatedTime:
packagesLibraries: rhdf5
topics: HDF5
languagesTool: R
datProduct:
code1: /R/openNeonH5_functions.R
tutorialSeries:
urlTitle: hdf5-neon-hyperspec-rs-functions
---

In this tutorial you will learn how to write functions to open and work with 
NEON Hyperspectral Remote Sensing data that are delivered in HDF5 format. You 
will then use those fuctions to open and access hyperspectral remote sensing data.
Note that this tutorial currently outlines out to access
the data in the HDF5 format used prior to 2017. The functions/pathways may need 
to be different for the reformatted data now available from NEON. 

There is also the 
<a href="https://github.com/lwasser/neon-aop-package" target="_blank"> neonAOP package developed by Leah Wasser </a> that contains these functions. Details at
the end of this tutorial. 

<div id="ds-objectives" markdown="1">

## Learning Objectives
After completing this tutorial, you will be able to:

* Access data in NEON H5 files using pre-written functions. 

## Things You’ll Need To Complete This Tutorial
To complete this tutorial you will need the most current version of R and, 
preferably, RStudio loaded on your computer.

### R Libraries to Install:

* **rhdf5** 
* **raster**
* **rgdal**

[ { { site.baseurl }} R/Packages-In-R/](More on Packages in R - Adapted from Software Carpentry.)

### Download Data

The data sets here is used for all the Data Institute 2016 tutorials. If you are
only completing this tutorial you will only need the TEAK data set. 

{% include/dataSubsets/_data_Data-Institute-16-TEAK.html %}
{% include/dataSubsets/_data_Data-Institute-16-SJER.html %}
{% include/dataSubsets/_data_Data-Institute-16-SOAP.html %}

## Recommended Background 

Consider reviewing the documentation for the <a href="http://www.bioconductor.org/packages/release/bioc/manuals/rhdf5/man/rhdf5.pdf" target="_blank">RHDF5 package</a>.

</div>


First, let's load the required libraries.

```{r load-libraries, warning=FALSE, results='hide', message=FALSE}

# load libraries
library(raster)
library(rhdf5)
library(rgdal)

# set wd
# setwd("~/Documents/data/NEONDI-2016") # Mac
# setwd("~/data/NEONDI-2016")  # Windows

```


## Get Reflectance Matrix Dimensions Function

Here we create a function that pulls the dimensions of the data from a H5 file. 
Note: if we had these as a **numeric** attribute it would be much easier to 
work with.

```{r get-data-dims }

#' Get Data Dimensions ####
#'
#' This function grabs the x,y and z dimemsions of an H5 dataset called "Reflectance"
#' It would be more robust if you could pass it the dataset name / path too
#' @param fileName a path to the H5 file that you'd like to open
#' @keywords hdf5, dimensions
#' @export
#' @examples
#' get_data_dims("filename.h5")

get_data_dims <- function(fileName){
  # make sure everything is closed
  H5close()
  # open the file for viewing
  fid <- H5Fopen(fileName)
  # open the reflectance dataset
  did <- H5Dopen(fid, "Reflectance")
  # grab the dimensions of the object
  sid <- H5Dget_space(did)
  dims <- H5Sget_simple_extent_dims(sid)$size

  # close everything
  H5Sclose(sid)
  H5Dclose(did)
  H5Fclose(fid)
  return(dims)
}


```

## Create Spatial Extent Object Function

Here we create a function that pulls the spatial extent of the data from a H5 
file. 

Note: once again if the xmin/max and ymin/max were in the H5 file as attributes,
this process would be more straight forward. NEON plans to add these attributes 
in the future.

```{r create-function-attrs }

#' Create h5 file extent ####
#'
#' This function uses a map tie point for an h5 file and data resolution to
#' create and return an object of class extent.
#' @param filename the path to the h5 file
#' @param res a vector of 2 objects - x resolution, y resolution
#' @keywords hdf5, extent
#' @export
#' @examples
#' create_extent(fileName, res=c(xres, yres))

create_extent <- function(fileName){
  # Grab upper LEFT corner coordinate from map info dataset
  mapInfo <- h5read(fileName, "map info")

  # create object with each value in the map info dataset
  mapInfo<-unlist(strsplit(mapInfo, ","))
  # grab the XY left corner coordinate (xmin,ymax)
  xMin <- as.numeric(mapInfo[4])
  yMax <- as.numeric(mapInfo[5])
  # get the x and y resolution
  res <- as.numeric(c(mapInfo[2], mapInfo[3]))
  # get dims to use to cal xMax, YMin
  dims <- get_data_dims(f)
  # calculate the xMAX value and the YMIN value
  xMax <- xMin + (dims[1]*res[1])
  yMin <- yMax - (dims[2]*res[2])

  # create extent object (left, right, top, bottom)
  rasExt <- extent(xMin, xMax, yMin, yMax)
  # return object of class extent
  return(rasExt)
}


```



## Clean Reflectance Data Function

Here we create a function that cleans the data from a H5 file by specifying the 
scale factor, ignore values, and returns a properly "projected" raster. 

```{r clean-refl-data-fun }
## FUNCTION - Clean Reflectance Data ####

#' Clean reflectance data
#'
#' This function reads in data from the "Reflecatnce" dataset, applies the data
#' ignore value, scales the data and returns a properly "projected" raster object.
#' @param filename the path to the h5 file.
#' @param reflMatrix , the matrix read in to be converted to a raster.
#' @param epsg - the epsg code for the CRS used to spatially locate the raster.
#' @keywords hdf5, extent
#' @export
#' @examples
#' clean_refl_data(fileName, reflMatrix, epsg)


clean_refl_data <- function(fileName, reflMatrix, epsg){
  # r  get attributes for the Reflectance dataset
  reflInfo <- h5readAttributes(fileName, "Reflectance")
  # grab noData value
  noData <- as.numeric(reflInfo$`data ignore value`)
  # set all values = 15,000 to NA
  reflMatrix[reflMatrix == noData] <- NA

  # apply the scale factor
  reflMatrix <- reflMatrix/(as.numeric(reflInfo$`Scale Factor`))

  # now we can create a raster and assign its spatial extent
  reflRast <- raster(reflMatrix,
                     crs=CRS(paste0("+init=epsg:", epsg)))

  # return a scaled and "cleaned" raster object
  return(reflRast)
}


```

## Read in Band of Choice Fuction

Here we create a function to open a band of choice from a H5 file. This function 
reads in data from the "Reflecatnce" dataset, applies the data ignore value, 
scales the data and returns a properly "projected" raster object.

We have to use the dims() function to grab the dimensions of the matrix. If the 
data were an attribute that would be much easier to access quickly.

```{r read-refl-data }

## FUNCTION - Read Band ####
#' read band
#'
#' This function reads in data from the "Reflecatnce" dataset, applies the data
#' ignore value, scales the data and returns a properly "projected" raster object.
#' @param filename the path to the h5 file.
#' @param index a list formated object  e.g. list(1:3, 1:6, bands)
#' @keywords hdf5, extent
#' @export
#' @examples
#' read_band(fileName, index)

read_band <- function(fileName, index){
  # Extract or "slice" data for band 34 from the HDF5 file
  aBand<- h5read(fileName, "Reflectance", index=index)
  # Convert from array to matrix so we can plot and convert to a raster
  aBand <- aBand[,,1]
  # transpose the data to account for columns being read in first
  # but R wants rows first.
  aBand<-t(aBand)
  return(aBand)
}


```

## Open Band Function

This function opens a band from an NEON H5 file using an input spatial extent. 

```{r function-open-band}
## FUNCTION - Open Band ####
#'
#' This function opens a band from an NEON H5 file using an input spatial extent. 
#' @param fileName the path to the h5 file that you wish to open. 
#' @param bandNum the band number in the reflectance data that you wish to open
#' @param epsg the epsg code for the CRS that the data are in.
#' @param subsetData, a boolean object. default is FALSE. If set to true, then
#' ... subset a slice out from the h5 file. otherwise take the entire xy extent.
#' @param dims, an optional object used if subsetData = TRUE that specifies the 
#' index extent to slice from the h5 file
#' @keywords hdf5, extent
#' @export
#' @examples
#' open_band(fileName, bandNum, epsg, subsetData=FALSE, dims=NULL)
#' 

open_band <- function(fileName, bandNum, epsg){
  # take the specified dims which may be a subset
  # note subtracting one because R indexes all values 1:3 whereas in a zero based system
  # that would yield one more value -- double check on this but it creates the proper
  # resolution
    dims <- get_data_dims(fileName)
    index <- list(1:dims[1], 1:dims[2], bandNum)
    aBand <- read_band(fileName, index)
    # clean data
    aBand <- clean_refl_data(fileName, aBand, epsg)
    extent(aBand) <- create_extent(fileName)
  
  # return raster object
  return(aBand)
}

```

## Create Raster Stack Function

This function calculates an index based subset to slice out data from an H5 file
using an input spatial extent. It returns a rasterStack object of bands.

```{r create-CIR-stack}

## FUNCTION - Open Bands, Create Stack ####
#'
#' This function calculates an index based subset to slice out data from an H5 file
#' using an input spatial extent. It returns a rasterStack object of bands.
#' @param fileName the path to the h5 file that you wish to open.
#' @param bandNum the band number in the reflectance data that you wish to open
#' @param epsg the epsg code for the CRS that the data are in.
#' @param subsetData, a boolean object. default is FALSE. If set to true, then
#' ... subset a slice out from the h5 file. otherwise take the entire xy extent.
#' @param dims, an optional object used if subsetData = TRUE that specifies the
#' index extent to slice from the h5 file
#' @keywords hdf5, extent
#' @export
#' @examples
#' open_band(fileName, bandNum, epsg, subsetData=FALSE, dims=NULL)
#'

#
create_stack <- function(file, bands, epsg, subset=FALSE, dims=NULL){
  # use lapply to run the band function across all three of the bands
  rgb_rast <- lapply(bands, open_band,
                     fileName=file,
                     epsg=epsg)

  # create a raster stack from the output
  rgb_rast <- stack(rgb_rast)
  # reassign band names
  names(rgb_rast) <- bands
  return(rgb_rast)
}
```

## Plot Raster Stack Function 

This function takes a stack and plots it with a title.

```{r plot-stack-function}

plot_stack <- function(aStack, title="3 band RGB Composite", theStretch='lin'){
  # takes a stack and plots it with a title
  # tricker to force the plot title to appear nicely
  # original_par <-par() #original par
  par(col.axis="white", col.lab="white", tck=0)
  # plot the output, use a linear stretch to make it look nice
  plotRGB(aStack,
          stretch=theStretch,
          axes=TRUE,
          main=title)
  box(col="white")
}

```


## Use The Functions

```{r define-data-wd, results='hide' }
# set wd - if you haven't done so already
# setwd("~/Documents/data/NEONDI-2016/")

# define the CRS definition by EPSG code
epsg <- 32611

# define the file you want to work with
f <- "NEONdata/D17-California/TEAK/2013/spectrometer/reflectance/Subset3NIS1_20130614_100459_atmcor.h5"

h5ls(f)
```

### View Wavelengths

View the associated band center in nanometers (um) per band. This is currently 
stored as a dataset.

```{r import-wavelength}
# import the center wavelength in um of each "band"
wavelengths<- h5read(f,"wavelength")

```

### Open and Plot Data

Now we can combine the functions and complete the same tasks as covered in 
<a href="{{ site.baseurl }}/resources/tutorials/hdf5-neon-hyperspec-rs-data" target="_blank"> *Work with Hyperspectral Remote Sensing data in R - HDF5* tutorial</a> 
in many fewer steps. 

```{r open-plot-band }

# find the dimensions of the data to help determine the slice range
# returns cols, rows, wavelengths
dims <- get_data_dims(fileName = f)

# open band, return cleaned and scaled raster
band <- open_band(fileName=f,
                  bandNum = 56,
                  epsg=epsg)

# simple plot of the data
plot(band,
     main="Raster for Lower Teakettle - B56")

```

```{r extract-many-bands}

# extract 3 bands
# create  alist of the bands
bands <- list(58, 34, 19)

# use lapply to run the band function across all three of the bands
rgb_rast <- lapply(bands, open_band,
                   fileName=f,
                   epsg=epsg)

# create a raster stack from the output
rgb_rast <- stack(rgb_rast)

# plot the output, use a linear stretch to make it look nice
plotRGB(rgb_rast,
        stretch='lin')

```

## Plot Various Band Combinations

```{r plot-band-combos}

# CIR create  alist of the bands
bands <- c(90, 34, 19)

CIRStack <- create_stack(f, 
                         bands, 
                         epsg)
plot_stack(CIRStack,
           title="Color Infrared (CIR) Image")

# create a list of the bands
bands <- list(152,90,58)
aStack <- create_stack(f, bands, epsg)
plot_stack(aStack,
           title="another combo")

# FALSE COLOR create a list of the bands
bands <- list(363, 246, 58)
falseStack <- create_stack(f, bands, epsg)
plot_stack(falseStack,
              title="False Color Image")

```

## Export the 3 Band Image as a GeoTIFF

Now that we have a nice 3 band raster stack of our area of interest, let's 
export it as a GeoTIFF.

```{r write-raster, eval=FALSE}

# export as a GeoTIFF
writeRaster(CIRStack,
            file="Outputs/TEAK/cirImage_2013.tif",
            format="GTiff",
            overwrite=TRUE) #caution using overwrite, you can loose data!
```


## Use the NEON AOP package

There is also the 
<a href="https://github.com/lwasser/neon-aop-package" target="_blank"> neonAOP package developed by Leah Wasser </a> 
that contains these functions. Details at the end of this tutorial. 

```{r import-neonAOP}

## import functions
# install devtools, if not previously installed
#install.packages("devtools")

# call devtools library, if you don't have neonAOP installed
#library(devtools)

# install from github, if not previously installed
#install_github("lwasser/neon-aop-package/neonAOP")

# call library
library(neonAOP)

```