---
layout: post
title: "Use NEON AOP Package"
date:   2016-06-18
dateCreated:  2016-05-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
authors: [Leah A. Wasser]
instructors: 
time: "1:00"
contributors: []
packagesLibraries: [neonAOP]
categories: [self-paced-tutorial]
mainTag: institute-day4
tags: [R, HDF5]
tutorialSeries: institute-day4
description: "Short vignettes that showcase the use of the NEON AOP Package."
code1:  /institute-materials/day4_thursday/use-neon-aop-package.R
image:
  feature: 
  credit: 
  creditlink:
permalink: /R/use-neon-aop-package/
comments: true
---

First, let's load the required libraries.

```{r load-libraries, warning=FALSE, results='hide', message=FALSE}

## import R library from github
# install devtools (if you have not previously intalled it)
# install.packages("devtools")
# call devtools library
# library(devtools)

# install package from github
# install_github("lwasser/neon-aop-package/neonAOP")
# call package
library(neonAOP)

# load packages
library(raster, rgeos, rgdao)
library(rhdf5)
library(ggplot2)

# set wd
# setwd("~/Documents/data/NEONDI-2016") # Mac
# setwd("~/data/NEONDI-2016")  # Windows

```


## Open a Single Band

```{r import-band }

# define the CRS definition by EPSG code
epsg <- 32611

# define the file you want to work with
f <- "NEONdata/D17-California/TEAK/2013/spectrometer/reflectance/Subset3NIS1_20130614_100459_atmcor.h5"

# open band, return cleaned and scaled raster
band56 <- open_band(fileName=f,
                  bandNum = 56,
                  epsg=epsg)

# plot data
plot(band56,
     main="Raster for Lower Teakettle - B56")

```


## Open Several Bands / Create Stack

```{r create-stack }

# extract 3 bands
# create  alist of the bands
# RGB Combo
bands <- c(58, 34, 19)
RGBStack <- create_stack(f, 
                         bands, 
                         epsg)
# plot the stack
plotRGB(RGBStack, 
        stretch='lin')


# CIR create  alist of the bands
bands <- c(90, 34, 19)

CIRStack <- create_stack(f, 
                         bands, 
                         epsg)
# plot the stack
plotRGB(CIRStack, 
        stretch='lin')

```

## Try some other band combos:

* `bands <- c(152,90,58)`
* FALSE COLOR: `bands <- c(363, 246, 58)`

Next, let's extract a spatial SUBSET from a H5 file.

```{r extract-subset }


# open file clipping extent
clip.extent <- readOGR("NEONdata/D17-California/TEAK/vector_data", 
                       "TEAK_plot")

# calculate extent of H5 file
h5.ext <- create_extent(f)
h5.ext

# turn the H5 extent into a polygon to check overlap
h5.ext.poly <- as(extent(h5.ext), 
                  "SpatialPolygons")

crs(h5.ext.poly) <- crs(clip.extent)

# test to see that the extents overlap
gIntersects(h5.ext.poly, 
            clip.extent)

# Use the clip extent to create the index extent that can be used to slice out data from the 
# H5 file
# xmin.index, xmax.index, ymin.index, ymax.index
# all units will be rounded which means the pixel must occupy a majority (.5 or greater)
# within the clipping extent
index.bounds <- calculate_index_extent(extent(clip.extent),
								h5.ext)
index.bounds


# open a band that is subsetted using the clipping extent
# if you set subsetData to true, then provide the dimensions that you wish to slice out
b58_clipped <- open_band(fileName=f,
								bandNum=58,
								epsg=32611,
								subsetData = TRUE,
								dims=index.bounds)

# plot clipped bands
plot(b58_clipped,
     main="Band 58 Clipped")

```


## Extract a Subset from a Stack

```{r extract-stack }

# create  alist of the bands
bands <- c(19,34,58)

# clip out raster
rgbRast.clip <- create_stack(file=f,
            bands=bands,
            epsg=epsg,
            subset=TRUE,
            dims=index.bounds)

plotRGB(rgbRast.clip,
        stretch="lin")


rgbRast <- create_stack(file=f,
              bands=bands,
              epsg=epsg,
              subset=FALSE)

plotRGB(rgbRast,
        stretch="lin")

plot(clip.extent,
     add=T,
     border="yellow",
     lwd=3)


```

## Extract Spectra

```{r extract-spectra }


# read in the wavelength information from the HDF5 file
wavelengths <- h5read(f, "wavelength")
head(wavelengths)

# NOTE: this currently doesn't work properly if the file is rotated
h5.ext <- create_extent(f)

# calculate the index subset dims to extract data from the H5 file
subset.dims <- calculate_index_extent(clip.extent, 
                    h5.ext, 
                    xscale = 1, yscale = 1)


# turn the H5 extent into a polygon to check overlap
h5.ext.poly <- as(extent(h5.ext),
                "SpatialPolygons")

# assign crs to new polygon
crs(h5.ext.poly) <- CRS("+proj=utm +zone=11 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")

# ensure the two extents overlap
gIntersects(h5.ext.poly, 
            clip.extent)

# finally determine the subset to extract from the h5 file
index.bounds <- calculate_index_extent(extent(clip.extent), 
                h5.ext)

# open a band
a.raster <- open_band(f, 
                      56, 
                      epsg)

# grab the average reflectance value for a band
refl <- extract_av_refl(a.raster, 
                        aFun = mean)

refl
```


## Get Spectra from Many Bands

```{r get-spectra-bands }

# grab all bands
bands <- c(1:426)

# get stack
all.bands.stack <- create_stack(f, 
                         bands, 
                         epsg)

# get spectra for each band
spectra <- extract_av_refl(all.bands.stack, 
                           aFun = mean)
spectra <- as.data.frame(spectra)

# read in the wavelength information from the HDF5 file
wavelengths<- h5read(f, "wavelength")

# convert wavelength to nanometers (nm)
wavelengths <- wavelengths * 1000

spectra$wavelength <- wavelengths[bands]

# plot spectra
qplot(x=spectra$wavelength,
      y=spectra$spectra,
      xlab="Wavelength (nm)",
      ylab="Reflectance",
      main="Spectra for all pixels",
      ylim = c(0, .35))

```


## Get Spectra From Masked Bands Only

In this instance, we are using a mask to extract only the pixels where NDVI > .6.
This represents vegetation.

```{r spectra-masked-bands }

index.bounds <- calculate_index_extent(extent(clip.extent),
								h5.ext)
head(bands)
epsg

# clip out raster
rgbRast.clip <- create_stack(file=f,
            bands=bands,
            epsg=epsg,
            subset=TRUE,
            dims=index.bounds)

# calculate NDVI
redBand <- rgbRast.clip[[60]]
nirBand <- rgbRast.clip[[83]]

ndvi <- (nirBand - redBand) / (nirBand + redBand)
plot(ndvi)
# create mask
ndvi[ndvi < .6] <- NA
plot(ndvi)

# mask stack
rgbRast.clip.mask <- mask(rgbRast.clip, ndvi)

# get spectra for each band
spectra.mask <- extract_av_refl(rgbRast.clip.mask, 
                           aFun = mean)

spectra.mask <- as.data.frame(spectra.mask)

# read in the wavelength information from the HDF5 file
wavelengths<- h5read(f, "wavelength")

# convert wavelength to nanometers (nm)
wavelengths <- wavelengths * 1000

spectra.mask$wavelength <- wavelengths[bands]

# plot spectra
qplot(x=spectra.mask$wavelength,
      y=spectra.mask$spectra,
      xlab="Wavelength (nm)",
      ylab="Reflectance",
      main="Spectra for pixels NDVI> .6",
      ylim = c(0, .35))

```

