---
syncID: 
title: "Working with NEON Site Management Data"
description: "Access, work with, and navigate data from Site management and event reporting (DP1.10111.001)."
dateCreated: 2025-09-04
authors: Claire K. Lunch
contributors: 
estimatedTime: 40 minutes
packagesLibraries: neonUtilities, geoNEON
topics: 
languageTool: R
dataProduct: 
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/NEON-general/neon-overview/site-management/site-management.R
tutorialSeries:
urlTitle: site-management-tutorial

---


This tutorial will guide you through accessing, navigating, and working with the NEON data product Site management and event reporting (DP1.10111.001), including how to use code tools in the `neonUtilities` and `geoNEON` R packages designed specifically for this data product.

If you are new to using NEON data, we recommend first following the <a href="https://www.neonscience.org/resources/learning-hub/tutorials/download-explore-neon-data" target="_blank">Download and Explore</a> tutorial to get familiar with NEON data formats and the `neonUtilities` package.

<div id="ds-objectives" markdown="1">

## Objectives

After completing this activity, you will be able to:

 * Download site event data by site or by event type
 * Access data related to the spatial extent of an event
 * Understand the metadata associated with site events

## Things You’ll Need To Complete This Tutorial

To complete this tutorial, you will need a version of R (4.0 or higher) and, 
preferably, RStudio loaded on your computer. This code may work with earlier 
versions but it hasn't been tested.

### Install R Packages

* **neonUtilities:** `install.packages("neonUtilities")`
* **devtools:** `install.packages("devtools")`
* **geoNEON:** `devtools::install_github("NEONScience/NEON-geolocation/geoNEON")`

## Additional Resources

* Site management and event reporting <a href="https://data.neonscience.org/data-products/DP1.10111.001" target="_blank">details page</a>

</div>


## Site management and event reporting: Background

Events often occur at NEON sites that have an impact on the local ecosystems, but that are not specifically captured in scheduled and planned NEON data collection. These events may be planned or unplanned - for example, controlled burns, wildfires, grazing rotations, and hurricanes. The Site management and event reporting data product is designed to record information about these events. Data are recorded on an ad hoc basis, driven by the occurrence of relevant events.

#### Event duration and recording

In order to match the publication model for NEON data products and to ensure events are published as quickly as possible, site management and event reporting data are recorded in one-month intervals. This means events longer than one month will have multiple data records for one `eventID`. The `eventID`, which includes the site code, original start date, and event type, will remain the same for all records related to the same event. The original start date embedded in the `eventID` compared to the `startDate` field for the record indicates if there are related records for previous months. The `ongoingEvent` field indicates if a record for the next subsequent month is expected.

Date uncertainty is captured in a record’s `estimatedOrActualDate` field. If the date is `estimated` additional comments are recorded in the `dateRemarks` field. The start and end date range includes any date uncertainty.

## Set up R environment

First install and load the necessary packages. Skip the install step if packages 
are already installed, and modify the working directory file path for your 
computer.

```{r setup, eval=FALSE}

install.packages("neonUtilities")
install.packages("devtools")
devtools::install_github("NEONScience/NEON-geolocation/geoNEON")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("lubridate")

library(neonUtilities)
library(geoNEON)
library(ggplot2)
library(dplyr)
library(lubridate)

setwd("~/data")

```

```{r libraries, include=FALSE, results="hide"}

library(neonUtilities)
library(geoNEON)
library(ggplot2)
library(dplyr)
library(lubridate)

```


## Downloading event data: All events at a site

The simplest way to access site event data is to download all the events for a particular site. Let's check out events at Jones Ecological Research Center in Georgia (JERC). We'll use `include.provisional=T` to get events up to the present.

``` {r load-data, results="hide"}

events <- loadByProduct(dpID="DP1.10111.001", 
                        site="JERC",
                        check.size=F,
                        include.provisional=T)

```

The download includes the usual metadata files associated with NEON data, and a single data table: `sim_eventData`. Let's take a look at it.

``` {r view-data}

list2env(events, .GlobalEnv)
head(sim_eventData)

```

Let's first check out the types of events recorded.

``` {r view-types}

unique(sim_eventData[,c("eventType", "methodTypeChoice")])

```

JERC is a managed site, so it's not surprising that many of the events recorded represent active management by the site hosts. When we look at an event like a controlled burn, we probably want to know exactly where this happened - which NEON sampling areas were affected?

## Finding the spatial extent of an event

How to approach the spatial distribution of an event depends on your scientific question. If you're working with data from NEON sampling plots, you might just want to know which plots are affected. You can find that information in the `locationID` field, which contains either one location or a comma-delimited list of locations for each record. Let's look at the locations associated with controlled burns in 2022.

``` {r herb-locs}

burn <- sim_eventData |>
  filter(methodTypeChoice=="fire-controlledBurn" & 
           year(endDate)=="2022")

burn.locs <- unique(trimws(unlist(lapply(burn$locationID, 
                                         strsplit, ",", 
                                         fixed=T))))
burn.locs

```

Now we have a list of NEON named locations. We can see if the plot(s) we are working with are on the list, and proceed accordingly.

But what if we're interested in a broader picture? Maybe we're working with remote sensing data, and we want to get a sense of the overall burned area.

### Stay tuned! There is an R package in the works that converts these plot locations to a shapefile

<div id="caveat-box" markdown="1">

**Caveats about locations:** 

NEON field staff do their best to record the extent of a disturbance in the list of locations. In the case of planned disturbances, such as controlled burns and release of animals for grazing, the site host may be able to provide a very clear assessment of the locations affected. But for unplanned events such as wildfires or bark beetle outbreaks, these data depend on staff being in the right place at the right time to observe the disturbance. At some NEON sites, sampling plots are very far apart, and if staff are on-site for a particular protocol and only sampling a subset of plots, they may not know the impact of disturbance on other plots.

These limitations primarily affect terrestrial sites. At aquatic sites, sampling locations are clustered in a fairly small area, and staff can more easily assess the extent of a disturbance. In addition, due to the close proximity of sampling, most significant events affect the entire aquatic sampling area.

Finally, we have limited information available about the impact of events outside of NEON sampling locations. A patchy event like a small wildfire may affect locations very unevenly. NEON AOP flights typically cover a 10 x 10 km box that includes the sampling locations, but also includes large areas where NEON does not sample. If you are analyzing a disturbance using remote sensing data, the locations identified in this data product may only be a small subset of the disturbed area within the AOP flight box.

</div>

## Downloading event data by event type

In many cases, you may be interested not in the full range of disturbances at a given site, but in a particular type of disturbance across sites. For example, if you are studying fires in the western US, you probably know that NEON sites have burned, but you may not know the specific fires at specific sites. Let's look for all the fires that have affected NEON sites in California. We can do this using the `byEventSIM()` function:

``` {r load-CA-fires, results="hide"}

fires <- byEventSIM(eventType="fire", 
                    site=c("SJER","SOAP","BIGC","TEAK","TECR"),
                    include.provisional=T)

```

The set of files returned by this function is the same as `loadByProduct()`, but the data in `sim_eventData` are filtered down to only the events where the `eventType` is `"fire"`.

Let's start by looking at the `siteID`, `startDate`, `endDate`, and `methodTypeChoice` fields to see which sites and years experienced fires.

``` {r fire-dates}

fires$sim_eventData |>
  select(siteID, startDate, endDate, methodTypeChoice)

```

We know 2020 was a terrible year for wildfires in the western US, and unsurprisingly many of the records here are from 2020. But there is also one controlled burn from early 2020, as well as fires in 2021 and 2025.

Let's use geoNEON to map these burned locations on top of AOP imagery

