---
layout: post
title: "Create a Canopy Height Model from LiDAR-derived Rasters in R"
date:   2014-7-18
createdDate:   2014-07-21
lastModified:`r format(Sys.time(), "%Y-%m-%d")`
estimatedTime: 1.0 - 1.5 Hours
packagesLibraries: [raster, sp, dplyr, maptools, rgeos]
authors: Edmund Hart, Leah A. Wasser
categories: [self-paced-tutorial]
mainTag: intro-lidar-r-series
tags: [lidar, R, raster, remote-sensing, spatial-data-gis]
tutorialSeries: [intro-lidar-r-series]
description: "In this tutorial, you will bring LiDAR-derived raster data 
(DSM and DTM) into R to create a  canopy height model (CHM). "
permalink: /lidar-data/lidar-data-rasters-in-R/
comments: true
code1: /R/intro-lidar-r/Create-Lidar-CHM-R.R
image:
  feature: textur2_pointsProfile.png
  credit: National Ecological Observatory Network (NEON)
  creditlink: http://www.neonscience.org
---

A common analysis using LiDAR data is to derive top of the canopy height values 
from the LiDAR data. These values are often used to track changes in forest 
structure over time, to calculate biomass, and even leaf area index (LAI). Let's 
dive into the basics of working with raster formatted LiDAR data in R! 

<div id="objectives" markdown="1">

# Goals/Objectives
After completing this activity, you will be able to:

* Work with digital terrain model (DTM) & digital surface model (DSM) raster files. 
* Create a canopy height model (CHM) raster from DTM & DSM rasters. 

 
## Things You’ll Need To Complete This Tutorial
You will need the most current version of `R` and, preferably, `RStudio` loaded 
on your computer to complete this tutorial.

### Install R Packages

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`

[More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}/R/Packages-In-R/)

## Download Data
{% include/dataSubsets/_data_Field-Site-Spatial-Data.html %}

This tutorial is designed for you to set your working directory to the directory
created by unzipping this file.

****

{% include/_greyBox-wd-rscript.html %}

***

## Recommended Reading
<a href="{{ site.baseurl }}/remote-sensing/2_LiDAR-Data-Concepts_Activity2/">
What is a CHM, DSM and DTM? About Gridded, Raster LiDAR Data</a>

</div>

## Create a LiDAR derived Canopy Height Model (CHM)

The National Ecological Observatory Network (NEON) will provide LiDAR-derived 
data products as one of its many free ecological data products. These products 
will come in the 
[GeoTIFF](http://trac.osgeo.org/geotiff/ "geotiff (read more)") 
format, which is a .tif raster format that is spatially located on the earth. 

In this tutorial, we create a Canopy Height Model. The 
[canopy height model (CHM)]({{ site.baseurl }}/remote-sensing/2_LiDAR-Data-Concepts_Activity2/),
represents the heights of the trees on the ground. We can derive the CHM 
by subtracting the ground elevation from the elevation of the top of the surface 
(or the tops of the trees). 

We will use the `raster` R package to work with the the lidar derived digital 
surface model (DSM) and the digital terrain model (DTM). 

```{r set-up}

# Load needed packages
library(raster)
library(rgdal)

# set working directory to data folder
#setwd("pathToDirHere")
```

First, we will import the Digital Surface Model (DSM). The 
[DSM]({{ base.url }}/remote-sensing/2_LiDAR-Data-Concepts_Activity2/)
represents the elevation of the top of the objects on the ground (trees, 
buildings, etc).

``` {r import-dsm}

# assign raster to object
dsm <- raster("SJER/DigitalSurfaceModel/SJER2013_DSM.tif")

# view info about the raster.
dsm

# plot the DSM
plot(dsm, main="LiDAR Digital Surface Model \n SJER, California")

```

Note the resolution, extent, and coordinate reference system (CRS) of the raster. 
To do later steps, our DTM will need to be the same. 

Next, we will import the Digital Terrain Model (DTM) for the same area. The 
[DTM]({{ base.url }} /remote-sensing/2_LiDAR-Data-Concepts_Activity2/)
represents the ground (terrain) elevation.

```{r plot-DTM }

# import the digital terrain model
dtm <- raster("SJER/DigitalTerrainModel/SJER2013_DTM.tif")

plot(dtm, main="LiDAR Digital Terrain Model \n SJER, California")

```

With both of these rasters now loaded, we can create the Canopy Height Model 
(CHM). The 
[CHM ]({{ base.url }} /remote-sensing/2_LiDAR-Data-Concepts_Activity2/)
represents the difference between the DSM and the DTM or the height of all objects
on the surface of the earth. 

To do this we perform some basic raster math to calculate the CHM. You can 
perform the same raster math in a GIS program like 
[QGIS](http://www.qgis.org/en/site/ "QGIS").

When you do the math, make sure to subtract the DTM from the DSM or you'll get 
trees with negative heights!

```{r calculate-plot-CHM }

# use raster math to create CHM
chm <- dsm - dtm

# view CHM attributes
chm

plot(chm, main="LiDAR Canopy Height Model \n SJER, California")

```

We've now created a CHM from our DSM and DTM. What do you notice about the 
canopy cover at this location in the San Joaquin Experimental Range? 

<div id="challenge" markdown="1">
## Challenge: Basic Raster Math 

Convert the CHM from meters to feet. Plot it. 
</div>

``` {r challenge-code-raster-math, include=TRUE, results="hide", echo=FALSE}
# conversion 1m = 3.28084 ft
chm_ft <- chm*3.28084

# plot 
plot(chm_ft, main="LiDAR Canopy Height Model \n in feet")

```

If, in your work you need to create lots of CHMs from different rasters, an 
efficient way to do this would be to create a function to create your CHMs. 

``` {r canopy-function}
# Create a function that subtracts one raster from another
# 
canopyCalc <- function(DTM, DSM) {
  return(DSM -DTM)
  }
    
# use the function to create the final CHM
chm2 <- canopyCalc(dsm,dtm)
chm2

# or use the overlay function
chm3 <- overlay(dsm,dtm,fun = canopyCalc) 
chm3 

```

As with any raster, we can write out the CHM as a GeoTiff using the 
`writeRaster()` function. 

```{r write-raster-to-geotiff, eval=FALSE }
# write out the CHM in tiff format. 
writeRaster(chm,"chm_SJER.tiff","GTiff")

```

We've now successfully created a canopy height model using basic raster math -- in 
R! We can bring the `chm_SJER.tiff` file into QGIS (or any GIS program) and look 
at it. 

***

Consider going onto the next tutorial 
<a href="{{ site.baseurl }}/R/extract-raster-data-R/" target="_blank">*Extract Values from a Raster in R*> to compare this lidar-derived CHM with data taken from in ground!

