---
syncID:
title: "Combining LiDAR pointclouds with observed vegetation structure"
description: "Align Vegetation Structure data with NEON LiDAR pointclouds in three-dimensional space. Identify different tree species in the pointcloud using the mapping and tagging dataset."
dateCreated: 2020-09-24
authors: Donal O'Leary
contributors: Maxwell Burner
estimatedTime: 1 Hour
packagesLibraries: neonUtilities, lidR, rgl, raster, geoNEON
topics: LiDAR
languagesTool: R
dataProduct: DP1.10098.001,DP1.30003.001
code1: 
tutorialSeries: 
urlTitle: 
---

This template should be used to start new tutorials. If your tutorial is in R 
or Jupyter Notebooks, or another format that will be converted to markdown. 
Please include the header above but create the tutorial in the appropriate 
application. 

All tutorials then have an Info Box with the information necessary for a new user
to NEON tutorials to complete the tutorial.  

<div id="ds-objectives" markdown="1">

## Learning Objectives 
After completing this tutorial you will be able to: 

* Create a new NEON Data Skills teaching module. 

## Things You’ll Need To Complete This Tutorial

### R Programming Language
You will need a current version of R to complete this tutorial. We also recommend 
the RStudio IDE to work with R. 

### R Packages to Install
Prior to starting the tutorial ensure that the following packages are installed. 
* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`
* **sp:** `install.packages("sp")`

<a href="/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

### Example Data Set

This tutorial will teach you how to download data directly from the NEON data 
portal.  This example data set is provided as an optional download and a backup. 

#### NEON Teaching Data Subset: Plant Phenology & Single Aspirated Air Temperature

The data used in this tutorial were collected at the 
<a href="http://www.neonscience.org" target="_blank"> National Ecological Observatory Network's</a> 
<a href="/field-sites/field-sites-map" target="_blank"> Domain 02 field sites</a>.  

* National Ecological Observatory Network. 2020. Data Product DP1.10055.001, Plant phenology observations. Provisional data downloaded from http://data.neonscience.org on February 5, 2020. Battelle, Boulder, CO, USA NEON. 2020.
* National Ecological Observatory Network. 2020. Data Product DP1.00002.001, Single aspirated air temperature. Provisional data downloaded from http://data.neonscience.org on February 5, 2020. Battelle, Boulder, CO, USA NEON. 2020.

<a href="https://ndownloader.figshare.com/files/9012085" class="link--button link--arrow">
Download Dataset</a>

### Working Directory
[Donal to add text]

## Additional Resources
You can list any additional resources you'd like to link to if needed. 

* <a href="URL Here" target="_blank"> Resource Name/a>,
 
</div>

```{r load-packages, warning=FALSE, message=FALSE}

library(neonUtilities)
library(geoNEON)
library(lidR)
library(sf)
library(rgl)
library(sp)
library(raster)

options(stringsAsFactors=F) #This line may be redundant for R version 4.0.0+

```

Choose your site of interest:

```{r select-site}

SITE="WREF"

```

First load veg structure dataset. For a detailed explanation of these steps, please see download-explore tutorial **LINK HERE**

```{r load-veg-str}

veglist <- loadByProduct(dpID="DP1.10098.001", site=SITE, check.size = F, package="basic")

# retrieve precise XY locations using geoNEON API
vegmap <- getLocTOS(veglist$vst_mappingandtagging, 
                    "vst_mappingandtagging")

## Merge tables
veg <- base::merge(veglist$vst_apparentindividual, vegmap, 
             by=c('individualID','namedLocation',
                  'domainID','siteID','plotID'))

```


## Download LiDAR and load into R

**ADD HERE** link to download this directly using the API

```{r load-LiDAR}


SITE_LAS=readLAS("~/Downloads/DP1.30003.001/2019/FullSite/D16/2019_WREF_1/L1/DiscreteLidar/ClassifiedPointCloud/NEON_D16_WREF_DP1_580000_5075000_classified_point_cloud_colorized.laz")
plot(SITE_LAS)

```

Oof, there are a LOT of extreneous points above and below the forest. Fortunately, there is an easy way to filter out points above and below specified elevations (on the Z-axis). Let's take a look at a histogram of point elevations to see where to filter out the extreneous points.

```{r make-LiDAR-histogram}

hist(SITE_LAS@data$Z)

```

We can see here that most points fall between approximately 1000 and 1300 meters. Therefore, we will use the `filter` argument of the `lidR::readLAS()` function to filter out the points above and below the 'bulk' of the points.

There are also many unneeded attributes assigned to each XYZ point in the LAS object. A LAS object stores the point cloud data in a table, so you can see all of the attributes using the `names()` functions:

```{r read-LAS-attributes}

names(SITE_LAS@data)

```

For this tutorial, we only need the "X" "Y" "Z" "R" "G" and "B" attributes (so that we can use "RGB" colors when visualizing later). We can also specify this subset using the `select` attribute in the `readLAS()` function. see `?readLAS()` for more details.

```{r load-LiDAR-with-filter}

SITE_LAS=readLAS("~/Downloads/DP1.30003.001/2019/FullSite/D16/2019_WREF_1/L1/DiscreteLidar/ClassifiedPointCloud/NEON_D16_WREF_DP1_580000_5075000_classified_point_cloud_colorized.laz",
                 filter = "-drop_z_below 350 -drop_z_above 550",
                 select = "xyzRGB")

```

**DATA TIP**
Finding the best elevations for the upper and lower filter limits can be a challenge, especially in mountainous terrain. If you filter too aggressively in the valleys, the trees on the hillsides might get chopped off! You may need to try a few different options to find your best result. In this case, we could remove many more extreneous points by first clipping the LAS object (SITE_LAS) to a smaller area, and then filtering specifically on that data subset.

Now that we have the LiDAR data subset loaded into a LAS object, let's plot the pointcloud while taking advantage of the "RGB" colorization:

```{r plot-LiDAR-subset}

x=plot(SITE_LAS, color="RGB", size=2)

```

So why did we assign the `plot(SITE_LAS)` function to a new variable, `x`? Because when you `plot()` a LAS object, this function returns the lower left corner cooridnates for the plot. This will become important when we start plotting other objects into this three-dimensional space.

from lasR::plot() documentation:
"clear_artifacts	[argument]
logical. It is a known and documented issue that the 3D visualisation with rgl displays artifacts. The points look aligned and/or regularly spaced in some view angles. This is because rgl computes with single precision float. To fix that the point cloud is shifted to (0,0) to reduce the number of digits needed to represent its coordinates. The drawback is that the point cloud is not plotted at its actual coordinates."
```{r add-base-plots-to-LiDAR}
x=plot(SITE_LAS, color="RGB", size=2)
setwd("~/Downloads/")
NEON_all_plots <- st_read('All_NEON_TOS_Plots_V8/All_NEON_TOS_Plot_Polygons_V8.shp')
base_plots_SPDF <- NEON_all_plots[(NEON_all_plots$siteID == SITE)&(NEON_all_plots$subtype == 'basePlot'),]

rm(NEON_all_plots)

#Plot base plots
#plot(base_plots_SPDF$geometry, border = 'blue')

base_plots_SPDF <- st_transform(base_plots_SPDF, as.character(SITE_LAS@proj4string))

base_crop=st_crop(base_plots_SPDF, extent(SITE_LAS))


###

coords=as.data.frame(st_coordinates(base_crop))
coords$X=coords$X-x[1]
coords$Y=coords$Y-x[2]
coords$Z=rep(base_crop$elevation, each=5)
c=1:nrow(coords)
coords=coords[!c%%5==0,]


for(i in 1:(nrow(coords)/4)){
  r=((i-1)*4)+1
  quads3d(x=coords$X[r:(r+3)], y=coords$Y[r:(r+3)], z=coords$Z[r:(r+3)]+50,
          add=T, col="red", lwd=2, lit=F)
}

for(i in 1:(nrow(coords)/4)){
  r=((i-1)*4)+1
  rgl.texts(x=coords$X[r], y=coords$Y[r], z=coords$Z[r]+50,
            text = base_crop$plotID[i],
          add=T, col="red", cex=1, lit=F)
}
```


```{r highlight-focus-plots}

for(i in c(3,4)){
  r=((i-1)*4)+1
  quads3d(x=coords$X[r:(r+3)], y=coords$Y[r:(r+3)], z=coords$Z[r:(r+3)]+50,
          add=T, col="blue", lwd=2, lit=F, alpha=.3)
}

# for(i in c(3,4)){
#   r=((i-1)*4)+1
#   rgl.texts(x=coords$X[r], y=coords$Y[r], z=coords$Z[r]+51,
#             text = base_crop$plotID[i],
#           add=T, col="blue", cex=1, lit=F)
# }

```

```{r clip-LAS-to-plots}


top_left=as.data.frame(st_coordinates(base_crop))
c=1:nrow(top_left)
top_left=top_left[c%%5==1,]

plots_LAS <- 
  lasclipRectangle(SITE_LAS,
                   xleft = (top_left$X), ybottom = (top_left$Y - 40),
                   xright = (top_left$X + 40), ytop = (top_left$Y))


plot(plots_LAS[[1]], color="RGB")


```


```{r plot-level-trees-generic}


## Loop through plots and add color boxes for each tree
#for(i in 1:length(plots_LAS)){
for(i in c(3,4)){
x=plot(plots_LAS[[i]], color="RGB", size=2)

tree_boxes=veg[veg$plotID==base_crop$plotID[i],]
tree_boxes=tree_boxes[!is.na(tree_boxes$adjElevation),]
tree_boxes$adjEasting=tree_boxes$adjEasting-x[1]
tree_boxes$adjNorthing=tree_boxes$adjNorthing-x[2]

# for(j in 1:(nrow(tree_boxes))){
#   print(j)
#   d=as.data.frame(tree_boxes[j,c("adjNorthing","adjEasting","adjElevation")])
#   d[,3]=as.numeric(d[,3])
#   d[2:4,]=d[1,]
#   d[1,1]=d[1,1]+2
#   d[1,2]=d[1,2]-2
#   
#   d[2,1]=d[2,1]+2
#   d[2,2]=d[2,2]+2
#   
#   d[3,1]=d[3,1]-2
#   d[3,2]=d[3,2]+2
#   
#   d[4,1]=d[4,1]-2
#   d[4,2]=d[4,2]-2
#   
#   d[,3]=d[,3]+tree_boxes[j,"height"]
#   
#   quads3d(x=d$adjEasting, y=d$adjNorthing, z=d$adjElevation, 
#           add=T, col="orange", lwd=2, lit=F)
# }

Thuja_boxes=tree_boxes[tree_boxes$taxonID=="TSHE",]

PSMEM_boxes=tree_boxes[tree_boxes$taxonID=="PSMEM",]


for(j in 1:(nrow(PSMEM_boxes))){
  print(j)
  d=as.data.frame(PSMEM_boxes[j,c("adjNorthing","adjEasting","adjElevation")])
  d[,3]=as.numeric(d[,3])
  d[2:4,]=d[1,]
  d[1,1]=d[1,1]+2
  d[1,2]=d[1,2]-2

  d[2,1]=d[2,1]+2
  d[2,2]=d[2,2]+2

  d[3,1]=d[3,1]-2
  d[3,2]=d[3,2]+2

  d[4,1]=d[4,1]-2
  d[4,2]=d[4,2]-2

  d[,3]=d[,3]+PSMEM_boxes[j,"height"]

  quads3d(x=d$adjEasting, y=d$adjNorthing, z=d$adjElevation,
          add=T, col="green", lwd=2, lit=F)
}


for(k in 1:(nrow(Thuja_boxes))){
  print(k)
  d=as.data.frame(Thuja_boxes[k,c("adjNorthing","adjEasting","adjElevation")])
  d[,3]=as.numeric(d[,3])
  d[2:4,]=d[1,]
  d[1,1]=d[1,1]+2
  d[1,2]=d[1,2]-2

  d[2,1]=d[2,1]+2
  d[2,2]=d[2,2]+2

  d[3,1]=d[3,1]-2
  d[3,2]=d[3,2]+2

  d[4,1]=d[4,1]-2
  d[4,2]=d[4,2]-2

  d[,3]=d[,3]+Thuja_boxes[k,"height"]

  quads3d(x=d$adjEasting, y=d$adjNorthing, z=d$adjElevation,
          add=T, col="red", lwd=2, lit=F)
}

}

```



```{r plot-level-trees}


## Loop through plots and add color boxes for each tree
#for(i in 1:length(plots_LAS)){
for(i in c(3,4)){
x=plot(plots_LAS[[i]], color="RGB", size=2)

tree_boxes=veg[veg$plotID==base_crop$plotID[i],]
tree_boxes=tree_boxes[!is.na(tree_boxes$adjElevation),]
tree_boxes$adjEasting=tree_boxes$adjEasting-x[1]
tree_boxes$adjNorthing=tree_boxes$adjNorthing-x[2]

# for(j in 1:(nrow(tree_boxes))){
#   print(j)
#   d=as.data.frame(tree_boxes[j,c("adjNorthing","adjEasting","adjElevation")])
#   d[,3]=as.numeric(d[,3])
#   d[2:4,]=d[1,]
#   d[1,1]=d[1,1]+2
#   d[1,2]=d[1,2]-2
#   
#   d[2,1]=d[2,1]+2
#   d[2,2]=d[2,2]+2
#   
#   d[3,1]=d[3,1]-2
#   d[3,2]=d[3,2]+2
#   
#   d[4,1]=d[4,1]-2
#   d[4,2]=d[4,2]-2
#   
#   d[,3]=d[,3]+tree_boxes[j,"height"]
#   
#   quads3d(x=d$adjEasting, y=d$adjNorthing, z=d$adjElevation, 
#           add=T, col="orange", lwd=2, lit=F)
# }

Thuja_boxes=tree_boxes[tree_boxes$taxonID=="TSHE",]

PSMEM_boxes=tree_boxes[tree_boxes$taxonID=="PSMEM",]


for(j in 1:(nrow(PSMEM_boxes))){
  print(j)
  d=as.data.frame(PSMEM_boxes[j,c("adjNorthing","adjEasting","adjElevation")])
  d[,3]=as.numeric(d[,3])
  d[2:4,]=d[1,]
  d[1,1]=d[1,1]+2
  d[1,2]=d[1,2]-2

  d[2,1]=d[2,1]+2
  d[2,2]=d[2,2]+2

  d[3,1]=d[3,1]-2
  d[3,2]=d[3,2]+2

  d[4,1]=d[4,1]-2
  d[4,2]=d[4,2]-2

  d[,3]=d[,3]+PSMEM_boxes[j,"height"]

  quads3d(x=d$adjEasting, y=d$adjNorthing, z=d$adjElevation,
          add=T, col="green", lwd=2, lit=F)
}


for(k in 1:(nrow(Thuja_boxes))){
  print(k)
  d=as.data.frame(Thuja_boxes[k,c("adjNorthing","adjEasting","adjElevation")])
  d[,3]=as.numeric(d[,3])
  d[2:4,]=d[1,]
  d[1,1]=d[1,1]+2
  d[1,2]=d[1,2]-2

  d[2,1]=d[2,1]+2
  d[2,2]=d[2,2]+2

  d[3,1]=d[3,1]-2
  d[3,2]=d[3,2]+2

  d[4,1]=d[4,1]-2
  d[4,2]=d[4,2]-2

  d[,3]=d[,3]+Thuja_boxes[k,"height"]

  quads3d(x=d$adjEasting, y=d$adjNorthing, z=d$adjElevation,
          add=T, col="red", lwd=2, lit=F)
}

}

```