---
syncID: 5defe34482a742b589ea841d0c3adeeb
title: "Vector 02: Plot Multiple Shapefiles and Create Custom Legends in Base Plot in R"
description: "This tutorial provides an overview of how to create a a plot of multiple shapefiles using base R plot. It also explores adding a legend with custom symbols that match your plot colors and symbols."
dateCreated:  2016-02-09
authors: Joseph Stachelek, Leah A. Wasser, Megan A. Jones
contributors: Sarah Newman, Maya R. Stahl
estimatedTime: 30 minutes
packagesLibraries: rgdal, raster
topics: vector-data, spatial-data-gis
languagesTool: R
dataProduct:
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/Geospatial-skills/intro-vector-r/02-plot-multiple-shapefiles-custom-legend/02-plot-multiple-shapefiles-custom-legend.R
tutorialSeries: vector-data-series
urlTitle: dc-plot-shapefiles-r
---


This tutorial builds upon 
<a href="https://www.neonscience.org/dc-shapefile-attributes-r" target="_blank">*the previous tutorial*</a>,
to work with shapefile attributes in R and explores how to plot multiple 
shapefiles using base R graphics. It then covers
how to create a custom legend with colors and symbols that match your plot.

<div id="ds-objectives" markdown="1">

## Learning Objectives
After completing this tutorial, you will be able to:

 * Plot multiple shapefiles using base R graphics.
 * Apply custom symbology to spatial objects in a plot in R.
 * Customize a baseplot legend in R.
 
## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and preferably `RStudio` loaded
on your computer to complete this tutorial.

### Install R Packages

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`
* **sp:** `install.packages("sp")`

<a href="https://www.neonscience.org/packages-in-r" target="_blank"> More on Packages in R </a>– Adapted from Software Carpentry.

## Download Data
<h3><a href="https://ndownloader.figshare.com/files/3708751" > NEON Teaching Data Subset: Site Layout Shapefiles</a></h3>

These vector data provide information on the site characterization and 
infrastructure at the 
<a href="https://www.neonscience.org/" target="_blank"> National Ecological Observatory Network's</a> 
<a href="https://www.neonscience.org/field-sites/field-sites-map/HARV" target="_blank"> Harvard Forest</a> 
field site.
The Harvard Forest shapefiles are from the 
 <a href="http://harvardforest.fas.harvard.edu/gis-maps/" target="_blank">Harvard Forest GIS & Map</a> 
archives. US Country and State Boundary layers are from the 
<a href="https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html" target="_blank">US Census Bureau</a>.

<a href="https://ndownloader.figshare.com/files/3708751" class="link--button link--arrow">
Download Dataset</a>




****

**Set Working Directory:** This lesson assumes that you have set your working 
directory to the location of the downloaded and unzipped data subsets. 

<a href="https://www.neonscience.org/set-working-directory-r" target="_blank"> An overview
of setting the working directory in R can be found here.</a>

**R Script & Challenge Code:** NEON data lessons often contain challenges that reinforce 
learned skills. If available, the code for challenge solutions is found in the
downloadable R script of the entire lesson, available in the footer of each lesson page.


</div>

## Load the Data
To work with vector data in R, we can use the `rgdal` library. The `raster` 
package also allows us to explore metadata using similar commands for both
raster and vector files. 

We will import three shapefiles. The first is our `AOI` or *area of
interest* boundary polygon that we worked with in 
<a href="https://www.neonscience.org/dc-open-shapefiles-r" target="_blank">*Open and Plot Shapefiles in R*</a>.
The second is a shapefile containing the location of roads and trails within the
field site. The third is a file containing the Harvard Forest Fisher tower
location. These latter two we worked with in the
<a href="https://www.neonscience.org/dc-shapefile-attributes-r" target="_blank">*Explore Shapefile Attributes & Plot Shapefile Objects by Attribute Value in R*</a> tutorial.

```{r load-packages-data }

# load packages
# rgdal: for vector work; sp package should always load with rgdal. 
library(rgdal)  
# raster: for metadata/attributes- vectors or rasters
library(raster)   

# set working directory to data folder
# setwd("pathToDirHere")

# Import a polygon shapefile 
aoiBoundary_HARV <- readOGR("NEON-DS-Site-Layout-Files/HARV",
                            "HarClip_UTMZ18", stringsAsFactors = T)

# Import a line shapefile
lines_HARV <- readOGR( "NEON-DS-Site-Layout-Files/HARV", "HARV_roads", stringsAsFactors = T)

# Import a point shapefile 
point_HARV <- readOGR("NEON-DS-Site-Layout-Files/HARV",
                      "HARVtower_UTM18N", stringsAsFactors = T)

```

## Plot Data

In the 
<a href="https://www.neonscience.org/dc-shapefile-attributes-r" target="_blank">*Explore Shapefile Attributes & Plot Shapefile Objects by Attribute Value in R*</a> tutorial
we created a plot where we customized the width of each line in a spatial object
according to a factor level or category. To do this, we create a vector of
colors containing a color value for EACH feature in our spatial object grouped
by factor level or category.

```{r plot-unique-lines, fig.cap="Roads and trails at NEON Harvard Forest Field Site with color and line width varied by attribute factor value." }

# view the factor levels
levels(lines_HARV$TYPE)
# create vector of line width values
lineWidth <- c(2,4,3,8)[lines_HARV$TYPE]
# view vector
lineWidth

# create a color palette of 4 colors - one for each factor level
roadPalette <- c("blue","green","grey","purple")
roadPalette
# create a vector of colors - one for each feature in our vector object
# according to its attribute value
roadColors <- c("blue","green","grey","purple")[lines_HARV$TYPE]
roadColors

# create vector of line width values
lineWidth <- c(2,4,3,8)[lines_HARV$TYPE]
# view vector
lineWidth

# in this case, boardwalk (the first level) is the widest.
plot(lines_HARV, 
     col=roadColors,
     main="NEON Harvard Forest Field Site\n Roads & Trails \nLine Width Varies by Type Attribute Value",
     lwd=lineWidth)

```

<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i> **Data Tip:** Given we have a factor with 4 levels, 
we can create a vector of numbers, each of which specifies the thickness of each
feature in our `SpatialLinesDataFrame` by factor level (category): `c(6,4,1,2)[lines_HARV$TYPE]`
</div>

## Add Plot Legend
In the 
<a href="https://www.neonscience.org/dc-shapefile-attributes-r" target="_blank">*the previous tutorial*</a>,
we also learned how to add a basic legend to our plot.

* `bottomright`: We specify the **location** of our legend by using a default 
keyword. We could also use `top`, `topright`, etc.
* `levels(objectName$attributeName)`: Label the **legend elements** using the
categories of `levels` in an attribute (e.g., levels(lines_HARV$TYPE) means use
the levels boardwalk, footpath, etc).
* `fill=`: apply unique **colors** to the boxes in our legend. `palette()` is 
the default set of colors that R applies to all plots. 

Let's add a legend to our plot.

```{r add-legend-to-plot, fig.cap="Roads and trails at NEON Harvard Forest Field Site with color varied by attribute factor value and with a default legend." }
plot(lines_HARV, 
     col=roadColors,
     main="NEON Harvard Forest Field Site\n Roads & Trails\n Default Legend")

# we can use the color object that we created above to color the legend objects
roadPalette

# add a legend to our map
legend("bottomright", 
       legend=levels(lines_HARV$TYPE), 
       fill=roadPalette, 
       bty="n", # turn off the legend border
       cex=.8) # decrease the font / legend size

```

However, what if we want to create a more complex plot with many shapefiles
and unique symbols that need to be represented clearly in a legend?

## Plot Multiple Vector Layers
Now, let's create a plot that combines our tower location (`point_HARV`), 
site boundary (`aoiBoundary_HARV`) and roads (`lines_HARV`) spatial objects. We
will need to build a **custom legend** as well.

To begin, create a plot with the site boundary as the first layer. Then layer 
the tower location and road data on top using `add=TRUE`.

```{r plot-many-shapefiles, fig.cap="Roads and tower location at NEON Harvard Forest Field Site with color varied by attribute type."}

# Plot multiple shapefiles
plot(aoiBoundary_HARV, 
     col = "grey93", 
     border="grey",
     main="NEON Harvard Forest Field Site")

plot(lines_HARV, 
     col=roadColors,
     add = TRUE)

plot(point_HARV, 
     add  = TRUE, 
     pch = 19, 
     col = "purple")

# assign plot to an object for easy modification!
plot_HARV<- recordPlot()

```


## Customize Your Legend

Next, let's build a custom legend using the symbology (the colors and symbols)
that we used to create the plot above. To do this,  we will need to build three 
things:

1. A list of all "labels" (the text used to describe each element in the legend
to use in the legend. 
2. A list of colors used to color each feature in our plot.
3. A list of symbols to use in the plot. NOTE: we have a combination of points,
lines and polygons in our plot. So we will need to customize our symbols!

Let's create objects for the labels, colors and symbols so we can easily reuse
them. We will start with the labels.

```{r create-custom-labels, fig.cap="Roads and tower location at NEON Harvard Forest Field Site with color varied by attribute type and with a modified legend." }

# create a list of all labels
labels <- c("Tower", "AOI", levels(lines_HARV$TYPE))
labels

# render plot
plot_HARV

# add a legend to our map
legend("bottomright", 
       legend=labels, 
       bty="n", # turn off the legend border
       cex=.8) # decrease the font / legend size
```

Now we have a legend with the labels identified. Let's add colors to each legend
element next. We can use the vectors of colors that we created earlier to do this.

```{r add-colors, fig.cap="Roads and tower location at NEON Harvard Forest Field Site with color and a modified legend varied by attribute type." }

# we have a list of colors that we used above - we can use it in the legend
roadPalette

# create a list of colors to use 
plotColors <- c("purple", "grey", roadPalette)
plotColors

# render plot
plot_HARV

# add a legend to our map
legend("bottomright", 
       legend=labels, 
       fill=plotColors,
       bty="n", # turn off the legend border
       cex=.8) # decrease the font / legend size

```

Great, now we have a legend! However, this legend uses boxes to symbolize each 
element in the plot. It might be better if the lines were symbolized as a line 
and the points were symbolized as a point symbol. We can customize this using
`pch=` in our legend: **16** is a point symbol, **15** is a box. 

<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i> **Data Tip:** To view a short list of `pch` symbols, 
type `?pch` into the R console. 
</div>


```{r custom-symbols, fig.cap="Roads and tower location at NEON Harvard Forest Field Site with color and a modified legend varied by attribute type; each symbol on the legend corresponds to the shapefile type (i.e., tower = point)."}

# create a list of pch values
# these are the symbols that will be used for each legend value
# ?pch will provide more information on values
plotSym <- c(16,15,15,15,15,15)
plotSym

# Plot multiple shapefiles
plot_HARV

# to create a custom legend, we need to fake it
legend("bottomright", 
       legend=labels,
       pch=plotSym, 
       bty="n", 
       col=plotColors,
       cex=.8)

```

Now we've added a point symbol to represent our point element in the plot. However
it might be more useful to use line symbols in our legend
rather than squares to represent the line data. We can create line symbols, 
using `lty = ()`. We have a total of 6 elements in our legend:

1. A Tower Location
2. An Area of Interest (AOI)
3. and **4** Road types (levels)

The `lty` list designates, in order, which of those elements should be
designated as a line (`1`) and which should be designated as a symbol (`NA`).
Our object will thus look like `lty = c(NA,NA,1,1,1,1)`. This tells R to only use a
line element for the 3-6 elements in our legend.

Once we do this, we still need to **modify** our `pch` element. Each **line** element
(3-6) should be represented by a `NA` value - this tells R to not use a
symbol, but to instead use a line.


```{r refine-legend, fig.cap="Roads and tower location at NEON Harvard Forest Field Site with color and a modified legend varied by attribute type; each symbol on the legend corresponds to the shapefile type [i.e., tower = point, roads = lines]." }
# create line object
lineLegend = c(NA,NA,1,1,1,1)
lineLegend
plotSym <- c(16,15,NA,NA,NA,NA)
plotSym

# plot multiple shapefiles
plot_HARV

# build a custom legend
legend("bottomright", 
       legend=labels, 
       lty = lineLegend,
       pch=plotSym, 
       bty="n", 
       col=plotColors,
       cex=.8)

```


<div id="ds-challenge" markdown="1">
### Challenge: Plot Polygon by Attribute

1. Using the `NEON-DS-Site-Layout-Files/HARV/PlotLocations_HARV.shp` shapefile, 
create a map of study plot locations, with each point colored by the soil type
(`soilTypeOr`). How many different soil types are there at this particular field 
site? Overlay this layer on top of the `lines_HARV` layer (the roads). Create a 
custom legend that applies line symbols to lines and point symbols to the points.

2. Modify the plot above. Tell R to plot each point, using a different
symbol of `pch` value. HINT: to do this, create a vector object of symbols by 
factor level using the syntax described above for line width: 
`c(15,17)[lines_HARV$soilTypeOr]`. Overlay this on top of the AOI Boundary. 
Create a custom legend.

</div>

``` {r challenge-code-plot-color, results="hide", warning= FALSE, echo=FALSE, fig.cap= c("Roads and study plots at NEON Harvard Forest Field Site with color and a modified legend varied by attribute type; each symbol on the legend corresponds to the shapefile type [i.e., soil plots = points, roads = lines].", "Roads and study plots at NEON Harvard Forest Field Site with color and a modified legend varied by attribute type; each symbol on the legend corresponds to the shapefile type [i.e., soil plots = points, roads = lines], and study plots symbols vary by soil type.")}
## 1
# open plot locations
plotLocations <- readOGR("NEON-DS-Site-Layout-Files/HARV",
          "PlotLocations_HARV")

# how many unique soils?  Two
unique(plotLocations$soilTypeOr)

# create new color palette -- topo.colors palette
blueGreen <- c("blue","darkgreen")
blueGreen

# plot roads
plot(lines_HARV, 
     col=roadColors,
     main="NEON Harvard Forest Field Site\n Study Plots by Soil Type\n One Symbol for All Types")

# plot the locations 
plot(plotLocations,
     col=(blueGreen)[plotLocations$soilTypeOr], 
     pch=18,
     add=TRUE)

# create line object
lineLegendElement = c(NA,NA,1,1,1,1)
lineLegendElement
plotSymElements <- c(18,18,NA,NA,NA,NA)
plotSymElements

# create vector of colors
colorElements <- c(blueGreen, roadPalette)
colorElements

# create legend 
legend("bottomright", 
       legend=c("Inceptisols","Histosols",levels(lines_HARV$TYPE)),
       pch=plotSymElements, 
       lty=lineLegendElement,
       col=colorElements,
       bty="n", 
       cex=1)

## 2
# create vector of DIFFERENT plot symbols
plSymbols <- c(15,17)[plotLocations$soilTypeOr]
plSymbols

# plot roads
plot(lines_HARV, 
     col=roadColors,
     main="NEON Harvard Forest Field Site\n Study Plots by Soil Type\n Different Symbols for Types")

# plot the locations 
plot(plotLocations,
     col=(blueGreen)[plotLocations$soilTypeOr], 
     pch=plSymbols,
     add=TRUE)

# create line object
lineLegendElement  <- c(NA,NA,1,1,1,1)
lineLegendElement
plotSymElementsMod <- c(15,17,NA,NA,NA,NA)
plotSymElementsMod

# create vector of colors
colorElements <- c(blueGreen, roadPalette)
colorElements
# create legend 
legend("bottomright", 
       legend=c("Inceptisols","Histosols",levels(lines_HARV$TYPE)),
       pch=plotSymElementsMod, 
       lty=lineLegendElement,
       col=colorElements,
       bty="n", 
       cex=1)
```

