---
syncID: add2081f81334a8194d5f00600430e25
title: "Time Series 01: Why Metadata Are Important: How to Work with Metadata in Text & EML Format"
description: "This tutorial covers what metadata are, and why we need to work with metadata. It covers the 3 most common metadata formats: text file format, web page format and Ecological Metadata Language (EML)."
dateCreated:  2015-10-22
authors: Leah A. Wasser, Megan A. Jones, Marisa Guarinello
contributors:
estimatedTime:
packagesLibraries: eml, ggmap
topics: metadata, time-series
languagesTool: R
dataProduct:
code1: /R/dc-tabular-time-series/01-Time-Series-Metadata.R
tutorialSeries: tabular-time-series
urlTitle: dc-metadata-importance-eml-r
---

This tutorial covers what metadata are, and why we need to work with
metadata. It covers the 3 most common metadata formats: text file format, 
web page format and Ecological Metadata Language (EML).

**R Skill Level:** Introduction - you've got the basics of R down and 
understand the general structure of tabular data.

<div id="ds-objectives" markdown="1">

## Learning Objectives
After completing this tutorial, you will be able to:

* Import a .csv file and examine the structure of the related R
object. 
* Use a metadata file to better understand the content of a dataset.
* Explain the importance of including metadata details in your R script.
* Describe what an EML file is. 

## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and, preferably, RStudio loaded on
your computer to complete this tutorial.

### Install R Packages

When presented in a workshop, the `EML` package will be presented as a
demonstration! 

If completing EML portion of tutorial on your own, you must
install `EML` directly from GitHub (the package is in development and is not 
yet on CRAN). You will need to install `devtools` to do this. 
 
* **ggmap:** `install.packages("ggmap")`
* **devtools:** `install.packages("devtools")` 
* **EML:** **must be installed AFTER `devtools` is loaded in R.
`install_github("ropensci/EML", build=FALSE, dependencies=c("DEPENDS", "IMPORTS"))`  

### Download Data 
{% include/dataSubsets/_data_Met-Time-Series.html %}

****

{% include/_greyBox-wd-rscript.html %}

</div>

## Understand Our Data

In order to work with any data, we need to understand three things about the
data:

* Structure 
* Format
* Processing methods

If the data are collected by other people and organizations, we might also need 
further information about:

* What metrics are included in the dataset
* The units those metrics were stored in
* Explanation of where the metrics are stored in the data and what they are "called"
(e.g. what are the column names in a spreadsheet)
* The time range that it covers
* The spatial extent that the data covers

The above information, and more are stored in **metadata** - data
about the data. Metadata is information that describes a dataset and is integral
to working with external data (data that we did not collect ourselves).

## Metadata Formats

Metadata come in different formats. We will discuss three of those in this tutorial:

* **Ecological Metadata Language (EML):** A standardized metadata format stored 
in `xml` format which is machine readable. Metadata has some standards however it's
common to encounter metadata stored differently in EML files created by different
organizations.
* **Text file:** Sometimes metadata files can be found in text files that are either
downloaded with a data product OR that are available separately for the data. 
* **Directly on a website (HTML / XML):** Sometimes data are documented directly
in text format, on a web page. 

<div id="ds-dataTip" markdown="1">
<i class="fa fa-star"></i> **Data Tip:** When you find metadata for a dataset 
that you are working with, immediately **DOWNLOAD AND SAVE IT** to the directory
on your computer where you saved the data. It is also a good idea to document
the URL where you found the metadata and data in a "readme" text file!
</div>


## Metadata Stored on a Web Page

The metadata for the data that we are working with for the Harvard Forest field 
site are stored in both `EML` format and on a web page. Let's explore the web 
page first.

* <a href="http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001" target="_blank">View Harvard Forest Fisher Tower webpage.</a>

Let's begin by visiting that page above. At the top of the page, there is a list
of data available for Harvard Forest. NOTE: **hf001-06: daily (metric) since 
2001 (preview)** is the data that we used in the
<a href="/dc-brief-tabular-time-series-qplot-r" target=_blank">previous tutorial.</a>

Scroll down to the **Overview** section on the website. Take note of the 
information provided in that section and answer the questions in the
Challenge below.

<div id="ds-challenge" markdown="1">
### Challenge: Explore Metadata

Explore the metadata stored on the Harvard Forest LTER web page. Answer the 
following questions. 

1. What is the time span of the data available for this site? 
2. You have some questions about these data. Who is the lead investigator / who 
do you contact for more information? And how do you contact them? 
3. Where is this field site located? How is the site location information stored
in the metadata? Is there any information that may be useful to you viewing the
location? HINT: What if you were not familiar with Harvard as a site / from 
another country, etc?
4. Field Site Information: What is the elevation for the site? What is the 
dominant vegetation cover for the site? HINT: Is dominant vegetation easy to 
find in the metadata?
5. How is snow dealt with in the precipitation data?
6. Are there some metadata attributes that might be useful to access in a script
in R or `Python` rather than viewed on a web page? HINT: Can you answer all of
the questions above from the information provided on this website? Is there 
additional information that you might want to find on the page?
</div>

```{r metadata-challenge-1, echo=FALSE}
# Metadata Notes from hf001_10-15-m_Metadata.txt
#  1. 2001-2015
#  2. Emery Boos - located at the top of the document, email is available
#  3. a lat long is available in the metadata at the top, we see the location described
# as Prospect Hill Tract (Harvard Forest). 
# 4. 342 m elevation, the veg type is not clear in the metadata
# 5. Found in the methods: Delayed melting of snow and ice (caused by problems with rain gage heater or heavy precipitation) is noted in log - daily values are corrected if necessary but 15-minute values are not. The gage may underestimate actual precipitation under windy or cold conditions.
# 6. this could be a discussion. things like units, time zone, etc are all useful
# if accessed programmatically

```

### View Metadata For Metrics of Interest

For this tutorial series, we are interested in the drivers of plant phenology -
specifically air and soil temperature, precipitation and photosynthetically
active radiation (PAR). Let's look for descriptions of these variables in the
metadata and determine several key attributes that we will need prior to working
with the data.

<div id="ds-challenge" markdown="1">
### Challenge: Metrics of Interest Metadata

View the metadata at the URL above. Answer the following questions about the 
Harvard Forest LTER data - **hf001-10: 15-minute (metric) since 2005**: 

1. What is the column heading name where each variable (air temperature, soil
temperature, precipitation and PAR) is stored?
2. What are the units that each variable are stored in? 
3. What is the frequency of measurement collected for each and how are noData
values stored?
4. Where is the date information stored (in what field) and what timezone are 
the dates stored in? 

</div>

```{r metadata-challenge-2, echo=FALSE}
# Metadata Notes from hf001_10-15-m_Metadata.txt
# 1. airt, s10t, prec, parr

# 2. units for quantitative variables: Celsius (both temps), millimeters,
# molePerMeterSquared

# 3. airt & s10t: average of 1-second measurements. (unit: celsius / missing
# value: NA)
# prec: Total value for 15-minute period. Measured in increments of 0.01 inch.
# (unit: millimeter / missing value: NA)
# parr: Average of 1-second measurements. (unit:
# micromolePerMeterSquaredPerSecond / missing value: NA)

# 4. datetime field, Eastern Standard Time. - note this is found in the methods
# towards the top of the document.
```

### Why Metadata on a Web Page Is Not Ideal

It is nice to have a web page that displays metadata information, however
accessing metadata on a web page is difficult:

* If the web page URL changes or the site goes down, the information is lost. 
* It's also more challenging to access metadata in text format on a web page
programatically - like using R as an interface - which we often
want to do when working with larger datasets. 

A machine readable metadata file is better - especially when we are working with
large data and we want to automate and carefully document workflows. The 
Ecological Metadata Language (EML) is one machine readable metadata format.

## Ecological Metadata Language (EML)

While  much of the documentation that we need to work with at the Harvard Forest
field site is available directly on the 
<a href="http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001" target="_blank">Harvard Forest Data Archive page</a>,
the website also offers metadata in `EML` format. 

## Introduction to EML

The Ecological Metadata Language (EML) is a data specification developed
specifically
to document ecological data. An EML file is created using a `XML` based format.
This means that content is embedded within hierarchical tags. For example,
the title of a dataset might be embedded in a `<title>` tag as follows:

`<title>Fisher Meteorological Station at Harvard Forest since 2001</title>`

Similarly, the creator of a dataset is also be found in a hierarchical tag
structure.

    <creator>
      <individualName>
        <givenName>Emery</givenName>
        <surName>Boose</surName>
      </individualName>
    </creator>

The `EML` package for R is designed to read and allow users to work with EML
formatted metadata. In this tutorial, we demonstrate how we can use EML in an
automated workflow. 

NOTE: The `EML` package is still being developed, therefore we will not 
explicitly teach all details of how to use it. Instead, we will provide
an example of how you can access EML files programmatically and background 
information so that you can further explore EML and the `EML` package if you
need to work with it further.

## EML Terminology

Let's first discuss some basic EML terminology. In the 
context of EML, each EML file documents a **dataset**. This **dataset** may 
consist of one or more files that contain the data in **data tables**. In the 
case of our tabular meteorology data, the structure of our EML file includes:

1. The **dataset**. A dataset may contain
one or more data tables associated with it that may contain different types of
related information. For this Harvard Forest meteorological data, the data 
tables contain tower measurements including precipitation and temperature, that
are aggregated at various time intervals (15 minute, daily, etc) and that date 
back to 2001.
2. The **data tables**. Data tables refer to the actual data that make up the
dataset. For the Harvard Forest data, each data table contains a suite of
meteorological metrics, including precipitation and temperature (and associated
quality flags), that are aggregated at a particular time interval (e.g. one 
data table contains monthly average data, another contains 15 minute 
averaged data, etc).

## Work With EML in R 

To begin, we will load the `EML` package directly from ROpenSci's Git repository.

```{r install-EML-package, results="hide", warning=FALSE }
# install R EML tool 
# load devtools (if you need to install "EML")
#library("devtools")
# IF YOU HAVE NOT DONE SO ALREADY: install EML from github -- package in
# development; not on CRAN
#install_github("ropensci/EML", build=FALSE, dependencies=c("DEPENDS", "IMPORTS"))

# load ROpenSci EML package
library("EML")
# load ggmap for mapping
library(ggmap)

```

Next, we will read in the LTER `EML` file - directly from the online URL using
`eml_read`. This file documents multiple data products that can be downloaded.
Check out the 
<a href="http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001" target="_blank"> Harvard Forest Data Archive Page for Fisher Meteorological Station</a>
for more on this dataset and to download the archive files directly.

Note that because this EML file is large, it takes a few seconds for the file to
load.

```{r read-eml }
# data location
# http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001
# table 4 http://harvardforest.fas.harvard.edu/data/p00/hf001/hf001-04-monthly-m.csv

# import EML from Harvard Forest Met Data
# note, for this particular tutorial, we will work with an abridged version of the file
# that you can access directly on the harvard forest website. (see comment below)
eml_HARV <- read_eml("http://harvardforest.fas.harvard.edu/data/eml/hf001.xml")

# import a truncated version of the eml file for quicker demonstration
# eml_HARV <- read_eml("http://neon-workwithdata.github.io/NEON-R-Tabular-Time-Series/hf001-revised.xml")

# view size of object
object.size(eml_HARV)

# view the object class
class(eml_HARV)

```

The `eml_read` function creates an `EML` class object. This object can be
accessed using `slots` in R (`@`) rather than a typical subset `[]` approach.

# Explore Metadata Attributes

We can begin to explore the contents of our EML file and associated data that it
describes. Let's start at the **dataset** level. We can use `slots` to view 
the contact information for the dataset and a description of the methods.

```{r view-eml-content }
# view the contact name listed in the file

eml_HARV@dataset@creator

# view information about the methods used to collect the data as described in EML
eml_HARV@dataset@methods


```

## Identify & Map Data Location

Looking at the coverage for our data, there is only one unique x and y value.
This suggests that our data were collected at (x,y) one point location. We know
this is a tower so a point location makes sense. Let's grab the x and y
coordinates and create a quick context map. We will use `ggmap` to create our
map.

NOTE: If this were a rectangular extent, we'd want the bounding box not just
the point. This is important if the data in raster, HDF5, or a similar format. 
We need the extent to properly geolocate and process the data.

```{r map-location, warning=FALSE, message=FALSE}

# grab x coordinate from the coverage information
XCoord <- eml_HARV@dataset@coverage@geographicCoverage[[1]]@boundingCoordinates@westBoundingCoordinate@.Data

# grab y coordinate from the coverage information
YCoord <- eml_HARV@dataset@coverage@geographicCoverage[[1]]@boundingCoordinates@northBoundingCoordinate@.Data

# map <- get_map(location='Harvard', maptype = "terrain")

# plot the NW corner of the site.
map <- get_map(location='massachusetts', maptype = "toner", zoom =8)

ggmap(map, extent=TRUE) +
  geom_point(aes(x=as.numeric(XCoord),y=as.numeric(YCoord)), 
             color="darkred", size=6, pch=18)

```

* Learn more about `ggmap`:
<a href="https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/ggmap/ggmapCheatsheet.pdf" target="_blank"> A nice cheatsheet created by NCEAS</a>

The above example, demonstrated how we can extract information from an `EML`
document and use it programatically in R! This is just the beginning of what
we can do!

## Metadata For Your Own Data 

Now, imagine that you are working with someone else's data and you don't have a 
metadata file associated with it? How do you know what units the data were in?
How the data were collected? The location that the data covers? It is equally 
important to create metadata for your own data, to make your data more
efficiently "shareable". 
