---
title: "Example workflow for processing NEON eddy-covariance turbulence data with eddy4R-Docker 0.2.0"
author: "David Durden, Stefan Metzger, Natchaya Pingintha-Durden, Claire Lunch, Megan Jones"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example workflow for processing NEON eddy-covariance turbulence data with eddy4R-Docker 0.2.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 0. Install and set up

This vignette assumes you are working with eddy4R within the Docker 
environment established by NEON. To get set up in that environment, 
follow the instructions <a href="https://github.com/NEONScience/NEON-Data-Skills/blob/fbd76ce778bc369fa9b94b7751546abb1ae56073/tutorials/R/eddy4r/dockerEddy4r.md" target="_blank">here</a>.
[CKL: Megan, I put a link to the page on GitHub, because I don't think we've 
put this up on the website yet]

At the end of those instructions, you should have an RStudio 
environment running within Docker. Use that workspace for this 
vignette.

## 1. Install packages and set up environment

First, check for required packages and install any that aren't 
already installed. Once all required packages are installed, 
load them into the environment.

Throughout this vignette, we use the `package::function()` 
syntax to explicitly refer to functions. This is done to avoid any 
possibility of ambiguity, in case there are functions in 
different packages with the same name.


```{r pack-install}

packReq <- c("DataCombine", "eddy4R.base", "eddy4R.qaqc",  
             "ff", "ffbase", "methods", "rhdf5","splus2R")
base::lapply(packReq, function(x) {
  print(x)
  if(base::require(x, character.only = TRUE) == FALSE) {
    utils::install.packages(x)
    base::library(x, character.only = TRUE)
  }})
base::rm(packReq)

```


## 2. Set environment variables

Here we set file paths and basic input parameters as environment 
variables for convenience.

```{r env-vars}

# file path for the directory containing input L0' data
base::Sys.setenv("DIRINP" = "/home/eddy/inpExmp")
#DD: This will be the case in the future, but the dp0p HDF5 file is not package into the deve docker, so you need to point it to a local file until we get ready to release the publication docker image.
# base::Sys.setenv("USER"="cklunch")
# base::Sys.setenv("DIRINP" = paste0("/home/", Sys.getenv("USER"), "/eddy/data/turbTow/inpRefe"))

#DD: This line is not needed in the new version 
#file path to the HDF5 file of input L0' data
#base::Sys.setenv("DIRFILEPARA" = 
 #                  "/home/eddy/inpExmp/ECTE_dp0p_CPER_2017-05-01.h5")

# file path for outputs
base::Sys.setenv("DIROUT" = "/home/eddy/out")

# date(s) of the input data, specified here for the ouput file name
base::Sys.setenv("DATEOUT" = "2017-05-01")

# NEON domain, site, and data product ID of the input data
# for the output file name:
# Domain 10 = D10
# Central Plains Experimental Range = CPER
# Bundled eddy covariance data = 00200.001
base::Sys.setenv("FILEOUTBASE" = "NEON.D10.CPER.00200.001")

# flag to indicate to the eddy4R.base::def.para.flow() 
# function that the variables above can be found as 
# environment variables, instead of provided as function 
# inputs
base::Sys.setenv("METHPARAFLOW" = "EnvVar")

```

## 3. Read in metadata

Pass along the environment variables defined in section 2, 
then read in additional parameters from the HDF5 file.

```{r metadata}

# start with an empty list of parameters
Para <- base::list()

# use the def.para.flow() function to specify the metadata 
# are stored in environment variables
Para$Flow <- eddy4R.base::def.para.flow(MethParaFlow = "EnvVar",                                          UrlInpRefe = "https://www.dropbox.com/s/d9ggsvepiypi9hb/inpRefe_20171220.zip?dl=1",                          UrlOutRefe = "https://www.dropbox.com/s/xf2viykgzajdfic/outRefe_20171220.zip?dl=1")

# [CKL: the error messages here are confusing. Is there bandwidth 
# on the EC team to modify? It would be helpful to add to the warning
# message, something like: "1: In FUN(X[[i]], ...) :
#  The variable ParaFlow$Deve is not specified as ENV variable; a 
#  default value has been assigned"
# Also, I'm now getting a new error: ""UrlInpRefe" is missing, with no default"


# [CKL: next few lines shown with error messages as comments. 
# Looks to me like some of the edits above resulted in 
# missing parameters]
  # Grab the NEON specific 4-letter code for the site location (Loc) from the dp0p input file   
  Para$Flow$Loc <- eddy4R.base::def.para.site(FileInp = Para$Flow$DirFilePara)$Loc 
  # Grab the tower top level (LvlTowr) from the dp0p input file 
  Para$Flow$LvlTowr <- eddy4R.base::def.para.site(FileInp = Para$Flow$DirFilePara)$LvlTowr
  
#Error: 'def.para.levl.towr' is not an exported object from 'namespace:eddy4R.base'
for(idx in c("soniAmrs", "irgaCo2", "irgaH2o", "soni")) {

        Para$Flow$dp01[[idx]] <- def.neon.read.hdf5.para(
            DirFileParaLoca = Para$Flow$DirFilePara,
            GrpName = paste0("/", Para$Flow$Loc, "/dp01/data/", idx),
            SetPara = c("PrdIncrAgrDflt", "PrdWndwAgrDflt")
          )
        
      }
#Error in def.neon.read.hdf5.para(DirFileParaLoca = Para$Flow$DirFilePara,  : 
#  unused argument (PosPara = c("PrdIncrAgrDflt", "PrdWndwAgrDflt"))

Para$Sci <- def.neon.read.hdf5.para(
    DirFileParaLoca = Para$Flow$DirFilePara,
    GrpName = "site",
    SetPara = c("Pf$AngEnuXaxs", "Pf$AngEnuYaxs", "Pf$Ofst", "ZoneTime")
  )
#Error in def.neon.read.hdf5.para(DirFileParaLoca = Para$Flow$DirFilePara,  : 
#  unused argument (PosPara = c("Pf$AngEnuXaxs", "Pf$AngEnuYaxs", "Pf$Ofst", "ZoneTime"))

```







[CKL: everything below comes with the vignette template, I 
kept it for my own reference, will delete in final version]

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
