---
syncID: 4a21c923ecc848e08af1222e5552bc89
title: "Create a Canopy Height Model from Lidar-derived rasters in R"
description: "In this tutorial, you will bring lidar-derived raster data (DSM and DTM) into R and difference them to create a canopy height model (CHM)."
dateCreated: 2014-07-21
authors: Edmund Hart, Leah A. Wasser
contributors: Donal O'Leary, Bridget Hass
estimatedTime: 0.5 Hours
packagesLibraries: terra, neonUtilities
topics: lidar, R, raster, remote-sensing, spatial-data-gis
languagesTool: R
dataProduct:
code1: https://raw.githubusercontent.com/NEONScience/NEON-Data-Skills/main/tutorials/R/AOP/Lidar/intro-to-lidar/make-chm-R/create-canopy-height-model-in-R.R
tutorialSeries: intro-lidar-r-series
urlTitle: create-chm-rasters-r
---


A common analysis using lidar data are to derive top of the canopy height values 
from the lidar data. These values are often used to track changes in forest 
structure over time, to calculate biomass, and even leaf area index (LAI). Let's 
dive into the basics of working with raster formatted lidar data in R! 

<div id="ds-objectives" markdown="1">

## Learning Objectives

After completing this tutorial, you will be able to:

* Work with digital terrain model (DTM) & digital surface model (DSM) raster files. 
* Create a canopy height model (CHM) raster from DTM & DSM rasters. 

 
## Things You’ll Need To Complete This Tutorial
You will need the most current version of R and, preferably, `RStudio` loaded 
on your computer to complete this tutorial.

### Install R Packages

* **terra:** `install.packages("terra")`
* **neonUtilities:** `install.packages("neonUtilities")`

<a href="https://www.neonscience.org/packages-in-r" target="_blank">More on Packages in R - Adapted from Software Carpentry.</a>

## Download Data

Lidar elevation raster data are downloaded using the R `neonUtilities::byTileAOP` function in the script.

These remote sensing data files provide information on the vegetation at the 
<a href="https://www.neonscience.org/" target="_blank"> National Ecological Observatory Network's</a> 
<a href="https://www.neonscience.org/field-sites/SJER" target="_blank"> San Joaquin Experimental Range</a> 
and 
<a href="https://www.neonscience.org/field-sites/SOAP" target="_blank"> Soaproot Saddle</a> 
field sites. The entire datasets can be accessed from the 
<a href="http://data.neonscience.org" target="_blank"> NEON Data Portal</a>.


This tutorial is designed for you to set your working directory to the directory
created by unzipping this file.

****

**Set Working Directory:** This lesson will walk you through setting the working 
directory before downloading the datasets from neonUtilities.

<a href="https://www.neonscience.org/set-working-directory-r" target="_blank"> An overview
of setting the working directory in R can be found here.</a>

**R Script & Challenge Code:** NEON data lessons often contain challenges to reinforce 
skills. If available, the code for challenge solutions is found in the downloadable R 
script of the entire lesson, available in the footer of each lesson page.


***

## Recommended Reading
<a href="https://www.neonscience.org/chm-dsm-dtm-gridded-lidar-data" target="_blank">
What is a CHM, DSM and DTM? About Gridded, Raster LiDAR Data</a>

</div>

## Create a lidar-derived Canopy Height Model (CHM)

The National Ecological Observatory Network (NEON) will provide lidar-derived 
data products as one of its many free ecological data products. These products 
will come in the 
<a href="http://trac.osgeo.org/geotiff/" target="_blank">GeoTIFF</a> 
format, which is a .tif raster format that is spatially located on the earth. 

In this tutorial, we create a Canopy Height Model. The 
<a href="https://www.neonscience.org/chm-dsm-dtm-gridded-lidar-data" target="_blank">Canopy Height Model (CHM)</a>,
represents the heights of the trees on the ground. We can derive the CHM 
by subtracting the ground elevation from the elevation of the top of the surface 
(or the tops of the trees). 

We will use the `terra` R package to work with the the lidar-derived Digital 
Surface Model (DSM) and the Digital Terrain Model (DTM). 

```{r load-libraries, message=FALSE, warning=FALSE}

# Load needed packages
library(terra)
library(neonUtilities)
```

Set the working directory so you know where to download data.

``` {r set-working-directory}

wd="~/data/" #This will depend on your local environment
setwd(wd)
```

We can use the `neonUtilities` function `byTileAOP` to download a single DTM and DSM tile at SJER. Both the DTM and DSM are delivered under the <a href="https://data.neonscience.org/data-products/DP3.30024.001" target="_blank">Elevation - LiDAR (DP3.30024.001)</a> data product.

You can run `help(byTileAOP)` to see more details on what the various inputs are. For this exercise, we'll specify the UTM Easting and Northing to be (257500, 4112500), which will download the tile with the lower left corner (257000,4112000). By default, the function will check the size total size of the download and ask you whether you wish to proceed (y/n). You can set `check.size=FALSE` if you want to download without a prompt. This example will not be very large (~8MB), since it is only downloading two single-band rasters (plus some associated metadata).

```{r download-refl, eval=FALSE}
byTileAOP(dpID='DP3.30024.001',
          site='SJER',
          year='2021',
          easting=257500,
          northing=4112500,
          check.size=TRUE, # set to FALSE if you don't want to enter y/n
          savepath = wd)
```

This file will be downloaded into a nested subdirectory under the `~/data` folder, inside a folder named `DP3.30024.001` (the Data Product ID). The files should show up in these locations:  `~/data/DP3.30024.001/neon-aop-products/2021/FullSite/D17/2021_SJER_5/L3/DiscreteLidar/DSMGtif/NEON_D17_SJER_DP3_257000_4112000_DSM.tif` and `~/data/DP3.30024.001/neon-aop-products/2021/FullSite/D17/2021_SJER_5/L3/DiscreteLidar/DTMGtif/NEON_D17_SJER_DP3_257000_4112000_DTM.tif`.

Now we can read in the files. You can move the files to a different location (eg. shorten the path), but make sure to change the path that points to the file accordingly.

```{r define-h5, results="hide" }
# Define the DSM and DTM file names, including the full path
dsm_file <- paste0(wd,"DP3.30024.001/neon-aop-products/2021/FullSite/D17/2021_SJER_5/L3/DiscreteLidar/DSMGtif/NEON_D17_SJER_DP3_257000_4112000_DSM.tif")
dtm_file <- paste0(wd,"DP3.30024.001/neon-aop-products/2021/FullSite/D17/2021_SJER_5/L3/DiscreteLidar/DTMGtif/NEON_D17_SJER_DP3_257000_4112000_DTM.tif")
```

First, we will read in the Digital Surface Model (DSM). The <a href="https://www.neonscience.org/resources/learning-hub/tutorials/chm-dsm-dtm" target="_blank">DSM</a> represents the elevation of the top of the objects on the ground (trees, buildings, etc).

```{r import-dsm}

# assign raster to object
dsm <- rast(dsm_file)

# view info about the raster.
dsm

# plot the DSM
plot(dsm, main="Lidar Digital Surface Model \n SJER, California")

```

Note the resolution, extent, and coordinate reference system (CRS) of the raster. 
To do later steps, our DTM will need to be the same. 

Next, we will import the Digital Terrain Model (DTM) for the same area. The 
<a href="{{ base.url }}/chm-dsm-dtm-gridded-lidar-data" target="_blank">DTM</a>
represents the ground (terrain) elevation.

```{r plot-DTM }

# import the digital terrain model
dtm <- rast(dtm_file)

plot(dtm, main="Lidar Digital Terrain Model \n SJER, California")

```

With both of these rasters now loaded, we can create the Canopy Height Model 
(CHM). The <a href="{{ base.url }}/chm-dsm-dtm" target="_blank">CHM</a>
represents the difference between the DSM and the DTM or the height of all objects
on the surface of the earth. 

To do this we perform some basic raster math to calculate the CHM. You can 
perform the same raster math in a GIS program like 
<a href="http://www.qgis.org/en/site/" target="_blank">QGIS</a>.

When you do the math, make sure to subtract the DTM from the DSM or you'll get 
trees with negative heights!

```{r calculate-plot-CHM }

# use raster math to create CHM
chm <- dsm - dtm

# view CHM attributes
chm

plot(chm, main="Lidar CHM - SJER, California")

```

We've now created a CHM from our DSM and DTM. What do you notice about the 
canopy cover at this location in the San Joaquin Experimental Range? 


<div id="ds-challenge" markdown="1">

### Challenge: Basic Raster Math 
Convert the CHM from meters to feet and plot it.
</div>

```{r challenge-code-raster-math, include=TRUE, results="hide", echo=FALSE}
# conversion 1m = 3.28084 ft
chm_ft <- chm*3.28084

# plot 
plot(chm_ft, main="Lidar Canopy Height Model (in feet)")
```

We can write out the CHM as a GeoTiff using the `writeRaster()` function. 

```{r write-raster-to-geotiff, eval=FALSE, comment=NA}
# write out the CHM in tiff format. 
writeRaster(chm,paste0(wd,"CHM_SJER.tif"),"GTiff")

```

We've now successfully created a canopy height model using basic raster math -- in 
R! We can bring the `CHM_SJER.tif` file into QGIS (or any GIS program) and look 
at it. 

***

Consider checking out the tutorial 
<a href="https://www.neonscience.org/tree-heights-veg-structure-chm/" target="_blank">*Compare tree height measured from the ground to a Lidar-based Canopy Height Model*</a> to compare a LiDAR-derived CHM with ground-based observations!

