---
title: "Download NEON water quality data"
description: "Tutorial for downloading NEON water quality data"
dateCreated: 2023-05-09
authors: Bobby Hensley
packagesLibraries: neonUtilities, ggplot2
---

NEON water quality sondes are deployed at stream sensor sets and lake and river buoys. At stream sites, the sondes are affixed to a post at a static depth relative to the stream bottom. The upstream sensor set #1 (S1) collects specific conductance, dissolved oxygen, pH, chlorophyll, and turbidity, but no fDOM. The downstream sensor set #2 (S2) collects specific conductance, dissolved oxygen, pH, chlorophyll, turbidity and fDOM. The buoy-deployed multisondes collect specific conductance, dissolved oxygen, pH, chlorophyll, turbidity, fDOM, and depth. At all but the Flint River, GA (FLNT) buoys, the water quality multisonde is fixed to a profiling winch to collect data from multiple depths every 4 hours and from 0.5 m parked depth when not profiling. Due to the high velocity of the Flint River (FLNT), one multisonde is deployed at a fixed depth of 0.5 m below the water surface.

```{r load-packages}
library(neonUtilities)
library(ggplot2)
```

For this example let's download a month of water quality data (DP1.20288.001) 
from the Flint River site (FLNT) from June 2022 (2022-06).  We'll download the
basic package and we can skip checking the file size because one month of data 
will not be very large.

```{r download-data-waq, results='hide'}

waq <- loadByProduct(dpID="DP1.20288.001", site="FLNT", 
                     startdate="2022-06", enddate="2022-06", 
                     package="expanded", 
                     release = "RELEASE-2023",
                     check.size = F)

```


```{r unlist-vars}
list2env(waq, .GlobalEnv)
```

Now we can plot the dissolved Oxygen time-series.  For plotting purposes we
can consider either the `startDateTime`or the `endDateTime` time stamp. From
the **variables_xxxxx** file we can see that these are the start and end times
of the interval over which measurements were collected.  Because water quality
is an instantaneous measurement that is not averaged, these will be the same.
Note all NEON data are always in UTC time. We will also consider the variable 
`dissolvedOxygen`, which the **variables_xxxxx** file tells us is the 
concentration in mg/L. 

```{r plot-wqual}

doPlot <- ggplot() +
	geom_line(data = waq_instantaneous, 
	          aes(endDateTime, dissolvedOxygen), 
	          na.rm=TRUE, color="blue") +
  geom_line( na.rm = TRUE) +
	ylim(6, 9) + ylab("DO (mg/L)") +
	xlab(" ") +
  ggtitle("Flint River - Dissolved Oxygen") 
  
  

doPlot

```
